---
title: "E10: the finalists on flu and covid"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  disease: "covid"
  hub_dir: "../covid19-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: how do the two finalists, REF-op and sqrt constant 0.018, do on
flu and covid, on the submitted ensemble and on the clean replay?

Part 1 runs them on both diseases and both bases, h0–h3, on identical rows.
Part 2 is the covid ensemble at h−1–h3 with REF-paper for reference, and the
standard views.

# Part 1: both diseases, both bases

Setup:

- Bases: the submitted `CMU-TimeSeries` ensemble (hub) and the clean
  `windowed_seasonal` replay, on the (round, location, horizon, level) rows
  both have, h0–h3 (the ensemble's h−1 is a different model).
- Learning truth: exact (re-run every round on the store's NHSN vintages),
  one week of extra delay. Scores against the store's finalized NHSN, states
  only, the two live seasons (August–July).
- Configs: REF-op warm (flu prod) and cold (covid prod); sqrt constant 0.018
  per 100k (E16), cold and warm (`slow_init = "burn_in_quantile"`). Warm
  runs see a 2023-24 burn-in season: for flu the rows both bases have (hub
  submissions and the HHS-stitched backfill, as E07); for covid only the
  replay's backfill, since covid submissions start 2024-11-23. So the covid
  ensemble has no warm runs. Cold runs see the live seasons only.

```{r fin-runs}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
key <- c("reference_date", "location", "horizon", "level")
fin_inputs <- function(d) {
  wi <- ws_inputs(d)
  # Right after ws_inputs(): both read the harness globals that ch_use() sets.
  burn <- ws_burn_in_forecasts(d)
  hub <- suppressMessages(hub_read_forecasts(
    hub_dir = here::here(HUB_DIRS[[d]]), target = c(flu = HUB_FLU_TARGET, covid = HUB_COVID_TARGET)[[d]]
  )) %>% filter(horizon %in% 0:3)
  ws <- bind_rows(burn, wi$fc)
  common <- inner_join(distinct(hub, across(all_of(key))), distinct(ws, across(all_of(key))), by = key)
  fcs <- list(ensemble = semi_join(hub, common, by = key), replay = semi_join(ws, common, by = key))
  warm <- fcs
  if (!any(season_of(fcs$ensemble$reference_date) == "2023-2024")) {
    warm <- list(replay = bind_rows(burn, fcs$replay))
  }
  list(wi = wi, cold = purrr::map(fcs, \(fc) filter(fc, season_of(reference_date) %in% LIVE)), warm = warm)
}
fin_cands <- function(rate_scales) {
  const <- ws_configs(rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
  sqrt_018 <- utils::modifyList(const, list(lr = 0.018, transform = "sqrt"))
  list(
    `REF-op warm` = list(cfg = list(ref = "op"), warm = TRUE),
    `REF-op cold` = list(cfg = list(ref = "op", burn_in_seasons = character(0), slow_init = NULL), warm = FALSE),
    `sqrt 0.018` = list(cfg = sqrt_018, warm = FALSE),
    `sqrt 0.018 warm` = list(cfg = utils::modifyList(sqrt_018, list(burn_in_seasons = "2023-2024", slow_init = "burn_in_quantile")), warm = TRUE)
  )
}
FIN <- purrr::map(c(flu = "flu", covid = "covid"), function(d) {
  inp <- fin_inputs(d)
  cands <- fin_cands(inp$wi$rate_scales)
  grid <- tidyr::expand_grid(base = c("ensemble", "replay"), cand = names(cands)) %>%
    filter(!cands[cand] %>% purrr::map_lgl("warm") | base %in% names(inp$warm))
  cals <- purrr::pmap(grid, function(base, cand) {
    cc <- cands[[cand]]
    fc <- if (cc$warm) inp$warm[[base]] else inp$cold[[base]]
    do.call(cal_run_cached, c(list(fc, inp$wi$truth), cc$cfg, list(learn = "exact", vintages = inp$wi$vintages, workers = params$workers)))
  })
  list(cals = setNames(cals, paste(grid$base, grid$cand, sep = ", ")), grid = grid, cands = names(cands))
})
fin_score <- function(cal) {
  ws_scores(cal$forecasts %>% mutate(season = season_of(reference_date)) %>% filter(location != "US")) %>%
    select(season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
FIN_M <- purrr::map(FIN, function(f) {
  purrr::map2(f$cals, seq_len(nrow(f$grid)), \(cal, i) fin_score(cal) %>% mutate(f$grid[i, ])) %>%
    bind_rows() %>%
    mutate(
      base = factor(base, levels = c("ensemble", "replay")), cand = factor(cand, levels = f$cands),
      season = factor(season, levels = c(LIVE, "both"), labels = c(LIVE, "both seasons")),
      h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
    )
})
BASE_COLS <- c(ensemble = "#0072B2", replay = "#D55E00")
fin_table <- function(m) {
  bind_rows(
    m %>% distinct(base, season, h, .keep_all = TRUE) %>% mutate(cand = "base", cell = sprintf("WIS %.1f", wis_base)),
    m %>% mutate(cell = sprintf("%+.1f", wis_pct))
  ) %>%
    mutate(config = paste(base, cand, sep = ", ")) %>%
    arrange(base, cand != "base", cand) %>%
    transmute(config = factor(config, levels = unique(config)), col = paste(season, h), cell) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell)
}
fin_plot <- function(m) {
  ggplot(m, aes(cand, wis_pct, fill = base)) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_col(position = position_dodge(width = 0.85, preserve = "single"), width = 0.8) +
    facet_grid(season ~ h) +
    scale_fill_manual(values = BASE_COLS, name = NULL) +
    labs(x = NULL, y = "WIS reduction % (positive = calibrated WIS lower than base)") +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
}
```

```{r fin-sections, results="asis", fig.height=6.5}
for (d in names(FIN)) {
  f <- FIN[[d]]
  m <- FIN_M[[d]]
  cat("\n\n", cal_h(2), " ", d, "\n\n", sep = "")
  cat("\n\n", cal_h(3), " WIS\n\n", sep = "")
  cat("WIS reduction % (positive = calibrated WIS lower than base), states only; the base rows give the base WIS.\n\n")
  print(knitr::kable(fin_table(m)))
  print(fin_plot(m))
  cat("\n\n", cal_h(3), " Coverage\n\n", sep = "")
  print(knitr::kable(cal_month_cov_table(cal_month_view(f$cals, seasons = LIVE)), caption = CAL_MONTH_COV_CAPTION))
  cat("\n\n", cal_h(3), " State by state\n\n", sep = "")
  cat("States share the national wave, so the counts are not independent tests.\n\n")
  op <- if (d == "flu") "REF-op warm" else "REF-op cold"
  pairs <- unlist(purrr::map(c("ensemble", "replay"), function(b) {
    nm <- function(x) paste(b, x, sep = ", ")
    ps <- list(c(nm(op), nm("sqrt 0.018")), c(nm("sqrt 0.018 warm"), nm("sqrt 0.018")))
    purrr::keep(ps, \(p) all(p %in% names(f$cals)))
  }), recursive = FALSE)
  print(knitr::kable(cal_pair_wins(cal_state_wis(f$cals), pairs), caption = CAL_PAIR_WINS_CAPTION))
}
```

```{r fin-month, results="asis"}
for (d in names(FIN)) {
  for (b in c("ensemble", "replay")) {
    mv <- cal_month_view(FIN[[d]]$cals[startsWith(names(FIN[[d]]$cals), paste0(b, ", "))], seasons = LIVE)
    mv_show_table <- FALSE
    cal_month_section(paste("Month view:", d, b), level = 2L)
  }
}
```

# Part 2: covid ensemble, h−1–h3

- Disease: covid. Base forecaster: the submitted `CMU-TimeSeries` ensemble.
- Learning truth: NHSN covid vintages, re-run at every round as prod does
  (E00). Scores against finalized covid NHSN.
- No burn-in season (covid submissions start 2024-11-23); both trackers start
  from zero offsets. REF-op runs without its warm start, as in covid prod.
- Rounds 2024-11-30 to 2026-09-19. The 2024-11-23 round is excluded (made
  from a bad NHSN release corrected the next day).
- Covid is submitted year-round; per-season views use August–July seasons.
- Horizons h−1 to h3. Headline tables: all hub locations; month view: states
  only.

| config | change from REF-paper |
|---|---|
| `REF-paper` | none (counts, adaptive learning rate) |
| `REF-op (no warm start)` | sqrt scale, learning-rate floor, slow term 0.003, leak 0.1; no warm start |
| `sqrt 0.018` | sqrt of the per-100k rate, constant learning rate 0.018 (E16), single term |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
no_burn_in <- list(burn_in_seasons = character(0), slow_init = NULL)
cals <- list(
  `REF-paper` = do.call(run, c(list("paper"), no_burn_in)),
  `REF-op (no warm start)` = do.call(run, c(list("op"), no_burn_in)),
  `sqrt 0.018` = do.call(run, c(list("paper", transform = "sqrt", lr = 0.018, scales = rate_scales), no_burn_in))
)
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
cals <- purrr::map(cals, function(cal) {
  cal$forecasts <- cal$forecasts %>% mutate(season = season_of(reference_date))
  cal
})
LIVE <- c("2024-2025", "2025-2026")
focus <- "REF-op (no warm start)"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```

# Findings, part 2

There are no uncertainty estimates; a WIS gap under 1 point or a month-avg
coverage gap under 0.01 is read as no clear difference. The pooled headline,
median-error and by-month tables run to 2026-09-19 to include the 2026 summer
wave; those Aug–Sep 2026 rounds are in no per-season table.

- **h−1:** both configs improve WIS by about 10–11% in each season
  (REF-paper 11.6/10.5, REF-op 8.9/10.8) and the absolute error of the median
  by about 20%. Which config is ahead flips between seasons.
- **REF-paper at h1–h3:** −4.5/−7.3/−10.3 pooled, almost all from 2025-26
  (−10.2/−14.4/−19.3). In 2024-25 it is within 0.3 points of the base.
- **REF-op at h1–h3:** 0.0/−0.6/−1.3 pooled; +2.4/+2.8/+1.9 in 2024-25 and
  −3.0/−4.0/−4.3 in 2025-26. No strong result: the sign depends on the
  season.
- **Month-avg coverage (states, h−1…h3):** base 0.148/0.102/0.125/0.143/0.157;
  REF-op 0.123/0.096/0.126/0.147/0.161 (lower at h−1 only); REF-paper
  0.111/0.094/0.126/0.155/0.171 (about 0.01 worse at h2/h3). Season-pooled,
  the base looks calibrated at h1–h3, but pooling lets the two seasons'
  errors cancel: within each season its bias is 0.072–0.078 (2024-25) and
  0.029–0.051 (2025-26).
- **By month at h3:** REF-op −16.5% in September and −12.0% in October
  (together 20% of base WIS); REF-paper −42% to −49% in April–May (about 6%).
