# Month view: WIS, coverage, median error

<!-- Expects `mv` from cal_month_view(). Optional: `mv_plot_cfgs` (configs to plot; the tables keep every config in `mv`), `mv_cols` (colours), `mv_colour_by` (a column of mv$month mapped to colour), `mv_linetype` (a column of mv$month mapped to linetype). -->

Scores by calendar month of the reference date, one figure per season,
**states only (US excluded)**. Rows are the three metrics, columns the
horizons:

- **WIS reduction %**: `100 × (WIS_base − WIS_cal) / WIS_base` within the
  month, so positive = calibrated WIS lower than base. Grey bars are the
  month's share of the season's base WIS, so a large swing in a month with a
  small bar costs little overall.
- **L1 coverage bias reduction %**: `100 × (bias_base − bias_cal) /
  bias_base`, so positive = calibrated coverage bias lower than base. The
  bias is computed within the month: the mean over the 23 levels of |share
  of outcomes at or below the quantile − level|. A season-pooled bias lets
  under-prediction on the ramp cancel over-prediction after the peak; the
  monthly one does not. When the base's bias is already small, a small
  absolute change shows as a large percentage; the month-average column of
  the coverage table below gives the absolute values.
- **Median AE reduction %**: the same as WIS reduction % for the absolute
  error of the 0.5 quantile, so positive = calibrated median closer to the
  outcome.

A month holds one to five rounds of about 52 states, and outcomes in the
same round move together, so a single month's coverage bias is noisy and
sits above zero even for a calibrated forecaster. That's why the plot shows
it relative to the base.

```{r mview-plots, results="asis", fig.height=7.5}
mv_plot_month <- mv$month
if (!is.null(get0("mv_plot_cfgs"))) {
  mv_plot_month <- mv_plot_month %>% filter(config %in% mv_plot_cfgs) %>% mutate(config = droplevels(config))
}
for (s in sort(unique(mv_plot_month$season))) {
  cat("\n\n", cal_h(2), " ", s, "\n\n", sep = "")
  print(cal_month_plot(mv_plot_month, s, colours = get0("mv_cols"), linetype = get0("mv_linetype"), colour_by = get0("mv_colour_by") %||% "config"))
}
```

```{r mview-tables, results="asis"}
cat("\n\n", cal_h(2), " By horizon\n\n", sep = "")
cat("WIS reduction % (positive = calibrated WIS lower than base), states only; `both` pools the seasons.\n\n")
print(knitr::kable(cal_month_wis_table(mv)))
print(knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION))
```
