# Standard views

<!-- Expects `cals` (a named list of runs) and `focus` (the run shown in the state panels and gallery). -->

Every score is over both live seasons, all
hub locations, against finalized truth; burn-in rounds are excluded.

## WIS and coverage

WIS reduction % (positive = calibrated WIS lower than base):

```{r v-head}
head_all <- cal_headline(cals)
knitr::kable(cal_wide(head_all, "pct_improvement"))
```

L1 coverage bias (lower is better; mean over levels of |coverage − level|), with the base first:

```{r v-head-cov}
knitr::kable(bind_rows(
  cal_wide(head_all %>% filter(variant == names(cals)[1]) %>% mutate(variant = "base"), "cal_error_base", 3),
  cal_wide(head_all %>% mutate(variant = as.character(variant)), "cal_error_cal", 3)
))
```

WIS reduction % by season (positive = calibrated WIS lower than base):

```{r v-head-season}
knitr::kable(
  cal_wide(cal_headline(cals, by = "season") %>% filter(season %in% LIVE), "pct_improvement", id = c("variant", "season")) %>%
    arrange(season, variant)
)
```

## Absolute error of the median

Summed |truth − median|, reduction % vs base (positive = calibrated absolute error lower than base):

```{r v-ae}
knitr::kable(cal_wide(cal_ae(cals), "ae_pct"))
```

## Where in the season

WIS reduction % (positive = calibrated WIS lower than base) by month of the reference date (clipped at −60), h0 and h3,
and each month's share of the base WIS at h3. Months run July to June; only
months with forecasts are shown.

```{r v-month, fig.height=4.5}
month_tbl <- purrr::imap(cals, function(cal, nm) cal_by_month(cal) %>% mutate(variant = nm)) %>%
  bind_rows() %>%
  mutate(variant = factor(variant, levels = names(cals)), month = droplevels(month))
ggplot(month_tbl %>% filter(horizon %in% c(0, 3)), aes(month, pct, colour = variant, group = variant)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_line() +
  geom_point() +
  facet_wrap(~ paste0("h", horizon)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  coord_cartesian(ylim = c(-60, 30)) +
  labs(x = "month of reference date", y = "WIS reduction % (positive = calibrated\nWIS lower than base)") +
  theme(legend.position = "bottom", legend.direction = "vertical")
knitr::kable(
  month_tbl %>%
    filter(horizon == 3) %>%
    transmute(variant, month, pct = round(pct, 1)) %>%
    tidyr::pivot_wider(names_from = month, values_from = pct) %>%
    bind_rows(month_tbl %>% filter(horizon == 3, variant == names(cals)[1]) %>%
      transmute(variant = "share of base WIS %", month, pct = round(share, 1)) %>%
      tidyr::pivot_wider(names_from = month, values_from = pct))
)
```

## Fixed states, `r focus` {.tabset}

The same states (US, CA, GA, MN, VT) in every experiment notebook. Every other
round's fan at h−1–h3: blue ribbon and line are the base 80% band and median,
orange lines the calibrated 10/50/90% quantiles, grey the finalized truth.

The black line is the last two reported weeks of NHSN data as published when
that round was forecast. Its dot is normally the round's h−1 week, where its
fan starts.
Plotted rounds are two weeks apart, so each black line also runs across the
first weeks of the previous round's fan. Black sits below grey where first
reports were later revised up. A black line with no fan in front of it, or a
gap between fans, is a round with no submission (for flu, 2025-10-25 to
2025-11-15 and 2026-04-18). Around Christmas and New Year 2024 the Wednesday
releases slipped to Thursday and Friday, so the 2024-12-28 and 2025-01-04
rounds saw one week less data than usual; their dot sits a week before the
fan starts.

```{r v-state, results="asis", fig.height=4}
for (nm in names(CAL_STATES)) {
  cat("\n\n### ", nm, "\n\n", sep = "")
  for (s in LIVE) print(cal_state_panel(cals[[focus]], CAL_STATES[[nm]], s, truth, vintages, title = paste(nm, s)))
}
```

## Gallery: `r focus` {.tabset}

For each season phase and location size (population above or below the
median), the forecast with the largest WIS gain and the one with the largest
loss, each from a different location. Forecasts are ranked by their
WIS reduction as a share of their own location-season's base WIS, so small states late in the season, where the base is
tiny, do not crowd the gallery. Ribbons are the 50%
and 90% bands, the dashed line is the forecast date.

```{r v-gallery, results="asis", fig.height=3.5}
picks <- cal_gallery_pick(cals[[focus]], population)
abbrev <- setNames(toupper(get_population_data()$state_id), get_population_data()$state_code)
for (i in seq_len(nrow(picks))) {
  p <- picks[i, ]
  if (i == 1L || p$phase != picks$phase[i - 1L]) cat("\n\n### ", as.character(p$phase), "\n\n", sep = "")
  sub <- sprintf("%s (%s), %s, %s: WIS reduction %+.1f%% (%+.1f%% of its season's base WIS; positive = calibrated WIS lower)", abbrev[[p$location]], p$size, p$reference_date, p$kind, p$wis_pct, p$share)
  print(cal_gallery_panel(cals[[focus]], p$location, p$reference_date, truth, vintages, subtitle = sub))
}
```
