---
title: "E01: adaptive learning-rate settings"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: how do the adaptive learning-rate settings change WIS and
coverage? The paper's tracker sets
`eta = mult × 90th percentile of |outcome − forecast|` over the last
`lr_window` observed rounds; the settings are the multiplier (`lr_mult`), the
window (`lr_window`), and the summer policy (`season_policy`: carry offsets
into the next season, or reset them to zero).

This re-asks the original E01 (a pre-fix, count-scale sweep on the submitted
ensemble with finalized truth, since deleted) on the current setup.

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (all 53
  locations, the 55 hub rounds; 27 in 2024-25, 28 in 2025-26). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Cold start: no burn-in season, offsets start at zero in November 2024. With
  no burn-in there is one summer to cross (2025), so the season policy only
  changes 2025-26.
- Tracker: REF-paper (single term, adaptive rate) on the square-root scale,
  learning-rate floor 0.001.
- Grid: `lr_mult` {0.3, 0.1, 0.03, 0.01} × `lr_window` {8, 20, Inf} ×
  `season_policy` {carry, reset}, 24 runs, plus one count-scale run at mult
  0.03, window 20, carry.
- Scores: states only (US excluded). Each season is scored on its
  own rounds.

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
MULTS <- c(0.3, 0.1, 0.03, 0.01)
WINDOWS <- c(8, 20, Inf)
POLICIES <- c("carry", "reset")
# The sqrt single-term tracker, cold; E02 builds its configs the same way so runs are shared.
sqrt_cfg <- function(mult = 0.03, window = 20, policy = "carry") {
  list(
    ref = "paper", transform = "sqrt", lr_args = list(mult = mult, floor = 1e-3),
    burn_in_seasons = character(0), slow_init = NULL, lr_window = window, season_policy = policy
  )
}
run <- function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
grid <- tidyr::expand_grid(mult = MULTS, window = WINDOWS, policy = POLICIES)
sweep <- purrr::pmap(grid, \(mult, window, policy) run(sqrt_cfg(mult, window, policy)))
count_ref <- run(list(ref = "paper", burn_in_seasons = character(0), slow_init = NULL))
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
label <- function(df) {
  df %>% mutate(
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
}
sweep_m <- purrr::map2(sweep, seq_len(nrow(grid)), \(cal, i) score(cal) %>% mutate(grid[i, ])) %>%
  bind_rows() %>%
  label() %>%
  mutate(window_f = factor(as.character(window), levels = as.character(WINDOWS)), policy = factor(policy, levels = POLICIES))
count_m <- score(count_ref) %>% label()
WINDOW_COLS <- c(`8` = "#E69F00", `20` = "#0072B2", `Inf` = "#009E73")
```

# 1. Coverage bias vs WIS over the multiplier

One line per window and season policy, tracing the multiplier from 0.01
(near the base, the open circle) to 0.3. Labels mark the multiplier on the
window-20 carry line. The cross is the count-scale run at mult 0.03, window
20, carry. The y axis is clipped at −15%.

```{r curve-fn}
curve_plot <- function(g) {
  d <- sweep_m %>% filter(geos == g) %>% arrange(mult)
  base <- d %>% distinct(h, season, cal_err_base)
  ggplot(d, aes(cal_err_cal, wis_pct, colour = window_f)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(aes(linetype = policy, group = interaction(window_f, policy)), linewidth = 0.7) +
    geom_point(size = 1) +
    geom_text(data = d %>% filter(window == 20, policy == "carry"), aes(label = mult),
      size = 2.4, vjust = -0.8, show.legend = FALSE) +
    geom_point(data = base, aes(cal_err_base, 0, shape = "base (uncalibrated)"), colour = "black", fill = "white", size = 2.2) +
    geom_point(data = count_m %>% filter(geos == g), aes(cal_err_cal, wis_pct, shape = "count scale, mult 0.03"), colour = "black", size = 2.4) +
    scale_shape_manual(values = c(`base (uncalibrated)` = 21, `count scale, mult 0.03` = 4), name = NULL) +
    scale_colour_manual(values = WINDOW_COLS, name = "lr_window") +
    scale_linetype_manual(values = c(carry = "solid", reset = "22"), name = "season policy") +
    facet_grid(season ~ h) +
    coord_cartesian(ylim = c(max(-15, min(d$wis_pct)), max(d$wis_pct) + 0.5)) +
    labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)", title = g) +
    theme(legend.position = "bottom")
}
```

```{r curve-states, fig.height=8}
curve_plot(GEOS[["states"]])
```

# 2. Which setting matters?

For each season and horizon, how far WIS reduction % moves when one setting
changes and the other two stay fixed: the range over the four multipliers,
over the three windows, and the carry-vs-reset gap, each averaged over the
settings held fixed. States only. The policy gap is zero in 2024-25 by
construction (no summer crossed yet).

```{r spread}
st <- sweep_m %>% filter(geos == GEOS[["states"]])
spread <- bind_rows(
  st %>% group_by(season, h, window, policy) %>% summarize(r = diff(range(wis_pct)), .groups = "drop") %>%
    group_by(season, h) %>% summarize(v = mean(r), .groups = "drop") %>% mutate(setting = "multiplier"),
  st %>% group_by(season, h, mult, policy) %>% summarize(r = diff(range(wis_pct)), .groups = "drop") %>%
    group_by(season, h) %>% summarize(v = mean(r), .groups = "drop") %>% mutate(setting = "window"),
  st %>% group_by(season, h, mult, window) %>% summarize(r = diff(range(wis_pct)), .groups = "drop") %>%
    group_by(season, h) %>% summarize(v = mean(r), .groups = "drop") %>% mutate(setting = "season policy")
)
spread %>%
  transmute(setting = factor(setting, levels = c("multiplier", "window", "season policy")), col = paste(season, h), v = round(v, 1)) %>%
  tidyr::pivot_wider(names_from = col, values_from = v) %>%
  arrange(setting) %>%
  knitr::kable(caption = "Mean range of WIS reduction % (points; positive = calibrated WIS lower than base) when one setting varies, states only")
```

# 3. Each metric against the multiplier

Top: WIS reduction % (positive = calibrated WIS lower than base), clipped at
−30% (mult 0.3 goes far below). Bottom: L1 coverage bias (lower is better), each season's base as a thin line. Carry only, states only;
colour is the window.

```{r vs-mult, fig.height=6.5}
long <- st %>%
  filter(policy == "carry", season != "both seasons") %>%
  transmute(h, season, mult, window_f, `WIS reduction % (positive = calibrated WIS lower than base)` = pmax(wis_pct, -30), `L1 coverage bias (lower is better)` = cal_err_cal) %>%
  tidyr::pivot_longer(c(`WIS reduction % (positive = calibrated WIS lower than base)`, `L1 coverage bias (lower is better)`), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("WIS reduction % (positive = calibrated WIS lower than base)", "L1 coverage bias (lower is better)")))
base_lines <- st %>%
  filter(season != "both seasons") %>%
  distinct(h, season, value = cal_err_base) %>%
  mutate(metric = factor("L1 coverage bias (lower is better)", levels = levels(long$metric)))
ggplot(long, aes(mult, value, colour = window_f, linetype = season)) +
  geom_hline(data = base_lines, aes(yintercept = value, linetype = season), colour = "grey40", linewidth = 0.3) +
  geom_line() +
  geom_point(size = 1) +
  scale_x_log10(breaks = MULTS) +
  scale_colour_manual(values = WINDOW_COLS, name = "lr_window") +
  scale_linetype_manual(values = c(`2024-2025` = "solid", `2025-2026` = "22"), name = NULL) +
  facet_grid(metric ~ h, scales = "free_y", labeller = labeller(metric = label_wrap_gen(28))) +
  labs(x = "lr_mult", y = NULL) +
  theme(legend.position = "bottom")
```

```{r month-view-cfgs}
# The plots show the multipliers around the elbow and the season policy at window 20, plus the count scale; the tables cover every config.
mv_cals <- c(
  setNames(sweep, with(grid, sprintf("sqrt mult %s, window %s, %s", mult, as.character(window), policy))),
  list(`count mult 0.03, window 20, carry` = count_ref)
)
mv <- cal_month_view(mv_cals, seasons = LIVE)
mv_plot_cfgs <- c(
  "sqrt mult 0.1, window 20, carry", "sqrt mult 0.03, window 20, carry", "sqrt mult 0.01, window 20, carry",
  "sqrt mult 0.03, window 20, reset", "count mult 0.03, window 20, carry"
)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
