---
title: "E01: adaptive learning-rate settings"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

The paper's tracker sets its learning rate from recent errors:
`eta = mult × 90th percentile of |outcome − forecast|` over the last
`lr_window` observed rounds. Three settings: the multiplier (`lr_mult`), the
window (`lr_window`), and what happens to the offsets over the summer
(`season_policy`: carry them into the next season, or reset them to zero).
**Which settings should we use, and which of the three matters?**

This asks the original E01's question (`calibration_qt_flu.Rmd`) on the
current setup. It is a new experiment, not a reproduction: four things
changed at once, so it can't say which change matters.

| | original E01 | this notebook |
|---|---|---|
| forecaster | submitted ensemble | clean `windowed_seasonal` replay |
| scale | count | sqrt |
| learning truth | finalized | as published each week |
| tracker code | before the 2026-09-17 fixes | after |

**Setup.** As E14–E17: flu, the clean `windowed_seasonal` replay (all 53
locations, the 55 hub rounds, h0–h3), learning from data as published each
week and re-run from scratch each week as prod does, one week of extra
revision delay, no burn-in season. The tracker is the plain single-term one
with the adaptive learning rate (REF-paper), **on the square-root scale**
with an eta floor of 0.001: that is the scale prod's tracker (REF-op) runs
on, and E03 found it better than counts at every horizon. One count-scale
run at the original operating point (mult 0.03, window 20, carry) links to
the old sweep.

Grid: `lr_mult` {0.3, 0.1, 0.03, 0.01} × `lr_window` {8, 20, Inf} ×
`season_policy` {carry, reset}, 24 runs. The old sweep also had window 50,
which it found indistinguishable from 20 and Inf; it is dropped here. With
no burn-in there is one summer to cross (2025 into 2025-26), so the season
policy only changes 2025-26.

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so
**positive = calibrated WIS lower than base**. "L1 coverage bias" (lower is
better) is the mean over the 23 levels of |empirical coverage − level|. Each season is scored on its own rounds.
Results lead with states only (US excluded).

# Findings

Numbers are from the render of 1 October 2026, states only, WIS reduction %
(positive = calibrated WIS lower than base) / L1 coverage bias (lower is better), h0/h1/h2/h3, both seasons pooled
unless stated. All-locations numbers point the same way.

- **The multiplier is the setting that matters.** Changing it moves WIS by
  16–81 points (pooled, averaged over the other settings); changing the
  window moves it by 1–2 points and the season policy by 0.3–1.7.
- **On the sqrt scale, 0.1 and 0.3 are far too large.** 0.1 loses WIS at
  every horizon (−1.3/−6.8/−11.3/−16.7 at window 20, carry) and 0.3 loses
  14–77%. **0.03** (REF-paper's value) gains at h0 and loses a little at
  h1–h3 (+1.9/−0.2/−1.1/−1.7), with coverage bias 0.032–0.045 against the
  base's 0.105–0.127. **0.01** is the only multiplier that never loses more
  than 0.2 (+1.7/+0.7/+0.5/+0.5), but from a cold start it barely moves
  2024-25's coverage (0.107–0.129 vs base 0.127–0.169). It catches up in
  2025-26 (0.015–0.041).
- **Window: keep 20.** It changes nothing in 2024-25 (under 0.2 points). In
  2025-26, Inf is up to a point worse than 20 at h1–h3 for mult ≤ 0.03; 8
  and 20 are within half a point.
- **Season policy: carry.** Reset only acts at the 2025 summer, and there it
  restarts the tracker from zero: at mult 0.03, window 20, 2025-26 goes from
  +2.2/−0.6/−0.7/−0.2 (carry) to +0.7/−0.9/−1.3/−1.6 (reset), and coverage
  bias from 0.008–0.016 to 0.039–0.050.
- **Count scale** at the same point loses more (−0.4/−2.5/−4.9/−7.0), as E03
  found.
- **Against the old E01** (count scale, ensemble, finalized truth, pre-fix):
  the same answer. The multiplier is the knob, the window is immaterial, 0.1
  overpays and 0.01 is near WIS parity but slow. One change: carry vs reset
  was "minor" before; here carry is clearly better in 2025-26.
- **Month view (2026-10-01):** scored within each month, the coverage
  ranking changes. Month-avg L1 coverage bias (both seasons, h0/h1/h2/h3,
  window 20, carry): base 0.112/0.127/0.153/0.177, mult 0.01
  0.093/0.114/0.142/0.168, 0.03 0.095/0.120/0.150/0.176, 0.1
  0.097/0.130/0.161/0.190. So 0.01, not 0.03, has the lowest month-level
  bias, and at h2–h3 0.03 is level with the base: most of its season-pooled
  coverage gain is ramp and post-peak errors cancelling. Carry vs reset
  flips slightly in 2025-26: reset 0.076/0.103/0.122/0.138, carry
  0.078/0.105/0.130/0.143.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
MULTS <- c(0.3, 0.1, 0.03, 0.01)
WINDOWS <- c(8, 20, Inf)
POLICIES <- c("carry", "reset")
# The sqrt single-term tracker, cold; E02 builds its configs the same way so runs are shared.
sqrt_cfg <- function(mult = 0.03, window = 20, policy = "carry") {
  list(
    ref = "paper", transform = "sqrt", lr_args = list(mult = mult, floor = 1e-3),
    burn_in_seasons = character(0), slow_init = NULL, lr_window = window, season_policy = policy
  )
}
run <- function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
grid <- tidyr::expand_grid(mult = MULTS, window = WINDOWS, policy = POLICIES)
sweep <- purrr::pmap(grid, \(mult, window, policy) run(sqrt_cfg(mult, window, policy)))
count_ref <- run(list(ref = "paper", burn_in_seasons = character(0), slow_init = NULL))
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
label <- function(df) {
  df %>% mutate(
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
}
sweep_m <- purrr::map2(sweep, seq_len(nrow(grid)), \(cal, i) score(cal) %>% mutate(grid[i, ])) %>%
  bind_rows() %>%
  label() %>%
  mutate(window_f = factor(as.character(window), levels = as.character(WINDOWS)), policy = factor(policy, levels = POLICIES))
count_m <- score(count_ref) %>% label()
WINDOW_COLS <- c(`8` = "#E69F00", `20` = "#0072B2", `Inf` = "#009E73")
```

# 1. Coverage bias vs WIS over the multiplier

One line per window and season policy, tracing the multiplier from 0.01
(near the base, the open circle) to 0.3. Labels mark the multiplier on the
window-20 carry line. The cross is the count-scale run at mult 0.03, window
20, carry. The y axis is clipped at −15%.

```{r curve-fn}
curve_plot <- function(g) {
  d <- sweep_m %>% filter(geos == g) %>% arrange(mult)
  base <- d %>% distinct(h, season, cal_err_base)
  ggplot(d, aes(cal_err_cal, wis_pct, colour = window_f)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(aes(linetype = policy, group = interaction(window_f, policy)), linewidth = 0.7) +
    geom_point(size = 1) +
    geom_text(data = d %>% filter(window == 20, policy == "carry"), aes(label = mult),
      size = 2.4, vjust = -0.8, show.legend = FALSE) +
    geom_point(data = base, aes(cal_err_base, 0, shape = "base (uncalibrated)"), colour = "black", fill = "white", size = 2.2) +
    geom_point(data = count_m %>% filter(geos == g), aes(cal_err_cal, wis_pct, shape = "count scale, mult 0.03"), colour = "black", size = 2.4) +
    scale_shape_manual(values = c(`base (uncalibrated)` = 21, `count scale, mult 0.03` = 4), name = NULL) +
    scale_colour_manual(values = WINDOW_COLS, name = "lr_window") +
    scale_linetype_manual(values = c(carry = "solid", reset = "22"), name = "season policy") +
    facet_grid(season ~ h) +
    coord_cartesian(ylim = c(max(-15, min(d$wis_pct)), max(d$wis_pct) + 0.5)) +
    labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)", title = g) +
    theme(legend.position = "bottom")
}
```

## States only

```{r curve-states, fig.height=8}
curve_plot(GEOS[["states"]])
```

## All locations

```{r curve-all, fig.height=8}
curve_plot(GEOS[["all"]])
```

# 2. Which setting matters?

For each season and horizon, how far WIS reduction % moves when one setting
changes and the other two stay fixed: the range over the four multipliers,
over the three windows, and the carry-vs-reset gap, each averaged over the
settings held fixed. States only. The policy gap is zero in 2024-25 by
construction (no summer crossed yet).

```{r spread}
st <- sweep_m %>% filter(geos == GEOS[["states"]])
spread <- bind_rows(
  st %>% group_by(season, h, window, policy) %>% summarize(r = diff(range(wis_pct)), .groups = "drop") %>%
    group_by(season, h) %>% summarize(v = mean(r), .groups = "drop") %>% mutate(setting = "multiplier"),
  st %>% group_by(season, h, mult, policy) %>% summarize(r = diff(range(wis_pct)), .groups = "drop") %>%
    group_by(season, h) %>% summarize(v = mean(r), .groups = "drop") %>% mutate(setting = "window"),
  st %>% group_by(season, h, mult, window) %>% summarize(r = diff(range(wis_pct)), .groups = "drop") %>%
    group_by(season, h) %>% summarize(v = mean(r), .groups = "drop") %>% mutate(setting = "season policy")
)
spread %>%
  transmute(setting = factor(setting, levels = c("multiplier", "window", "season policy")), col = paste(season, h), v = round(v, 1)) %>%
  tidyr::pivot_wider(names_from = col, values_from = v) %>%
  arrange(setting) %>%
  knitr::kable(caption = "Mean range of WIS reduction % (points; positive = calibrated WIS lower than base) when one setting varies, states only")
```

# 3. Each metric against the multiplier

Top: WIS reduction % (positive = calibrated WIS lower than base), clipped at
−30% (mult 0.3 goes far below). Bottom: L1 coverage bias (lower is better), each season's base as a thin line. Carry only, states only;
colour is the window.

```{r vs-mult, fig.height=6.5}
long <- st %>%
  filter(policy == "carry", season != "both seasons") %>%
  transmute(h, season, mult, window_f, `WIS reduction % (positive = calibrated WIS lower than base)` = pmax(wis_pct, -30), `L1 coverage bias (lower is better)` = cal_err_cal) %>%
  tidyr::pivot_longer(c(`WIS reduction % (positive = calibrated WIS lower than base)`, `L1 coverage bias (lower is better)`), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("WIS reduction % (positive = calibrated WIS lower than base)", "L1 coverage bias (lower is better)")))
base_lines <- st %>%
  filter(season != "both seasons") %>%
  distinct(h, season, value = cal_err_base) %>%
  mutate(metric = factor("L1 coverage bias (lower is better)", levels = levels(long$metric)))
ggplot(long, aes(mult, value, colour = window_f, linetype = season)) +
  geom_hline(data = base_lines, aes(yintercept = value, linetype = season), colour = "grey40", linewidth = 0.3) +
  geom_line() +
  geom_point(size = 1) +
  scale_x_log10(breaks = MULTS) +
  scale_colour_manual(values = WINDOW_COLS, name = "lr_window") +
  scale_linetype_manual(values = c(`2024-2025` = "solid", `2025-2026` = "22"), name = NULL) +
  facet_grid(metric ~ h, scales = "free_y", labeller = labeller(metric = label_wrap_gen(28))) +
  labs(x = "lr_mult", y = NULL) +
  theme(legend.position = "bottom")
```

# 4. Tables

WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage
bias (lower is better) per config (rows) and season × horizon (columns). The base row shows the base's coverage bias; the count-scale
run is last.

```{r tables}
season_table <- function(g) {
  rows <- bind_rows(
    sweep_m %>% filter(mult == MULTS[1], window == 20, policy == "carry") %>% mutate(config = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    sweep_m %>% arrange(desc(mult), window, policy) %>% mutate(config = sprintf("sqrt mult %s, window %s, %s", mult, as.character(window), policy)),
    count_m %>% mutate(config = "count mult 0.03, window 20, carry")
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(config, levels = unique(config)))
  rows %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

## States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r table-states-month}
mv_cals <- c(
  setNames(sweep, with(grid, sprintf("sqrt mult %s, window %s, %s", mult, as.character(window), policy))),
  list(`count mult 0.03, window 20, carry` = count_ref)
)
mv <- cal_month_view(mv_cals, seasons = LIVE)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

## All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```

## 50% and 90% interval coverage, states only, carry

```{r table-cov}
bind_rows(
  st %>% filter(mult == MULTS[1], window == 20, policy == "carry") %>% mutate(config = "base", cov50_cal = cov50_base, cov90_cal = cov90_base),
  st %>% filter(policy == "carry") %>% arrange(desc(mult), window) %>% mutate(config = sprintf("mult %s, window %s", mult, as.character(window)))
) %>%
  filter(h %in% c("h0", "h2"), season != "both seasons") %>%
  mutate(config = factor(config, levels = unique(config))) %>%
  transmute(config, col = paste(season, h), cell = sprintf("%.2f / %.2f", cov50_cal, cov90_cal)) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  arrange(config) %>%
  knitr::kable()
```

```{r month-view-cfgs}
# The multipliers and the season policy the Findings compare, at window 20, plus the count-scale run.
mv_plot_cfgs <- c(
  "sqrt mult 0.1, window 20, carry", "sqrt mult 0.03, window 20, carry", "sqrt mult 0.01, window 20, carry",
  "sqrt mult 0.03, window 20, reset", "count mult 0.03, window 20, carry"
)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
