---
title: "E02: switch-off date and learning-rate variants"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Variants of the adaptive tracker, each changed alone from the reference.
**Does any of them beat the reference?**

- **Switch off in spring** (`off_after`): from a calendar date until July,
  play the base forecast and stop learning; the offsets carry unchanged into
  the next season. Dates: 1 April, 1 March, 15 February.
- **Per-level learning rate** (`lr_args$per_level`): each of the 23
  quantile levels gets its own eta, instead of one eta pooled over levels.
- **Seasonal window** (`lr_seasonal`): eta is sized from the last 10 rounds
  plus the rounds within 5 weeks of the same point in earlier seasons, so the
  start of a ramp is sized by last year's ramp. Run with carry and with
  reset at the summer.

This asks the original E02's question (`calibration_qt_seasons_flu.Rmd`
and the "Method variants" table in `notes/CALIBRATION.md`) on the current
setup. It is a new experiment, not a reproduction: four things changed at
once, so it can't say which change matters.

| | original E02 | this notebook |
|---|---|---|
| forecaster | submitted ensemble | clean `windowed_seasonal` replay |
| scale | count | sqrt |
| learning truth | finalized | as published each week |
| tracker code | before the 2026-09-17 fixes | after |

The original also
had a geo-pooled eta (one eta shared across locations, population-weighted).
That option never beat the baseline and was removed from the code on
2026-09-17, so it is not run here.

**Setup.** As E01: flu, the clean `windowed_seasonal` replay (all 53
locations, the 55 hub rounds, h0–h3), learning from data as published each
week and re-run from scratch each week, one week of extra revision delay.
The reference is the sqrt single-term tracker with the adaptive learning
rate (mult 0.03, floor 0.001, window 20, carry), no burn-in: E01's center
point. The seasonal window needs an earlier season to look back on, so it
runs with the 2023-24 burn-in (the HHS-stitched backfill from E13), which
only warms eta; the reference is re-run with the same burn-in as a bridge.

"WIS change %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so **positive
is better**. "L1 coverage bias" is the mean over the 23 levels of
|empirical coverage − level|. Results lead with states only.

# Findings

Numbers are from the render of 1 October 2026, states only, WIS change %
(positive is better) / L1 coverage bias, h0/h1/h2/h3, both seasons pooled
unless stated. All-locations numbers point the same way.

- **Switching off in spring is the only variant with a large WIS gain.**
  The reference is +1.9/−0.2/−1.1/−1.7. `off_after` 1 March gives
  +2.8/+1.7/+2.0/+2.5 and 15 February +2.7/+2.1/+2.5/+3.2, positive at every
  horizon in both seasons. 1 April only halves the h2–h3 loss
  (+1.6/−0.2/−0.7/−0.8).
- **It costs coverage.** Coverage bias rises from 0.032–0.045 to
  0.080–0.091 (1 March) and 0.086–0.100 (15 February), against the base's
  0.105–0.127. From the switch-off date the base forecast is played, and the
  over-prediction after the peak no longer offsets the under-prediction on
  the ramp (E16).
- **Where the gain comes from.** It removes the reference's March–April
  losses (2024-25 h2: −31% in March, −41% in April). The offsets frozen in
  February also carry into autumn 2025 and help November–December (h2 +14%
  vs +5–6%) but hurt October (−60%, on under 1% of the season's WIS).
- **Same as E17.** E17 found `off_after = "03-01"` raised WIS and gave back
  most of the coverage gain for constant rates (sqrt 0.018:
  +2.3/+1.2/+1.2/+1.3, coverage bias 0.094–0.116). The adaptive tracker
  gains more from it because it loses more after the peak.
- **Per-level eta: a small gain at h1–h3** (+2.1/+0.5/0.0/−0.3), mostly from
  smaller March–April losses, with slightly higher coverage bias
  (0.043–0.055).
- **Seasonal window: no gain.** Against the reference with the same burn-in
  (+2.0/0.0/−0.6/−1.0), carry gives +2.0/−0.1/−0.8/−1.3 with lower coverage
  bias at h2–h3 (0.025–0.032 vs 0.040–0.044); reset gives
  +1.4/−0.2/−0.9/−1.6. The burn-in alone (warming eta on 2023-24) is worth
  +0.5–0.7 at h2–h3 over the cold reference.
- **Against the old E02** (count scale, ensemble, finalized truth, pre-fix):
  the same ranking. Switching off on 1 March or 15 February removes the WIS
  cost (15 February best at h1–h3), per-level eta is a modest gain and the
  seasonal window does not help. Two changes: reset now trails carry for the
  seasonal window (it used to recover some ground), and the coverage cost of
  switching off, not reported before, is large.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
fc_burn_in <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Same builder as E01, so the reference run is shared.
sqrt_cfg <- function(mult = 0.03, window = 20, policy = "carry") {
  list(
    ref = "paper", transform = "sqrt", lr_args = list(mult = mult, floor = 1e-3),
    burn_in_seasons = character(0), slow_init = NULL, lr_window = window, season_policy = policy
  )
}
ref <- sqrt_cfg()
burn <- list(burn_in_seasons = "2023-2024")
seasonal <- list(lr_window = 10, lr_seasonal = list(half_width_weeks = 5))
CFGS <- list(
  reference = list(cfg = ref),
  `off after 1 Apr` = list(cfg = c(ref, list(off_after = "04-01"))),
  `off after 1 Mar` = list(cfg = c(ref, list(off_after = "03-01"))),
  `off after 15 Feb` = list(cfg = c(ref, list(off_after = "02-15"))),
  `per-level eta` = list(cfg = utils::modifyList(ref, list(lr_args = list(mult = 0.03, floor = 1e-3, per_level = TRUE)))),
  `reference + burn-in` = list(cfg = utils::modifyList(ref, burn), burn_in = TRUE),
  `seasonal window, carry` = list(cfg = utils::modifyList(ref, c(burn, seasonal)), burn_in = TRUE),
  `seasonal window, reset` = list(cfg = utils::modifyList(ref, c(burn, seasonal, list(season_policy = "reset"))), burn_in = TRUE)
)
run <- function(spec) {
  fc <- if (isTRUE(spec$burn_in)) fc_burn_in else wi$fc
  do.call(cal_run_cached, c(list(fc, wi$truth), spec$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
cals <- purrr::map(CFGS, run)
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
m <- purrr::imap(cals, \(cal, nm) score(cal) %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(
    config = factor(config, levels = names(CFGS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
CFG_COLS <- setNames(c("black", "#56B4E9", "#0072B2", "#003F6B", "#CC79A7", "grey55", "#E69F00", "#D55E00"), names(CFGS))
```

# 1. Overall, by season

WIS change % per config, one panel per season and horizon, states only. The
seasonal-window bars compare with "reference + burn-in", the others with
"reference".

```{r head-plot, fig.height=6}
ggplot(m %>% filter(geos == GEOS[["states"]]), aes(config, wis_pct, fill = config)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = CFG_COLS, name = NULL) +
  labs(x = NULL, y = "WIS change % (positive is better)") +
  theme(legend.position = "bottom", axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

WIS change % / L1 coverage bias.

```{r tables}
season_table <- function(g) {
  bind_rows(
    m %>% filter(config == "reference") %>% mutate(config = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    m
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(config, levels = c("base", names(CFGS)))) %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

## States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

## All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```

## 50% and 90% interval coverage, states only

```{r table-cov}
bind_rows(
  m %>% filter(config == "reference") %>% mutate(config = "base", cov50_cal = cov50_base, cov90_cal = cov90_base),
  m
) %>%
  filter(geos == GEOS[["states"]], h %in% c("h0", "h2"), season != "both seasons") %>%
  mutate(config = factor(config, levels = c("base", names(CFGS)))) %>%
  transmute(config, col = paste(season, h), cell = sprintf("%.2f / %.2f", cov50_cal, cov90_cal)) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  arrange(config) %>%
  knitr::kable()
```

# 2. Month by month

WIS change % by calendar month of the reference date; rows are seasons,
columns horizons. Grey bars are each month's share of the season's base WIS.
States only, clipped at −40%. The burn-in configs are left out except the
seasonal window with reset.

```{r month, fig.height=6}
MONTH_CFGS <- c("reference", "off after 1 Mar", "off after 15 Feb", "per-level eta", "seasonal window, reset")
fc_m <- purrr::imap(cals[MONTH_CFGS], \(cal, nm) cal$forecasts %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), config = factor(config, levels = MONTH_CFGS)) %>%
  filter(season %in% LIVE, location != "US")
month_m <- ws_month_scores(fc_m) %>% mutate(h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
share <- month_m %>% filter(config == "reference")
ggplot(month_m, aes(month, pct, colour = config, group = config)) +
  geom_col(data = share, aes(month, share), inherit.aes = FALSE, fill = "grey85", width = 0.7) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_line() +
  geom_point(size = 1.1) +
  facet_grid(season ~ h, scales = "free_x", space = "free_x") +
  scale_colour_manual(values = CFG_COLS[MONTH_CFGS], name = NULL) +
  coord_cartesian(ylim = c(-40, 30)) +
  labs(x = "month of reference date", y = "WIS change %; grey: % of season base WIS") +
  theme(legend.position = "bottom")
```

h2 by month: WIS change % (share of outcomes below the calibrated median).

```{r month-table}
month_m %>%
  filter(h == "h2") %>%
  transmute(season, month, config, v = sprintf("%+.1f (%.2f)", pct, below_med_cal)) %>%
  tidyr::pivot_wider(names_from = config, values_from = v) %>%
  left_join(share %>% filter(h == "h2") %>% transmute(season, month, `base share %` = round(share, 1), `base below` = round(below_med_base, 2)),
    by = c("season", "month")) %>%
  arrange(season, month) %>%
  knitr::kable()
```
