---
title: "E02: switch-off date and learning-rate variants"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: how do three variants of the adaptive tracker, each changed alone
from a reference, change WIS and coverage?

- **Switch off in spring** (`off_after`): from a calendar date until July,
  play the base forecast and stop learning; the offsets carry unchanged into
  the next season. Dates: 1 April, 1 March, 15 February.
- **Per-level learning rate** (`lr_args$per_level`): each of the 23 quantile
  levels gets its own eta, instead of one eta pooled over levels.
- **Seasonal window** (`lr_seasonal`): eta is sized from the last 10 rounds
  plus the rounds within 5 weeks of the same point in earlier seasons. Run
  with carry and with reset at the summer. This also shortens the window
  from 20 to 10, so it changes two things at once.

This re-asks the original E02 (a pre-fix, count-scale comparison on the
submitted ensemble with finalized truth, since deleted) on the current setup.

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (all 53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Reference: E01's center point, the sqrt single-term tracker with the
  adaptive rate (mult 0.03, floor 0.001, window 20, carry), cold start (no
  burn-in).
- The seasonal window needs an earlier season to look back on, so it runs
  with the 2023-24 burn-in (the HHS-stitched backfill from E13), which only
  warms eta. The reference is re-run with the same burn-in
  (`reference + burn-in`) for comparison.
- Scores: states only (US excluded).

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
fc_burn_in <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Same builder as E01, so the reference run is shared.
sqrt_cfg <- function(mult = 0.03, window = 20, policy = "carry") {
  list(
    ref = "paper", transform = "sqrt", lr_args = list(mult = mult, floor = 1e-3),
    burn_in_seasons = character(0), slow_init = NULL, lr_window = window, season_policy = policy
  )
}
ref <- sqrt_cfg()
burn <- list(burn_in_seasons = "2023-2024")
seasonal <- list(lr_window = 10, lr_seasonal = list(half_width_weeks = 5))
CFGS <- list(
  reference = list(cfg = ref),
  `off after 1 Apr` = list(cfg = c(ref, list(off_after = "04-01"))),
  `off after 1 Mar` = list(cfg = c(ref, list(off_after = "03-01"))),
  `off after 15 Feb` = list(cfg = c(ref, list(off_after = "02-15"))),
  `per-level eta` = list(cfg = utils::modifyList(ref, list(lr_args = list(mult = 0.03, floor = 1e-3, per_level = TRUE)))),
  `reference + burn-in` = list(cfg = utils::modifyList(ref, burn), burn_in = TRUE),
  `seasonal window, carry` = list(cfg = utils::modifyList(ref, c(burn, seasonal)), burn_in = TRUE),
  `seasonal window, reset` = list(cfg = utils::modifyList(ref, c(burn, seasonal, list(season_policy = "reset"))), burn_in = TRUE)
)
run <- function(spec) {
  fc <- if (isTRUE(spec$burn_in)) fc_burn_in else wi$fc
  do.call(cal_run_cached, c(list(fc, wi$truth), spec$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
cals <- purrr::map(CFGS, run)
```

```{r month-view-cfgs}
CFG_COLS <- setNames(c("black", "#56B4E9", "#0072B2", "#003F6B", "#CC79A7", "grey55", "#E69F00", "#D55E00"), names(CFGS))
mv <- cal_month_view(cals, seasons = LIVE)
# The plots leave out the burn-in configs except the seasonal window with reset; the tables keep every config.
mv_plot_cfgs <- c("reference", "off after 1 Mar", "off after 15 Feb", "per-level eta", "seasonal window, reset")
mv_cols <- CFG_COLS
```

The seasonal-window configs compare with "reference + burn-in", the others
with "reference".

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
