---
title: "E02: switch-off date and learning-rate variants"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: how do three variants of the adaptive tracker, each changed alone
from a reference, change WIS and coverage?

- **Switch off in spring** (`off_after`): from a calendar date until July,
  play the base forecast and stop learning; the offsets carry unchanged into
  the next season. Dates: 1 April, 1 March, 15 February.
- **Per-level learning rate** (`lr_args$per_level`): each of the 23 quantile
  levels gets its own eta, instead of one eta pooled over levels.
- **Seasonal window** (`lr_seasonal`): eta is sized from the last 10 rounds
  plus the rounds within 5 weeks of the same point in earlier seasons. Run
  with carry and with reset at the summer. This also shortens the window
  from 20 to 10, so it changes two things at once.

This re-asks the original E02 (a pre-fix, count-scale comparison on the
submitted ensemble with finalized truth, since deleted) on the current setup.

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (all 53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Reference: E01's center point, the sqrt single-term tracker with the
  adaptive rate (mult 0.03, floor 0.001, window 20, carry), cold start (no
  burn-in).
- The seasonal window needs an earlier season to look back on, so it runs
  with the 2023-24 burn-in (the HHS-stitched backfill from E13), which only
  warms eta. The reference is re-run with the same burn-in
  (`reference + burn-in`) for comparison.
- Tables lead with states only (US excluded).

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
fc_burn_in <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Same builder as E01, so the reference run is shared.
sqrt_cfg <- function(mult = 0.03, window = 20, policy = "carry") {
  list(
    ref = "paper", transform = "sqrt", lr_args = list(mult = mult, floor = 1e-3),
    burn_in_seasons = character(0), slow_init = NULL, lr_window = window, season_policy = policy
  )
}
ref <- sqrt_cfg()
burn <- list(burn_in_seasons = "2023-2024")
seasonal <- list(lr_window = 10, lr_seasonal = list(half_width_weeks = 5))
CFGS <- list(
  reference = list(cfg = ref),
  `off after 1 Apr` = list(cfg = c(ref, list(off_after = "04-01"))),
  `off after 1 Mar` = list(cfg = c(ref, list(off_after = "03-01"))),
  `off after 15 Feb` = list(cfg = c(ref, list(off_after = "02-15"))),
  `per-level eta` = list(cfg = utils::modifyList(ref, list(lr_args = list(mult = 0.03, floor = 1e-3, per_level = TRUE)))),
  `reference + burn-in` = list(cfg = utils::modifyList(ref, burn), burn_in = TRUE),
  `seasonal window, carry` = list(cfg = utils::modifyList(ref, c(burn, seasonal)), burn_in = TRUE),
  `seasonal window, reset` = list(cfg = utils::modifyList(ref, c(burn, seasonal, list(season_policy = "reset"))), burn_in = TRUE)
)
run <- function(spec) {
  fc <- if (isTRUE(spec$burn_in)) fc_burn_in else wi$fc
  do.call(cal_run_cached, c(list(fc, wi$truth), spec$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
cals <- purrr::map(CFGS, run)
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
m <- purrr::imap(cals, \(cal, nm) score(cal) %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(
    config = factor(config, levels = names(CFGS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
CFG_COLS <- setNames(c("black", "#56B4E9", "#0072B2", "#003F6B", "#CC79A7", "grey55", "#E69F00", "#D55E00"), names(CFGS))
```

# 1. Overall, by season

WIS reduction % (positive = calibrated WIS lower than base) per config, one panel per season and horizon, states only. The
seasonal-window bars compare with "reference + burn-in", the others with
"reference".

```{r head-plot, fig.height=6}
ggplot(m %>% filter(geos == GEOS[["states"]]), aes(config, wis_pct, fill = config)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = CFG_COLS, name = NULL) +
  labs(x = NULL, y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom", axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

WIS reduction % against L1 coverage bias, one point per config, states only.
The open circle is the base (at its coverage bias and zero WIS reduction); up
and to the left is better. The seasonal-window configs compare with
"reference + burn-in", the others with "reference".

```{r scatter, fig.height=7}
d <- m %>% filter(geos == GEOS[["states"]])
base <- d %>% filter(config == "reference") %>% distinct(season, h, cal_err_base)
ggplot(d, aes(cal_err_cal, wis_pct, colour = config)) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_point(data = base, aes(cal_err_base, 0, shape = "base (uncalibrated)"), colour = "black", fill = "white", size = 2.2, inherit.aes = FALSE) +
  geom_point(size = 2) +
  facet_grid(season ~ h) +
  scale_colour_manual(values = CFG_COLS, name = NULL) +
  scale_shape_manual(values = c(`base (uncalibrated)` = 21), name = NULL) +
  expand_limits(x = 0) +
  labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom")
```

Per config (rows) and season × horizon (columns): first WIS reduction %
(positive = calibrated WIS lower than base), then L1 coverage bias (lower is
better), with the base's coverage bias in the base row.

```{r tables}
season_table <- function(g, value, fmt) {
  bind_rows(
    m %>% filter(config == "reference") %>% mutate(config = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    m
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(config, levels = c("base", names(CFGS)))) %>%
    transmute(config, col = paste(season, h), cell = sprintf(fmt, .data[[value]])) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
wis_table <- function(g) {
  knitr::kable(season_table(g, "wis_pct", "%+.1f"), caption = paste0("WIS reduction %, ", g, "; positive = calibrated WIS lower than base"))
}
cov_table <- function(g) {
  knitr::kable(season_table(g, "cal_err_cal", "%.3f"), caption = paste0("L1 coverage bias (lower is better), ", g))
}
```

## States only

```{r table-states}
wis_table(GEOS[["states"]])
cov_table(GEOS[["states"]])
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r table-states-month}
mv <- cal_month_view(cals, seasons = LIVE)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

## All locations

```{r table-all}
wis_table(GEOS[["all"]])
cov_table(GEOS[["all"]])
```

## 50% and 90% interval coverage, states only

```{r table-cov}
bind_rows(
  m %>% filter(config == "reference") %>% mutate(config = "base", cov50_cal = cov50_base, cov90_cal = cov90_base),
  m
) %>%
  filter(geos == GEOS[["states"]], h %in% c("h0", "h2"), season != "both seasons") %>%
  mutate(config = factor(config, levels = c("base", names(CFGS)))) %>%
  transmute(config, col = paste(season, h), cell = sprintf("%.2f / %.2f", cov50_cal, cov90_cal)) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  arrange(config) %>%
  knitr::kable()
```

# 2. Month by month

The month plots are in "Month view" below. h2 by month, states only: WIS
reduction % (positive = calibrated WIS lower than base) (share of outcomes
below the calibrated median). The burn-in configs are left out except the
seasonal window with reset.

```{r month-table}
MONTH_CFGS <- c("reference", "off after 1 Mar", "off after 15 Feb", "per-level eta", "seasonal window, reset")
fc_m <- purrr::imap(cals[MONTH_CFGS], \(cal, nm) cal$forecasts %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), config = factor(config, levels = MONTH_CFGS)) %>%
  filter(season %in% LIVE, location != "US")
month_m <- ws_month_scores(fc_m) %>% mutate(h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
share <- month_m %>% filter(config == "reference")
month_m %>%
  filter(h == "h2") %>%
  transmute(season, month, config, v = sprintf("%+.1f (%.2f)", pct, below_med_cal)) %>%
  tidyr::pivot_wider(names_from = config, values_from = v) %>%
  left_join(share %>% filter(h == "h2") %>% transmute(season, month, `base share %` = round(share, 1), `base below` = round(below_med_base, 2)),
    by = c("season", "month")) %>%
  arrange(season, month) %>%
  knitr::kable()
```

```{r month-view-cfgs}
# Same configs as the h2 table.
mv_plot_cfgs <- MONTH_CFGS
mv_cols <- CFG_COLS
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Findings

There are no uncertainty estimates; a WIS gap under 1 point or a month-avg
coverage gap under 0.01 is read as no clear difference. The cold start leaves
Nov–Dec 2024 equal to the base for the cold configs. States only, pooled over
both seasons unless stated.

- **Switching off from 1 March or 15 February:** +2.8/+1.7/+2.0/+2.5 and
  +2.7/+2.1/+2.5/+3.2 at h0–h3, against the reference's
  +1.9/−0.2/−1.1/−1.7, ahead at every horizon in both seasons. The two dates
  differ by 0.1–0.7 (no clear difference). Both dates were chosen with these
  seasons in view.
- **Where that gain sits:** in 2024-25 it is the March–April months, where
  the reference loses (h2: −31% in March, −41% in April). In 2025-26 about
  half of the h2 gain is December, where offsets frozen in February 2025 are
  still in play (+13.6% vs +6.3%), partly offset by January and October.
- **Switching off from 1 April:** +1.6/−0.2/−0.7/−0.8, within 1 point of the
  reference at h0–h2.
- **Per-level eta:** +2.1/+0.5/−0.0/−0.3, about 1 point above the reference
  at h1–h3 in both seasons.
- **Seasonal window and burn-in:** against `reference + burn-in`
  (+2.0/−0.0/−0.6/−1.0), the seasonal window is +2.0/−0.1/−0.8/−1.3 with carry
  and +1.4/−0.2/−0.9/−1.6 with reset. The burn-in alone is +0.5 to +0.7 at
  h2–h3 over the cold reference, in 2024-25 only. No strong result.
- **Month-avg coverage (h0…h3):** base 0.112/0.127/0.153/0.177; reference
  0.095/0.120/0.150/0.176; off 1 March 0.104/0.124/0.153/0.170; off 15
  February 0.106/0.126/0.153/0.169; per-level 0.092/0.117/0.147/0.173. No
  clear difference between any variant and the reference. Season-pooled the
  switch-offs look much worse than the reference (0.080–0.100 vs
  0.032–0.045), but pooling lets ramp and post-peak errors cancel.
