---
title: "E03: working scale (count, sqrt, rate)"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: does the scale the tracker works on (count, sqrt, rate) change
WIS and coverage?

Setup:

- Disease: flu. Base forecaster: the submitted hub ensemble (spoiled
  submissions dropped).
- Reference: REF-paper. One offset per level; adaptive learning rate of 0.03
  × the 90th percentile of the last 20 rounds' absolute errors; offsets
  carried across seasons; one week of extra delay (`settle_days = 14`).
- Burn-in: 2023-24, used only to warm up the learning rate. Offsets start at
  zero at the first 2024-25 round (cold start).
- Learning truth: exact (re-run at every round on the data published by then,
  as prod does; E00). Scores against finalized truth.
- Seasons scored: 2024-25 and 2025-26. Horizons h−1 to h3. States only
  (US excluded).

| config | change from REF-paper |
|---|---|
| `count` | none (additive offsets in admissions) |
| `sqrt` | offsets on the square root of admissions, mapped back; learning-rate floor 0.001 |
| `rate` | offsets in admissions per 100k; learning-rate floor 0.001 |

The learning-rate floor (0.1 by default) is in the tracker's working units,
so the sqrt and rate configs use 0.001.

```{r runs}
cals <- list(
  count = run("paper"),
  sqrt = run("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3)),
  rate = run("paper", scales = rate_scales, lr_args = list(mult = 0.03, floor = 1e-3))
)
```

```{r month-view-data}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
