---
title: "E03: working scale (count, sqrt, rate)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Ledger entry **E03** (`notes/calibration-ledger.md`). One axis: the scale
the tracker works on. Everything else is REF-paper (single term, adaptive
learning rate with `mult` 0.03 and window 20, carry, 2023-24 burn-in,
`settle_days` 14), learning from vintages the way prod does (re-run each
week; E00).

- **count**: REF-paper itself; additive offsets in admissions.
- **sqrt**: offsets on the square root of admissions, mapped back.
- **rate**: offsets in admissions per 100k, so one location's step size is
  comparable to another's.

The adaptive learning rate has a floor (0.1 by default) in the tracker's
working units, so the sqrt and rate variants use a floor of 0.001; that floor
change goes with the scale.

```{r runs}
cals <- list(
  count = run("paper"),
  sqrt = run("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3)),
  rate = run("paper", scales = rate_scales, lr_args = list(mult = 0.03, floor = 1e-3))
)
focus <- "sqrt"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
