---
title: "E03: working scale (count, sqrt, rate)"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

One axis: the scale the tracker works on. Everything else is REF-paper (one
offset per level, adaptive learning rate of 0.03 × the 90th percentile of the
last 20 rounds' absolute errors, offsets carried across seasons, 2023-24
burn-in, one week of extra delay), learning from vintages the way prod does
(re-run each week; E00).

- **count**: REF-paper itself; additive offsets in admissions.
- **sqrt**: offsets on the square root of admissions, mapped back.
- **rate**: offsets in admissions per 100k, so one location's step size is
  comparable to another's.

The adaptive learning rate has a floor (0.1 by default) in the tracker's
working units, so the sqrt and rate variants use a floor of 0.001; that floor
change goes with the scale.

# Findings

Numbers are from the render of 30 September 2026, pooled over both live seasons and all locations; WIS and absolute-error changes are % vs base, positive is better.

- **sqrt beats count at every horizon.** WIS +6.4/+4.1/+0.8/−0.6/−1.3 vs
  +4.2/+2.4/−0.5/−2.9/−5.2 at h−1…h3, with similar coverage bias
  (0.054–0.026 vs 0.047–0.032). The median improves too: absolute error
  +12.0/+8.6/+2.6/+2.4/+1.7 vs +6.3/+5.7/+1.9/−2.1/−4.8.
- **The gain is in the spring.** At h3, count loses 164% of April's and 92% of
  May's base WIS; sqrt loses 33% and gains 20%. An offset learned at the peak
  is sized in admissions, so in count space it is far too large for the tail.
  Both lose 43–44% in October (0.2% of the base WIS).
- **rate equals count with the adaptive learning rate.** The adaptive rate is
  a quantile of the location's own residuals, so dividing a location by its
  population rescales steps and offsets together. The rate scale matters only
  with a constant learning rate (E09).
- **2025-26 is harder for both.** sqrt loses 2.5–3.3% at h1–h3 there (count
  5.6–6.9%), against gains in 2024-25.

```{r runs}
cals <- list(
  count = run("paper"),
  sqrt = run("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3)),
  rate = run("paper", scales = rate_scales, lr_args = list(mult = 0.03, floor = 1e-3))
)
focus <- "sqrt"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
