---
title: "E04: offset structure (leak and slow term)"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: does adding a leak or a second, slow offset term to the tracker
change WIS and coverage?

Setup:

- Disease: flu. Base forecaster: the submitted hub ensemble (spoiled
  submissions dropped).
- Reference: REF-paper on the sqrt scale (E03's `sqrt`): adaptive learning
  rate of 0.03 × the 90th percentile of the last 20 rounds' absolute errors,
  floor 0.001; offsets carried across seasons; one week of extra delay
  (`settle_days = 14`).
- Burn-in: 2023-24, used only to warm up the learning rate. Offsets start at
  zero at the first 2024-25 round (cold start; no warm start for the slow
  term).
- Learning truth: exact (re-run at every round on the data published by then,
  as prod does; E00). Scores against finalized truth.
- Seasons scored: 2024-25 and 2025-26. Horizons h−1 to h3. States only
  (US excluded).

| config | change from the sqrt reference |
|---|---|
| `single` | none (one offset per level) |
| `leak 0.1` | the offset shrinks by 10% each round (half-life about 6.6 rounds) |
| `slow 0.003` | adds a second offset term with learning-rate multiplier 0.003 (vs 0.03 for the main term), no leak |
| `slow 0.003 + leak 0.1` | both; this is REF-op without its warm start (E05) |

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
cals <- list(
  single = do.call(run, sqrt_args),
  `leak 0.1` = do.call(run, c(sqrt_args, fast_decay = 0.1)),
  `slow 0.003` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)))),
  `slow 0.003 + leak 0.1` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)), fast_decay = 0.1))
)
```

```{r month-view-data}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
