---
title: "E04: offset structure (leak and slow term)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Ledger entry **E04** (`notes/calibration-ledger.md`). One axis: the
structure of the offset. Everything else is REF-paper on the sqrt scale (the
E03 sqrt variant), learning from vintages the way prod does (E00).

- **single**: the paper's tracker, one offset per level.
- **leak 0.1**: the offset shrinks by 10% each round (`fast_decay = 0.1`), so
  a correction that new outcomes do not refresh fades (half-life about 6.6
  rounds).
- **slow 0.003**: a second, persistent offset term with a small learning rate
  (`lr_slow = list(mult = 0.003)`) added to the paper's one.
- **slow 0.003 + leak 0.1**: both. This is REF-op without its warm start
  (E05).

# Findings

Numbers are from the render of 30 September 2026, pooled over both live seasons and all locations; WIS and absolute-error changes are % vs base, positive is better.

- **The leak buys WIS at h1–h3 with coverage.** `leak 0.1` moves WIS from
  +0.8/−0.6/−1.3 to +1.8/+1.0/+0.4 at h1–h3, but coverage bias nearly
  triples (0.026–0.028 → 0.075–0.082), close to the base's 0.095–0.103.
  A correction that fades between refreshes cannot hold coverage through a
  3–5 round reveal lag.
- **The leak is what repairs the tail and the second season.** At h3 it cuts
  the October loss from 44% to 11% and March from 46% to 16%, and turns April
  from −33% to +9%. In 2025-26, h1–h3 go from −2.5/−3.3/−2.7 to
  +1.7/+1.1/+0.8.
- **A slow term alone does nothing.** `slow 0.003` is within 0.3 points of
  `single` everywhere, and `slow 0.003 + leak 0.1` is within 0.5 points of
  `leak 0.1` (slightly better coverage, 0.068–0.089). At this learning rate
  the slow term never accumulates enough to matter without a warm start
  (E05).
- The median follows WIS: the leak improves absolute error at h1–h3
  (+3.7/+3.1/+3.3 vs +2.6/+2.4/+1.7) and costs about a point at h−1/h0.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
cals <- list(
  single = do.call(run, sqrt_args),
  `leak 0.1` = do.call(run, c(sqrt_args, fast_decay = 0.1)),
  `slow 0.003` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)))),
  `slow 0.003 + leak 0.1` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)), fast_decay = 0.1))
)
focus <- "leak 0.1"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
