---
title: "E05: warm start from the burn-in season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: does starting the persistent offset from the burn-in season's
errors, rather than from zero, change WIS and coverage for each tracker?

**Setup.**

- Flu, the submitted hub ensemble (`CMU-TimeSeries`), horizons h−1 to h3.
- 2023-24 is the burn-in season; 2024-25 and 2025-26 are scored. Burn-in
  outcomes never move the offsets, so a cold run plays the base forecast
  until its first live outcomes arrive.
- Square-root scale, REF-paper otherwise (adaptive learning rate, mult
  0.03, floor 0.001, window 20, offsets carried across seasons).
- Learning truth: NHSN data as published each week, re-run from scratch
  each week (prod's design, E00). Scoring: finalized truth.
- Scores cover states only (US excluded).
- The warm start sets the slow offset, per horizon and level, to the
  level-quantile of the burn-in season's errors pooled over locations, on
  the sqrt scale.

| config | warm start | tracking after the start |
|---|---|---|
| `warm only` | yes | none (learning rate ≈ 0) |
| `single` | no | REF-paper tracker |
| `warm + single` | yes | REF-paper tracker |
| `leak 0.1` | no | fast offset shrinks 10% per round |
| `warm + leak 0.1` | yes | as `leak 0.1` |
| `slow + leak` | no | leak 0.1 plus a slow offset (mult 0.003) |
| `warm + slow + leak (REF-op)` | yes | as `slow + leak`; prod's config |

WIS reduction % is `100 × (WIS_base − WIS_cal) / WIS_base`, so positive =
calibrated WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
warm <- list(slow_init = "burn_in_quantile")
cals <- list(
  `warm only` = do.call(run, c(sqrt_args[1:2], warm, list(lr_args = list(mult = 1e-6, floor = 0)))),
  single = do.call(run, sqrt_args),
  `warm + single` = do.call(run, c(sqrt_args, warm)),
  `leak 0.1` = do.call(run, c(sqrt_args, fast_decay = 0.1)),
  `warm + leak 0.1` = do.call(run, c(sqrt_args, warm, fast_decay = 0.1)),
  `slow + leak` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)), fast_decay = 0.1)),
  `warm + slow + leak (REF-op)` = run("op")
)
# The month view drops the two leak-only runs to keep it readable.
mv_plot_cfgs <- c("warm only", "single", "warm + single", "slow + leak", "warm + slow + leak (REF-op)")
```

```{r month-view-data}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
