---
title: "E05: warm start from the burn-in season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

One axis: whether the persistent offset starts from the burn-in season's
conformal offset (per horizon and level, the level-quantile of the burn-in
season's errors pooled over locations, on the sqrt scale) or from zero. Everything else is REF-paper on the sqrt scale, learning from vintages
the way prod does (E00). Each warm-started run is paired with its cold twin
from E04:

- **warm only**: the warm start, then no tracking (learning rate ≈ 0).
- **single** / **warm + single**: the paper's tracker.
- **leak 0.1** / **warm + leak 0.1**.
- **slow + leak** / **warm + slow + leak**: the second is REF-op, the
  operating point in prod.

# Findings

Numbers are from the render of 30 September 2026, pooled over both live
seasons and all locations. WIS reduction % is 100 × (base − calibrated) / base,
so positive = calibrated WIS lower than base; the absolute-error change of the
median is defined the same way. L1 coverage bias: lower is better.

- **The warm start helps every tracker, and never hurts.** Each warm run is
  at least as good as its cold twin on WIS at every horizon, and better on
  coverage: `single` +5.1/+4.0/+0.7/−1.3/−1.4 → `warm + single`
  +5.1/+4.7/+1.3/−0.2/+0.3, coverage bias 0.049–0.026 → 0.046–0.014.
- **`warm + single` is the best-calibrated config so far.** Its coverage bias
  is 0.013–0.020 at h0–h3, a fifth of the base's, with WIS within 0.2 points
  of the base or better at every horizon. REF-op (`warm + slow + leak`) has
  the best WIS (+5.4/+4.9/+2.4/+1.7/+2.2) at about three times the coverage
  bias (0.074–0.041). These two mark the ends of the WIS/coverage trade-off.
- **The warm start's gain is in the first live season.** In 2024-25 it adds
  1.7–3.2 points at h2/h3 to every tracker; in 2025-26 warm and cold runs are
  within 0.5 points, since by then the tracker has learned the level itself.
  `warm + single` still loses 1.8–4.3% at h1–h3 in 2025-26; only the leaky
  variants gain there.
- **The warm start alone** (no tracking) gains 0.5–2.1% WIS and cuts
  coverage bias by about a third at h0–h3 (12% at h−1): the persistent low
  bias is real and cheap to correct.
- **Where REF-op loses** is unchanged from the finalized-truth runs: at h3,
  October (−30%) and March (−29%), 6% of the base WIS between them;
  December–January (71% of it) gain about 5%.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
warm <- list(slow_init = "burn_in_quantile")
cals <- list(
  `warm only` = do.call(run, c(sqrt_args[1:2], warm, list(lr_args = list(mult = 1e-6, floor = 0)))),
  single = do.call(run, sqrt_args),
  `warm + single` = do.call(run, c(sqrt_args, warm)),
  `leak 0.1` = do.call(run, c(sqrt_args, fast_decay = 0.1)),
  `warm + leak 0.1` = do.call(run, c(sqrt_args, warm, fast_decay = 0.1)),
  `slow + leak` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)), fast_decay = 0.1)),
  `warm + slow + leak (REF-op)` = run("op")
)
focus <- "warm + slow + leak (REF-op)"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
