---
title: "E05: warm start from the burn-in season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: does starting the persistent offset from the burn-in season's
errors, rather than from zero, change WIS and coverage for each tracker?

**Setup.**

- Flu, the submitted hub ensemble (`CMU-TimeSeries`), horizons h−1 to h3.
- 2023-24 is the burn-in season; 2024-25 and 2025-26 are scored. Burn-in
  outcomes never move the offsets, so a cold run plays the base forecast
  until its first live outcomes arrive.
- Square-root scale, REF-paper otherwise (adaptive learning rate, mult
  0.03, floor 0.001, window 20, offsets carried across seasons).
- Learning truth: NHSN data as published each week, re-run from scratch
  each week (prod's design, E00). Scoring: finalized truth.
- Headline tables cover all hub locations; the month view and its tables
  cover states only.
- The warm start sets the slow offset, per horizon and level, to the
  level-quantile of the burn-in season's errors pooled over locations, on
  the sqrt scale.

| config | warm start | tracking after the start |
|---|---|---|
| `warm only` | yes | none (learning rate ≈ 0) |
| `single` | no | REF-paper tracker |
| `warm + single` | yes | REF-paper tracker |
| `leak 0.1` | no | fast offset shrinks 10% per round |
| `warm + leak 0.1` | yes | as `leak 0.1` |
| `slow + leak` | no | leak 0.1 plus a slow offset (mult 0.003) |
| `warm + slow + leak (REF-op)` | yes | as `slow + leak`; prod's config |

WIS reduction % is `100 × (WIS_base − WIS_cal) / WIS_base`, so positive =
calibrated WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
warm <- list(slow_init = "burn_in_quantile")
cals <- list(
  `warm only` = do.call(run, c(sqrt_args[1:2], warm, list(lr_args = list(mult = 1e-6, floor = 0)))),
  single = do.call(run, sqrt_args),
  `warm + single` = do.call(run, c(sqrt_args, warm)),
  `leak 0.1` = do.call(run, c(sqrt_args, fast_decay = 0.1)),
  `warm + leak 0.1` = do.call(run, c(sqrt_args, warm, fast_decay = 0.1)),
  `slow + leak` = do.call(run, c(sqrt_args, list(lr_slow = list(mult = 0.003)), fast_decay = 0.1)),
  `warm + slow + leak (REF-op)` = run("op")
)
focus <- "warm + slow + leak (REF-op)"
# The month view drops the two leak-only runs to keep it readable.
mv_plot_cfgs <- c("warm only", "single", "warm + single", "slow + leak", "warm + slow + leak (REF-op)")
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```

# Findings

WIS reduction % (positive = calibrated WIS lower than base) differences are
in points. L1 coverage bias (lower is better) leads with the month-avg,
states only.

- **Warm start, WIS: depends on the season.** In 2024-25 it adds 1.7–3.2
  points at h2/h3 and 0.1–1.2 at h−1…h1 to every tracker. In 2025-26 warm
  and cold differ by −0.4 to +0.4 at every horizon: no clear difference.
  Part of the 2024-25 gain is against cold runs that play the base
  forecast for the first live rounds.
- **Warm start, coverage.** In 2024-25 month-avg bias is 0.007–0.012 lower
  for `warm + single` than `single`, and 0.011–0.017 lower for REF-op than
  `slow + leak`, at h0–h3. In 2025-26 the pairs are within 0.008 in either
  direction (`warm + single` is 0.002–0.004 higher than `single`): no clear
  difference. Season-pooled bias makes `warm + single` look lowest
  (0.013–0.020 at h0–h3, all locations), but pooling lets ramp
  under-prediction cancel post-peak over-prediction.
- **Leak vs plain tracker.** In 2025-26 at h1–h3, `single` is
  −1.8/−4.6/−3.5 and `leak 0.1` +1.5/+1.0/+0.9, 3.3–5.6 points apart.
  In 2024-25 `single` is ahead by 1.3 points at h−1 and by under 1 point
  elsewhere. Month-avg bias, both seasons: within 0.007 at every horizon.
- **Slow term (REF-op vs `warm + leak 0.1`).** No strong result: WIS within
  0.4 points at every horizon (`warm + leak 0.1` ahead at h2/h3),
  month-avg bias within 0.003.
- **Warm start alone.** Pooled WIS +0.5 to +2.1 (2024-25 +0.7 to +3.5,
  2025-26 +0.1 to +1.3). Month-avg bias 0.007–0.017 below the base at
  h−1…h3.
- **By month, h3, both seasons.** Every config loses in October (−12 to
  −49%) and March (−11 to −51%), months holding 0.3% and 5.4% of the base
  WIS; December–January (71% of it) gain 1–6%.
