---
title: "E07: the submitted ensemble vs one forecaster"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

The hub-era results (E00–E05, E09) calibrate our submitted FluSight
forecasts: an ensemble whose members changed between seasons. **Are the
gains an artifact of that ensemble, or does the tracker help a single,
fixed forecaster the same way?** The fixed forecaster is flu prod's
`windowed_seasonal`.

This rebuilds the NHSN half of the original E07
(`scripts/calibration/calibration_ws_experiments.R`, finalized truth, the
2026-09-17 replay with prod's data substitutions). Its third setup, C (the
ILI+ decade), was invalid and is not rebuilt.

**Setup.** Two forecasters, everything else shared:

- **A**: the submitted ensemble, read from the hub as in E00 and E03–E05.
- **B**: the clean `windowed_seasonal` replay (E12–E17), with the 2023-24
  HHS backfill (E13) as its burn-in season.

Both are cut to the same (round, location, horizon) rows: the hub's
submission rounds, all 53 locations, h0–h3 (the ensemble's h−1 is a
different model, so it is left out), and the submissions excluded from the
hub reads are dropped from B too. Both learn from and are scored on the
same NHSN data, as published each week and re-run from scratch each week,
one week of extra revision delay, with 2023-24 as a burn-in.

The trackers are the original's three, all on the square-root scale with
the adaptive learning rate (mult 0.03, floor 0.001, window 20, carry):

- **single**: the paper's tracker (REF-paper at sqrt).
- **leaky**: single with the offset shrinking 10% per round (`fast_decay`).
- **operating**: REF-op, prod's tracker: leaky plus a slow term and a warm
  start from the burn-in season's errors.

"WIS change %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so **positive
is better**. "L1 coverage bias" is the mean over the 23 levels of
|empirical coverage − level|. Results lead with states only.

# Findings

Numbers are from the render of 1 October 2026, states only, WIS change %
(positive is better) / L1 coverage bias, h0/h1/h2/h3. All-locations numbers
point the same way.

- **The two bases are equally miscalibrated in 2024-25** (coverage bias
  0.13–0.17, 68–77% of outcomes above the median for both). In 2025-26 the
  ensemble is better calibrated (0.046–0.078 vs 0.081–0.090). Base WIS is
  similar.
- **Calibration does the same thing to both.** The trackers rank the same
  way on each: operating gains the most WIS, single gives the best coverage
  and loses at h2–h3. Pooled over both seasons:

  | tracker | A ensemble | B `windowed_seasonal` |
  |---|---|---|
  | single | +2.3/0.0/−1.3/−1.8 | +2.0/0.0/−0.6/−1.0 |
  | leaky | +2.2/+0.8/0.0/−0.3 | +1.7/+0.6/+0.4/+0.2 |
  | operating | +3.9/+1.9/+1.5/+2.3 | +2.8/+1.1/+0.8/+1.2 |

- **The ensemble gains about 1 point more with REF-op**, mostly at h2–h3 in
  2024-25 (+2.8/+4.3 vs +1.4/+1.7) and h0 in 2025-26 (+4.3 vs +2.2). The
  single tracker loses more on the ensemble in 2025-26 (h2 −3.1 vs −0.7),
  where the ensemble was already close to calibrated.
- **Answer: the gains are not an artifact of the ensemble.** On the same
  rows and data, one fixed forecaster gains the same way, a little less.
- **Against the old E07** (finalized truth, the replay with prod's data
  substitutions, 50 locations, h−1 included): the verdict holds. The old
  explanation for B's smaller gains does not: at h0–h3 the gap is about a
  point, not up to 11 (that was h−1, a different model), and B is not less
  miscalibrated than the ensemble (equal in 2024-25, worse in 2025-26).
  B single still has much the best 2025-26 coverage (0.010–0.013).

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
hub_fc <- suppressMessages(hub_read_forecasts(hub_dir = here::here(params$hub_dir), target = HUB_FLU_TARGET)) %>%
  filter(horizon %in% 0:3)
ws_fc <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Keep only (round, location, horizon, level) rows both forecasters have.
key <- c("reference_date", "location", "horizon", "level")
common <- inner_join(distinct(hub_fc, across(all_of(key))), distinct(ws_fc, across(all_of(key))), by = key)
FCS <- list(
  `A ensemble` = semi_join(hub_fc, common, by = key),
  `B windowed_seasonal` = semi_join(ws_fc, common, by = key)
)
sqrt_single <- list(ref = "paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3), burn_in_seasons = "2023-2024")
TRACKERS <- list(
  single = sqrt_single,
  leaky = c(sqrt_single, list(fast_decay = 0.1)),
  operating = list(ref = "op")
)
grid <- tidyr::expand_grid(forecaster = names(FCS), tracker = names(TRACKERS))
cals <- purrr::pmap(grid, function(forecaster, tracker) {
  do.call(cal_run_cached, c(list(FCS[[forecaster]], wi$truth), TRACKERS[[tracker]], list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
```

Rows per forecaster after the cut, by season:

```{r rows}
purrr::imap(FCS, \(fc, nm) fc %>% mutate(season = season_of(reference_date)) %>% count(season, name = nm)) %>%
  purrr::reduce(full_join, by = "season") %>%
  left_join(FCS[[1]] %>% mutate(season = season_of(reference_date)) %>% group_by(season) %>% summarize(rounds = n_distinct(reference_date)), by = "season") %>%
  knitr::kable()
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90, above_median) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
m <- purrr::map2(cals, seq_len(nrow(grid)), \(cal, i) score(cal) %>% mutate(grid[i, ])) %>%
  bind_rows() %>%
  mutate(
    forecaster = factor(forecaster, levels = names(FCS)), tracker = factor(tracker, levels = names(TRACKERS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
FC_COLS <- c(`A ensemble` = "#0072B2", `B windowed_seasonal` = "#D55E00")
```

# 1. The two bases

How miscalibrated each forecaster is before calibration, states only: WIS,
L1 coverage bias, and the share of outcomes above the median (0.5 is
calibrated; above 0.5 means the forecaster under-predicts).

```{r bases}
m %>%
  filter(geos == GEOS[["states"]], tracker == "single") %>%
  transmute(forecaster, season, h, cell = sprintf("%.1f / %.3f / %.2f", wis_base, cal_err_base, above_median)) %>%
  tidyr::pivot_wider(names_from = h, values_from = cell) %>%
  arrange(season, forecaster) %>%
  knitr::kable(caption = "Base WIS / L1 coverage bias / share above the median, states only")
```

# 2. What calibration does to each

WIS change % per tracker, one panel per season and horizon, states only.

```{r head-plot, fig.height=6}
ggplot(m %>% filter(geos == GEOS[["states"]]), aes(tracker, wis_pct, fill = forecaster)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = FC_COLS, name = NULL) +
  labs(x = NULL, y = "WIS change % (positive is better)") +
  theme(legend.position = "bottom")
```

L1 coverage bias, calibrated (filled) against the base (open), states only.

```{r cov-plot, fig.height=6}
d <- m %>% filter(geos == GEOS[["states"]])
ggplot(d, aes(tracker, cal_err_cal, colour = forecaster)) +
  geom_point(aes(y = cal_err_base), shape = 1, size = 2.4, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(season ~ h) +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "L1 coverage bias (open: base)") +
  theme(legend.position = "bottom")
```

WIS change % / L1 coverage bias:

```{r tables}
season_table <- function(g) {
  bind_rows(
    m %>% filter(tracker == "single") %>% mutate(tracker = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    m
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(paste(forecaster, tracker, sep = ", "), levels = paste(rep(names(FCS), each = 4), c("base", names(TRACKERS)), sep = ", "))) %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

## States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

## All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```
