---
title: "E07: the submitted ensemble vs one forecaster"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: does calibration change WIS and coverage the same way for the
submitted ensemble as for one fixed forecaster?

**Setup.** Two forecasters, everything else shared:

- **A**: the submitted flu ensemble, read from the hub as in E00 and
  E03–E05.
- **B**: the clean `windowed_seasonal` replay (E12–E17), with the 2023-24
  HHS backfill (E13) as its burn-in season.

| | |
|---|---|
| rows | identical (round, location, horizon, level) rows for A and B: hub submission rounds, submissions excluded from the hub reads dropped from both |
| seasons | 2023-24 burn-in (27 rounds), 2024-25 (27) and 2025-26 (28) scored |
| locations | all 53; results lead with states only |
| horizons | h0–h3 (the ensemble's h−1 is a different model) |
| learning truth | NHSN as published each week, re-run from scratch each week, one week of extra revision delay |
| scoring truth | the store's NHSN truth |

Trackers (section 2), all sqrt scale, adaptive learning rate (mult 0.03,
floor 0.001, window 20, carry):

- **single**: REF-paper at sqrt.
- **leaky**: single with the fast offset shrinking 10% per round.
- **operating**: REF-op (leaky plus a slow offset and a warm start from the
  burn-in season's errors).

Candidates (section 3): REF-op warm (= operating) and cold; constant rate
0.018 on the sqrt of the per-100k rate (E16); constant 0.032 per 100k on
the rate scale (E14); sqrt 0.018 switched off from 1 March
(`off_after = "03-01"`, E17). The cold candidates see the live seasons
only and start from zero offsets.

This rebuilds the original E07 (a script since deleted),
which learned from finalized data, used the replay with prod's data
substitutions and included h−1.

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so
**positive = calibrated WIS lower than base**. "L1 coverage bias" (lower is
better) is the mean over the 23 levels of |empirical coverage − level|.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
hub_fc <- suppressMessages(hub_read_forecasts(hub_dir = here::here(params$hub_dir), target = HUB_FLU_TARGET)) %>%
  filter(horizon %in% 0:3)
ws_fc <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Keep only (round, location, horizon, level) rows both forecasters have.
key <- c("reference_date", "location", "horizon", "level")
common <- inner_join(distinct(hub_fc, across(all_of(key))), distinct(ws_fc, across(all_of(key))), by = key)
FCS <- list(
  `A ensemble` = semi_join(hub_fc, common, by = key),
  `B windowed_seasonal` = semi_join(ws_fc, common, by = key)
)
sqrt_single <- list(ref = "paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3), burn_in_seasons = "2023-2024")
TRACKERS <- list(
  single = sqrt_single,
  leaky = c(sqrt_single, list(fast_decay = 0.1)),
  operating = list(ref = "op")
)
grid <- tidyr::expand_grid(forecaster = names(FCS), tracker = names(TRACKERS))
cals <- purrr::pmap(grid, function(forecaster, tracker) {
  do.call(cal_run_cached, c(list(FCS[[forecaster]], wi$truth), TRACKERS[[tracker]], list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
# Operating-point candidates (E14, E16, E17). Cold ones see only the live seasons, so they start from zero offsets.
const_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
sqrt_018 <- utils::modifyList(const_cfg, list(lr = 0.018, transform = "sqrt"))
CANDS <- list(
  `REF-op warm` = list(cfg = list(ref = "op"), warm = TRUE),
  `REF-op cold` = list(cfg = list(ref = "op", burn_in_seasons = character(0), slow_init = NULL), warm = FALSE),
  `sqrt 0.018` = list(cfg = sqrt_018, warm = FALSE),
  `rate 0.032` = list(cfg = utils::modifyList(const_cfg, list(lr = 0.032)), warm = FALSE),
  `sqrt 0.018, off 1 Mar` = list(cfg = c(sqrt_018, list(off_after = "03-01")), warm = FALSE)
)
FCS_LIVE <- purrr::map(FCS, \(fc) filter(fc, season_of(reference_date) %in% LIVE))
cgrid <- tidyr::expand_grid(forecaster = names(FCS), cand = names(CANDS))
ccals <- purrr::pmap(cgrid, function(forecaster, cand) {
  fc <- if (CANDS[[cand]]$warm) FCS[[forecaster]] else FCS_LIVE[[forecaster]]
  do.call(cal_run_cached, c(list(fc, wi$truth), CANDS[[cand]]$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
```

Rows per forecaster after the cut, by season:

```{r rows}
purrr::imap(FCS, \(fc, nm) fc %>% mutate(season = season_of(reference_date)) %>% count(season, name = nm)) %>%
  purrr::reduce(full_join, by = "season") %>%
  left_join(FCS[[1]] %>% mutate(season = season_of(reference_date)) %>% group_by(season) %>% summarize(rounds = n_distinct(reference_date)), by = "season") %>%
  knitr::kable()
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90, above_median) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
m <- purrr::map2(cals, seq_len(nrow(grid)), \(cal, i) score(cal) %>% mutate(grid[i, ])) %>%
  bind_rows() %>%
  mutate(
    forecaster = factor(forecaster, levels = names(FCS)), tracker = factor(tracker, levels = names(TRACKERS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
FC_COLS <- c(`A ensemble` = "#0072B2", `B windowed_seasonal` = "#D55E00")
```

# 1. The two bases

## Base WIS

Uncalibrated WIS (2 × mean pinball, in admissions) on identical rows, and
the ratio B/A (below 1: `windowed_seasonal` is better).

```{r base-wis}
m %>%
  filter(tracker == "single") %>%
  select(geos, season, h, forecaster, wis_base) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = wis_base) %>%
  transmute(geos, season, h, A = .data[[names(FCS)[1]]], B = .data[[names(FCS)[2]]], `B/A` = B / A) %>%
  arrange(geos, season, h) %>%
  knitr::kable(digits = c(0, 0, 0, 1, 1, 3), caption = "Base WIS, A ensemble vs B windowed_seasonal")
```

Per state: base WIS pooled over h0–h3, A against B (log scales; below the
line, B is better).

```{r base-wis-state, fig.height=5}
base_fc <- purrr::map(which(grid$tracker == "single"), \(i) cals[[i]]$forecasts %>% mutate(forecaster = grid$forecaster[i])) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date)) %>%
  filter(season %in% LIVE, !is.na(truth), !is.na(value_base))
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
loc_wis <- base_fc %>%
  filter(location != "US") %>%
  group_by(forecaster, season, location) %>%
  summarize(wis = 2 * mean(pinball(truth, value_base, level)), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = wis) %>%
  transmute(season, location, A = .data[[names(FCS)[1]]], B = .data[[names(FCS)[2]]], ratio = B / A)
ggplot(loc_wis, aes(A, B)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  geom_point(alpha = 0.8) +
  facet_wrap(~season) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "A ensemble base WIS", y = "B windowed_seasonal base WIS")
```

```{r base-wis-ratio}
loc_wis %>%
  group_by(season) %>%
  summarize(
    states = n(), `B better` = sum(ratio < 1), q10 = quantile(ratio, 0.1), q25 = quantile(ratio, 0.25),
    median = median(ratio), q75 = quantile(ratio, 0.75), q90 = quantile(ratio, 0.9)
  ) %>%
  knitr::kable(digits = 2, caption = "Per-state B/A base WIS ratio, h0–h3 pooled")
```

## Base miscalibration

How miscalibrated each forecaster is before calibration, states only: WIS,
L1 coverage bias, and the share of outcomes above the median (0.5 is
calibrated; above 0.5 means the forecaster under-predicts).

```{r bases}
m %>%
  filter(geos == GEOS[["states"]], tracker == "single") %>%
  transmute(forecaster, season, h, cell = sprintf("%.1f / %.3f / %.2f", wis_base, cal_err_base, above_median)) %>%
  tidyr::pivot_wider(names_from = h, values_from = cell) %>%
  arrange(season, forecaster) %>%
  knitr::kable(caption = "Base WIS / L1 coverage bias (lower is better) / share above the median, states only")
```

## Base bias by month

Share of outcomes above the base median and L1 coverage bias, by calendar
month of the reference date, states only.

```{r base-month, fig.height=7}
month_base <- function(d) {
  d %>%
    group_by(across(c(any_of(c("forecaster", "season", "horizon", "month")), level))) %>%
    summarize(gap = mean(truth <= value_base) - first(level), above = mean(truth > value_base), n = n(), .groups = "drop") %>%
    group_by(across(any_of(c("forecaster", "season", "horizon", "month")))) %>%
    summarize(above = above[level == 0.5], cal_err = mean(abs(gap)), n = first(n), .groups = "drop")
}
base_fc_states <- base_fc %>%
  filter(location != "US") %>%
  mutate(month = factor(format(reference_date, "%b"), levels = month.abb[c(8:12, 1:7)]))
mb <- month_base(base_fc_states) %>% mutate(h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
mb %>%
  tidyr::pivot_longer(c(above, cal_err), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("above", "cal_err"), labels = c("share above median", "L1 coverage bias (lower is better)"))) %>%
  ggplot(aes(month, value, colour = forecaster, group = forecaster)) +
  geom_hline(data = data.frame(metric = factor("share above median", levels = c("share above median", "L1 coverage bias (lower is better)")), y = 0.5), aes(yintercept = y), colour = "grey50") +
  geom_line() +
  geom_point(size = 1.1) +
  facet_grid(metric + season ~ h, scales = "free", space = "free_x", labeller = labeller(metric = label_wrap_gen(20))) +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  labs(x = "month of reference date", y = NULL) +
  theme(legend.position = "bottom")
```

Pooled over h0–h3: share above the base median / L1 coverage bias (lower is better), A vs B.

```{r base-month-table}
month_base(base_fc_states %>% select(-horizon)) %>%
  transmute(season, month, forecaster, cell = sprintf("%.2f / %.3f", above, cal_err)) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = cell) %>%
  left_join(base_fc_states %>% filter(level == 0.5) %>% count(season, month, forecaster) %>% filter(forecaster == names(FCS)[1]) %>% select(season, month, forecasts = n), by = c("season", "month")) %>%
  arrange(season, month) %>%
  knitr::kable(caption = "Share above base median / L1 coverage bias (lower is better), states only, h0–h3 pooled")
```

# 2. What calibration does to each

WIS reduction % (positive = calibrated WIS lower than base) per tracker, one panel per season and horizon, states only.

```{r head-plot, fig.height=6}
ggplot(m %>% filter(geos == GEOS[["states"]]), aes(tracker, wis_pct, fill = forecaster)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = FC_COLS, name = NULL) +
  labs(x = NULL, y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom")
```

L1 coverage bias (lower is better), calibrated (filled) against the base (open), states only.

```{r cov-plot, fig.height=6}
d <- m %>% filter(geos == GEOS[["states"]])
ggplot(d, aes(tracker, cal_err_cal, colour = forecaster)) +
  geom_point(aes(y = cal_err_base), shape = 1, size = 2.4, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(season ~ h) +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "L1 coverage bias (lower is better; open: base)") +
  theme(legend.position = "bottom")
```

WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage bias (lower is better):

```{r tables}
season_table <- function(g) {
  bind_rows(
    m %>% filter(tracker == "single") %>% mutate(tracker = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    m
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(paste(forecaster, tracker, sep = ", "), levels = paste(rep(names(FCS), each = 4), c("base", names(TRACKERS)), sep = ", "))) %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

## States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r table-states-month}
mv_trk <- cal_month_view(setNames(cals, paste(grid$forecaster, grid$tracker, sep = ", ")), seasons = LIVE)
knitr::kable(cal_month_cov_table(mv_trk), caption = CAL_MONTH_COV_CAPTION)
```

## All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```

# 3. Operating-point candidates on both

The candidates from E14, E16 and E17, each run on both forecasters on the
same rows: REF-op warm (as in section 2) and cold, sqrt-scale constant
0.018 (E16), rate-scale constant 0.032 per 100k (E14), and sqrt 0.018
switched off from 1 March (`off_after = "03-01"`, E17). The cold ones learn
from the live seasons only.

```{r cand-metrics}
cm <- purrr::map2(ccals, seq_len(nrow(cgrid)), \(cal, i) score(cal) %>% mutate(cgrid[i, ])) %>%
  bind_rows() %>%
  mutate(
    forecaster = factor(forecaster, levels = names(FCS)), cand = factor(cand, levels = names(CANDS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
```

WIS reduction % (positive = calibrated WIS lower than base) per candidate, states only.

```{r cand-plot, fig.height=6.5}
ggplot(cm %>% filter(geos == GEOS[["states"]]), aes(cand, wis_pct, fill = forecaster)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = FC_COLS, name = NULL) +
  labs(x = NULL, y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

L1 coverage bias (lower is better), calibrated (filled) against the base (open), states only.

```{r cand-cov, fig.height=6.5}
ggplot(cm %>% filter(geos == GEOS[["states"]]), aes(cand, cal_err_cal, colour = forecaster)) +
  geom_point(aes(y = cal_err_base), shape = 1, size = 2.4, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(season ~ h) +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "L1 coverage bias (lower is better; open: base)") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

## Rankings on A and B

Candidates ordered by mean WIS reduction % (positive = calibrated WIS lower than base) over h0–h3, separately on each
forecaster, states only. `rank A` / `rank B` are the positions (1 is the
largest WIS reduction; `cov rank` 1 is the lowest L1 coverage bias); Spearman ρ compares the two orders per season, over all
(candidate, horizon) cells.

```{r cand-rank}
rk <- cm %>%
  filter(geos == GEOS[["states"]]) %>%
  group_by(forecaster, season, cand) %>%
  summarize(wis = mean(wis_pct), cov = mean(cal_err_cal), .groups = "drop") %>%
  group_by(forecaster, season) %>%
  mutate(rank = rank(-wis), cov_rank = rank(cov)) %>%
  ungroup() %>%
  mutate(fc = ifelse(forecaster == names(FCS)[1], "A", "B")) %>%
  select(season, cand, fc, wis, rank, cov, cov_rank) %>%
  tidyr::pivot_wider(names_from = fc, values_from = c(wis, rank, cov, cov_rank)) %>%
  arrange(season, rank_A)
rk %>%
  transmute(season, cand, `WIS reduction % A` = wis_A, `rank A` = rank_A, `WIS reduction % B` = wis_B, `rank B` = rank_B,
            `cov bias A` = cov_A, `cov rank A` = cov_rank_A, `cov bias B` = cov_B, `cov rank B` = cov_rank_B) %>%
  knitr::kable(digits = c(0, 0, 2, 0, 2, 0, 3, 0, 3, 0), caption = "Mean over h0–h3, states only. WIS reduction %: positive = calibrated WIS lower than base. Coverage bias is L1 coverage bias (lower is better)")
cm %>%
  filter(geos == GEOS[["states"]]) %>%
  select(season, cand, h, forecaster, wis_pct, cal_err_cal) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = c(wis_pct, cal_err_cal)) %>%
  group_by(season) %>%
  summarize(
    `ρ WIS cells` = cor(.data[[paste0("wis_pct_", names(FCS)[1])]], .data[[paste0("wis_pct_", names(FCS)[2])]], method = "spearman"),
    `ρ coverage cells` = cor(.data[[paste0("cal_err_cal_", names(FCS)[1])]], .data[[paste0("cal_err_cal_", names(FCS)[2])]], method = "spearman")
  ) %>%
  knitr::kable(digits = 2, caption = "Spearman correlation of A vs B over (candidate, horizon) cells")
```

## Tables

WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage bias (lower is better).

```{r cand-tables}
cand_table <- function(g) {
  bind_rows(
    cm %>% filter(cand == names(CANDS)[1]) %>% mutate(cand = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    cm
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(paste(forecaster, cand, sep = ", "), levels = paste(rep(names(FCS), each = length(CANDS) + 1), c("base", names(CANDS)), sep = ", "))) %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

### States only

```{r cand-table-states}
knitr::kable(cand_table(GEOS[["states"]]))
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r cand-table-states-month}
mv <- cal_month_view(setNames(ccals, paste(cgrid$forecaster, cgrid$cand, sep = ", ")), seasons = LIVE)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

### All locations

```{r cand-table-all}
knitr::kable(cand_table(GEOS[["all"]]))
```

```{r month-view-cfgs}
# Three candidates on both forecasters: colour is the candidate, linetype the forecaster (also for the grey base lines).
MV_CANDS <- c(`REF-op warm` = "black", `sqrt 0.018` = "#0072B2", `sqrt 0.018, off 1 Mar` = "#D55E00")
mv_plot_cfgs <- paste(rep(names(FCS), each = length(MV_CANDS)), names(MV_CANDS), sep = ", ")
mv_cols <- MV_CANDS
mv$month <- mv$month %>% mutate(forecaster = sub(",.*", "", config), cand = factor(sub("^[^,]*, ", "", config), levels = names(CANDS)))
mv_colour_by <- "cand"
mv_linetype <- "forecaster"
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Findings

States only. WIS reduction % (positive = calibrated WIS lower than base)
differences are in points; L1 coverage bias is lower-is-better and leads
with the month-avg.

- **Bases.** B/A base WIS is 0.93–1.01 in 2024-25 and 1.05–1.33 in 2025-26
  (largest at h0); per state B is lower in 29 of 52 states, then 7 of 52.
  Month-avg base bias over h0–h3 is 0.116–0.151 for A and 0.112–0.177 for
  B (both seasons), 0.089–0.117 vs 0.097–0.146 in 2025-26. Both bases have
  over 86% of outcomes above the median in December and under 40% in March
  2025-26. They differ in November 2024-25 (51% vs 73% above), March–April
  2024-25 (59/75% vs 45/56%) and February 2025-26 (60% vs 71%).
- **Trackers.** On both, pooled WIS at h1–h3 orders operating > leaky >
  single. Operating gains 0.7–1.1 points more on A than on B (A
  +3.9/+1.9/+1.5/+2.3, B +2.8/+1.1/+0.8/+1.2), mostly 2024-25 h2/h3
  (+2.8/+4.3 vs +1.4/+1.7) and 2025-26 h0 (+4.3 vs +2.2). On B, leaky and
  single are under 1 point apart except at h3 (1.2).
- **Candidates, WIS: the order flips between seasons.** Mean over h0–h3,
  REF-op warm leads in 2024-25 on both (A +3.44, B +2.07; next +0.79 and
  +0.90). In 2025-26 the 1 March switch-off leads on both (A +1.96, B
  +2.37) and REF-op warm is 4th on A, 5th on B (+0.69, +0.60). Pooled over
  both seasons the other four candidates are within 0.2 points of each
  other on A (+0.64 to +0.83) and within 0.9 on B (+0.63 to +1.48).
- **1 March switch-off in 2024-25.** No strong result: 0.47 points below
  sqrt 0.018 on A and 0.24 above it on B. The rates (0.018, 0.032) and the
  1 March date were picked in E14/E16/E17 on these same seasons, on B.
- **Coverage, month-avg.** No strong result among candidates: over h0–h3,
  A ranges 0.115–0.126 for four candidates (base 0.132) and B 0.130–0.136
  (base 0.142). Rate 0.032 is no better than the base on either (A 0.133,
  B 0.142), and in 2025-26 it is above the base at every horizon (A by
  0.016–0.024, B by 0.007–0.017). Season-pooled, rate 0.032 has the lowest
  2025-26 bias on B (0.010–0.014), because pooling lets ramp and post-peak
  errors cancel. Season-pooled coverage ranks agree between A and B over
  both seasons (Spearman 0.97) but not in 2025-26 (0.31).
