---
title: "E07: the submitted ensemble vs one forecaster"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

The hub-era results (E00–E05, E09) calibrate our submitted FluSight
forecasts: an ensemble whose members changed between seasons. **Are the
gains an artifact of that ensemble, or does the tracker help a single,
fixed forecaster the same way?** The fixed forecaster is flu prod's
`windowed_seasonal`.

This asks the NHSN half of the original E07's question
(`scripts/calibration/calibration_ws_experiments.R`) on the current setup.
It is a new experiment, not a reproduction: the original learned from
finalized data and replayed `windowed_seasonal` with prod's data
substitutions; here both learn from data as published each week, and B is
the clean replay. The original's third setup, C (the ILI+ decade), was
invalid and is not repeated.

**Setup.** Two forecasters, everything else shared:

- **A**: the submitted ensemble, read from the hub as in E00 and E03–E05.
- **B**: the clean `windowed_seasonal` replay (E12–E17), with the 2023-24
  HHS backfill (E13) as its burn-in season.

Both are cut to the same (round, location, horizon) rows: the hub's
submission rounds, all 53 locations, h0–h3 (the ensemble's h−1 is a
different model, so it is left out), and the submissions excluded from the
hub reads are dropped from B too. Both learn from and are scored on the
same NHSN data, as published each week and re-run from scratch each week,
one week of extra revision delay, with 2023-24 as a burn-in.

The trackers are the original's three, all on the square-root scale with
the adaptive learning rate (mult 0.03, floor 0.001, window 20, carry):

- **single**: the paper's tracker (REF-paper at sqrt).
- **leaky**: single with the offset shrinking 10% per round (`fast_decay`).
- **operating**: REF-op, prod's tracker: leaky plus a slow term and a warm
  start from the burn-in season's errors.

Section 3 also runs the operating-point candidates (E14, E16, E17) on both
forecasters.

"WIS change %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so **positive
is better**. "L1 coverage bias" is the mean over the 23 levels of
|empirical coverage − level|. Results lead with states only.

# Findings

Numbers are from the render of 1 October 2026, states only, WIS change %
(positive is better) / L1 coverage bias, h0/h1/h2/h3. All-locations numbers
point the same way.

- **Base WIS: equal in 2024-25, B worse in 2025-26.** B/A base WIS is
  0.93/0.99/1.01/1.01 in 2024-25 and 1.33/1.22/1.11/1.05 in 2025-26
  (1.06/1.07/1.05/1.02 pooled). Per state, pooled over h0–h3, the median
  ratio is 0.99 in 2024-25 (B better in 29 of 52 states) and 1.12 in
  2025-26 (B better in 7 of 52). So in 2025-26 the ensemble beats B almost
  everywhere, by about 10%, most at h0–h1.
- **The two bases are equally miscalibrated in 2024-25** (coverage bias
  0.13–0.17, 68–77% of outcomes above the median for both). In 2025-26 the
  ensemble is better calibrated (0.046–0.078 vs 0.081–0.090).
- **By month, both bases have the same shape.** Both under-predict on the
  ramp (December: 96–97% of outcomes above the median in 2024-25, 86–94% in
  2025-26, coverage bias 0.24–0.34) and over-predict after the peak (March
  2025-26: 32% above for A, 38% for B). B's ramp bias is larger in 2025-26
  (November–December coverage bias 0.19/0.31 vs 0.14/0.24). The one place
  they differ is spring 2024-25: A still under-predicts in March–April (59%
  and 75% above the median), while B is close to calibrated (45% and 56%).
- **The operating-point candidates rank mostly the same on both, with one
  exception.** Pooled over both seasons, h0/h1/h2/h3 WIS change %:

  | candidate | A ensemble | B `windowed_seasonal` |
  |---|---|---|
  | REF-op warm | +3.9/+1.9/+1.5/+2.3 | +2.8/+1.1/+0.8/+1.2 |
  | REF-op cold | +2.8/+0.7/−0.3/−0.7 | +2.0/+0.6/+0.1/−0.2 |
  | sqrt 0.018 | +2.4/+0.6/+0.1/0.0 | +2.2/+0.8/+0.6/+0.6 |
  | rate 0.032 | +1.5/+0.7/+0.5/+0.2 | +1.3/+0.7/+0.5/+0.4 |
  | sqrt 0.018, off 1 Mar | +1.7/+0.6/+0.4/+0.6 | +2.3/+1.2/+1.2/+1.3 |

  Ordered by the mean over h0–h3, sqrt 0.018 > rate 0.032 > REF-op cold on
  both. REF-op warm is first on A (+2.4 mean) and tied first with
  off-1-March on B (both +1.5). Coverage bias ranks identically on both,
  pooled: REF-op warm best (A 0.035–0.047, B 0.037–0.061), then rate 0.032,
  sqrt 0.018, REF-op cold, and off-1-March worst (A 0.084–0.092, B
  0.094–0.116).
- **The exception is the March switch-off, and it follows the base's spring
  bias.** In 2024-25, off-1-March is last of the five on A (+0.6/+0.2/0.0/+0.1)
  but second on B (+1.6/+0.8/+0.7/+0.5): A was still under-predicting that
  spring, so the offsets it switches off were still helping. In 2025-26,
  when both over-predict in March, it is first on both. So whether to
  switch off in spring depends on the forecaster, and B's result for it
  should not be carried over to the ensemble without checking.
- **sqrt 0.018 is WIS-positive at every horizon in both seasons only on B.**
  On A it is −0.3/−0.1 at h2/h3 in 2025-26, where the ensemble was already
  close to calibrated.
- **Calibration does the same thing to both.** The trackers rank the same
  way on each: operating gains the most WIS, single gives the best coverage
  and loses at h2–h3. Pooled over both seasons:

  | tracker | A ensemble | B `windowed_seasonal` |
  |---|---|---|
  | single | +2.3/0.0/−1.3/−1.8 | +2.0/0.0/−0.6/−1.0 |
  | leaky | +2.2/+0.8/0.0/−0.3 | +1.7/+0.6/+0.4/+0.2 |
  | operating | +3.9/+1.9/+1.5/+2.3 | +2.8/+1.1/+0.8/+1.2 |

- **The ensemble gains about 1 point more with REF-op**, mostly at h2–h3 in
  2024-25 (+2.8/+4.3 vs +1.4/+1.7) and h0 in 2025-26 (+4.3 vs +2.2). The
  single tracker loses more on the ensemble in 2025-26 (h2 −3.1 vs −0.7),
  where the ensemble was already close to calibrated.
- **Answer: the gains are not an artifact of the ensemble.** On the same
  rows and data, one fixed forecaster gains the same way, a little less.
  B's base is about 5% worse on WIS pooled and its bias has the same
  monthly shape, so experiments on B transfer to the ensemble for the
  choice of tracker. They do not transfer for date-based switches, which
  depend on where each base is biased in spring.
- **Against the old E07** (finalized truth, the replay with prod's data
  substitutions, 50 locations, h−1 included): the verdict holds. The old
  explanation for B's smaller gains does not: at h0–h3 the gap is about a
  point, not up to 11 (that was h−1, a different model), and B is not less
  miscalibrated than the ensemble (equal in 2024-25, worse in 2025-26).
  B single still has much the best 2025-26 coverage (0.010–0.013).

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
hub_fc <- suppressMessages(hub_read_forecasts(hub_dir = here::here(params$hub_dir), target = HUB_FLU_TARGET)) %>%
  filter(horizon %in% 0:3)
ws_fc <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Keep only (round, location, horizon, level) rows both forecasters have.
key <- c("reference_date", "location", "horizon", "level")
common <- inner_join(distinct(hub_fc, across(all_of(key))), distinct(ws_fc, across(all_of(key))), by = key)
FCS <- list(
  `A ensemble` = semi_join(hub_fc, common, by = key),
  `B windowed_seasonal` = semi_join(ws_fc, common, by = key)
)
sqrt_single <- list(ref = "paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3), burn_in_seasons = "2023-2024")
TRACKERS <- list(
  single = sqrt_single,
  leaky = c(sqrt_single, list(fast_decay = 0.1)),
  operating = list(ref = "op")
)
grid <- tidyr::expand_grid(forecaster = names(FCS), tracker = names(TRACKERS))
cals <- purrr::pmap(grid, function(forecaster, tracker) {
  do.call(cal_run_cached, c(list(FCS[[forecaster]], wi$truth), TRACKERS[[tracker]], list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
# Operating-point candidates (E14, E16, E17). Cold ones see only the live seasons, so they start from zero offsets.
const_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
sqrt_018 <- utils::modifyList(const_cfg, list(lr = 0.018, transform = "sqrt"))
CANDS <- list(
  `REF-op warm` = list(cfg = list(ref = "op"), warm = TRUE),
  `REF-op cold` = list(cfg = list(ref = "op", burn_in_seasons = character(0), slow_init = NULL), warm = FALSE),
  `sqrt 0.018` = list(cfg = sqrt_018, warm = FALSE),
  `rate 0.032` = list(cfg = utils::modifyList(const_cfg, list(lr = 0.032)), warm = FALSE),
  `sqrt 0.018, off 1 Mar` = list(cfg = c(sqrt_018, list(off_after = "03-01")), warm = FALSE)
)
FCS_LIVE <- purrr::map(FCS, \(fc) filter(fc, season_of(reference_date) %in% LIVE))
cgrid <- tidyr::expand_grid(forecaster = names(FCS), cand = names(CANDS))
ccals <- purrr::pmap(cgrid, function(forecaster, cand) {
  fc <- if (CANDS[[cand]]$warm) FCS[[forecaster]] else FCS_LIVE[[forecaster]]
  do.call(cal_run_cached, c(list(fc, wi$truth), CANDS[[cand]]$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
```

Rows per forecaster after the cut, by season:

```{r rows}
purrr::imap(FCS, \(fc, nm) fc %>% mutate(season = season_of(reference_date)) %>% count(season, name = nm)) %>%
  purrr::reduce(full_join, by = "season") %>%
  left_join(FCS[[1]] %>% mutate(season = season_of(reference_date)) %>% group_by(season) %>% summarize(rounds = n_distinct(reference_date)), by = "season") %>%
  knitr::kable()
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90, above_median) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
m <- purrr::map2(cals, seq_len(nrow(grid)), \(cal, i) score(cal) %>% mutate(grid[i, ])) %>%
  bind_rows() %>%
  mutate(
    forecaster = factor(forecaster, levels = names(FCS)), tracker = factor(tracker, levels = names(TRACKERS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
FC_COLS <- c(`A ensemble` = "#0072B2", `B windowed_seasonal` = "#D55E00")
```

# 1. The two bases

## Base WIS

Uncalibrated WIS (2 × mean pinball, in admissions) on identical rows, and
the ratio B/A (below 1: `windowed_seasonal` is better).

```{r base-wis}
m %>%
  filter(tracker == "single") %>%
  select(geos, season, h, forecaster, wis_base) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = wis_base) %>%
  transmute(geos, season, h, A = .data[[names(FCS)[1]]], B = .data[[names(FCS)[2]]], `B/A` = B / A) %>%
  arrange(geos, season, h) %>%
  knitr::kable(digits = c(0, 0, 0, 1, 1, 3), caption = "Base WIS, A ensemble vs B windowed_seasonal")
```

Per state: base WIS pooled over h0–h3, A against B (log scales; below the
line, B is better).

```{r base-wis-state, fig.height=5}
base_fc <- purrr::map(which(grid$tracker == "single"), \(i) cals[[i]]$forecasts %>% mutate(forecaster = grid$forecaster[i])) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date)) %>%
  filter(season %in% LIVE, !is.na(truth), !is.na(value_base))
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
loc_wis <- base_fc %>%
  filter(location != "US") %>%
  group_by(forecaster, season, location) %>%
  summarize(wis = 2 * mean(pinball(truth, value_base, level)), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = wis) %>%
  transmute(season, location, A = .data[[names(FCS)[1]]], B = .data[[names(FCS)[2]]], ratio = B / A)
ggplot(loc_wis, aes(A, B)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  geom_point(alpha = 0.8) +
  facet_wrap(~season) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "A ensemble base WIS", y = "B windowed_seasonal base WIS")
```

```{r base-wis-ratio}
loc_wis %>%
  group_by(season) %>%
  summarize(
    states = n(), `B better` = sum(ratio < 1), q10 = quantile(ratio, 0.1), q25 = quantile(ratio, 0.25),
    median = median(ratio), q75 = quantile(ratio, 0.75), q90 = quantile(ratio, 0.9)
  ) %>%
  knitr::kable(digits = 2, caption = "Per-state B/A base WIS ratio, h0–h3 pooled")
```

## Base miscalibration

How miscalibrated each forecaster is before calibration, states only: WIS,
L1 coverage bias, and the share of outcomes above the median (0.5 is
calibrated; above 0.5 means the forecaster under-predicts).

```{r bases}
m %>%
  filter(geos == GEOS[["states"]], tracker == "single") %>%
  transmute(forecaster, season, h, cell = sprintf("%.1f / %.3f / %.2f", wis_base, cal_err_base, above_median)) %>%
  tidyr::pivot_wider(names_from = h, values_from = cell) %>%
  arrange(season, forecaster) %>%
  knitr::kable(caption = "Base WIS / L1 coverage bias / share above the median, states only")
```

## Base bias by month

Share of outcomes above the base median and L1 coverage bias, by calendar
month of the reference date, states only.

```{r base-month, fig.height=7}
month_base <- function(d) {
  d %>%
    group_by(across(c(any_of(c("forecaster", "season", "horizon", "month")), level))) %>%
    summarize(gap = mean(truth <= value_base) - first(level), above = mean(truth > value_base), n = n(), .groups = "drop") %>%
    group_by(across(any_of(c("forecaster", "season", "horizon", "month")))) %>%
    summarize(above = above[level == 0.5], cal_err = mean(abs(gap)), n = first(n), .groups = "drop")
}
base_fc_states <- base_fc %>%
  filter(location != "US") %>%
  mutate(month = factor(format(reference_date, "%b"), levels = month.abb[c(8:12, 1:7)]))
mb <- month_base(base_fc_states) %>% mutate(h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
mb %>%
  tidyr::pivot_longer(c(above, cal_err), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("above", "cal_err"), labels = c("share above median", "L1 coverage bias"))) %>%
  ggplot(aes(month, value, colour = forecaster, group = forecaster)) +
  geom_hline(data = data.frame(metric = factor("share above median", levels = c("share above median", "L1 coverage bias")), y = 0.5), aes(yintercept = y), colour = "grey50") +
  geom_line() +
  geom_point(size = 1.1) +
  facet_grid(metric + season ~ h, scales = "free", space = "free_x") +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  labs(x = "month of reference date", y = NULL) +
  theme(legend.position = "bottom")
```

Pooled over h0–h3: share above the base median / L1 coverage bias, A vs B.

```{r base-month-table}
month_base(base_fc_states %>% select(-horizon)) %>%
  transmute(season, month, forecaster, cell = sprintf("%.2f / %.3f", above, cal_err)) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = cell) %>%
  left_join(base_fc_states %>% filter(level == 0.5) %>% count(season, month, forecaster) %>% filter(forecaster == names(FCS)[1]) %>% select(season, month, forecasts = n), by = c("season", "month")) %>%
  arrange(season, month) %>%
  knitr::kable(caption = "Share above base median / L1 coverage bias, states only, h0–h3 pooled")
```

# 2. What calibration does to each

WIS change % per tracker, one panel per season and horizon, states only.

```{r head-plot, fig.height=6}
ggplot(m %>% filter(geos == GEOS[["states"]]), aes(tracker, wis_pct, fill = forecaster)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = FC_COLS, name = NULL) +
  labs(x = NULL, y = "WIS change % (positive is better)") +
  theme(legend.position = "bottom")
```

L1 coverage bias, calibrated (filled) against the base (open), states only.

```{r cov-plot, fig.height=6}
d <- m %>% filter(geos == GEOS[["states"]])
ggplot(d, aes(tracker, cal_err_cal, colour = forecaster)) +
  geom_point(aes(y = cal_err_base), shape = 1, size = 2.4, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(season ~ h) +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "L1 coverage bias (open: base)") +
  theme(legend.position = "bottom")
```

WIS change % / L1 coverage bias:

```{r tables}
season_table <- function(g) {
  bind_rows(
    m %>% filter(tracker == "single") %>% mutate(tracker = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    m
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(paste(forecaster, tracker, sep = ", "), levels = paste(rep(names(FCS), each = 4), c("base", names(TRACKERS)), sep = ", "))) %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

## States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

## All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```

# 3. Operating-point candidates on both

The candidates from E14, E16 and E17, each run on both forecasters on the
same rows: REF-op warm (as in section 2) and cold, sqrt-scale constant
0.018 (E16), rate-scale constant 0.032 per 100k (E14), and sqrt 0.018
switched off from 1 March (`off_after = "03-01"`, E17). The cold ones learn
from the live seasons only.

```{r cand-metrics}
cm <- purrr::map2(ccals, seq_len(nrow(cgrid)), \(cal, i) score(cal) %>% mutate(cgrid[i, ])) %>%
  bind_rows() %>%
  mutate(
    forecaster = factor(forecaster, levels = names(FCS)), cand = factor(cand, levels = names(CANDS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
```

WIS change % per candidate, states only.

```{r cand-plot, fig.height=6.5}
ggplot(cm %>% filter(geos == GEOS[["states"]]), aes(cand, wis_pct, fill = forecaster)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ h) +
  scale_fill_manual(values = FC_COLS, name = NULL) +
  labs(x = NULL, y = "WIS change % (positive is better)") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

L1 coverage bias, calibrated (filled) against the base (open), states only.

```{r cand-cov, fig.height=6.5}
ggplot(cm %>% filter(geos == GEOS[["states"]]), aes(cand, cal_err_cal, colour = forecaster)) +
  geom_point(aes(y = cal_err_base), shape = 1, size = 2.4, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(season ~ h) +
  scale_colour_manual(values = FC_COLS, name = NULL) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "L1 coverage bias (open: base)") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

## Rankings on A and B

Candidates ordered by mean WIS change % over h0–h3, separately on each
forecaster, states only. `rank A` / `rank B` are the positions (1 is the
largest gain); Spearman ρ compares the two orders per season, over all
(candidate, horizon) cells.

```{r cand-rank}
rk <- cm %>%
  filter(geos == GEOS[["states"]]) %>%
  group_by(forecaster, season, cand) %>%
  summarize(wis = mean(wis_pct), cov = mean(cal_err_cal), .groups = "drop") %>%
  group_by(forecaster, season) %>%
  mutate(rank = rank(-wis), cov_rank = rank(cov)) %>%
  ungroup() %>%
  mutate(fc = ifelse(forecaster == names(FCS)[1], "A", "B")) %>%
  select(season, cand, fc, wis, rank, cov, cov_rank) %>%
  tidyr::pivot_wider(names_from = fc, values_from = c(wis, rank, cov, cov_rank)) %>%
  arrange(season, rank_A)
rk %>%
  transmute(season, cand, `WIS % A` = wis_A, `rank A` = rank_A, `WIS % B` = wis_B, `rank B` = rank_B,
            `cov bias A` = cov_A, `cov rank A` = cov_rank_A, `cov bias B` = cov_B, `cov rank B` = cov_rank_B) %>%
  knitr::kable(digits = c(0, 0, 2, 0, 2, 0, 3, 0, 3, 0), caption = "Mean over h0–h3, states only")
cm %>%
  filter(geos == GEOS[["states"]]) %>%
  select(season, cand, h, forecaster, wis_pct, cal_err_cal) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = c(wis_pct, cal_err_cal)) %>%
  group_by(season) %>%
  summarize(
    `ρ WIS cells` = cor(.data[[paste0("wis_pct_", names(FCS)[1])]], .data[[paste0("wis_pct_", names(FCS)[2])]], method = "spearman"),
    `ρ coverage cells` = cor(.data[[paste0("cal_err_cal_", names(FCS)[1])]], .data[[paste0("cal_err_cal_", names(FCS)[2])]], method = "spearman")
  ) %>%
  knitr::kable(digits = 2, caption = "Spearman correlation of A vs B over (candidate, horizon) cells")
```

## Tables

WIS change % / L1 coverage bias.

```{r cand-tables}
cand_table <- function(g) {
  bind_rows(
    cm %>% filter(cand == names(CANDS)[1]) %>% mutate(cand = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    cm
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(paste(forecaster, cand, sep = ", "), levels = paste(rep(names(FCS), each = length(CANDS) + 1), c("base", names(CANDS)), sep = ", "))) %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

### States only

```{r cand-table-states}
knitr::kable(cand_table(GEOS[["states"]]))
```

### All locations

```{r cand-table-all}
knitr::kable(cand_table(GEOS[["all"]]))
```
