---
title: "E07b: operating-point candidates on the ensemble and the replay"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: do the operating-point candidates rank the same way on the
submitted ensemble as on one fixed forecaster?

**Setup.** As in E07: A is the submitted flu ensemble,
B the clean `windowed_seasonal` replay with the 2023-24 HHS backfill as its
burn-in season, on identical (round, location, horizon, level) rows, h0–h3,
exact learning truth with one week of extra delay.

Candidates: REF-op warm and cold; constant rate 0.018 on the sqrt of the
per-100k rate (E16); constant 0.032 per 100k on the rate scale (E14); sqrt
0.018 switched off from 1 March (`off_after = "03-01"`, E17). The cold
candidates see the live seasons only and start from zero offsets.

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so
**positive = calibrated WIS lower than base**. "L1 coverage bias" (lower is
better) is the mean over the 23 levels of |empirical coverage − level|.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
hub_fc <- suppressMessages(hub_read_forecasts(hub_dir = here::here(params$hub_dir), target = HUB_FLU_TARGET)) %>%
  filter(horizon %in% 0:3)
ws_fc <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Keep only (round, location, horizon, level) rows both forecasters have.
key <- c("reference_date", "location", "horizon", "level")
common <- inner_join(distinct(hub_fc, across(all_of(key))), distinct(ws_fc, across(all_of(key))), by = key)
FCS <- list(
  `A ensemble` = semi_join(hub_fc, common, by = key),
  `B windowed_seasonal` = semi_join(ws_fc, common, by = key)
)
# Cold candidates see only the live seasons, so they start from zero offsets.
const_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
sqrt_018 <- utils::modifyList(const_cfg, list(lr = 0.018, transform = "sqrt"))
CANDS <- list(
  `REF-op warm` = list(cfg = list(ref = "op"), warm = TRUE),
  `REF-op cold` = list(cfg = list(ref = "op", burn_in_seasons = character(0), slow_init = NULL), warm = FALSE),
  `sqrt 0.018` = list(cfg = sqrt_018, warm = FALSE),
  `rate 0.032` = list(cfg = utils::modifyList(const_cfg, list(lr = 0.032)), warm = FALSE),
  `sqrt 0.018, off 1 Mar` = list(cfg = c(sqrt_018, list(off_after = "03-01")), warm = FALSE)
)
FCS_LIVE <- purrr::map(FCS, \(fc) filter(fc, season_of(reference_date) %in% LIVE))
cgrid <- tidyr::expand_grid(forecaster = names(FCS), cand = names(CANDS))
ccals <- purrr::pmap(cgrid, function(forecaster, cand) {
  fc <- if (CANDS[[cand]]$warm) FCS[[forecaster]] else FCS_LIVE[[forecaster]]
  do.call(cal_run_cached, c(list(fc, wi$truth), CANDS[[cand]]$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)")
SEASONS <- c(LIVE, "both")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
```

# Candidates on both

The candidates from E14, E16 and E17, each run on both forecasters on the
same rows: REF-op warm (as in E07) and cold, sqrt-scale constant
0.018 (E16), rate-scale constant 0.032 per 100k (E14), and sqrt 0.018
switched off from 1 March (`off_after = "03-01"`, E17). The cold ones learn
from the live seasons only.

```{r cand-metrics}
cm <- purrr::map2(ccals, seq_len(nrow(cgrid)), \(cal, i) score(cal) %>% mutate(cgrid[i, ])) %>%
  bind_rows() %>%
  mutate(
    forecaster = factor(forecaster, levels = names(FCS)), cand = factor(cand, levels = names(CANDS)),
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
```

```{r month-view-cfgs}
mv <- cal_month_view(setNames(ccals, paste(cgrid$forecaster, cgrid$cand, sep = ", ")), seasons = LIVE)
# Three candidates on both forecasters: colour is the candidate, linetype the forecaster (also for the grey base lines).
MV_CANDS <- c(`REF-op warm` = "black", `sqrt 0.018` = "#0072B2", `sqrt 0.018, off 1 Mar` = "#D55E00")
mv_plot_cfgs <- paste(rep(names(FCS), each = length(MV_CANDS)), names(MV_CANDS), sep = ", ")
mv_cols <- MV_CANDS
mv$month <- mv$month %>% mutate(forecaster = sub(",.*", "", config), cand = factor(sub("^[^,]*, ", "", config), levels = names(CANDS)))
mv_colour_by <- "cand"
mv_linetype <- "forecaster"
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Rankings on A and B

Candidates ordered by mean WIS reduction % (positive = calibrated WIS lower than base) over h0–h3, separately on each
forecaster, states only. `rank A` / `rank B` are the positions (1 is the
largest WIS reduction; `cov rank` 1 is the lowest L1 coverage bias); Spearman ρ compares the two orders per season, over all
(candidate, horizon) cells.

```{r cand-rank}
rk <- cm %>%
  filter(geos == GEOS[["states"]]) %>%
  group_by(forecaster, season, cand) %>%
  summarize(wis = mean(wis_pct), cov = mean(cal_err_cal), .groups = "drop") %>%
  group_by(forecaster, season) %>%
  mutate(rank = rank(-wis), cov_rank = rank(cov)) %>%
  ungroup() %>%
  mutate(fc = ifelse(forecaster == names(FCS)[1], "A", "B")) %>%
  select(season, cand, fc, wis, rank, cov, cov_rank) %>%
  tidyr::pivot_wider(names_from = fc, values_from = c(wis, rank, cov, cov_rank)) %>%
  arrange(season, rank_A)
rk %>%
  transmute(season, cand, `WIS reduction % A` = wis_A, `rank A` = rank_A, `WIS reduction % B` = wis_B, `rank B` = rank_B,
            `cov bias A` = cov_A, `cov rank A` = cov_rank_A, `cov bias B` = cov_B, `cov rank B` = cov_rank_B) %>%
  knitr::kable(digits = c(0, 0, 2, 0, 2, 0, 3, 0, 3, 0), caption = "Mean over h0–h3, states only. WIS reduction %: positive = calibrated WIS lower than base. Coverage bias is L1 coverage bias (lower is better)")
cm %>%
  filter(geos == GEOS[["states"]]) %>%
  select(season, cand, h, forecaster, wis_pct, cal_err_cal) %>%
  tidyr::pivot_wider(names_from = forecaster, values_from = c(wis_pct, cal_err_cal)) %>%
  group_by(season) %>%
  summarize(
    `ρ WIS cells` = cor(.data[[paste0("wis_pct_", names(FCS)[1])]], .data[[paste0("wis_pct_", names(FCS)[2])]], method = "spearman"),
    `ρ coverage cells` = cor(.data[[paste0("cal_err_cal_", names(FCS)[1])]], .data[[paste0("cal_err_cal_", names(FCS)[2])]], method = "spearman")
  ) %>%
  knitr::kable(digits = 2, caption = "Spearman correlation of A vs B over (candidate, horizon) cells")
```
