---
title: "E09: constant learning rate × revision delay (rate scale)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
  wis_max: 20
  smoke: false
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 11, fig.height = 11,
  message = FALSE, warning = FALSE,
  cache.path = here::here("cache/calibration/experiments/knitr_e09/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Ledger entry **E09** (`notes/calibration-ledger.md`). The question, from a
collaborator: for the plain MultiQT tracker applied on the **rate scale**
(admissions per 100k) with a **constant** learning rate, how do coverage and
WIS trade off across learning rates, and how does that change with the
**extra delay** before an outcome is learned from?

**Extra delay.** The first NHSN report of the week ending on Saturday `X` is
published on Wednesday `X + 4`, in time for the round with reference date
`X + 7`. Learning from it then is `settle_days = 7` (`h + 1` rounds), zero
extra delay. Each further week adds 7 days: our reference `settle_days = 14`
is one extra week. The sweep is 0–3 extra weeks.

**Learning truth.** The tracker learns from NHSN vintages, not finalized
values (E00). Two deployment designs are shown, since E00 found they respond
to the delay in opposite directions:

- **re-run** (`exact`, what prod does): the tracker is re-run from scratch each
  week on that week's data, so a value learned early is replaced by its
  revision the next week.
- **stateful**: the tracker steps once per outcome, on the value available at
  the round it is revealed, and never revisits it.

Everything else is REF-paper (`cal_ref_args("paper")`): single term, carry
across seasons, 2023-24 burn-in (no steps; the constant rate needs no
warm-up). Scores are on counts against finalized truth, over the two live
seasons, all 53 hub locations. "WIS inflation %" is `100 × (WIS_cal / WIS_base − 1)`,
so negative is better than the base. "L1 coverage bias" is the mean over the
23 levels of |empirical coverage − level|.

# Findings

Numbers are from the render of 30 September 2026, pooled over locations.
"Knee" below means a constant learning rate of 0.056 admissions per 100k per
step.

- **The trade-off is L-shaped.** Raising the constant learning rate cuts the
  coverage bias from about 0.10 to 0.03 at almost no WIS cost, up to a knee
  near 0.05–0.1 (h1–h3; up to 0.3 at h−1). Past the knee WIS climbs steeply
  and coverage stops improving. The per-location lines spread widely: past
  0.3 many locations inflate WIS by 20% or more.
- **On the rate scale, the knee beats the paper's adaptive tracker.** At the
  reference delay (+1 week, re-run design), the knee gives WIS inflation
  −3.9/−2.3/−1.0/−0.6/−0.1% and coverage bias 0.043/0.031/0.026/0.026/0.026
  at h−1…h3. REF-paper gives −4.2/−2.4/+0.5/+2.9/+5.2% and
  0.047/0.030/0.032/0.034/0.032, so the knee is better on both at h1–h3 and
  within 0.3 points at h−1/h0.
- **The rate scale only matters with a constant learning rate.** "rate,
  adaptive" matches REF-paper exactly. The adaptive rate is a quantile of the
  location's own residuals, so dividing a location by its population rescales
  the steps and the offsets together.
- **REF-op makes a different trade.** Its WIS gain (−6.8/−5.1/−2.5/−1.8/−1.9%)
  is larger than any constant rate reaches at h0–h3, but at about twice the
  coverage bias (0.080/0.054/0.045/0.046/0.041). The leak and the warm start
  buy WIS; the plain tracker buys coverage.
- **Extra delay, re-run design (prod): shorter is better, but it matters only
  at h−1/h0.** At a learning rate of 0.1, h−1 WIS inflation is −6.2/−4.6/−2.8%
  at +0/+1/+2 weeks. At h1–h3 the curves nearly coincide, and coverage bias
  barely moves with delay. A value learned early is replaced by its revision
  in the next week's run, so waiting only delays the signal.
- **Extra delay, stateful design: one week is needed.** At +0 weeks the
  tracker learns once from the first report, which under-reports, so the
  offsets are pushed down: at the knee, h−1 coverage bias is 0.143, worse than
  the uncalibrated 0.118. +1 week recovers most of it (0.067); +2 and +3
  weeks change little at h1–h3 and lose WIS at h−1.

```{r libs, cache=FALSE}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))
LRS <- 10^seq(-3, 0.5, by = 0.25)
SETTLE <- c(7L, 14L, 21L, 28L)
# smoke = TRUE checks the notebook end to end in a few minutes: a small grid,
# and the re-run design computed with the fast stateful mode.
if (params$smoke) {
  LRS <- LRS[c(1, 9, 13)]
  SETTLE <- c(7L, 14L)
}
LEARN_LABELS <- c(exact = "re-run (prod)", vintage = "stateful")
```

```{r load, cache=TRUE}
archive <- get_nhsn_data_archive("flu")
truth <- nhsn_read_truth("flu", archive)
vintages <- nhsn_read_vintages("flu", archive)
hub_fc <- hub_read_forecasts(hub_dir = here::here(params$hub_dir))
rate_scales <- get_population_data() %>%
  distinct(state_code, .keep_all = TRUE) %>%
  filter(state_code %in% unique(hub_fc$location)) %>%
  transmute(location = state_code, from = as.Date("2000-01-01"), scale = population / 1e5)
stopifnot(setequal(rate_scales$location, unique(hub_fc$location)))
```

```{r runs, cache=FALSE}
# Each run is cached to its own file, so a change to the grid only computes the
# new runs. Clear cache/calibration/experiments/e09/ after changing R/calibration/.
run_dir <- here::here("cache/calibration/experiments", if (params$smoke) "e09_smoke" else "e09")
dir.create(run_dir, showWarnings = FALSE, recursive = TRUE)
cached_run <- function(key, ..., learn) {
  path <- file.path(run_dir, paste0(key, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  if (params$smoke) learn <- "vintage"
  cal <- cal_run(hub_fc, truth, ..., learn = learn, vintages = vintages, workers = params$workers)
  # Only what the metrics need, to keep the cache small.
  cal <- list(forecasts = cal$forecasts %>%
    select(location, horizon, reference_date, season, level, value_base, value_cal, truth, is_burn_in))
  saveRDS(cal, path)
  cal
}
rate <- list(transform = "identity", scales = rate_scales)

sweep_grid <- tidyr::expand_grid(learn = c("exact", "vintage"), settle_days = SETTLE, lr = LRS)
sweep <- purrr::pmap(sweep_grid, function(learn, settle_days, lr) {
  key <- sprintf("sweep_%s_s%d_lr%.4f", learn, settle_days, lr)
  do.call(cached_run, c(list(key, "paper", settle_days = settle_days, lr = lr, learn = learn), rate))
})

# Reference points and the bridge from REF-paper, per (learn, settle_days).
ref_grid <- tidyr::expand_grid(learn = c("exact", "vintage"), settle_days = SETTLE)
refs <- purrr::pmap(ref_grid, function(learn, settle_days) {
  s <- sprintf("%s_s%d", learn, settle_days)
  list(
    `REF-paper (count, adaptive)` = cached_run(paste0("paper_", s), "paper", settle_days = settle_days, learn = learn),
    `rate, adaptive` = do.call(cached_run, c(list(paste0("rate_adaptive_", s), "paper",
      settle_days = settle_days, learn = learn, lr_args = list(mult = 0.03, floor = 1e-3)), rate)),
    `REF-op` = cached_run(paste0("op_", s), "op", settle_days = settle_days, learn = learn)
  )
})
```

```{r metrics, cache=FALSE}
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
# Per (location, horizon) and pooled over locations: WIS inflation and L1
# coverage bias, for the calibrated forecast and the base.
metrics <- function(cal) {
  fc <- cal$forecasts %>% filter(!is_burn_in, !is.na(truth), !is.na(value_base))
  one <- function(df, by) {
    df %>%
      group_by(across(all_of(c(by, "horizon", "level")))) %>%
      summarize(
        cov_base = mean(truth <= value_base), cov_cal = mean(truth <= value_cal),
        wis_base = sum(pinball(truth, value_base, level)), wis_cal = sum(pinball(truth, value_cal, level)),
        .groups = "drop"
      ) %>%
      group_by(across(all_of(c(by, "horizon")))) %>%
      summarize(
        bias_base = mean(abs(cov_base - level)), bias_cal = mean(abs(cov_cal - level)),
        wis_infl = 100 * (sum(wis_cal) / sum(wis_base) - 1),
        .groups = "drop"
      )
  }
  bind_rows(one(fc, "location"), one(fc, character(0)) %>% mutate(location = "pooled"))
}
sweep_m <- purrr::map2(sweep, seq_along(sweep), function(cal, i) metrics(cal) %>% bind_cols(sweep_grid[i, ])) %>%
  bind_rows()
ref_m <- purrr::map2(refs, seq_along(refs), function(r, i) {
  purrr::imap(r, function(cal, nm) metrics(cal) %>% mutate(ref = nm)) %>% bind_rows() %>% bind_cols(ref_grid[i, ])
}) %>%
  bind_rows() %>%
  filter(location == "pooled")
label_panels <- function(df) {
  df %>% mutate(
    extra = factor(sprintf("+%d wk extra delay", (settle_days - 7L) %/% 7L)),
    horizon = factor(paste0("h", horizon), levels = paste0("h", -1:3))
  )
}
sweep_m <- label_panels(sweep_m)
ref_m <- label_panels(ref_m)
```

# Bridge from REF-paper

One axis at a time at `settle_days = 14` (the reference), re-run design: first
the rate scale with the adaptive learning rate (its `floor` moves from 0.1
count to 0.001 rate units, since the floor is a scale-dependent constant),
then the constant learning rate with the lowest pooled WIS at each horizon.
REF-op is for orientation only; it differs in several axes.

```{r bridge}
best_const <- sweep_m %>%
  filter(location == "pooled", learn == "exact", settle_days == 14L) %>%
  group_by(horizon) %>%
  slice_min(wis_infl, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(ref = "rate, constant lr (best WIS per horizon)")
bridge <- bind_rows(
  ref_m %>% filter(learn == "exact", settle_days == 14L),
  best_const
) %>%
  mutate(ref = factor(ref, levels = c("REF-paper (count, adaptive)", "rate, adaptive", "rate, constant lr (best WIS per horizon)", "REF-op")))
base_row <- bridge %>% filter(ref == "REF-paper (count, adaptive)") %>% mutate(ref = "base", wis_infl = 0, bias_cal = bias_base)
tbl <- bind_rows(base_row, bridge) %>%
  transmute(config = ref, horizon, cell = sprintf("%+.1f%% / %.3f", wis_infl, bias_cal)) %>%
  tidyr::pivot_wider(names_from = horizon, values_from = cell)
knitr::kable(tbl, caption = "WIS inflation % / L1 coverage bias, pooled over locations")
knitr::kable(
  best_const %>% select(horizon, lr) %>% mutate(lr = signif(lr, 2)) %>% tidyr::pivot_wider(names_from = horizon, values_from = lr),
  caption = "Constant learning rate with the lowest pooled WIS (rate units per step)"
)
```

# Coverage bias vs WIS, per location

Each thin grey line is one location traced across learning rates; the bold
line is the pooled result, labelled with the learning rate at each decade. Learning rates whose pooled WIS inflation exceeds
`r params$wis_max`% in a panel are dropped from that panel; the axes are
clipped to the bulk of the per-location points, so a few location lines run
off the edge. The open circle is the base forecast (no calibration) and the
triangles are the reference configs at the same delay.

Per-location coverage comes from about 50 forecasts per horizon, so a
location's L1 bias has a noise floor of a few points even for a perfectly
calibrated forecaster. Read the shape of the pooled line; the grey lines show
spread.

```{r curve-fn}
ACCENT <- "#1f5fa8"
curve_plot <- function(which_learn) {
  d <- sweep_m %>% filter(learn == which_learn)
  keep <- d %>%
    filter(location == "pooled", wis_infl <= params$wis_max) %>%
    select(horizon, extra, lr)
  d <- d %>% semi_join(keep, by = c("horizon", "extra", "lr")) %>% arrange(lr)
  pooled <- d %>% filter(location == "pooled")
  locs <- d %>% filter(location != "pooled")
  base <- pooled %>% distinct(horizon, extra, bias_base)
  marks <- ref_m %>% filter(learn == which_learn, ref %in% c("REF-paper (count, adaptive)", "REF-op"))
  lab <- pooled %>% filter(abs(log10(lr) - round(log10(lr))) < 1e-9)
  xlim <- c(0, max(stats::quantile(locs$bias_cal, 0.97), base$bias_base) * 1.05)
  ylim <- range(stats::quantile(locs$wis_infl, c(0.03, 0.97)), pooled$wis_infl, marks$wis_infl, 0)
  ggplot() +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(data = locs, aes(bias_cal, wis_infl, group = location), colour = "grey45", alpha = 0.25, linewidth = 0.3) +
    geom_path(data = pooled, aes(bias_cal, wis_infl), colour = ACCENT, linewidth = 1.1) +
    geom_point(data = pooled, aes(bias_cal, wis_infl), colour = ACCENT, size = 1.2) +
    geom_text(data = lab, aes(bias_cal, wis_infl, label = format(lr, scientific = FALSE, drop0trailing = TRUE)),
      colour = "grey15", size = 2.6, vjust = -0.9) +
    geom_point(data = base, aes(bias_base, 0), shape = 21, fill = "white", colour = "black", size = 2.2) +
    geom_point(data = marks, aes(bias_cal, wis_infl, shape = ref), colour = "black", size = 2) +
    scale_shape_manual(values = c(`REF-paper (count, adaptive)` = 2, `REF-op` = 17), name = NULL) +
    facet_grid(horizon ~ extra) +
    coord_cartesian(xlim = xlim, ylim = ylim) +
    labs(x = "L1 coverage bias", y = "WIS inflation % (negative = better than base)") +
    theme(legend.position = "bottom")
}
```

## Re-run design (prod)

```{r curve-exact}
curve_plot("exact")
```

## Stateful design

```{r curve-vintage}
curve_plot("vintage")
```

# Pooled curves by delay

The bold lines from above, one colour per extra delay (darker is longer), so
the delays can be compared directly. Solid is the re-run design, dashed the
stateful one.

```{r pooled-overlay, fig.height=6}
DELAY_COLS <- setNames(RColorBrewer::brewer.pal(9, "Blues")[c(4, 6, 8, 9)], levels(sweep_m$extra))
pooled_all <- sweep_m %>%
  filter(location == "pooled") %>%
  semi_join(sweep_m %>% filter(location == "pooled", wis_infl <= params$wis_max) %>% select(horizon, extra, lr, learn),
    by = c("horizon", "extra", "lr", "learn")) %>%
  arrange(lr) %>%
  mutate(design = factor(LEARN_LABELS[learn], levels = LEARN_LABELS))
ggplot(pooled_all, aes(bias_cal, wis_infl, colour = extra, linetype = design, group = interaction(extra, design))) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_path(linewidth = 0.8) +
  geom_point(data = pooled_all %>% distinct(horizon, bias_base), aes(bias_base, 0), inherit.aes = FALSE,
    shape = 21, fill = "white", colour = "black", size = 2) +
  scale_colour_manual(values = DELAY_COLS, name = NULL) +
  scale_linetype_manual(values = c(`re-run (prod)` = "solid", stateful = "22"), name = NULL) +
  facet_wrap(~horizon, nrow = 1) +
  labs(x = "L1 coverage bias", y = "WIS inflation %") +
  theme(legend.position = "bottom", legend.box = "vertical")
```

# Tables

Pooled WIS inflation % / L1 coverage bias for every learning rate, by horizon
(columns) and extra delay (row blocks).

```{r table-fn}
sweep_table <- function(which_learn) {
  sweep_m %>%
    filter(location == "pooled", learn == which_learn) %>%
    transmute(extra, lr = signif(lr, 2), horizon, cell = sprintf("%+.1f / %.3f", wis_infl, bias_cal)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = cell) %>%
    arrange(extra, lr) %>%
    knitr::kable()
}
```

## Re-run design (prod)

```{r table-exact}
sweep_table("exact")
```

## Stateful design

```{r table-vintage}
sweep_table("vintage")
```
