---
title: "E09: constant learning rate × revision delay (rate scale)"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
  wis_max: 20
  smoke: false
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 11, fig.height = 11,
  message = FALSE, warning = FALSE,
  cache.path = here::here("cache/calibration/experiments/knitr_e09/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: for the plain quantile tracker on the rate scale (admissions per
100k) with a constant learning rate, how do WIS and coverage change with
the learning rate and with the extra delay before an outcome is learned?

**Setup.**

| | |
|---|---|
| disease, base | flu, the submitted hub ensemble (`CMU-TimeSeries`) |
| seasons | 2023-24 burn-in (its outcomes take no steps; offsets start at zero); 2024-25 and 2025-26 scored |
| locations, horizons | all 53 hub locations, h−1 to h3; the month view is states only |
| scoring truth | finalized NHSN |
| tracker | REF-paper (one offset per level, carried across seasons) on the rate scale |
| learning rate | constant, 0.001 to about 3.2 per 100k per step (15 values, a quarter-decade apart) |
| extra delay | 0, 1, 2 or 3 weeks (`settle_days` 7/14/21/28) |
| learning truth | two designs, below |

**The tracker.** For each location, horizon and quantile level, the offset
moves by `lr × (1{y > q} − (1 − τ))` when an outcome is learned. With a
constant `lr` the step does not depend on the forecast errors. The
reference configs use the adaptive rate instead: 0.03 × the 90th
percentile of the location's absolute base errors over its last 20
observed rounds.

**Extra delay.** The first NHSN report of the week ending Saturday `X` is
published Wednesday `X + 4`, in time for the round with reference date
`X + 7`. Learning from it at that round is zero extra delay; each further
week of waiting is one week of extra delay. The reference configs use one
week.

**Learning truth designs.**

- **re-run** (prod): each week the tracker is re-run from the first round
  on that week's vintage, so an outcome is learned from its latest value.
- **stateful**: the tracker runs once and learns each outcome from the
  value published at the round it becomes available.

**Reference configs** (marked on the plots):

- **REF-paper**: the paper's tracker on counts, adaptive learning rate.
- **rate, adaptive**: REF-paper on the rate scale (floor 0.001 per 100k).
- **REF-op**: prod's config: REF-paper on the sqrt scale, plus a 10%
  per-round leak, a slow offset and a warm start from the burn-in season.

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so positive =
calibrated WIS lower than base. "L1 coverage bias" (lower is better) is the
mean over the 23 quantile levels of |empirical coverage − level|. "Knee"
below is the constant rate 0.056.

```{r libs, cache=FALSE}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))
LRS <- 10^seq(-3, 0.5, by = 0.25)
SETTLE <- c(7L, 14L, 21L, 28L)
# smoke = TRUE checks the notebook end to end in a few minutes: a small grid,
# and the re-run design computed with the fast stateful mode.
if (params$smoke) {
  LRS <- LRS[c(1, 9, 13)]
  SETTLE <- c(7L, 14L)
}
LEARN_LABELS <- c(exact = "re-run (prod)", vintage = "stateful")
```

```{r load, cache=TRUE}
archive <- get_nhsn_data_archive("flu")
truth <- nhsn_read_truth("flu", archive)
vintages <- nhsn_read_vintages("flu", archive)
hub_fc <- hub_read_forecasts(hub_dir = here::here(params$hub_dir))
rate_scales <- get_population_data() %>%
  distinct(state_code, .keep_all = TRUE) %>%
  filter(state_code %in% unique(hub_fc$location)) %>%
  transmute(location = state_code, from = as.Date("2000-01-01"), scale = population / 1e5)
stopifnot(setequal(rate_scales$location, unique(hub_fc$location)))
```

```{r runs, cache=FALSE}
# Each run is cached to its own file, so a change to the grid only computes the
# new runs. Clear cache/calibration/experiments/e09/ after changing R/calibration/.
run_dir <- here::here("cache/calibration/experiments", if (params$smoke) "e09_smoke" else "e09")
dir.create(run_dir, showWarnings = FALSE, recursive = TRUE)
cached_run <- function(key, ..., learn) {
  path <- file.path(run_dir, paste0(key, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  if (params$smoke) learn <- "vintage"
  cal <- cal_run(hub_fc, truth, ..., learn = learn, vintages = vintages, workers = params$workers)
  # Only what the metrics need, to keep the cache small.
  cal <- list(forecasts = cal$forecasts %>%
    select(location, horizon, reference_date, season, level, value_base, value_cal, truth, is_burn_in))
  saveRDS(cal, path)
  cal
}
rate <- list(transform = "identity", scales = rate_scales)

sweep_grid <- tidyr::expand_grid(learn = c("exact", "vintage"), settle_days = SETTLE, lr = LRS)
sweep <- purrr::pmap(sweep_grid, function(learn, settle_days, lr) {
  key <- sprintf("sweep_%s_s%d_lr%.4f", learn, settle_days, lr)
  do.call(cached_run, c(list(key, "paper", settle_days = settle_days, lr = lr, learn = learn), rate))
})

# Reference points and the bridge from REF-paper, per (learn, settle_days).
ref_grid <- tidyr::expand_grid(learn = c("exact", "vintage"), settle_days = SETTLE)
refs <- purrr::pmap(ref_grid, function(learn, settle_days) {
  s <- sprintf("%s_s%d", learn, settle_days)
  list(
    `REF-paper (count, adaptive)` = cached_run(paste0("paper_", s), "paper", settle_days = settle_days, learn = learn),
    `rate, adaptive` = do.call(cached_run, c(list(paste0("rate_adaptive_", s), "paper",
      settle_days = settle_days, learn = learn, lr_args = list(mult = 0.03, floor = 1e-3)), rate)),
    `REF-op` = cached_run(paste0("op_", s), "op", settle_days = settle_days, learn = learn)
  )
})
```

```{r metrics, cache=FALSE}
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
# Per (location, horizon) and pooled over locations: WIS reduction % and L1
# coverage bias, for the calibrated forecast and the base.
metrics <- function(cal) {
  fc <- cal$forecasts %>% filter(!is_burn_in, !is.na(truth), !is.na(value_base))
  one <- function(df, by) {
    df %>%
      group_by(across(all_of(c(by, "horizon", "level")))) %>%
      summarize(
        cov_base = mean(truth <= value_base), cov_cal = mean(truth <= value_cal),
        wis_base = sum(pinball(truth, value_base, level)), wis_cal = sum(pinball(truth, value_cal, level)),
        .groups = "drop"
      ) %>%
      group_by(across(all_of(c(by, "horizon")))) %>%
      summarize(
        bias_base = mean(abs(cov_base - level)), bias_cal = mean(abs(cov_cal - level)),
        wis_pct = 100 * (1 - sum(wis_cal) / sum(wis_base)),
        .groups = "drop"
      )
  }
  bind_rows(one(fc, "location"), one(fc, character(0)) %>% mutate(location = "pooled"))
}
sweep_m <- purrr::map2(sweep, seq_along(sweep), function(cal, i) metrics(cal) %>% bind_cols(sweep_grid[i, ])) %>%
  bind_rows()
ref_m <- purrr::map2(refs, seq_along(refs), function(r, i) {
  purrr::imap(r, function(cal, nm) metrics(cal) %>% mutate(ref = nm)) %>% bind_rows() %>% bind_cols(ref_grid[i, ])
}) %>%
  bind_rows() %>%
  filter(location == "pooled")
label_panels <- function(df) {
  df %>% mutate(
    extra = factor(sprintf("+%d wk extra delay", (settle_days - 7L) %/% 7L)),
    horizon = factor(paste0("h", horizon), levels = paste0("h", -1:3))
  )
}
sweep_m <- label_panels(sweep_m)
ref_m <- label_panels(ref_m)
```

# Bridge from REF-paper

One change at a time at one week of extra delay, re-run design: first the
rate scale with the adaptive learning rate (its minimum step drops from 0.1
admissions to 0.001 per 100k, since that minimum is in the working units),
then the constant learning rate with the lowest pooled WIS at each horizon.
REF-op is for orientation only; it differs in several settings.

```{r bridge}
best_const <- sweep_m %>%
  filter(location == "pooled", learn == "exact", settle_days == 14L) %>%
  group_by(horizon) %>%
  slice_max(wis_pct, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(ref = "rate, constant lr (highest WIS per horizon)")
bridge <- bind_rows(
  ref_m %>% filter(learn == "exact", settle_days == 14L),
  best_const
) %>%
  mutate(ref = factor(ref, levels = c("REF-paper (count, adaptive)", "rate, adaptive", "rate, constant lr (highest WIS per horizon)", "REF-op")))
base_row <- bridge %>% filter(ref == "REF-paper (count, adaptive)") %>% mutate(ref = "base", wis_pct = 0, bias_cal = bias_base)
tbl <- bind_rows(base_row, bridge) %>%
  transmute(config = ref, horizon, cell = sprintf("%+.1f%% / %.3f", wis_pct, bias_cal)) %>%
  tidyr::pivot_wider(names_from = horizon, values_from = cell)
knitr::kable(tbl, caption = "WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage bias (lower is better), pooled over locations")
knitr::kable(
  best_const %>% select(horizon, lr) %>% mutate(lr = signif(lr, 2)) %>% tidyr::pivot_wider(names_from = horizon, values_from = lr),
  caption = "Constant learning rate with the lowest pooled WIS (rate units per step)"
)
```

L1 coverage bias for states only, both seasons, season-pooled against
month-average (see "Month view" below), for the reference configs, the knee
(0.056) and 0.1 at several delays and both designs:

```{r bridge-month}
sweep_run <- function(learn, settle_days, lr) {
  sweep[[which(sweep_grid$learn == learn & sweep_grid$settle_days == settle_days & abs(log10(sweep_grid$lr / lr)) < 1e-6)]]
}
ref_run <- function(learn, settle_days, nm) refs[[which(ref_grid$learn == learn & ref_grid$settle_days == settle_days)]][[nm]]
mv_cals <- list(
  `REF-paper, re-run +1 wk` = ref_run("exact", 14L, "REF-paper (count, adaptive)"),
  `REF-op, re-run +1 wk` = ref_run("exact", 14L, "REF-op"),
  `knee 0.056, re-run +0 wk` = sweep_run("exact", 7L, 10^-1.25),
  `knee 0.056, re-run +1 wk` = sweep_run("exact", 14L, 10^-1.25),
  `knee 0.056, re-run +2 wk` = sweep_run("exact", 21L, 10^-1.25),
  `0.1, re-run +1 wk` = sweep_run("exact", 14L, 10^-1),
  `knee 0.056, stateful +0 wk` = sweep_run("vintage", 7L, 10^-1.25),
  `knee 0.056, stateful +1 wk` = sweep_run("vintage", 14L, 10^-1.25)
)
mv <- cal_month_view(mv_cals, seasons = c("2024-2025", "2025-2026"), horizons = NULL)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv, seasons = "both"), caption = CAL_MONTH_COV_CAPTION)
```

# Coverage bias vs WIS, per location

Each thin grey line is one location traced across learning rates; the bold
line is the pooled result, labelled with the learning rate at each decade. Learning rates whose pooled WIS is more than
`r params$wis_max`% worse than the base in a panel are dropped from that panel; the axes are
clipped to the bulk of the per-location points, so a few location lines run
off the edge. The open circle is the base forecast (no calibration) and the
triangles are the reference configs at the same delay.

Per-location coverage comes from about 50 forecasts per horizon, so a
location's L1 bias has a noise floor of a few points even for a perfectly
calibrated forecaster. The grey lines show the spread across locations.

```{r curve-fn}
ACCENT <- "#1f5fa8"
curve_plot <- function(which_learn) {
  d <- sweep_m %>% filter(learn == which_learn)
  keep <- d %>%
    filter(location == "pooled", wis_pct >= -params$wis_max) %>%
    select(horizon, extra, lr)
  d <- d %>% semi_join(keep, by = c("horizon", "extra", "lr")) %>% arrange(lr)
  pooled <- d %>% filter(location == "pooled")
  locs <- d %>% filter(location != "pooled")
  base <- pooled %>% distinct(horizon, extra, bias_base)
  marks <- ref_m %>% filter(learn == which_learn, ref %in% c("REF-paper (count, adaptive)", "REF-op"))
  lab <- pooled %>% filter(abs(log10(lr) - round(log10(lr))) < 1e-9)
  xlim <- c(0, max(stats::quantile(locs$bias_cal, 0.97), base$bias_base) * 1.05)
  ylim <- range(stats::quantile(locs$wis_pct, c(0.03, 0.97)), pooled$wis_pct, marks$wis_pct, 0)
  ggplot() +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(data = locs, aes(bias_cal, wis_pct, group = location), colour = "grey45", alpha = 0.25, linewidth = 0.3) +
    geom_path(data = pooled, aes(bias_cal, wis_pct), colour = ACCENT, linewidth = 1.1) +
    geom_point(data = pooled, aes(bias_cal, wis_pct), colour = ACCENT, size = 1.2) +
    geom_text(data = lab, aes(bias_cal, wis_pct, label = format(lr, scientific = FALSE, drop0trailing = TRUE)),
      colour = "grey15", size = 2.6, vjust = -0.9) +
    geom_point(data = base, aes(bias_base, 0), shape = 21, fill = "white", colour = "black", size = 2.2) +
    geom_point(data = marks, aes(bias_cal, wis_pct, shape = ref), colour = "black", size = 2) +
    scale_shape_manual(values = c(`REF-paper (count, adaptive)` = 2, `REF-op` = 17), name = NULL) +
    facet_grid(horizon ~ extra) +
    coord_cartesian(xlim = xlim, ylim = ylim) +
    labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)") +
    theme(legend.position = "bottom")
}
```

## Re-run design (prod)

```{r curve-exact}
curve_plot("exact")
```

## Stateful design

```{r curve-vintage}
curve_plot("vintage")
```

# Pooled curves by delay

The bold lines from above, one colour per extra delay (darker is longer), so
the delays can be compared directly. Solid is the re-run design, dashed the
stateful one.

```{r pooled-overlay, fig.height=6}
DELAY_COLS <- setNames(RColorBrewer::brewer.pal(9, "Blues")[c(4, 6, 8, 9)], levels(sweep_m$extra))
pooled_all <- sweep_m %>%
  filter(location == "pooled") %>%
  semi_join(sweep_m %>% filter(location == "pooled", wis_pct >= -params$wis_max) %>% select(horizon, extra, lr, learn),
    by = c("horizon", "extra", "lr", "learn")) %>%
  arrange(lr) %>%
  mutate(design = factor(LEARN_LABELS[learn], levels = LEARN_LABELS))
ggplot(pooled_all, aes(bias_cal, wis_pct, colour = extra, linetype = design, group = interaction(extra, design))) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_path(linewidth = 0.8) +
  geom_point(data = pooled_all %>% distinct(horizon, bias_base), aes(bias_base, 0), inherit.aes = FALSE,
    shape = 21, fill = "white", colour = "black", size = 2) +
  scale_colour_manual(values = DELAY_COLS, name = NULL) +
  scale_linetype_manual(values = c(`re-run (prod)` = "solid", stateful = "22"), name = NULL) +
  facet_wrap(~horizon, nrow = 1) +
  labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom", legend.box = "vertical")
```

# Tables

Pooled WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage bias (lower is better) for every learning rate, by horizon
(columns) and extra delay (row blocks).

```{r table-fn}
sweep_table <- function(which_learn) {
  sweep_m %>%
    filter(location == "pooled", learn == which_learn) %>%
    transmute(extra, lr = signif(lr, 2), horizon, cell = sprintf("%+.1f / %.3f", wis_pct, bias_cal)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = cell) %>%
    arrange(extra, lr) %>%
    knitr::kable()
}
```

## Re-run design (prod)

```{r table-exact}
sweep_table("exact")
```

## Stateful design

```{r table-vintage}
sweep_table("vintage")
```

```{r month-view-cfgs}
# The configs the Findings compare: the references and the knee at one week of delay, and the knee in the stateful design.
mv_plot_cfgs <- c(
  "REF-paper, re-run +1 wk", "REF-op, re-run +1 wk", "knee 0.056, re-run +1 wk", "0.1, re-run +1 wk",
  "knee 0.056, stateful +0 wk", "knee 0.056, stateful +1 wk"
)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Findings

WIS reduction % (positive = calibrated WIS lower than base) differences
are in points. L1 coverage bias is lower-is-better; month-avg is states
only, season-pooled is all locations. Unless stated, +1 week, re-run.

- **Constant rate.** WIS rises with the rate up to about 0.056–0.1 (knee:
  +3.2/+2.5/+1.1/+0.5/+0.2 at h−1…h3) and falls fast past 0.3 (at 1.0:
  −12 to −23 at h0–h3). The rate with the highest WIS per horizon is
  0.1/0.056/0.056/0.018/0.018, picked on the scored seasons; at h2,
  0.018–0.056 are within 0.1 point.
- **Knee vs REF-paper.** WIS: knee ahead by 3.1–5.5 points at h1–h3, 0.6 at
  h0, tied at h−1. Coverage: no clear difference season-pooled (within
  0.007); month-avg REF-paper is lower at h−1…h1 by 0.007/0.015/0.009 and
  the knee at h2/h3 by 0.001/0.004.
- **Month-avg coverage vs base.** Against the base
  (0.134/0.116/0.123/0.137/0.151 at h−1…h3), the knee differs by
  −0.012/+0.001/+0.006/+0.005/+0.013 and 0.1 by
  −0.016/−0.002/+0.004/+0.010/+0.021: no lower than the base at h1–h3. REF-op is 0.012–0.027 below the base at every horizon. REF-op
  differs from the sweep in scale, leak, slow offset and warm start, so
  the comparison does not isolate any one of them. Season-pooled, the
  knee's bias (0.026–0.040) is about half REF-op's (0.041–0.074), but
  pooling lets ramp and post-peak errors cancel.
- **Extra delay, re-run.** At the knee, h−1 WIS is +4.1/+3.2/+2.1/+1.0 at
  +0/+1/+2/+3 weeks and h0 +2.9/+2.5/+2.0/+1.5; at h1–h3 the delays are
  within 1 point. Month-avg bias at the knee moves by under 0.01 between
  +0 and +2 weeks.
- **Extra delay, stateful.** At +0 weeks the knee makes h−1 worse than the
  base: WIS −0.9, season-pooled bias 0.142 vs 0.111, month-avg 0.151 vs
  0.134. At +1 week h−1 is +3.7, 0.066 pooled, 0.118 month-avg. At h0–h3,
  +0 weeks has WIS within 0.6 points of +1 week.
- **Known issues.** `rate, adaptive` matches REF-paper to every printed
  digit because the adaptive rate is scale invariant (`scales` is applied;
  checked 2026-10-02). Runs are cached
  by name, not by inputs or code (clear `cache/calibration/experiments/e09/`
  after changes). In October 2025 (one round) the knee and 0.1 carry spring
  offsets forward: h0 month bias 0.232/0.218 vs the base's 0.053.
