---
title: "E09: constant learning rate × revision delay (rate scale)"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
  wis_max: 20
  smoke: false
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 11, fig.height = 11,
  message = FALSE, warning = FALSE,
  cache.path = here::here("cache/calibration/experiments/knitr_e09/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

The question: for the plain quantile tracker applied on the **rate scale**
(admissions per 100k) with a **constant** learning rate, how do coverage and
WIS trade off across learning rates, and how does that change with the
**extra delay** before the tracker learns from an outcome?

**The tracker.** For each location, horizon and quantile level, the tracker
keeps an additive offset that is added to the submitted quantile. When an
outcome is learned, each offset moves by `lr × (1{y > q} − (1 − τ))`: up by
`lr × τ` if the outcome landed above the played quantile `q`, down by
`lr × (1 − τ)` if below. `lr` is the learning rate (the step size).

- **Constant learning rate** (the sweep in this notebook): `lr` is a fixed
  number, the same for every location and round, and does not depend on the
  forecast errors. On the rate scale it is in admissions per 100k, so one
  step moves an offset by at most `lr` per 100k. The curves below trace this
  one number from 0.001 to about 3.2.
- **Adaptive learning rate** (the paper's default, used by both reference
  configs): `lr` is recomputed every round as 0.03 × the 90th percentile of
  the base forecast's absolute errors over the location's last 20 observed
  rounds. The paper uses 0.1 and 50 rounds; an earlier sweep found 0.03 gives
  similar coverage at a much smaller WIS cost, and that the window length
  (8, 20, 50 or all history) barely matters. The window has no role with a
  constant learning rate, so it is not varied here; the reference marks use
  20.

**Extra delay.** The first NHSN report of the week ending on Saturday `X` is
published on Wednesday `X + 4`, in time for the round with reference date
`X + 7`. Learning from it at that round is zero extra delay. Each further
week of waiting is one week of extra delay, giving later revisions time to
come in. The reference configs wait one extra week; the sweep covers 0–3.

**What the tracker learns from.** At each round the tracker learns from the
NHSN data as published by then (a "vintage"), not from finalized values,
because finalized values are not available in real time (see E00). There are
two ways to deploy that, and E00 found they respond to the delay in opposite
directions, so both are shown:

- **re-run** (what prod does): each week the tracker is run again from the
  first round over the whole history, on that week's vintage. An outcome
  first learned from its early report is learned from its revised value in
  every later week's run.
- **stateful**: the tracker runs once, stepping forward a round at a time,
  and learns each outcome once, from the value published at the round it
  becomes available. It never sees later revisions of that outcome.

This is a separate choice from what happens to the offsets between seasons.
In both designs the offsets carry over the summer off-season rather than
resetting to zero.

**Other settings.** Everything else follows the REF-paper reference config:
one offset per level, carried across seasons, and a 2023-24 burn-in season
whose outcomes take no steps (the adaptive rate uses it to warm up; the
constant rate needs no warm-up). Scores are on counts against finalized
truth, over the two live seasons (2024-25 and 2025-26), all 53 hub
locations. "WIS change %" is `100 × (1 − WIS_cal / WIS_base)`, so positive
is better than the base, as in the other notebooks. "L1 coverage bias" is the mean over the 23 quantile
levels of |empirical coverage − level|.

**Reference configs** (marked on the plots):

- **REF-paper**: the paper's tracker on counts, adaptive learning rate as
  above.
- **REF-op**: the configuration running in prod. REF-paper on the
  square-root scale, plus a 10% per-round shrink of the offset toward zero, a
  second slow-moving offset, and a starting offset taken from the burn-in
  season's errors (E03–E05).

# Findings

Numbers are from the render of 30 September 2026, pooled over locations.
"Knee" below means a constant learning rate of 0.056 admissions per 100k per
step.

- **The trade-off is L-shaped.** Raising the constant learning rate cuts the
  coverage bias from about 0.10 to 0.03 at almost no WIS cost, up to a knee
  near 0.05–0.1. Past the knee WIS worsens steeply and coverage barely
  improves. At the reference delay, h−1 has its best WIS at 0.1 (at 0.32
  with no extra delay). The per-location lines spread widely: past 0.3 many
  locations lose 20% of WIS or more.
- **On the rate scale, the knee matches or beats the paper's adaptive
  tracker at every horizon.** At the reference delay (+1 week, re-run
  design), the knee gives WIS change +3.2/+2.5/+1.1/+0.5/+0.2% and coverage
  bias 0.040/0.031/0.026/0.027/0.027 at h−1…h3. REF-paper gives
  +3.2/+1.9/−2.0/−2.7/−5.3% and 0.044/0.030/0.032/0.034/0.032: the same WIS
  at h−1, worse WIS from h0 on, and worse coverage at h−1 and h1–h3.
- **The rate scale only matters with a constant learning rate.** "rate,
  adaptive" matches REF-paper exactly. The adaptive rate is a quantile of the
  location's own residuals, so dividing a location by its population rescales
  the steps and the offsets together.
- **REF-op makes a different trade.** Its WIS gain (+5.4/+4.9/+2.4/+1.7/+2.2%)
  is larger than any constant rate reaches at h0–h3, but at about twice the
  coverage bias (0.074/0.054/0.045/0.046/0.041). The leak and the warm start
  buy WIS; the plain tracker buys coverage.
- **Extra delay, re-run design (prod): shorter is better, but it matters only
  at h−1/h0.** At a learning rate of 0.1, h−1 WIS change is +5.3/+3.6/+1.0%
  at +0/+1/+2 weeks. At h1–h3 the curves nearly coincide, and coverage bias
  barely moves with delay. A value learned early is replaced by its revision
  in the next week's run, so waiting only delays the signal.
- **Extra delay, stateful design: one week is needed.** At +0 weeks the
  tracker learns once from the first report, which under-reports, so the
  offsets are pushed down: at the knee, h−1 coverage bias is 0.142, worse than
  the uncalibrated 0.111, and h−1 WIS is 0.9% worse than the base. +1 week
  recovers most of it (0.066); +2 and +3 weeks change little at h1–h3 and
  lose WIS at h−1.

```{r libs, cache=FALSE}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))
LRS <- 10^seq(-3, 0.5, by = 0.25)
SETTLE <- c(7L, 14L, 21L, 28L)
# smoke = TRUE checks the notebook end to end in a few minutes: a small grid,
# and the re-run design computed with the fast stateful mode.
if (params$smoke) {
  LRS <- LRS[c(1, 9, 13)]
  SETTLE <- c(7L, 14L)
}
LEARN_LABELS <- c(exact = "re-run (prod)", vintage = "stateful")
```

```{r load, cache=TRUE}
archive <- get_nhsn_data_archive("flu")
truth <- nhsn_read_truth("flu", archive)
vintages <- nhsn_read_vintages("flu", archive)
hub_fc <- hub_read_forecasts(hub_dir = here::here(params$hub_dir))
rate_scales <- get_population_data() %>%
  distinct(state_code, .keep_all = TRUE) %>%
  filter(state_code %in% unique(hub_fc$location)) %>%
  transmute(location = state_code, from = as.Date("2000-01-01"), scale = population / 1e5)
stopifnot(setequal(rate_scales$location, unique(hub_fc$location)))
```

```{r runs, cache=FALSE}
# Each run is cached to its own file, so a change to the grid only computes the
# new runs. Clear cache/calibration/experiments/e09/ after changing R/calibration/.
run_dir <- here::here("cache/calibration/experiments", if (params$smoke) "e09_smoke" else "e09")
dir.create(run_dir, showWarnings = FALSE, recursive = TRUE)
cached_run <- function(key, ..., learn) {
  path <- file.path(run_dir, paste0(key, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  if (params$smoke) learn <- "vintage"
  cal <- cal_run(hub_fc, truth, ..., learn = learn, vintages = vintages, workers = params$workers)
  # Only what the metrics need, to keep the cache small.
  cal <- list(forecasts = cal$forecasts %>%
    select(location, horizon, reference_date, season, level, value_base, value_cal, truth, is_burn_in))
  saveRDS(cal, path)
  cal
}
rate <- list(transform = "identity", scales = rate_scales)

sweep_grid <- tidyr::expand_grid(learn = c("exact", "vintage"), settle_days = SETTLE, lr = LRS)
sweep <- purrr::pmap(sweep_grid, function(learn, settle_days, lr) {
  key <- sprintf("sweep_%s_s%d_lr%.4f", learn, settle_days, lr)
  do.call(cached_run, c(list(key, "paper", settle_days = settle_days, lr = lr, learn = learn), rate))
})

# Reference points and the bridge from REF-paper, per (learn, settle_days).
ref_grid <- tidyr::expand_grid(learn = c("exact", "vintage"), settle_days = SETTLE)
refs <- purrr::pmap(ref_grid, function(learn, settle_days) {
  s <- sprintf("%s_s%d", learn, settle_days)
  list(
    `REF-paper (count, adaptive)` = cached_run(paste0("paper_", s), "paper", settle_days = settle_days, learn = learn),
    `rate, adaptive` = do.call(cached_run, c(list(paste0("rate_adaptive_", s), "paper",
      settle_days = settle_days, learn = learn, lr_args = list(mult = 0.03, floor = 1e-3)), rate)),
    `REF-op` = cached_run(paste0("op_", s), "op", settle_days = settle_days, learn = learn)
  )
})
```

```{r metrics, cache=FALSE}
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
# Per (location, horizon) and pooled over locations: WIS change and L1
# coverage bias, for the calibrated forecast and the base.
metrics <- function(cal) {
  fc <- cal$forecasts %>% filter(!is_burn_in, !is.na(truth), !is.na(value_base))
  one <- function(df, by) {
    df %>%
      group_by(across(all_of(c(by, "horizon", "level")))) %>%
      summarize(
        cov_base = mean(truth <= value_base), cov_cal = mean(truth <= value_cal),
        wis_base = sum(pinball(truth, value_base, level)), wis_cal = sum(pinball(truth, value_cal, level)),
        .groups = "drop"
      ) %>%
      group_by(across(all_of(c(by, "horizon")))) %>%
      summarize(
        bias_base = mean(abs(cov_base - level)), bias_cal = mean(abs(cov_cal - level)),
        wis_pct = 100 * (1 - sum(wis_cal) / sum(wis_base)),
        .groups = "drop"
      )
  }
  bind_rows(one(fc, "location"), one(fc, character(0)) %>% mutate(location = "pooled"))
}
sweep_m <- purrr::map2(sweep, seq_along(sweep), function(cal, i) metrics(cal) %>% bind_cols(sweep_grid[i, ])) %>%
  bind_rows()
ref_m <- purrr::map2(refs, seq_along(refs), function(r, i) {
  purrr::imap(r, function(cal, nm) metrics(cal) %>% mutate(ref = nm)) %>% bind_rows() %>% bind_cols(ref_grid[i, ])
}) %>%
  bind_rows() %>%
  filter(location == "pooled")
label_panels <- function(df) {
  df %>% mutate(
    extra = factor(sprintf("+%d wk extra delay", (settle_days - 7L) %/% 7L)),
    horizon = factor(paste0("h", horizon), levels = paste0("h", -1:3))
  )
}
sweep_m <- label_panels(sweep_m)
ref_m <- label_panels(ref_m)
```

# Bridge from REF-paper

One change at a time at one week of extra delay, re-run design: first the
rate scale with the adaptive learning rate (its minimum step drops from 0.1
admissions to 0.001 per 100k, since that minimum is in the working units),
then the constant learning rate with the lowest pooled WIS at each horizon.
REF-op is for orientation only; it differs in several settings.

```{r bridge}
best_const <- sweep_m %>%
  filter(location == "pooled", learn == "exact", settle_days == 14L) %>%
  group_by(horizon) %>%
  slice_max(wis_pct, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(ref = "rate, constant lr (best WIS per horizon)")
bridge <- bind_rows(
  ref_m %>% filter(learn == "exact", settle_days == 14L),
  best_const
) %>%
  mutate(ref = factor(ref, levels = c("REF-paper (count, adaptive)", "rate, adaptive", "rate, constant lr (best WIS per horizon)", "REF-op")))
base_row <- bridge %>% filter(ref == "REF-paper (count, adaptive)") %>% mutate(ref = "base", wis_pct = 0, bias_cal = bias_base)
tbl <- bind_rows(base_row, bridge) %>%
  transmute(config = ref, horizon, cell = sprintf("%+.1f%% / %.3f", wis_pct, bias_cal)) %>%
  tidyr::pivot_wider(names_from = horizon, values_from = cell)
knitr::kable(tbl, caption = "WIS change % (positive is better) / L1 coverage bias, pooled over locations")
knitr::kable(
  best_const %>% select(horizon, lr) %>% mutate(lr = signif(lr, 2)) %>% tidyr::pivot_wider(names_from = horizon, values_from = lr),
  caption = "Constant learning rate with the lowest pooled WIS (rate units per step)"
)
```

# Coverage bias vs WIS, per location

Each thin grey line is one location traced across learning rates; the bold
line is the pooled result, labelled with the learning rate at each decade. Learning rates whose pooled WIS is more than
`r params$wis_max`% worse than the base in a panel are dropped from that panel; the axes are
clipped to the bulk of the per-location points, so a few location lines run
off the edge. The open circle is the base forecast (no calibration) and the
triangles are the reference configs at the same delay.

Per-location coverage comes from about 50 forecasts per horizon, so a
location's L1 bias has a noise floor of a few points even for a perfectly
calibrated forecaster. Read the shape of the pooled line; the grey lines show
spread.

```{r curve-fn}
ACCENT <- "#1f5fa8"
curve_plot <- function(which_learn) {
  d <- sweep_m %>% filter(learn == which_learn)
  keep <- d %>%
    filter(location == "pooled", wis_pct >= -params$wis_max) %>%
    select(horizon, extra, lr)
  d <- d %>% semi_join(keep, by = c("horizon", "extra", "lr")) %>% arrange(lr)
  pooled <- d %>% filter(location == "pooled")
  locs <- d %>% filter(location != "pooled")
  base <- pooled %>% distinct(horizon, extra, bias_base)
  marks <- ref_m %>% filter(learn == which_learn, ref %in% c("REF-paper (count, adaptive)", "REF-op"))
  lab <- pooled %>% filter(abs(log10(lr) - round(log10(lr))) < 1e-9)
  xlim <- c(0, max(stats::quantile(locs$bias_cal, 0.97), base$bias_base) * 1.05)
  ylim <- range(stats::quantile(locs$wis_pct, c(0.03, 0.97)), pooled$wis_pct, marks$wis_pct, 0)
  ggplot() +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(data = locs, aes(bias_cal, wis_pct, group = location), colour = "grey45", alpha = 0.25, linewidth = 0.3) +
    geom_path(data = pooled, aes(bias_cal, wis_pct), colour = ACCENT, linewidth = 1.1) +
    geom_point(data = pooled, aes(bias_cal, wis_pct), colour = ACCENT, size = 1.2) +
    geom_text(data = lab, aes(bias_cal, wis_pct, label = format(lr, scientific = FALSE, drop0trailing = TRUE)),
      colour = "grey15", size = 2.6, vjust = -0.9) +
    geom_point(data = base, aes(bias_base, 0), shape = 21, fill = "white", colour = "black", size = 2.2) +
    geom_point(data = marks, aes(bias_cal, wis_pct, shape = ref), colour = "black", size = 2) +
    scale_shape_manual(values = c(`REF-paper (count, adaptive)` = 2, `REF-op` = 17), name = NULL) +
    facet_grid(horizon ~ extra) +
    coord_cartesian(xlim = xlim, ylim = ylim) +
    labs(x = "L1 coverage bias", y = "WIS change % (positive is better)") +
    theme(legend.position = "bottom")
}
```

## Re-run design (prod)

```{r curve-exact}
curve_plot("exact")
```

## Stateful design

```{r curve-vintage}
curve_plot("vintage")
```

# Pooled curves by delay

The bold lines from above, one colour per extra delay (darker is longer), so
the delays can be compared directly. Solid is the re-run design, dashed the
stateful one.

```{r pooled-overlay, fig.height=6}
DELAY_COLS <- setNames(RColorBrewer::brewer.pal(9, "Blues")[c(4, 6, 8, 9)], levels(sweep_m$extra))
pooled_all <- sweep_m %>%
  filter(location == "pooled") %>%
  semi_join(sweep_m %>% filter(location == "pooled", wis_pct >= -params$wis_max) %>% select(horizon, extra, lr, learn),
    by = c("horizon", "extra", "lr", "learn")) %>%
  arrange(lr) %>%
  mutate(design = factor(LEARN_LABELS[learn], levels = LEARN_LABELS))
ggplot(pooled_all, aes(bias_cal, wis_pct, colour = extra, linetype = design, group = interaction(extra, design))) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_path(linewidth = 0.8) +
  geom_point(data = pooled_all %>% distinct(horizon, bias_base), aes(bias_base, 0), inherit.aes = FALSE,
    shape = 21, fill = "white", colour = "black", size = 2) +
  scale_colour_manual(values = DELAY_COLS, name = NULL) +
  scale_linetype_manual(values = c(`re-run (prod)` = "solid", stateful = "22"), name = NULL) +
  facet_wrap(~horizon, nrow = 1) +
  labs(x = "L1 coverage bias", y = "WIS change % (positive is better)") +
  theme(legend.position = "bottom", legend.box = "vertical")
```

# Tables

Pooled WIS change % (positive is better) / L1 coverage bias for every learning rate, by horizon
(columns) and extra delay (row blocks).

```{r table-fn}
sweep_table <- function(which_learn) {
  sweep_m %>%
    filter(location == "pooled", learn == which_learn) %>%
    transmute(extra, lr = signif(lr, 2), horizon, cell = sprintf("%+.1f / %.3f", wis_pct, bias_cal)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = cell) %>%
    arrange(extra, lr) %>%
    knitr::kable()
}
```

## Re-run design (prod)

```{r table-exact}
sweep_table("exact")
```

## Stateful design

```{r table-vintage}
sweep_table("vintage")
```
