---
title: "E10: covid at the reference configs"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  disease: "covid"
  hub_dir: "../covid19-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Ledger entry **E10** (`notes/calibration-ledger.md`). One axis: the
disease. Both reference configs run on the covid hub's `CMU-TimeSeries`
submissions, learning from NHSN covid vintages the way prod does (E00), and
scored against finalized covid NHSN.

Covid submissions start 2024-11-23, so there is no burn-in season. As in the
covid prod pipeline, both configs run with `burn_in_seasons = character(0)`,
and REF-op runs without its warm start (`slow_init = NULL`), leaving sqrt +
slow term 0.003 + leak 0.1. The tracker therefore starts from zero offsets at
the first round.

Covid is submitted year-round, so the hub round axis has no off-season gap
and `hub_label_seasons()` puts every round in one season. For the per-season
views below, seasons are relabelled as August–July years.

```{r runs}
no_burn_in <- list(burn_in_seasons = character(0), slow_init = NULL)
cals <- list(
  `REF-paper` = do.call(run, c(list("paper"), no_burn_in)),
  `REF-op (no warm start)` = do.call(run, c(list("op"), no_burn_in))
)
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
cals <- purrr::map(cals, function(cal) {
  cal$forecasts <- cal$forecasts %>% mutate(season = season_of(reference_date))
  cal
})
LIVE <- c("2024-2025", "2025-2026")
focus <- "REF-op (no warm start)"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
