---
title: "E10: the finalists on flu and covid"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  disease: "covid"
  hub_dir: "../covid19-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
# The month view is knit several times, so its chunk labels repeat.
options(knitr.duplicate.label = "allow")
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: how do the two finalists, REF-op and sqrt constant 0.018, do on
flu and covid, on the submitted ensemble and on the clean replay?

Part 1 runs them on both diseases and both bases, h0–h3, on identical rows.
Part 2 is the covid ensemble at h−1–h3 with REF-paper for reference.

# Part 1: both diseases, both bases

Setup:

- Bases: the submitted `CMU-TimeSeries` ensemble (hub) and the clean
  `windowed_seasonal` replay, on the (round, location, horizon, level) rows
  both have, h0–h3 (the ensemble's h−1 is a different model).
- Learning truth: exact (re-run every round on the store's NHSN vintages),
  one week of extra delay. Scores against the store's finalized NHSN, states
  only, the two live seasons (August–July).
- Configs: REF-op warm (flu prod) and cold (covid prod); sqrt constant 0.018
  per 100k (E16), cold and warm (`slow_init = "burn_in_quantile"`). Warm
  runs see a 2023-24 burn-in season: for flu the rows both bases have (hub
  submissions and the HHS-stitched backfill, as E07); for covid only the
  replay's backfill, since covid submissions start 2024-11-23. So the covid
  ensemble has no warm runs. Cold runs see the live seasons only.

```{r fin-runs}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
key <- c("reference_date", "location", "horizon", "level")
fin_inputs <- function(d) {
  wi <- ws_inputs(d)
  # Right after ws_inputs(): both read the harness globals that ch_use() sets.
  burn <- ws_burn_in_forecasts(d)
  hub <- suppressMessages(hub_read_forecasts(
    hub_dir = here::here(HUB_DIRS[[d]]), target = c(flu = HUB_FLU_TARGET, covid = HUB_COVID_TARGET)[[d]]
  )) %>% filter(horizon %in% 0:3)
  ws <- bind_rows(burn, wi$fc)
  common <- inner_join(distinct(hub, across(all_of(key))), distinct(ws, across(all_of(key))), by = key)
  fcs <- list(ensemble = semi_join(hub, common, by = key), replay = semi_join(ws, common, by = key))
  warm <- fcs
  if (!any(season_of(fcs$ensemble$reference_date) == "2023-2024")) {
    warm <- list(replay = bind_rows(burn, fcs$replay))
  }
  list(wi = wi, cold = purrr::map(fcs, \(fc) filter(fc, season_of(reference_date) %in% LIVE)), warm = warm)
}
fin_cands <- function(rate_scales) {
  const <- ws_configs(rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
  sqrt_018 <- utils::modifyList(const, list(lr = 0.018, transform = "sqrt"))
  list(
    `REF-op warm` = list(cfg = list(ref = "op"), warm = TRUE),
    `REF-op cold` = list(cfg = list(ref = "op", burn_in_seasons = character(0), slow_init = NULL), warm = FALSE),
    `sqrt 0.018` = list(cfg = sqrt_018, warm = FALSE),
    `sqrt 0.018 warm` = list(cfg = utils::modifyList(sqrt_018, list(burn_in_seasons = "2023-2024", slow_init = "burn_in_quantile")), warm = TRUE)
  )
}
FIN <- purrr::map(c(flu = "flu", covid = "covid"), function(d) {
  inp <- fin_inputs(d)
  cands <- fin_cands(inp$wi$rate_scales)
  grid <- tidyr::expand_grid(base = c("ensemble", "replay"), cand = names(cands)) %>%
    filter(!cands[cand] %>% purrr::map_lgl("warm") | base %in% names(inp$warm))
  cals <- purrr::pmap(grid, function(base, cand) {
    cc <- cands[[cand]]
    fc <- if (cc$warm) inp$warm[[base]] else inp$cold[[base]]
    do.call(cal_run_cached, c(list(fc, inp$wi$truth), cc$cfg, list(learn = "exact", vintages = inp$wi$vintages, workers = params$workers)))
  })
  list(cals = setNames(cals, paste(grid$base, grid$cand, sep = ", ")), grid = grid, cands = names(cands))
})
```

Per disease, a month view for each base. The ensemble and replay rows are
the same (round, location, horizon, level) cells, so the two bases' tables
compare like for like.

```{r fin-month, results="asis"}
for (d in names(FIN)) {
  cat("\n\n", cal_h(2), " ", d, "\n\n", sep = "")
  for (b in c("ensemble", "replay")) {
    mv <- cal_month_view(FIN[[d]]$cals[startsWith(names(FIN[[d]]$cals), paste0(b, ", "))], seasons = LIVE)
    cal_month_section(paste(d, b), level = 3L)
  }
}
```

# Part 2: covid ensemble, h−1–h3

- Disease: covid. Base forecaster: the submitted `CMU-TimeSeries` ensemble.
- Learning truth: NHSN covid vintages, re-run at every round as prod does
  (E00). Scores against finalized covid NHSN.
- No burn-in season (covid submissions start 2024-11-23); both trackers start
  from zero offsets. REF-op runs without its warm start, as in covid prod.
- Rounds 2024-11-30 to 2026-09-19. The 2024-11-23 round is excluded (made
  from a bad NHSN release corrected the next day).
- Covid is submitted year-round; per-season views use August–July seasons.
- Horizons h−1 to h3. Scores: states only (US excluded), the two live
  seasons.

| config | change from REF-paper |
|---|---|
| `REF-paper` | none (counts, adaptive learning rate) |
| `REF-op (no warm start)` | sqrt scale, learning-rate floor, slow term 0.003, leak 0.1; no warm start |
| `sqrt 0.018` | sqrt of the per-100k rate, constant learning rate 0.018 (E16), single term |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
no_burn_in <- list(burn_in_seasons = character(0), slow_init = NULL)
cals <- list(
  `REF-paper` = do.call(run, c(list("paper"), no_burn_in)),
  `REF-op (no warm start)` = do.call(run, c(list("op"), no_burn_in)),
  `sqrt 0.018` = do.call(run, c(list("paper", transform = "sqrt", lr = 0.018, scales = rate_scales), no_burn_in))
)
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
cals <- purrr::map(cals, function(cal) {
  cal$forecasts <- cal$forecasts %>% mutate(season = season_of(reference_date))
  cal
})
LIVE <- c("2024-2025", "2025-2026")
```

```{r p2-month-view, results="asis"}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
cal_month_section("Month view", level = 2L)
```
