---
title: "E10: covid at the reference configs"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  disease: "covid"
  hub_dir: "../covid19-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

One axis: the disease. Both reference configs run on the covid hub's `CMU-TimeSeries`
submissions, learning from NHSN covid vintages the way prod does (E00), and
scored against finalized covid NHSN.

Covid submissions start 2024-11-23, so there is no burn-in season. As in the
covid prod pipeline, both configs run with no burn-in season,
and REF-op runs without its warm start, leaving sqrt + slow term 0.003 +
leak 0.1. The tracker therefore starts from zero offsets at
the first round.

Covid is submitted year-round, so there is no off-season gap between
seasons. For the per-season views below, seasons are August–July years.

# Findings

Numbers are from the render of 30 September 2026, pooled over all rounds
(2024-11-23 to 2026-09-19) and locations; WIS changes are % vs base, positive
is better.

- **The covid base is already calibrated at h1–h3,** so there is little to
  fix there. Its coverage bias is 0.094/0.063/0.043/0.034/0.027 at h−1…h3,
  against about 0.1 at every horizon for flu.
- **h−1 is a clear win.** WIS +9.5% (REF-op) and +9.2% (REF-paper), absolute
  error of the median +18% and +16%. At h−1 the submission comes entirely
  from one revision-aware component model, and h−1 is where the base is
  least calibrated.
- **At h1–h3, REF-op is neutral and REF-paper is harmful.** REF-op: WIS
  −0.1/−0.2/−0.8%, coverage bias slightly worse than base at h2/h3
  (0.037/0.035 vs 0.034/0.027). REF-paper: WIS −5.6/−6.8/−9.4%, coverage bias
  worse than base from h2 on (0.043/0.048). An offset learned where the base
  is calibrated is noise.
- **The losses are in 2025-26 and around September.** REF-op gains
  +1.8/+3.0/+2.4% at h1–h3 in 2024-25 and loses 3.0–4.1% in 2025-26. By month
  at h3, REF-op loses 16% in September and 12% in October (19% of the base
  WIS between them), around the late-summer wave.
- **Implication for prod:** for covid, calibrating h−1 (and perhaps h0) and
  passing h1–h3 through would keep the gain and drop the losses. The covid prod
  pipeline calibrates every horizon today.

```{r runs}
no_burn_in <- list(burn_in_seasons = character(0), slow_init = NULL)
cals <- list(
  `REF-paper` = do.call(run, c(list("paper"), no_burn_in)),
  `REF-op (no warm start)` = do.call(run, c(list("op"), no_burn_in))
)
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
cals <- purrr::map(cals, function(cal) {
  cal$forecasts <- cal$forecasts %>% mutate(season = season_of(reference_date))
  cal
})
LIVE <- c("2024-2025", "2025-2026")
focus <- "REF-op (no warm start)"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
