---
title: "E10: covid at the reference configs"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  disease: "covid"
  hub_dir: "../covid19-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: how do the two reference configs do on covid?

Setup:

- Disease: covid. Base forecaster: the submitted `CMU-TimeSeries` ensemble.
- Learning truth: NHSN covid vintages, re-run at every round as prod does
  (E00). Scores against finalized covid NHSN.
- No burn-in season (covid submissions start 2024-11-23); both trackers start
  from zero offsets. REF-op runs without its warm start, as in covid prod.
- Rounds 2024-11-30 to 2026-09-19. The 2024-11-23 round is excluded (made
  from a bad NHSN release corrected the next day).
- Covid is submitted year-round; per-season views use August–July seasons.
- Horizons h−1 to h3. Headline tables: all hub locations; month view: states
  only.

| config | change from REF-paper |
|---|---|
| `REF-paper` | none (counts, adaptive learning rate) |
| `REF-op (no warm start)` | sqrt scale, learning-rate floor, slow term 0.003, leak 0.1; no warm start |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
no_burn_in <- list(burn_in_seasons = character(0), slow_init = NULL)
cals <- list(
  `REF-paper` = do.call(run, c(list("paper"), no_burn_in)),
  `REF-op (no warm start)` = do.call(run, c(list("op"), no_burn_in))
)
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
cals <- purrr::map(cals, function(cal) {
  cal$forecasts <- cal$forecasts %>% mutate(season = season_of(reference_date))
  cal
})
LIVE <- c("2024-2025", "2025-2026")
focus <- "REF-op (no warm start)"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```

# Findings

There are no uncertainty estimates; a WIS gap under 1 point or a month-avg
coverage gap under 0.01 is read as no clear difference. The pooled headline,
median-error and by-month tables run to 2026-09-19 to include the 2026 summer
wave; those Aug–Sep 2026 rounds are in no per-season table.

- **h−1:** both configs improve WIS by about 10–11% in each season
  (REF-paper 11.6/10.5, REF-op 8.9/10.8) and the absolute error of the median
  by about 20%. Which config is ahead flips between seasons.
- **REF-paper at h1–h3:** −4.5/−7.3/−10.3 pooled, almost all from 2025-26
  (−10.2/−14.4/−19.3). In 2024-25 it is within 0.3 points of the base.
- **REF-op at h1–h3:** 0.0/−0.6/−1.3 pooled; +2.4/+2.8/+1.9 in 2024-25 and
  −3.0/−4.0/−4.3 in 2025-26. No strong result: the sign depends on the
  season.
- **Month-avg coverage (states, h−1…h3):** base 0.148/0.102/0.125/0.143/0.157;
  REF-op 0.123/0.096/0.126/0.147/0.161 (lower at h−1 only); REF-paper
  0.111/0.094/0.126/0.155/0.171 (about 0.01 worse at h2/h3). Season-pooled,
  the base looks calibrated at h1–h3, but pooling lets the two seasons'
  errors cancel: within each season its bias is 0.072–0.078 (2024-25) and
  0.029–0.051 (2025-26).
- **By month at h3:** REF-op −16.5% in September and −12.0% in October
  (together 20% of base WIS); REF-paper −42% to −49% in April–May (about 6%).
