---
title: "E10: the finalists on flu and covid"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  disease: "covid"
  hub_dir: "../covid19-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
# The month view is knit several times, so its chunk labels repeat.
options(knitr.duplicate.label = "allow")
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: how do the two finalists, REF-op and sqrt constant 0.018, do on
flu and covid, on the submitted ensemble and on the clean replay?

Part 1 runs them on both diseases and both bases, h0–h3, on identical rows.
Part 1 also checks whether the finalists rank the same way on both bases.
Part 2 is the covid ensemble at h−1–h3.

# Part 1: both diseases, both bases

Setup:

- Bases: the submitted `CMU-TimeSeries` ensemble (hub) and the clean
  `windowed_seasonal` replay, on the (round, location, horizon, level) rows
  both have, h0–h3 (the ensemble's h−1 is a different model).
- Learning truth: exact (re-run every round on the store's NHSN vintages),
  one week of extra delay. Scores against the store's finalized NHSN, states
  only, the two live seasons (August–July).
- Configs: REF-op warm (flu prod) and cold (covid prod); sqrt constant 0.018
  per 100k (E16), cold and warm (`slow_init = "burn_in_quantile"`). Warm
  runs see a 2023-24 burn-in season: for flu the rows both bases have (hub
  submissions and the HHS-stitched backfill, as E07); for covid only the
  replay's backfill, since covid submissions start 2024-11-23. So the covid
  ensemble has no warm runs. Cold runs see the live seasons only.

```{r fin-runs}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
FIN <- purrr::map(c(flu = "flu", covid = "covid"), \(d) fin_runs(d, workers = params$workers))
```

Per disease, a month view for each base. The ensemble and replay rows are
the same (round, location, horizon, level) cells, so the two bases' tables
compare like for like.

```{r fin-month, results="asis"}
for (d in names(FIN)) {
  cat("\n\n", cal_h(2), " ", d, "\n\n", sep = "")
  for (b in c("ensemble", "replay")) {
    mv <- cal_month_view(FIN[[d]]$cals[startsWith(names(FIN[[d]]$cals), paste0(b, ", "))], seasons = LIVE)
    cal_month_section(paste(d, b), level = 3L)
  }
}
```

## Do the finalists rank the same way on both bases?

Per disease, the candidates run on both bases (covid has no warm runs on the
ensemble), ordered by mean WIS reduction % (positive = calibrated WIS lower
than base) over h0–h3, separately on each base, states only. `rank` 1 is the
largest WIS reduction; `cov rank` 1 is the lowest month-avg L1 coverage
bias. Spearman ρ compares the two bases over all (candidate, horizon) cells
of a season.

```{r fin-rank, results="asis"}
for (d in names(FIN)) {
  mv_all <- cal_month_view(FIN[[d]]$cals, seasons = LIVE)
  wis <- bind_rows(mv_all$month, mv_all$month %>% mutate(season = "both")) %>%
    group_by(config, season, horizon) %>%
    summarize(wis = 100 * (sum(wis_base) - sum(wis_cal)) / sum(wis_base), .groups = "drop")
  cells <- mv_all$cov %>%
    transmute(config, season = as.character(season), horizon, cov = mavg_cal) %>%
    inner_join(wis, by = c("config", "season", "horizon")) %>%
    tidyr::separate(config, c("base", "cand"), sep = ", ", extra = "merge")
  shared <- cells %>% distinct(base, cand) %>% count(cand) %>% filter(n == 2) %>% pull(cand)
  cells <- cells %>% filter(cand %in% shared)
  cat("\n\n", cal_h(3), " ", d, "\n\n", sep = "")
  cells %>%
    group_by(base, season, cand) %>%
    summarize(wis = mean(wis), cov = mean(cov), .groups = "drop") %>%
    group_by(base, season) %>%
    mutate(rank = rank(-wis), cov_rank = rank(cov)) %>%
    ungroup() %>%
    tidyr::pivot_wider(names_from = base, values_from = c(wis, rank, cov, cov_rank)) %>%
    arrange(season, rank_ensemble) %>%
    transmute(season, cand,
      `WIS % ensemble` = round(wis_ensemble, 2), `rank ensemble` = rank_ensemble,
      `WIS % replay` = round(wis_replay, 2), `rank replay` = rank_replay,
      `cov ensemble` = round(cov_ensemble, 3), `cov rank ensemble` = cov_rank_ensemble,
      `cov replay` = round(cov_replay, 3), `cov rank replay` = cov_rank_replay
    ) %>%
    knitr::kable(caption = "Mean over h0–h3, states only. Coverage is month-avg L1 coverage bias (lower is better)") %>%
    print()
  cells %>%
    select(season, cand, horizon, base, wis, cov) %>%
    tidyr::pivot_wider(names_from = base, values_from = c(wis, cov)) %>%
    group_by(season) %>%
    summarize(
      `ρ WIS cells` = cor(wis_ensemble, wis_replay, method = "spearman"),
      `ρ coverage cells` = cor(cov_ensemble, cov_replay, method = "spearman")
    ) %>%
    knitr::kable(digits = 2, caption = "Spearman correlation of ensemble vs replay over (candidate, horizon) cells") %>%
    print()
}
```

# Part 2: covid ensemble, h−1–h3

- Disease: covid. Base forecaster: the submitted `CMU-TimeSeries` ensemble.
- Learning truth: NHSN covid vintages, re-run at every round as prod does
  (E00). Scores against finalized covid NHSN.
- No burn-in season (covid submissions start 2024-11-23); both trackers start
  from zero offsets. REF-op runs without its warm start, as in covid prod.
- Rounds 2024-11-30 to 2026-09-19. The 2024-11-23 round is excluded (made
  from a bad NHSN release corrected the next day).
- Covid is submitted year-round; per-season views use August–July seasons.
- Horizons h−1 to h3. Scores: states only (US excluded), the two live
  seasons.

| config | tracker |
|---|---|
| `REF-op cold` | covid prod: sqrt scale, adaptive learning rate with floor, slow term 0.003, leak 0.1; no warm start |
| `sqrt 0.018` | sqrt of the per-100k rate, constant learning rate 0.018 (E16), single term |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
no_burn_in <- list(burn_in_seasons = character(0), slow_init = NULL)
cals <- list(
  `REF-op cold` = do.call(run, c(list("op"), no_burn_in)),
  `sqrt 0.018` = do.call(run, c(list("paper", transform = "sqrt", lr = 0.018, scales = rate_scales), no_burn_in))
)
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
cals <- purrr::map(cals, function(cal) {
  cal$forecasts <- cal$forecasts %>% mutate(season = season_of(reference_date))
  cal
})
LIVE <- c("2024-2025", "2025-2026")
```

```{r p2-month-view, results="asis"}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
cal_month_section("Month view", level = 2L)
```
