---
title: "E11: constant learning rate vs adaptive"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: does a constant learning rate on the rate scale do better than the
adaptive learning rate?

Setup:

- Disease: flu. Base forecaster: the submitted hub ensemble (spoiled
  submissions dropped).
- Burn-in: 2023-24, used only to warm up the adaptive learning rate. Offsets
  start at zero at the first 2024-25 round unless a config has a warm start.
  The constant-rate config has no learning rate to warm up, so it is
  effectively cold.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay. Scores against finalized truth.
- Seasons scored: 2024-25 and 2025-26. Horizons h−1 to h3. Headline tables:
  all hub locations; month view: states only.

| config | change from REF-paper (counts, adaptive 0.03 × p90 of the last 20 errors) |
|---|---|
| `REF-paper` | none |
| `sqrt, adaptive` | sqrt scale, learning-rate floor 0.001 (E03 `sqrt`) |
| `warm + sqrt, adaptive` | as above, plus a warm start from the burn-in season (E05 `warm + single`) |
| `rate, constant 0.1` | rate scale (per 100k), constant learning rate 0.1 (E09, +1 week, re-run) |
| `REF-op` | the prod config; differs in several settings |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
cals <- list(
  `REF-paper` = run("paper"),
  `sqrt, adaptive` = do.call(run, sqrt_args),
  `warm + sqrt, adaptive` = do.call(run, c(sqrt_args, slow_init = "burn_in_quantile")),
  `rate, constant 0.1` = run("paper", scales = rate_scales, lr = 0.1),
  `REF-op` = run("op")
)
focus <- "rate, constant 0.1"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```

# Findings

There are no uncertainty estimates; a WIS gap under 1 point or a month-avg
coverage gap under 0.01 is read as no clear difference.

- **Constant 0.1 vs `sqrt, adaptive`:** WIS +3.6/+2.4/+0.7/+0.3/−0.5 vs
  +5.1/+4.0/+0.7/−1.3/−1.4 at h−1…h3 pooled. The adaptive tracker is ahead
  at h−1/h0 (1.5/1.6 points), the constant at h2/h3 (1.6/0.9). In 2025-26 the
  gaps are larger in both directions (adaptive +4.6 at h0, constant +2.8 at
  h2). The order depends on horizon.
- **Constant 0.1 vs `warm + sqrt, adaptive`:** the warm tracker is ahead in
  every 2024-25 cell (by 0.8–2.7). In 2025-26 the constant is ahead at h2/h3
  (by 2.5/1.9). This compares a cold config with a warm one.
- **REF-op:** pooled WIS +5.4/+4.9/+2.4/+1.7/+2.2, the highest at every
  horizon, by up to about 2 points over `warm + sqrt, adaptive`. Its 2025-26
  gains at h1–h3 are under 1.5 points.
- **Month-avg coverage (states, h−1…h3):** base 0.134/0.116/0.123/0.137/0.151;
  constant 0.118/0.114/0.127/0.147/0.172; `sqrt, adaptive`
  0.108/0.102/0.117/0.130/0.150; `warm + sqrt, adaptive`
  0.107/0.100/0.116/0.128/0.146; REF-op 0.107/0.091/0.111/0.125/0.136. The
  adaptive configs are within 0.01 of each other at most horizons. About half
  of the constant's h0 gap comes from October 2025, a single sparse month
  (0.3% of base WIS) with spring offsets carried over. Season-pooled the
  constant has the lowest bias (0.017–0.032), but pooling lets errors cancel.
