---
title: "E11: constant learning rate vs adaptive"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

One axis: the learning rate. The constant rate is the corner of E09's
trade-off curve, re-run design, one week of extra delay: 0.1 admissions per
100k per step on the rate scale. Below that rate, coverage bias falls with
little change in WIS. Above it, coverage stops improving and WIS gets worse
at every horizon. It is compared with the adaptive learning rate (0.03 × the
90th percentile of the last 20 rounds' absolute errors) in the configs from
E03 and E05. Everything else is REF-paper, learning from vintages the way
prod does (re-run each week; E00).

- **REF-paper**: counts, adaptive.
- **sqrt, adaptive**: the E03 winner.
- **warm + sqrt, adaptive**: the E05 `warm + single`, the best-calibrated
  adaptive config.
- **rate, constant 0.1**: the E09 corner.
- **REF-op**: prod, for orientation; it differs in several settings.

# Findings

Numbers are from the render of 1 October 2026, pooled over both live
seasons and all locations. WIS reduction % is 100 × (base − calibrated) / base,
so positive = calibrated WIS lower than base; the absolute-error change of the
median is defined the same way. L1 coverage bias: lower is better. The constant run matches E09's table at 0.1, +1 week,
re-run.

- **Against the cold adaptive trackers, the constant rate trades WIS at
  h−1/h0 for better coverage everywhere.** WIS +3.6/+2.4/+0.7/+0.3/−0.5 at
  h−1…h3, against +5.1/+4.0/+0.7/−1.3/−1.4 for `sqrt, adaptive`; coverage
  bias 0.032/0.022/0.018/0.017/0.018 against 0.049/0.032/0.026/0.028/0.026.
  It beats REF-paper on both at every horizon.
- **The warm-started adaptive tracker still beats it at h0–h1.**
  `warm + sqrt, adaptive` has better WIS at h−1–h1 (+5.1/+4.7/+1.3) and
  better coverage at h0–h3 (0.013–0.020). The constant rate keeps an edge only
  in h−1 coverage (0.032 vs 0.046) and is within 0.8 points on WIS at h2/h3.
- **It is the most stable tracker across seasons and months.** In 2025-26 it
  loses 1.8–2.0% at h1–h3, against up to 4.6% for the adaptive sqrt trackers
  (REF-op is the only config that gains there). By month at h3 it has the
  smallest March loss of the plain trackers (16% vs 42–51% for sqrt), but is
  nearly neutral in December–January (about +1%, 71% of the base WIS), where
  the warm start and REF-op gain 4–6%. Its h−1 is the weakest in 2025-26
  (−1.6%).
- **The median gains less near the present.** Absolute error improves
  +6.3/+4.9% at h−1/h0, against about +10/+8% for the sqrt trackers; it is
  better than them at h2 (+1.8%).
- **Untested:** a warm start on the constant-rate tracker. The warm start
  is the largest single gain for the adaptive tracker (E05), and the constant
  rate keeps coverage without a leak.
- **Month view (2026-10-01):** scored within each month (states only), the
  constant rate has the worst coverage, not the best. Month-avg L1 coverage
  bias at h−1…h3: constant 0.1 0.118/0.114/0.127/0.147/0.172, `sqrt,
  adaptive` 0.108/0.102/0.117/0.130/0.150, REF-paper
  0.115/0.102/0.120/0.143/0.168, `warm + sqrt, adaptive`
  0.107/0.100/0.116/0.128/0.146, REF-op 0.107/0.091/0.111/0.125/0.136, base
  0.134/0.116/0.123/0.137/0.151. It is highest of the five at every horizon
  and worse than the base at h1–h3; its h−1 edge over the warm tracker
  disappears.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
cals <- list(
  `REF-paper` = run("paper"),
  `sqrt, adaptive` = do.call(run, sqrt_args),
  `warm + sqrt, adaptive` = do.call(run, c(sqrt_args, slow_init = "burn_in_quantile")),
  `rate, constant 0.1` = run("paper", scales = rate_scales, lr = 0.1),
  `REF-op` = run("op")
)
focus <- "rate, constant 0.1"
```

```{r child=here::here("reports/writeups/calibration_experiments/_standard_views.Rmd")}
```
