---
title: "E11: constant learning rate vs adaptive"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

```{r child=here::here("reports/writeups/calibration_experiments/_load.Rmd")}
```

Question: does a constant learning rate on the rate scale do better than the
adaptive learning rate?

Setup:

- Disease: flu. Base forecaster: the submitted hub ensemble (spoiled
  submissions dropped).
- Burn-in: 2023-24, used only to warm up the adaptive learning rate. Offsets
  start at zero at the first 2024-25 round unless a config has a warm start.
  The constant-rate config has no learning rate to warm up, so it is
  effectively cold.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay. Scores against finalized truth.
- Seasons scored: 2024-25 and 2025-26. Horizons h−1 to h3. States only
  (US excluded).

| config | change from REF-paper (counts, adaptive 0.03 × p90 of the last 20 errors) |
|---|---|
| `REF-paper` | none |
| `sqrt, adaptive` | sqrt scale, learning-rate floor 0.001 (E03 `sqrt`) |
| `warm + sqrt, adaptive` | as above, plus a warm start from the burn-in season (E05 `warm + single`) |
| `rate, constant 0.1` | rate scale (per 100k), constant learning rate 0.1 (E09, +1 week, re-run) |
| `REF-op` | the prod config; differs in several settings |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r runs}
sqrt_args <- list("paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3))
cals <- list(
  `REF-paper` = run("paper"),
  `sqrt, adaptive` = do.call(run, sqrt_args),
  `warm + sqrt, adaptive` = do.call(run, c(sqrt_args, slow_init = "burn_in_quantile")),
  `rate, constant 0.1` = run("paper", scales = rate_scales, lr = 0.1),
  `REF-op` = run("op")
)
```

```{r month-view-data}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
