---
title: "E14: constant learning rate, season by season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

E09 traced how coverage and WIS trade off as the constant learning rate of
the plain tracker changes, pooled over both live seasons. This notebook
splits those curves by season. **Is there a
range of learning rates that is good in both 2024-25 and 2025-26?**

**Setup.** The clean `windowed_seasonal` replay for flu (E12: substitutions
off, all 53 locations, the 55 hub rounds, h0–h3; 27 rounds in 2024-25 and
28 in 2025-26), learning from data as published each week and re-run from
scratch each week as prod does, one week of extra revision delay (E09's
best re-run setting), no burn-in season. One axis: the constant learning
rate of the plain tracker on the rate scale (admissions per 100k per step),
12 values from 0.0032 to 1.8, a quarter decade apart. This is E11's
config with its learning rate swept; 0.1 is E11 itself.

Reference points, run on the same data:

- **REF-op cold**: prod's tracker without its warm start.
- **sqrt, adaptive**: the plain tracker on the square-root scale with the
  adaptive learning rate (E03's winner; E13's "E05 single cold").

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so
**positive = calibrated WIS lower than base**. "L1 coverage bias" (lower is
better) is the mean over the 23 levels of |empirical coverage − level|. Each season is
scored on its own rounds; the tracker still runs over both seasons, so
2025-26 starts from the offsets 2024-25 left behind. Results lead with states
only (US excluded), since US alone is about 45% of the all-locations WIS.

# Findings

Numbers are from the render of 1 October 2026, states only, WIS reduction %
(positive = calibrated WIS lower than base) / L1 coverage bias (lower is better). All-locations numbers point the
same way.

- **Yes: 0.018–0.032 per 100k is good in both seasons.** At those rates WIS
  improves at every horizon in both seasons (2024-25 +0.2 to +1.0, 2025-26
  +0.8 to +1.9) and coverage bias falls in both. At 0.056 WIS is still
  non-negative everywhere (2024-25 h3 −0.0). From 0.1 up WIS gets worse at
  h1–h3 in both seasons, and past about 0.18 it falls fast.
- **The seasons disagree about coverage, not WIS.** In 2025-26 coverage
  bias reaches its floor (about 0.01–0.015) by 0.032. In 2024-25 it keeps
  falling until about 0.3, and 0.032 leaves it at 0.10–0.12 (base
  0.13–0.17). The extra coverage 2024-25 gets from a larger rate costs WIS
  in both seasons. Part of the gap is the cold start: with no burn-in, the
  tracker starts at zero in November 2024, so 2024-25 is also the season it
  is still learning in.
- **The knee is lower than E09's.** E09 (submitted ensemble, 2023-24
  burn-in) put the knee at 0.05–0.1. Here the WIS-best rate is 0.018–0.056
  at every horizon and season, and E11's 0.1 is already past it: −0.5/−0.9
  at h2/h3 in 2024-25 and −1.0/−1.2 in 2025-26.
- **The references.** REF-op cold has the most WIS gain at h0 (+2.2/+1.8)
  but keeps 2024-25's coverage bias near the base (0.106–0.124). The adaptive
  sqrt tracker reaches a 2025-26 coverage bias of 0.008–0.016 with +2.2 at
  h0, but loses WIS at h2/h3 in 2024-25 (−1.4/−2.6). Neither is good in both
  seasons at every horizon; a constant rate of 0.018–0.032 is, at a smaller
  gain.
- **The 50% interval.** At 0.018–0.032 the 50% interval covers 0.47–0.51
  at h0/h2, close to nominal; 0.1 and above under-cover it in 2024-25
  (0.39–0.45 at h2).

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
LRS <- signif(10^seq(-2.5, 0.25, by = 0.25), 2)
all_cfgs <- ws_configs(wi$rate_scales, burn_in = TRUE)
sweep_cfgs <- purrr::map(LRS, function(lr) utils::modifyList(all_cfgs[["E11 constant 0.1"]], list(lr = lr)))
ref_cfgs <- list(`REF-op cold` = all_cfgs[["REF-op cold"]], `sqrt, adaptive` = all_cfgs[["E05 single cold"]])
# Cached under cache/calibration/experiments/runs; lr 0.1 and the references are E12/E13's runs.
run <- function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
sweep <- purrr::map(sweep_cfgs, run)
refs <- purrr::map(ref_cfgs, run)
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
# ws_scores() per location subset; wis_pct is the WIS reduction % vs base (positive = calibrated WIS lower).
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
label <- function(df) {
  df %>% mutate(
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
}
sweep_m <- purrr::map2(sweep, LRS, function(cal, lr) score(cal) %>% mutate(lr = lr)) %>% bind_rows() %>% label()
ref_m <- purrr::imap(refs, function(cal, nm) score(cal) %>% mutate(ref = nm)) %>% bind_rows() %>% label()
SEASON_COLS <- c(`2024-2025` = "#0072B2", `2025-2026` = "#D55E00", `both seasons` = "grey45")
```

# Coverage bias vs WIS, by season

One line per season, tracing the learning rate from 0.0032 (near the
base, the open circle) to 1.8. Labels mark every half decade. The grey
dashed line pools both seasons, as E09 did. Triangles are REF-op cold,
squares the adaptive sqrt tracker, in the season's colour. The y axis is
clipped at −10%.

```{r curve-fn}
LABELLED <- c(0.01, 0.032, 0.1, 0.32, 1)
curve_plot <- function(g) {
  d <- sweep_m %>% filter(geos == g) %>% arrange(lr)
  r <- ref_m %>% filter(geos == g)
  base <- d %>% distinct(h, season, cal_err_base)
  ggplot(d, aes(cal_err_cal, wis_pct, colour = season)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(aes(linetype = season == "both seasons"), linewidth = 0.8) +
    geom_point(size = 1) +
    geom_text(data = d %>% filter(lr %in% LABELLED, season != "both seasons"),
      aes(label = format(lr, drop0trailing = TRUE)), size = 2.6, vjust = -0.8, show.legend = FALSE) +
    geom_point(data = base, aes(cal_err_base, 0), shape = 21, fill = "white", size = 2.2) +
    geom_point(data = r %>% filter(season != "both seasons"), aes(shape = ref), size = 2.4) +
    scale_colour_manual(values = SEASON_COLS, name = NULL) +
    scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "22"), guide = "none") +
    scale_shape_manual(values = c(`REF-op cold` = 17, `sqrt, adaptive` = 15), name = NULL) +
    facet_wrap(~h, nrow = 1) +
    coord_cartesian(ylim = c(max(-10, min(d$wis_pct)), max(c(d$wis_pct, r$wis_pct)) + 0.5)) +
    labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)", title = g) +
    theme(legend.position = "bottom")
}
```

## States only

```{r curve-states}
curve_plot(GEOS[["states"]])
```

## All locations

```{r curve-all}
curve_plot(GEOS[["all"]])
```

# Each metric against the learning rate

The same runs with the learning rate on the x axis, so the seasons can be
compared at the same rate. Top: WIS reduction % (positive = calibrated WIS lower than
base). Bottom: L1 coverage bias (lower is better), with each season's base as a dotted line. States only.

```{r vs-lr, fig.height=6.5}
long <- sweep_m %>%
  filter(geos == GEOS[["states"]], season != "both seasons") %>%
  transmute(h, season, lr, `WIS reduction % (positive = calibrated WIS lower than base)` = wis_pct, `L1 coverage bias (lower is better)` = cal_err_cal) %>%
  tidyr::pivot_longer(c(`WIS reduction % (positive = calibrated WIS lower than base)`, `L1 coverage bias (lower is better)`), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("WIS reduction % (positive = calibrated WIS lower than base)", "L1 coverage bias (lower is better)")))
base_lines <- sweep_m %>%
  filter(geos == GEOS[["states"]], season != "both seasons") %>%
  distinct(h, season, value = cal_err_base) %>%
  mutate(metric = factor("L1 coverage bias (lower is better)", levels = levels(long$metric)))
zero <- tibble(metric = factor("WIS reduction % (positive = calibrated WIS lower than base)", levels = levels(long$metric)), value = 0)
ggplot(long, aes(lr, value, colour = season)) +
  geom_hline(data = zero, aes(yintercept = value), colour = "grey60", linewidth = 0.3) +
  geom_hline(data = base_lines, aes(yintercept = value, colour = season), linetype = "dotted") +
  geom_line() +
  geom_point(size = 1) +
  scale_x_log10(breaks = c(0.003, 0.01, 0.03, 0.1, 0.3, 1), labels = c("0.003", "0.01", "0.03", "0.1", "0.3", "1")) +
  scale_colour_manual(values = SEASON_COLS, name = NULL) +
  facet_grid(metric ~ h, scales = "free_y", labeller = labeller(metric = label_wrap_gen(28))) +
  labs(x = "constant learning rate (per 100k per step)", y = NULL) +
  theme(legend.position = "bottom")
```

# Tables

WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage
bias (lower is better), per learning rate (rows) and season × horizon
(columns). The base row shows the base's coverage bias. The
references follow the sweep.

```{r tables}
season_table <- function(g) {
  rows <- bind_rows(
    sweep_m %>% filter(lr == LRS[1]) %>% mutate(config = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    sweep_m %>% mutate(config = format(lr, drop0trailing = TRUE)),
    ref_m %>% mutate(config = ref)
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(config, levels = unique(config)))
  rows %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

## States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r table-states-month}
mv <- cal_month_view(c(setNames(sweep, paste("rate", format(LRS, drop0trailing = TRUE))), refs), seasons = LIVE)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

## All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```

## 50% and 90% interval coverage, states only

E12 found the constant rate at 0.1 under-covers the 50% interval. Coverage
of the 50% / 90% intervals, per season, at h0 and h2.

```{r table-cov}
bind_rows(
  sweep_m %>% filter(lr == LRS[1]) %>% mutate(config = "base", cov50_cal = cov50_base, cov90_cal = cov90_base),
  sweep_m %>% mutate(config = format(lr, drop0trailing = TRUE)),
  ref_m %>% mutate(config = ref)
) %>%
  filter(geos == GEOS[["states"]], h %in% c("h0", "h2"), season != "both seasons") %>%
  mutate(config = factor(config, levels = unique(config))) %>%
  transmute(config, col = paste(season, h), cell = sprintf("%.2f / %.2f", cov50_cal, cov90_cal)) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  arrange(config) %>%
  knitr::kable()
```

```{r month-view-cfgs}
# The good range (0.018–0.032), the knee and E11's 0.1, and the two references.
mv_plot_cfgs <- c("rate 0.018", "rate 0.032", "rate 0.056", "rate 0.1", "REF-op cold", "sqrt, adaptive")
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
