---
title: "E14: constant learning rate, season by season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: how do WIS and coverage change with a constant learning rate,
season by season?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (E12:
  substitutions off, all 53 locations, the 55 hub rounds; 27 in 2024-25 and
  28 in 2025-26). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Cold start: no burn-in season, offsets start at zero in November 2024. The
  tracker runs over both seasons, so 2025-26 starts from the offsets 2024-25
  left behind; each season is scored on its own rounds.
- Swept: the constant learning rate of the plain tracker on the rate scale
  (admissions per 100k per step), 12 values from 0.0032 to 1.8, a quarter
  decade apart. 0.1 is E11's config, though E11 used the submitted ensemble.
- References on the same data: `REF-op cold` (prod's tracker without its warm
  start) and `sqrt, adaptive` (E13's "E05 single cold").
- Scores: states only (US excluded; it is about 45% of the all-locations
  WIS).

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
LRS <- signif(10^seq(-2.5, 0.25, by = 0.25), 2)
all_cfgs <- ws_configs(wi$rate_scales, burn_in = TRUE)
sweep_cfgs <- purrr::map(LRS, function(lr) utils::modifyList(all_cfgs[["E11 constant 0.1"]], list(lr = lr)))
ref_cfgs <- list(`REF-op cold` = all_cfgs[["REF-op cold"]], `sqrt, adaptive` = all_cfgs[["E05 single cold"]])
# Cached under cache/calibration/experiments/runs; lr 0.1 and the references are E12/E13's runs.
run <- function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
sweep <- purrr::map(sweep_cfgs, run)
refs <- purrr::map(ref_cfgs, run)
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)")
SEASONS <- c(LIVE, "both")
# ws_scores() per location subset; wis_pct is the WIS reduction % vs base (positive = calibrated WIS lower).
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
label <- function(df) {
  df %>% mutate(
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
}
sweep_m <- purrr::map2(sweep, LRS, function(cal, lr) score(cal) %>% mutate(lr = lr)) %>% bind_rows() %>% label()
ref_m <- purrr::imap(refs, function(cal, nm) score(cal) %>% mutate(ref = nm)) %>% bind_rows() %>% label()
SEASON_COLS <- c(`2024-2025` = "#0072B2", `2025-2026` = "#D55E00", `both seasons` = "grey45")
```

# Coverage bias vs WIS, by season

One line per season, tracing the learning rate from 0.0032 (near the
base, the open circle) to 1.8. Labels mark every half decade. The grey
dashed line pools both seasons, as E09 did. Triangles are REF-op cold,
squares the adaptive sqrt tracker, in the season's colour. The y axis is
clipped at −10%.

```{r curve-fn}
LABELLED <- c(0.01, 0.032, 0.1, 0.32, 1)
curve_plot <- function(g) {
  d <- sweep_m %>% filter(geos == g) %>% arrange(lr)
  r <- ref_m %>% filter(geos == g)
  base <- d %>% distinct(h, season, cal_err_base)
  ggplot(d, aes(cal_err_cal, wis_pct, colour = season)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(aes(linetype = season == "both seasons"), linewidth = 0.8) +
    geom_point(size = 1) +
    geom_text(data = d %>% filter(lr %in% LABELLED, season != "both seasons"),
      aes(label = format(lr, drop0trailing = TRUE)), size = 2.6, vjust = -0.8, show.legend = FALSE) +
    geom_point(data = base, aes(cal_err_base, 0), shape = 21, fill = "white", size = 2.2) +
    geom_point(data = r %>% filter(season != "both seasons"), aes(shape = ref), size = 2.4) +
    scale_colour_manual(values = SEASON_COLS, name = NULL) +
    scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "22"), guide = "none") +
    scale_shape_manual(values = c(`REF-op cold` = 17, `sqrt, adaptive` = 15), name = NULL) +
    facet_wrap(~h, nrow = 1) +
    coord_cartesian(ylim = c(max(-10, min(d$wis_pct)), max(c(d$wis_pct, r$wis_pct)) + 0.5)) +
    labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)", title = g) +
    theme(legend.position = "bottom")
}
```

```{r curve-states}
curve_plot(GEOS[["states"]])
```

# Each metric against the learning rate

The same runs with the learning rate on the x axis, so the seasons can be
compared at the same rate. Top: WIS reduction % (positive = calibrated WIS lower than
base). Bottom: L1 coverage bias (lower is better), with each season's base as a dotted line. States only.

```{r vs-lr, fig.height=6.5}
long <- sweep_m %>%
  filter(geos == GEOS[["states"]], season != "both seasons") %>%
  transmute(h, season, lr, `WIS reduction % (positive = calibrated WIS lower than base)` = wis_pct, `L1 coverage bias (lower is better)` = cal_err_cal) %>%
  tidyr::pivot_longer(c(`WIS reduction % (positive = calibrated WIS lower than base)`, `L1 coverage bias (lower is better)`), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("WIS reduction % (positive = calibrated WIS lower than base)", "L1 coverage bias (lower is better)")))
base_lines <- sweep_m %>%
  filter(geos == GEOS[["states"]], season != "both seasons") %>%
  distinct(h, season, value = cal_err_base) %>%
  mutate(metric = factor("L1 coverage bias (lower is better)", levels = levels(long$metric)))
zero <- tibble(metric = factor("WIS reduction % (positive = calibrated WIS lower than base)", levels = levels(long$metric)), value = 0)
ggplot(long, aes(lr, value, colour = season)) +
  geom_hline(data = zero, aes(yintercept = value), colour = "grey60", linewidth = 0.3) +
  geom_hline(data = base_lines, aes(yintercept = value, colour = season), linetype = "dotted") +
  geom_line() +
  geom_point(size = 1) +
  scale_x_log10(breaks = c(0.003, 0.01, 0.03, 0.1, 0.3, 1), labels = c("0.003", "0.01", "0.03", "0.1", "0.3", "1")) +
  scale_colour_manual(values = SEASON_COLS, name = NULL) +
  facet_grid(metric ~ h, scales = "free_y", labeller = labeller(metric = label_wrap_gen(28))) +
  labs(x = "constant learning rate (per 100k per step)", y = NULL) +
  theme(legend.position = "bottom")
```

```{r month-view-cfgs}
mv <- cal_month_view(c(setNames(sweep, paste("rate", format(LRS, drop0trailing = TRUE))), refs), seasons = LIVE)
# The plots show the rates around the elbow and REF-op cold; the tables cover every rate.
mv_plot_cfgs <- c("rate 0.018", "rate 0.032", "rate 0.056", "rate 0.1", "REF-op cold")
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
