---
title: "E15: where the WIS gains come from"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: for each finalist, how does the WIS change split across quantile
levels, how large are the offsets next to the base forecast's spread, and
how much of the change comes from moving the median versus changing widths?

Setup:

- Diseases: flu and covid, one section each.
- Base forecaster: the clean `windowed_seasonal` replay, on the (round,
  location, horizon, level) rows the submitted ensemble also has, h0–h3
  (E10's replay runs).
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Scores against finalized truth over 2024-25 and 2025-26, pooled; states
  only (US excluded).
- Configs, E10's finalists:
  - **REF-op warm**: flu prod; warm start from the 2023-24 burn-in season.
  - **REF-op cold**: covid prod; no burn-in season.
  - **sqrt 0.018**: constant learning rate 0.018 on the square root of
    admissions per 100k (E16), cold.
  - **sqrt 0.018 warm**: the same with the burn-in season and a warm start.

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base.

**Three ways to cut the WIS reduction.**

1. **By quantile level.** WIS is the average pinball loss over the 23
   quantile levels, so the total change splits exactly into one piece per
   level. A level's piece is its change in pinball loss as a share of the
   base WIS; the pieces add up to the total.
2. **Offset size against the forecast's own spread.** For each forecast,
   the offset at a level is `calibrated quantile − base quantile` (in
   admissions, after every projection step). It is compared with the base
   forecast's 90% interval width (q95 − q05) and 50% interval width
   (q75 − q25). A ratio of 0.1 means the quantile moved by a tenth of the
   interval width.
3. **Median shift vs width change.** Each calibrated forecast is split into
   two parts and each part is scored on its own:
   - **median shift only**: every base quantile moves by the same amount,
     the calibrated median minus the base median. The base forecast's
     shape and widths are kept.
   - **width change only**: the calibrated forecast is moved so that its
     median sits on the base median. The calibrated widths and shape are
     kept, the base centre is kept.

   The interaction is the full change minus the two parts. Both parts are
   floored at zero, as the calibrated forecast is.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
FIN_COLS <- c(`REF-op warm` = "#000000", `REF-op cold` = "#D55E00", `sqrt 0.018` = "#0072B2", `sqrt 0.018 warm` = "#56B4E9")
fin_fc <- function(d) {
  f <- fin_runs(d, bases = "replay", workers = params$workers)
  purrr::imap(f$cals, \(cal, nm) cal$forecasts %>% mutate(config = sub("^replay, ", "", nm))) %>%
    bind_rows() %>%
    mutate(season = season_of(reference_date), config = factor(config, levels = names(FIN_COLS))) %>%
    filter(season %in% LIVE, !is_burn_in, !is.na(truth), !is.na(value_base), location != "US")
}
FC <- purrr::map(c(flu = "flu", covid = "covid"), fin_fc)
```

```{r fns}
# 1. Each level's change in pinball loss as a share of the base WIS.
by_level <- function(fc) {
  fc %>%
    group_by(config, horizon, level) %>%
    summarize(base = sum(pinball(truth, value_base, level)), cal = sum(pinball(truth, value_cal, level)), .groups = "drop") %>%
    group_by(config, horizon) %>%
    mutate(contrib = 100 * (base - cal) / sum(base), total = sum(contrib)) %>%
    ungroup()
}
level_plot <- function(bl) {
  d <- bl %>% mutate(panel = sprintf("h%d, total %+.1f%%", horizon, total))
  ggplot(d, aes(factor(level), contrib, fill = config)) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_col(width = 0.75) +
    facet_wrap(~ paste(config, panel, sep = ": "), ncol = 4) +
    scale_fill_manual(values = FIN_COLS, guide = "none") +
    scale_x_discrete(labels = function(x) ifelse(x %in% c("0.01", "0.05", "0.25", "0.5", "0.75", "0.95", "0.99"), x, "")) +
    labs(x = "quantile level", y = "share of WIS reduction, % of base WIS\n(positive = calibrated WIS lower than base)") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 7))
}
group_tbl <- function(bl) {
  grouped <- bl %>%
    mutate(
      band = case_when(level <= 0.05 | level >= 0.95 ~ "tails", level <= 0.3 | level >= 0.7 ~ "shoulders", TRUE ~ "centre"),
      side = case_when(level < 0.5 ~ "lower", level > 0.5 ~ "upper", TRUE ~ "median")
    )
  bind_rows(
    grouped %>% group_by(config, horizon, part = band) %>% summarize(v = sum(contrib), .groups = "drop"),
    grouped %>% group_by(config, horizon, part = paste(side, "side")) %>% summarize(v = sum(contrib), .groups = "drop"),
    grouped %>% group_by(config, horizon) %>% summarize(v = sum(contrib), .groups = "drop") %>% mutate(part = "total")
  ) %>%
    mutate(part = factor(part, levels = c("tails", "shoulders", "centre", "lower side", "median side", "upper side", "total"),
      labels = c("tails", "shoulders", "centre", "below the median", "the median level", "above the median", "total"))) %>%
    transmute(config, part, horizon = paste0("h", horizon), v = sprintf("%+.2f", v)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = v) %>%
    arrange(config, part)
}
# 2. Signed offset ÷ the base 90% interval width, per forecast and level.
offset_plot <- function(fc) {
  ref <- levels(droplevels(fc$config))[1]
  widths <- fc %>%
    filter(config == ref, level %in% c(0.05, 0.95)) %>%
    select(location, horizon, reference_date, level, value_base) %>%
    tidyr::pivot_wider(names_from = level, values_from = value_base, names_prefix = "q") %>%
    transmute(location, horizon, reference_date, w90 = q0.95 - q0.05)
  off_q <- fc %>%
    inner_join(widths, by = c("location", "horizon", "reference_date")) %>%
    filter(w90 > 0) %>%
    mutate(r90 = (value_cal - value_base) / w90) %>%
    group_by(config, horizon, level) %>%
    summarize(med = median(r90), lo = quantile(r90, 0.1), hi = quantile(r90, 0.9), .groups = "drop")
  ggplot(off_q, aes(level, med, colour = config, fill = config)) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.12, colour = NA) +
    geom_line() +
    geom_point(size = 0.8) +
    facet_wrap(~ paste0("h", horizon), nrow = 2) +
    scale_colour_manual(values = FIN_COLS, name = NULL, aesthetics = c("colour", "fill")) +
    labs(x = "quantile level", y = "offset / base (q95 − q05)") +
    theme(legend.position = "bottom")
}
# 3. Score the median shift and the width change on their own.
split_tbl <- function(fc) {
  med <- fc %>%
    filter(level == 0.5) %>%
    select(config, location, horizon, reference_date, med_base = value_base, med_cal = value_cal)
  scores <- fc %>%
    left_join(med, by = c("config", "location", "horizon", "reference_date")) %>%
    mutate(median_only = pmax(value_base + (med_cal - med_base), 0), width_only = pmax(med_base + (value_cal - med_cal), 0)) %>%
    group_by(config, horizon) %>%
    summarize(
      base = sum(pinball(truth, value_base, level)),
      full = sum(pinball(truth, value_cal, level)),
      `median shift only` = sum(pinball(truth, median_only, level)),
      `width change only` = sum(pinball(truth, width_only, level)),
      .groups = "drop"
    ) %>%
    tidyr::pivot_longer(c(full, `median shift only`, `width change only`), names_to = "variant", values_to = "wis") %>%
    mutate(pct = 100 * (base - wis) / base)
  bind_rows(
    scores %>% select(config, variant, horizon, pct),
    scores %>%
      group_by(config, horizon) %>%
      summarize(pct = pct[variant == "full"] - pct[variant == "median shift only"] - pct[variant == "width change only"], .groups = "drop") %>%
      mutate(variant = "interaction (full − the two parts)")
  ) %>%
    mutate(variant = factor(variant, levels = c("full", "median shift only", "width change only", "interaction (full − the two parts)"))) %>%
    transmute(config, variant, horizon = paste0("h", horizon), pct = sprintf("%+.2f", pct)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = pct) %>%
    arrange(config, variant)
}
```

```{r sections, results="asis", fig.height=9}
for (d in names(FC)) {
  fc <- FC[[d]]
  bl <- by_level(fc)
  cat("\n\n# ", d, "\n\n", sep = "")
  cat("\n\n## WIS reduction by quantile level\n\n")
  cat("Each bar is one level's share of the total WIS reduction %; the bars in a panel add up to the panel's total, printed in the strip.\n\n")
  print(level_plot(bl))
  cat("\n\nThe same pieces summed over groups of levels: **tails** are the six levels at or beyond 0.05/0.95, **shoulders** the ten levels 0.1–0.3 and 0.7–0.9, **centre** the seven levels 0.35–0.65. Below and above the median are split, to show whether the gain comes from moving the lower or the upper side.\n\n")
  print(knitr::kable(group_tbl(bl), caption = paste0(d, ": WIS reduction % by group of levels (positive = calibrated WIS lower than base)")))
  cat("\n\n## Offset size against the forecast's spread\n\n")
  cat("Offsets are signed: positive means the calibrated quantile is above the base one. The line is the median over forecasts (location × round) and the band the middle 80%, offset ÷ base 90% interval width. Forecasts with a zero-width base interval are left out.\n\n")
  print(offset_plot(fc))
  cat("\n\n## Median shift vs width change\n\n")
  print(knitr::kable(split_tbl(fc), caption = paste0(d, ": WIS reduction % (positive = calibrated WIS lower than base)")))
}
```
