---
title: "E15: where the WIS gains come from"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 10, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: how does the WIS change split across quantile levels, how large
are the offsets next to the base forecast's spread, and how much of the
change comes from moving the median versus changing widths?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (E12:
  substitutions off, all 53 locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay. Cold start, no burn-in season.
- Scores against finalized truth over 2024-25 and 2025-26, pooled; states
  only (US excluded).
- Configs, both from E12:
  - **REF-op cold**: prod's tracker (sqrt, adaptive rate, slow term, leak)
    without its warm start.
  - **E11 constant 0.1**: the plain tracker on the rate scale (per 100k),
    constant learning rate 0.1.

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

**Three ways to cut the WIS reduction.**

1. **By quantile level.** WIS is the average pinball loss over the 23
   quantile levels, so the total change splits exactly into one piece per
   level. A level's piece is its change in pinball loss as a share of the
   base WIS; the pieces add up to the total.
2. **Offset size against the forecast's own spread.** For each forecast,
   the offset at a level is `calibrated quantile − base quantile` (in
   admissions, after every projection step). It is compared with the base
   forecast's 90% interval width (q95 − q05) and 50% interval width
   (q75 − q25). A ratio of 0.1 means the quantile moved by a tenth of the
   interval width.
3. **Median shift vs width change.** Each calibrated forecast is split into
   two parts and each part is scored on its own:
   - **median shift only**: every base quantile moves by the same amount,
     the calibrated median minus the base median. The base forecast's
     shape and widths are kept.
   - **width change only**: the calibrated forecast is moved so that its
     median sits on the base median. The calibrated widths and shape are
     kept, the base centre is kept.

   The interaction is the full change minus the two parts. Both parts are
   floored at zero, as the calibrated forecast is.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
cfgs <- ws_configs(wi$rate_scales, burn_in = FALSE)[c("REF-op cold", "E11 constant 0.1")]
cals <- purrr::map(cfgs, function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
GEOS <- c(states = "states only (US excluded)")
fc <- purrr::imap(cals, function(cal, nm) cal$forecasts %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), config = factor(config, levels = names(cals))) %>%
  filter(season %in% LIVE, !is.na(truth), !is.na(value_base))
fc <- fc %>%
  filter(location != "US") %>%
  mutate(geos = "states") %>%
  mutate(geos = factor(geos, levels = names(GEOS), labels = GEOS))
```

# 1. WIS reduction by quantile level

Each bar is one level's share of the total WIS reduction % (positive =
calibrated WIS lower than base); the bars in a panel add up to the panel's total, printed in the
strip.

```{r by-level}
by_level <- fc %>%
  group_by(geos, config, horizon, level) %>%
  summarize(base = sum(pinball(truth, value_base, level)), cal = sum(pinball(truth, value_cal, level)), .groups = "drop") %>%
  group_by(geos, config, horizon) %>%
  mutate(contrib = 100 * (base - cal) / sum(base), level_pct = 100 * (base - cal) / base, total = sum(contrib)) %>%
  ungroup()
level_plot <- function(g) {
  d <- by_level %>%
    filter(geos == g) %>%
    mutate(panel = sprintf("h%d, total %+.1f%%", horizon, total))
  ggplot(d, aes(factor(level), contrib, fill = config)) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_col(position = position_dodge(width = 0.8), width = 0.75) +
    facet_wrap(~ paste(config, panel, sep = ": "), ncol = 4) +
    scale_fill_manual(values = c(`REF-op cold` = "#D55E00", `E11 constant 0.1` = "#0072B2"), guide = "none") +
    scale_x_discrete(labels = function(x) ifelse(x %in% c("0.01", "0.05", "0.25", "0.5", "0.75", "0.95", "0.99"), x, "")) +
    labs(x = "quantile level", y = "share of WIS reduction, % of base WIS\n(positive = calibrated WIS lower than base)", title = g) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 7))
}
```

```{r by-level-states, fig.height=6}
level_plot(GEOS[["states"]])
```

## Grouped: tails, shoulders, centre

The same pieces summed over groups of levels: **tails** are the six levels
at or beyond 0.05/0.95, **shoulders** the ten levels 0.1–0.3 and 0.7–0.9,
**centre** the seven levels 0.35–0.65. Below and above the median are
split, to show whether the gain comes from moving the lower or the upper
side.

```{r by-group}
grouped <- by_level %>%
  mutate(
    band = case_when(level <= 0.05 | level >= 0.95 ~ "tails", level <= 0.3 | level >= 0.7 ~ "shoulders", TRUE ~ "centre"),
    side = case_when(level < 0.5 ~ "lower", level > 0.5 ~ "upper", TRUE ~ "median")
  )
group_tbl <- function(g) {
  bind_rows(
    grouped %>% group_by(config, horizon, part = band) %>% filter(geos == g) %>% summarize(v = sum(contrib), .groups = "drop"),
    grouped %>% group_by(config, horizon, part = paste(side, "side")) %>% filter(geos == g) %>% summarize(v = sum(contrib), .groups = "drop"),
    grouped %>% group_by(config, horizon) %>% filter(geos == g) %>% summarize(v = sum(contrib), .groups = "drop") %>% mutate(part = "total")
  ) %>%
    mutate(part = factor(part, levels = c("tails", "shoulders", "centre", "lower side", "median side", "upper side", "total"),
      labels = c("tails", "shoulders", "centre", "below the median", "the median level", "above the median", "total"))) %>%
    transmute(config, part, horizon = paste0("h", horizon), v = sprintf("%+.2f", v)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = v) %>%
    arrange(config, part)
}
knitr::kable(group_tbl(GEOS[["states"]]), caption = "States only: WIS reduction % by group of levels (positive = calibrated WIS lower than base)")
```

# 2. Offset size against the forecast's spread

```{r offsets}
widths <- fc %>%
  filter(config == names(cals)[1], level %in% c(0.05, 0.25, 0.75, 0.95)) %>%
  select(geos, location, horizon, reference_date, level, value_base) %>%
  tidyr::pivot_wider(names_from = level, values_from = value_base, names_prefix = "q") %>%
  transmute(geos, location, horizon, reference_date, w90 = q0.95 - q0.05, w50 = q0.75 - q0.25)
off <- fc %>%
  inner_join(widths, by = c("geos", "location", "horizon", "reference_date")) %>%
  filter(w90 > 0, w50 > 0) %>%
  mutate(offset = value_cal - value_base, r90 = offset / w90, r50 = offset / w50)
n_dropped <- fc %>% distinct(geos, location, horizon, reference_date) %>% nrow() - off %>% distinct(geos, location, horizon, reference_date) %>% nrow()
```

Offsets are signed: positive means the calibrated quantile is above the
base one. The line is the median over forecasts (location × round) and the
band the middle 80%, offset ÷ base 90% interval width, states only.
`r n_dropped` forecast rows with a zero-width base interval are left out.

```{r offset-plot, fig.height=6}
off_q <- off %>%
  filter(geos == GEOS[["states"]]) %>%
  group_by(config, horizon, level) %>%
  summarize(med = median(r90), lo = quantile(r90, 0.1), hi = quantile(r90, 0.9), .groups = "drop")
ggplot(off_q, aes(level, med, colour = config, fill = config)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line() +
  geom_point(size = 0.8) +
  facet_wrap(~ paste0("h", horizon), nrow = 1) +
  scale_colour_manual(values = c(`REF-op cold` = "#D55E00", `E11 constant 0.1` = "#0072B2"), name = NULL) +
  scale_fill_manual(values = c(`REF-op cold` = "#D55E00", `E11 constant 0.1` = "#0072B2"), name = NULL) +
  labs(x = "quantile level", y = "offset / base (q95 − q05)") +
  theme(legend.position = "bottom")
```

# 3. Median shift vs width change

```{r split}
med <- fc %>%
  filter(level == 0.5) %>%
  select(geos, config, location, horizon, reference_date, med_base = value_base, med_cal = value_cal)
split <- fc %>%
  left_join(med, by = c("geos", "config", "location", "horizon", "reference_date")) %>%
  mutate(
    median_only = pmax(value_base + (med_cal - med_base), 0),
    width_only = pmax(med_base + (value_cal - med_cal), 0)
  )
split_scores <- split %>%
  group_by(geos, config, horizon) %>%
  summarize(
    base = sum(pinball(truth, value_base, level)),
    full = sum(pinball(truth, value_cal, level)),
    `median shift only` = sum(pinball(truth, median_only, level)),
    `width change only` = sum(pinball(truth, width_only, level)),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(c(full, `median shift only`, `width change only`), names_to = "variant", values_to = "wis") %>%
  mutate(pct = 100 * (base - wis) / base)
split_tbl <- function(g) {
  d <- split_scores %>% filter(geos == g)
  bind_rows(
    d %>% select(config, variant, horizon, pct),
    d %>%
      group_by(config, horizon) %>%
      summarize(pct = pct[variant == "full"] - pct[variant == "median shift only"] - pct[variant == "width change only"], .groups = "drop") %>%
      mutate(variant = "interaction (full − the two parts)")
  ) %>%
    mutate(variant = factor(variant, levels = c("full", "median shift only", "width change only", "interaction (full − the two parts)"))) %>%
    transmute(config, variant, horizon = paste0("h", horizon), pct = sprintf("%+.2f", pct)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = pct) %>%
    arrange(config, variant)
}
knitr::kable(split_tbl(GEOS[["states"]]), caption = "States only: WIS reduction % (positive = calibrated WIS lower than base)")
```
