---
title: "E16: constant learning rate on the square-root scale, and month by month"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: how does a constant-rate tracker on the square-root scale compare
with the same tracker on the rate scale and with REF-op cold, across learning
rates and by month?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (all 53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay. Cold start, no burn-in season.
- The tracker runs over both seasons, so 2025-26 starts from the offsets
  2024-25 left behind; each season is scored on its own rounds. Scores:
  states only (US excluded).
- Configs:
  - `sqrt <lr>`: the plain tracker with a constant learning rate on the
    square root of admissions per 100k; 11 rates from 0.001 to 0.32, a
    quarter decade apart, in sqrt(per-100k) units (not comparable number for
    number with the rate scale). The grid was set after a first look at
    results.
  - `rate <lr>`: E14's rate-scale runs; 0.018 and 0.032 are shown as
    references.
  - `REF-op cold`: prod's tracker without its warm start.
- A sqrt rate is selected below by E14's rule (highest mean gain with no loss
  in any season × horizon). The selection is in-sample.

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
RATE_LRS <- signif(10^seq(-2.5, 0.25, by = 0.25), 2)
SQRT_LRS <- signif(10^seq(-3, -0.5, by = 0.25), 2)
all_cfgs <- ws_configs(wi$rate_scales, burn_in = FALSE)
rate_cfg <- function(lr) utils::modifyList(all_cfgs[["E11 constant 0.1"]], list(lr = lr))
sqrt_cfg <- function(lr) utils::modifyList(all_cfgs[["E11 constant 0.1"]], list(lr = lr, transform = "sqrt"))
# Cached under cache/calibration/experiments/runs; the rate sweep and REF-op cold are E14's runs.
run <- function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
rate_runs <- purrr::map(RATE_LRS, \(lr) run(rate_cfg(lr)))
sqrt_runs <- purrr::map(SQRT_LRS, \(lr) run(sqrt_cfg(lr)))
refop <- run(all_cfgs[["REF-op cold"]])
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)")
SEASONS <- c(LIVE, "both")
SEASON_COLS <- c(`2024-2025` = "#0072B2", `2025-2026` = "#D55E00", `both seasons` = "grey45")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
label <- function(df) {
  df %>% mutate(
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
}
sweep_m <- bind_rows(
  purrr::map2(rate_runs, RATE_LRS, \(cal, lr) score(cal) %>% mutate(scale = "rate", lr = lr)),
  purrr::map2(sqrt_runs, SQRT_LRS, \(cal, lr) score(cal) %>% mutate(scale = "sqrt", lr = lr))
) %>% label()
ref_m <- score(refop) %>% mutate(ref = "REF-op cold") %>% label()
# Best sqrt rate: the highest mean WIS reduction among rates that improve WIS in both seasons at h0–h3 (states only).
best_tbl <- sweep_m %>%
  filter(scale == "sqrt", geos == GEOS[["states"]], season %in% LIVE) %>%
  group_by(lr) %>%
  summarize(worst = min(wis_pct), mean = mean(wis_pct), .groups = "drop")
BEST_SQRT <- best_tbl %>% filter(worst > 0) %>% slice_max(mean, n = 1) %>% pull(lr)
```

# 1. Coverage bias vs WIS, by season and scale

One curve per season and scale, tracing the learning rate from smallest
(near the base, the open circle) to largest. Solid lines are the rate scale
(E14), dashed lines the sqrt scale. Labels mark a few rates on each curve.
Triangles are REF-op cold. The y axis is clipped at −10%. The sqrt rate
selected by the rule below is **`r BEST_SQRT`**.

```{r curve-fn}
LABELLED <- list(rate = c(0.01, 0.032, 0.1, 0.32), sqrt = c(0.0032, 0.01, 0.032, 0.1))
curve_plot <- function(g) {
  d <- sweep_m %>% filter(geos == g, season != "both seasons") %>% arrange(scale, lr)
  r <- ref_m %>% filter(geos == g, season != "both seasons")
  lab <- d %>% filter((scale == "rate" & lr %in% LABELLED$rate) | (scale == "sqrt" & lr %in% LABELLED$sqrt))
  base <- d %>% distinct(h, season, cal_err_base)
  ggplot(d, aes(cal_err_cal, wis_pct, colour = season)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(aes(linetype = scale, group = interaction(season, scale)), linewidth = 0.7) +
    geom_point(aes(shape = scale), size = 1.1) +
    geom_text(data = lab, aes(label = paste0(ifelse(scale == "sqrt", "s", "r"), format(lr, drop0trailing = TRUE))),
      size = 2.3, vjust = -0.8, show.legend = FALSE) +
    geom_point(data = base, aes(cal_err_base, 0), shape = 21, fill = "white", size = 2.2) +
    geom_point(data = r, shape = 17, size = 2.6) +
    scale_colour_manual(values = SEASON_COLS, name = NULL) +
    scale_linetype_manual(values = c(rate = "solid", sqrt = "22"), name = "scale") +
    scale_shape_manual(values = c(rate = 16, sqrt = 1), name = "scale") +
    facet_wrap(~h, nrow = 1) +
    coord_cartesian(ylim = c(max(-10, min(d$wis_pct)), max(c(d$wis_pct, r$wis_pct)) + 0.5)) +
    labs(x = "L1 coverage bias (lower is better)", y = "WIS reduction % (positive = calibrated WIS lower than base)", title = g,
      caption = "labels: r = rate-scale learning rate (per 100k), s = sqrt-scale learning rate; triangles: REF-op cold") +
    theme(legend.position = "bottom")
}
```

```{r curve-states}
curve_plot(GEOS[["states"]])
```

## Each metric against the learning rate, sqrt scale

Top: WIS reduction % (positive = calibrated WIS lower than base). Bottom: L1 coverage bias (lower is better), each season's base dotted.
States only.

```{r vs-lr, fig.height=6.5}
long <- sweep_m %>%
  filter(scale == "sqrt", geos == GEOS[["states"]], season != "both seasons") %>%
  transmute(h, season, lr, `WIS reduction % (positive = calibrated WIS lower than base)` = wis_pct, `L1 coverage bias (lower is better)` = cal_err_cal) %>%
  tidyr::pivot_longer(c(`WIS reduction % (positive = calibrated WIS lower than base)`, `L1 coverage bias (lower is better)`), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("WIS reduction % (positive = calibrated WIS lower than base)", "L1 coverage bias (lower is better)")))
base_lines <- sweep_m %>%
  filter(scale == "sqrt", geos == GEOS[["states"]], season != "both seasons") %>%
  distinct(h, season, value = cal_err_base) %>%
  mutate(metric = factor("L1 coverage bias (lower is better)", levels = levels(long$metric)))
zero <- tibble(metric = factor("WIS reduction % (positive = calibrated WIS lower than base)", levels = levels(long$metric)), value = 0)
ggplot(long, aes(lr, value, colour = season)) +
  geom_hline(data = zero, aes(yintercept = value), colour = "grey60", linewidth = 0.3) +
  geom_hline(data = base_lines, aes(yintercept = value, colour = season), linetype = "dotted") +
  geom_line() +
  geom_point(size = 1) +
  scale_x_log10(breaks = c(0.001, 0.003, 0.01, 0.03, 0.1, 0.3), labels = c("0.001", "0.003", "0.01", "0.03", "0.1", "0.3")) +
  scale_colour_manual(values = SEASON_COLS, name = NULL) +
  facet_grid(metric ~ h, scales = "free_y", labeller = labeller(metric = label_wrap_gen(28))) +
  labs(x = "constant learning rate (sqrt of per-100k units per step)", y = NULL) +
  theme(legend.position = "bottom")
```

## Selecting a sqrt rate

For each sqrt rate, the worst and the mean WIS reduction % (positive = calibrated WIS lower than base) over the two seasons
and h0–h3 (states only). The selected rate has the highest mean among those
whose worst case is still a gain (E14's rule).

```{r best}
knitr::kable(best_tbl %>% mutate(across(c(worst, mean), \(x) round(x, 2))))
```

```{r month-view-cfgs}
lr_lab <- function(lr) vapply(lr, \(x) format(x, drop0trailing = TRUE), character(1))
rate_keep <- RATE_LRS %in% c(0.018, 0.032, 0.056, 0.1)
mv <- cal_month_view(c(
  setNames(sqrt_runs, paste("sqrt", lr_lab(SQRT_LRS))),
  setNames(rate_runs[rate_keep], paste("rate", lr_lab(RATE_LRS[rate_keep]))),
  list(`REF-op cold` = refop)
), seasons = LIVE)
# The plots show the sqrt rates around the elbow, rate 0.032 (E14) and REF-op cold; the tables cover every sqrt rate.
MONTH_COLS <- c(`sqrt 0.01` = "#56B4E9", `sqrt 0.018` = "#CC79A7", `sqrt 0.032` = "#0072B2", `rate 0.032` = "#009E73", `REF-op cold` = "#D55E00")
mv_plot_cfgs <- names(MONTH_COLS)
mv_cols <- MONTH_COLS
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Mean offset at the median over time

The calibrated minus base median, in admissions per 100k, averaged over
states at each round, h0 and h2. The bottom row is the mean base median per
100k at h0, to place the season peak.

```{r offset-data}
OFFSET_CFGS <- list(`sqrt 0.018` = sqrt_runs[[which(SQRT_LRS == 0.018)]], `rate 0.032` = rate_runs[[which(RATE_LRS == 0.032)]], `REF-op cold` = refop)
fc_pick <- purrr::imap(OFFSET_CFGS, \(cal, nm) cal$forecasts %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), config = factor(config, levels = names(OFFSET_CFGS))) %>%
  filter(season %in% LIVE, location != "US", !is.na(value_base))
```

```{r offset-time, fig.height=7}
pop <- wi$rate_scales %>% select(location, scale)
med <- fc_pick %>%
  filter(level == 0.5, horizon %in% c(0, 2), config %in% names(OFFSET_CFGS)) %>%
  left_join(pop, by = "location") %>%
  group_by(config, season, horizon, reference_date) %>%
  summarize(off = mean((value_cal - value_base) / scale), base = mean(value_base / scale), truth = mean(truth / scale, na.rm = TRUE), .groups = "drop")
off_long <- bind_rows(
  med %>% transmute(config, season, reference_date, row = paste0("offset, h", horizon), value = off),
  med %>% filter(horizon == 0, config == names(OFFSET_CFGS)[1]) %>% transmute(config = "base median", season, reference_date, row = "base median h0", value = base),
  med %>% filter(horizon == 0, config == names(OFFSET_CFGS)[1]) %>% transmute(config = "truth", season, reference_date, row = "base median h0", value = truth)
) %>%
  mutate(config = factor(config, levels = c(names(OFFSET_CFGS), "base median", "truth")), row = factor(row, levels = c("offset, h0", "offset, h2", "base median h0")))
ggplot(off_long, aes(reference_date, value, colour = config)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line() +
  geom_point(size = 0.7) +
  facet_grid(row ~ season, scales = "free") +
  scale_colour_manual(values = c(MONTH_COLS[names(OFFSET_CFGS)], `base median` = "grey30", truth = "grey65"), name = NULL) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = "reference date", y = "admissions per 100k (mean over states)") +
  theme(legend.position = "bottom")
```
