---
title: "E16: constant learning rate on the square-root scale, and month by month"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

E14 found a constant learning rate of 0.018–0.032 per 100k good in both
seasons. E15 found that E11's rate-scale tracker (0.1 per 100k) loses WIS
through a tail of offsets that are large next to narrow intervals: a fixed
step per 100k is big for a small state or an off-season week. Two questions:

1. **Does the same tracker on the square-root scale do better?** On the
   square root of the rate, a fixed step moves the forecast less where the
   rate is low.
2. **Do the offsets go stale late in the season?** Offsets learned on the
   ramp up point upward; after the peak they may linger and push the
   forecast above outcomes that are now falling.

**Setup.** As E14: the clean `windowed_seasonal` replay for flu (all 53
locations, the 55 hub rounds, h0–h3), learning from data as published each
week and re-run from scratch each week as prod does, one week of extra
revision delay, no burn-in season. The tracker is the plain one with a
constant learning rate. The new axis is the scale: **sqrt** runs it on the
square root of admissions per 100k, so the learning rate is in
sqrt(per-100k) units and the grid is not comparable number for number with
the rate scale's. The sqrt grid is 11 values from 0.001 to 0.32, a quarter
decade apart, picked after a first look so the knee sits inside it. The rate
curve is E14's runs.

Reference points: **REF-op cold** (prod's tracker without its warm start)
and the rate-scale rates **0.018** and **0.032** (E14's best range).

"WIS change %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so **positive
is better**. "L1 coverage bias" is the mean over the 23 levels of
|empirical coverage − level|. Each season is scored on its own rounds; the
tracker runs over both, so 2025-26 starts from the offsets 2024-25 left
behind. Results lead with states only (US excluded).

# Findings

Numbers are from the render of 1 October 2026, states only, WIS change %
(positive is better) / L1 coverage bias, h0/h1/h2/h3. All-locations numbers
point the same way.

- **The sqrt scale works, with a knee at 0.01–0.018.** Every sqrt rate from
  0.001 to 0.018 improves WIS at every horizon in both seasons. At 0.032 h3
  of 2024-25 turns negative (−0.3), and from 0.056 up WIS falls fast. The
  best rate by E14's criterion (highest mean gain, no loss anywhere) is
  **0.018**.
- **sqrt 0.018 vs rate 0.032 (E14's best range).** 2024-25:
  +1.7/+0.6/+0.3/+0.1 vs +1.0/+0.5/+0.3/+0.2. 2025-26: +3.0/+1.0/+1.1/+1.5
  vs +1.9/+1.0/+0.8/+0.8. Coverage bias is about the same: 0.099–0.137 vs
  0.098–0.123 in 2024-25, 0.021–0.024 vs 0.010–0.014 in 2025-26. So the sqrt
  scale buys more WIS at h0 in both seasons and at every horizon in
  2025-26, for no coverage gain. It does not help h2–h3 in 2024-25.
- **Against REF-op cold**, sqrt 0.018 is ahead in 2025-26 at every horizon
  and avoids REF-op's 2024-25 h3 loss (−0.8). REF-op has more WIS at h0–h1
  in 2024-25 (+2.2/+0.9) but keeps 2024-25 coverage bias near the base.
- **The tail shrinks, but not everywhere.** The 90th percentile of
  |offset| ÷ base 90% width, over all levels, is 0.092–0.111 for sqrt 0.018
  vs 0.124–0.153 for rate 0.032 (rate 0.1: 0.24–0.34). At the low levels it
  is much smaller (0.05 level: 0.008–0.025 vs 0.055–0.064). At the 0.95 level
  it is larger (0.115–0.180 vs 0.096–0.152), because a fixed step on the
  square root stretches the upper quantiles more where counts are high.
  sqrt 0.01 has the smallest tail (0.056–0.071).
- **Yes, offsets go stale after the peak, and March is where it costs.** All
  four trackers lose WIS in March of both seasons at h1–h3: rate 0.018
  −1.2 / −10.6 at h2 (2024-25 / 2025-26), rate 0.032 −2.2 / −18.7, sqrt
  0.018 −7.2 / −27.9, REF-op cold −19.8 / −24.3. In March the base
  already over-predicts (58% and 67% of outcomes below its median at h2)
  and the calibrated median sits higher still (62–77%). March is 4–8% of a
  season's base WIS at h2, so a −20% March costs about 1 point of season
  WIS.
- **April and May carry little WIS (1–3% of the season) and go both ways.**
  In 2024-25 the base under-predicts in May and every tracker gains
  (+6 to +18%). In 2025-26 rate 0.032 loses at h3 in May (−14%, 79% of
  outcomes below the median).
- **Rate-scale offsets do not come down after the peak.** The mean offset
  at the median stays at 0.07–0.2 per 100k from the peak through May, while
  the base median falls from 10–12 to under 1 per 100k. On the sqrt scale and
  for REF-op the offset falls with the level, but too slowly to avoid March.
- **Season-level coverage hides month-level errors.** In December the base's
  monthly coverage bias is 0.25–0.38 and no tracker moves it; in March the
  trackers make it worse. Part of the season-level coverage gain comes from
  under-prediction on the ramp and over-prediction after the peak cancelling
  out.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
RATE_LRS <- signif(10^seq(-2.5, 0.25, by = 0.25), 2)
SQRT_LRS <- signif(10^seq(-3, -0.5, by = 0.25), 2)
all_cfgs <- ws_configs(wi$rate_scales, burn_in = FALSE)
rate_cfg <- function(lr) utils::modifyList(all_cfgs[["E11 constant 0.1"]], list(lr = lr))
sqrt_cfg <- function(lr) utils::modifyList(all_cfgs[["E11 constant 0.1"]], list(lr = lr, transform = "sqrt"))
# Cached under cache/calibration/experiments/runs; the rate sweep and REF-op cold are E14's runs.
run <- function(cfg) {
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
rate_runs <- purrr::map(RATE_LRS, \(lr) run(rate_cfg(lr)))
sqrt_runs <- purrr::map(SQRT_LRS, \(lr) run(sqrt_cfg(lr)))
refop <- run(all_cfgs[["REF-op cold"]])
```

```{r metrics}
GEOS <- c(states = "states only (US excluded)", all = "all locations")
SEASONS <- c(LIVE, "both")
SEASON_COLS <- c(`2024-2025` = "#0072B2", `2025-2026` = "#D55E00", `both seasons` = "grey45")
score <- function(cal) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date))
  purrr::imap(WS_GEOS[names(GEOS)], function(keep, g) ws_scores(fc %>% filter(keep(location))) %>% mutate(geos = g)) %>%
    bind_rows() %>%
    select(geos, season, horizon, which, wis, cal_err, cov50, cov90) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err, cov50, cov90)) %>%
    mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}
label <- function(df) {
  df %>% mutate(
    geos = factor(geos, levels = names(GEOS), labels = GEOS),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
}
sweep_m <- bind_rows(
  purrr::map2(rate_runs, RATE_LRS, \(cal, lr) score(cal) %>% mutate(scale = "rate", lr = lr)),
  purrr::map2(sqrt_runs, SQRT_LRS, \(cal, lr) score(cal) %>% mutate(scale = "sqrt", lr = lr))
) %>% label()
ref_m <- score(refop) %>% mutate(ref = "REF-op cold") %>% label()
# Best sqrt rate: the highest mean WIS change among rates that improve WIS in both seasons at h0–h3 (states only).
best_tbl <- sweep_m %>%
  filter(scale == "sqrt", geos == GEOS[["states"]], season %in% LIVE) %>%
  group_by(lr) %>%
  summarize(worst = min(wis_pct), mean = mean(wis_pct), .groups = "drop")
BEST_SQRT <- best_tbl %>% filter(worst > 0) %>% slice_max(mean, n = 1) %>% pull(lr)
```

# 1. Coverage bias vs WIS, by season and scale

One curve per season and scale, tracing the learning rate from smallest
(near the base, the open circle) to largest. Solid lines are the rate scale
(E14), dashed lines the sqrt scale. Labels mark a few rates on each curve.
Triangles are REF-op cold. The y axis is clipped at −10%. The best sqrt
rate by the rule below is **`r BEST_SQRT`**.

```{r curve-fn}
LABELLED <- list(rate = c(0.01, 0.032, 0.1, 0.32), sqrt = c(0.0032, 0.01, 0.032, 0.1))
curve_plot <- function(g) {
  d <- sweep_m %>% filter(geos == g, season != "both seasons") %>% arrange(scale, lr)
  r <- ref_m %>% filter(geos == g, season != "both seasons")
  lab <- d %>% filter((scale == "rate" & lr %in% LABELLED$rate) | (scale == "sqrt" & lr %in% LABELLED$sqrt))
  base <- d %>% distinct(h, season, cal_err_base)
  ggplot(d, aes(cal_err_cal, wis_pct, colour = season)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_path(aes(linetype = scale, group = interaction(season, scale)), linewidth = 0.7) +
    geom_point(aes(shape = scale), size = 1.1) +
    geom_text(data = lab, aes(label = paste0(ifelse(scale == "sqrt", "s", "r"), format(lr, drop0trailing = TRUE))),
      size = 2.3, vjust = -0.8, show.legend = FALSE) +
    geom_point(data = base, aes(cal_err_base, 0), shape = 21, fill = "white", size = 2.2) +
    geom_point(data = r, shape = 17, size = 2.6) +
    scale_colour_manual(values = SEASON_COLS, name = NULL) +
    scale_linetype_manual(values = c(rate = "solid", sqrt = "22"), name = "scale") +
    scale_shape_manual(values = c(rate = 16, sqrt = 1), name = "scale") +
    facet_wrap(~h, nrow = 1) +
    coord_cartesian(ylim = c(max(-10, min(d$wis_pct)), max(c(d$wis_pct, r$wis_pct)) + 0.5)) +
    labs(x = "L1 coverage bias", y = "WIS change % (positive is better)", title = g,
      caption = "labels: r = rate-scale learning rate (per 100k), s = sqrt-scale learning rate; triangles: REF-op cold") +
    theme(legend.position = "bottom")
}
```

## States only

```{r curve-states}
curve_plot(GEOS[["states"]])
```

## All locations

```{r curve-all}
curve_plot(GEOS[["all"]])
```

## Each metric against the learning rate, sqrt scale

Top: WIS change %. Bottom: L1 coverage bias, each season's base dotted.
States only.

```{r vs-lr, fig.height=6.5}
long <- sweep_m %>%
  filter(scale == "sqrt", geos == GEOS[["states"]], season != "both seasons") %>%
  transmute(h, season, lr, `WIS change %` = wis_pct, `L1 coverage bias` = cal_err_cal) %>%
  tidyr::pivot_longer(c(`WIS change %`, `L1 coverage bias`), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = c("WIS change %", "L1 coverage bias")))
base_lines <- sweep_m %>%
  filter(scale == "sqrt", geos == GEOS[["states"]], season != "both seasons") %>%
  distinct(h, season, value = cal_err_base) %>%
  mutate(metric = factor("L1 coverage bias", levels = levels(long$metric)))
zero <- tibble(metric = factor("WIS change %", levels = levels(long$metric)), value = 0)
ggplot(long, aes(lr, value, colour = season)) +
  geom_hline(data = zero, aes(yintercept = value), colour = "grey60", linewidth = 0.3) +
  geom_hline(data = base_lines, aes(yintercept = value, colour = season), linetype = "dotted") +
  geom_line() +
  geom_point(size = 1) +
  scale_x_log10(breaks = c(0.001, 0.003, 0.01, 0.03, 0.1, 0.3), labels = c("0.001", "0.003", "0.01", "0.03", "0.1", "0.3")) +
  scale_colour_manual(values = SEASON_COLS, name = NULL) +
  facet_grid(metric ~ h, scales = "free_y") +
  labs(x = "constant learning rate (sqrt of per-100k units per step)", y = NULL) +
  theme(legend.position = "bottom")
```

## Tables

WIS change % / L1 coverage bias per config (rows) and season × horizon
(columns). The sqrt sweep, then the rate-scale rates around E14's best range
and REF-op cold.

```{r tables}
season_table <- function(g, rate_keep = c(0.018, 0.032, 0.056, 0.1)) {
  rows <- bind_rows(
    sweep_m %>% filter(scale == "sqrt", lr == SQRT_LRS[1]) %>% mutate(config = "base", wis_pct = 0, cal_err_cal = cal_err_base),
    sweep_m %>% filter(scale == "sqrt") %>% mutate(config = paste("sqrt", format(lr, drop0trailing = TRUE))),
    sweep_m %>% filter(scale == "rate", lr %in% rate_keep) %>% mutate(config = paste("rate", format(lr, drop0trailing = TRUE))),
    ref_m %>% mutate(config = ref)
  ) %>%
    filter(geos == g) %>%
    mutate(config = factor(config, levels = unique(config)))
  rows %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(config)
}
```

### States only

```{r table-states}
knitr::kable(season_table(GEOS[["states"]]))
```

### All locations

```{r table-all}
knitr::kable(season_table(GEOS[["all"]]))
```

### Picking the best sqrt rate

For each sqrt rate, the worst and the mean WIS change % over the two seasons
and h0–h3 (states only). The best rate is the one with the highest mean
among those whose worst case is still a gain (E14's criterion).

```{r best}
knitr::kable(best_tbl %>% mutate(across(c(worst, mean), \(x) round(x, 2))))
```

### 50% and 90% interval coverage, states only

```{r table-cov}
bind_rows(
  sweep_m %>% filter(scale == "sqrt", lr == SQRT_LRS[1]) %>% mutate(config = "base", cov50_cal = cov50_base, cov90_cal = cov90_base),
  sweep_m %>% filter(scale == "sqrt", lr >= 0.0032, lr <= 0.056) %>% mutate(config = paste("sqrt", format(lr, drop0trailing = TRUE))),
  sweep_m %>% filter(scale == "rate", lr %in% c(0.018, 0.032)) %>% mutate(config = paste("rate", format(lr, drop0trailing = TRUE))),
  ref_m %>% mutate(config = ref)
) %>%
  filter(geos == GEOS[["states"]], h %in% c("h0", "h2"), season != "both seasons") %>%
  mutate(config = factor(config, levels = unique(config))) %>%
  transmute(config, col = paste(season, h), cell = sprintf("%.2f / %.2f", cov50_cal, cov90_cal)) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  arrange(config) %>%
  knitr::kable()
```

# 2. Does the offset tail shrink?

E15's statistic: for each forecast and level, |offset| ÷ the base 90%
interval width (q95 − q05), and its 90th percentile over forecasts (location
× round, both seasons, states only). "all levels" pools the 23 levels. A
smaller number means fewer offsets that are large next to the interval.

```{r tail}
pick <- list(
  `sqrt best` = sqrt_runs[[which(SQRT_LRS == BEST_SQRT)]],
  `rate 0.018` = rate_runs[[which(RATE_LRS == 0.018)]],
  `rate 0.032` = rate_runs[[which(RATE_LRS == 0.032)]],
  `rate 0.1` = rate_runs[[which(RATE_LRS == 0.1)]],
  `REF-op cold` = refop
)
names(pick)[1] <- paste("sqrt", BEST_SQRT)
fc_pick <- purrr::imap(pick, \(cal, nm) cal$forecasts %>% mutate(config = nm)) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), config = factor(config, levels = names(pick))) %>%
  filter(season %in% LIVE, location != "US", !is.na(value_base))
w90 <- fc_pick %>%
  filter(config == names(pick)[1], level %in% c(0.05, 0.95)) %>%
  select(location, horizon, reference_date, level, value_base) %>%
  tidyr::pivot_wider(names_from = level, values_from = value_base, names_prefix = "q") %>%
  transmute(location, horizon, reference_date, w90 = q0.95 - q0.05)
ratio <- fc_pick %>%
  inner_join(w90, by = c("location", "horizon", "reference_date")) %>%
  filter(w90 > 0) %>%
  mutate(r = abs(value_cal - value_base) / w90)
tail_tbl <- bind_rows(
  ratio %>% filter(level %in% c(0.05, 0.25, 0.5, 0.75, 0.95)) %>% mutate(lv = as.character(level)),
  ratio %>% mutate(lv = "all levels")
) %>%
  group_by(config, horizon, lv) %>%
  summarize(p90 = quantile(r, 0.9), .groups = "drop") %>%
  mutate(lv = factor(lv, levels = c("0.05", "0.25", "0.5", "0.75", "0.95", "all levels")))
knitr::kable(
  tail_tbl %>%
    transmute(config, lv, h = paste0("h", horizon), p90 = round(p90, 3)) %>%
    tidyr::pivot_wider(names_from = h, values_from = p90) %>%
    arrange(lv, config),
  caption = "States only: 90th percentile of |offset| / base 90% width"
)
```

# 3. Month by month

Five forecasts per state-round are compared: the base, the rate-scale rates
0.018 and 0.032, the best sqrt rate and REF-op cold. Months are calendar
months of the reference date. States only, h0–h3. Each cell holds about
150–260 state forecasts, so single months are noisy, but the March losses
below repeat across seasons and horizons.

```{r month-data}
MONTH_COLS <- setNames(c("#009E73", "#0072B2", "#CC79A7", "#D55E00"), names(pick)[c(2, 3, 1, 5)])
month_m <- ws_month_scores(fc_pick %>% filter(config %in% names(MONTH_COLS)) %>% mutate(config = factor(config, levels = names(MONTH_COLS)))) %>%
  mutate(h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
```

## WIS change % by month

Each season a row, each horizon a column. The grey bars (right axis
scaled to the left) are each month's share of the season's base WIS, so a
large loss in a month with a small share costs little overall. The y axis is
clipped at −40% and +30%, which also cuts the tallest bars.

```{r month-wis, fig.height=6}
share <- month_m %>% filter(config == names(MONTH_COLS)[1])
ggplot(month_m, aes(month, pct, colour = config, group = config)) +
  geom_col(data = share, aes(month, share), inherit.aes = FALSE, fill = "grey85", width = 0.7) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_line() +
  geom_point(size = 1.2) +
  facet_grid(season ~ h, scales = "free_x", space = "free_x") +
  scale_colour_manual(values = MONTH_COLS, name = NULL) +
  coord_cartesian(ylim = c(-40, 30)) +
  labs(x = "month of reference date", y = "WIS change % (positive is better); grey: % of season base WIS") +
  theme(legend.position = "bottom")
```

## Coverage by month

Top: L1 coverage bias per month, base in black. Bottom: share of outcomes
below the median; 0.5 is calibrated, above 0.5 means the median sits too
high (over-prediction). Base in black.

```{r month-cov, fig.height=8}
cov_long <- bind_rows(
  month_m %>% transmute(config = as.character(config), season, h, month, `L1 coverage bias` = cal_err_cal, `share below the median` = below_med_cal),
  month_m %>% filter(config == names(MONTH_COLS)[1]) %>%
    transmute(config = "base", season, h, month, `L1 coverage bias` = cal_err_base, `share below the median` = below_med_base)
) %>%
  tidyr::pivot_longer(c(`L1 coverage bias`, `share below the median`), names_to = "metric") %>%
  mutate(config = factor(config, levels = c("base", names(MONTH_COLS))))
half <- tibble(metric = "share below the median", y = 0.5)
ggplot(cov_long, aes(month, value, colour = config, group = config)) +
  geom_hline(data = half, aes(yintercept = y), colour = "grey50", linetype = "dashed") +
  geom_line() +
  geom_point(size = 1) +
  facet_grid(metric + season ~ h, scales = "free", space = "free_x") +
  scale_colour_manual(values = c(base = "black", MONTH_COLS), name = NULL) +
  labs(x = "month of reference date", y = NULL) +
  theme(legend.position = "bottom")
```

## Mean offset at the median over time

The calibrated minus base median, in admissions per 100k, averaged over
states at each round, h0 and h2. The bottom row is the mean base median per
100k at h0, to place the season peak.

```{r offset-time, fig.height=7}
pop <- wi$rate_scales %>% select(location, scale)
med <- fc_pick %>%
  filter(level == 0.5, horizon %in% c(0, 2), config %in% names(MONTH_COLS)) %>%
  left_join(pop, by = "location") %>%
  group_by(config, season, horizon, reference_date) %>%
  summarize(off = mean((value_cal - value_base) / scale), base = mean(value_base / scale), truth = mean(truth / scale, na.rm = TRUE), .groups = "drop")
off_long <- bind_rows(
  med %>% transmute(config, season, reference_date, row = paste0("offset, h", horizon), value = off),
  med %>% filter(horizon == 0, config == names(MONTH_COLS)[1]) %>% transmute(config = "base median", season, reference_date, row = "base median h0", value = base),
  med %>% filter(horizon == 0, config == names(MONTH_COLS)[1]) %>% transmute(config = "truth", season, reference_date, row = "base median h0", value = truth)
) %>%
  mutate(config = factor(config, levels = c(names(MONTH_COLS), "base median", "truth")), row = factor(row, levels = c("offset, h0", "offset, h2", "base median h0")))
ggplot(off_long, aes(reference_date, value, colour = config)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line() +
  geom_point(size = 0.7) +
  facet_grid(row ~ season, scales = "free") +
  scale_colour_manual(values = c(MONTH_COLS, `base median` = "grey30", truth = "grey65"), name = NULL) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = "reference date", y = "admissions per 100k (mean over states)") +
  theme(legend.position = "bottom")
```

## Month table, h2

WIS change % per month (states only, h2), with the base's share of the
season's WIS and the share of outcomes below the base and calibrated medians.

```{r month-table}
month_m %>%
  filter(h == "h2") %>%
  transmute(season, month, config, cell = sprintf("%+.1f (%.2f)", pct, below_med_cal)) %>%
  tidyr::pivot_wider(names_from = config, values_from = cell) %>%
  left_join(
    month_m %>% filter(h == "h2", config == names(MONTH_COLS)[1]) %>%
      transmute(season, month, `base share %` = round(share, 1), `base below median` = round(below_med_base, 2)),
    by = c("season", "month")
  ) %>%
  arrange(season, month) %>%
  knitr::kable(caption = "WIS change % (share of outcomes below the calibrated median)")
```
