---
title: "E17: shrinking the offset late in the season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: does shrinking the offset from 1 March, or switching calibration
off from 1 March, change WIS and coverage compared with the same tracker
without it?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (all 53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay. Cold start, no burn-in season. States only.
- Three constant-rate trackers: rate 0.018 and rate 0.032 per 100k (E14),
  and sqrt 0.018 (E16). Each crossed with five settings:
  - **no decay**.
  - **`late_decay`** with factor 0.85, 0.7 or 0.5: from 1 March to 30 June
    the stored offset is multiplied by the factor at each round before it is
    played. Learning continues, and because offsets carry across seasons,
    the shrink carries into the next season.
  - **`off_after = "03-01"`**: from 1 March to 30 June the base is played,
    those rounds produce no updates, and the offset carries unchanged into
    the next season.
- The 1 March date was chosen after E16, on these same seasons.
  (`fast_decay` and `season_policy` are other ways to shrink offsets; they
  are not run here.)

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
base_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
SQRT_BEST <- 0.018
RATES <- list(
  `rate 0.018` = list(lr = 0.018),
  `rate 0.032` = list(lr = 0.032),
  sqrt = list(lr = SQRT_BEST, transform = "sqrt")
)
names(RATES)[3] <- paste("sqrt", SQRT_BEST)
DECAYS <- list(
  `no decay` = list(),
  `decay 0.85` = list(late_decay = list(from = "03-01", factor = 0.85)),
  `decay 0.7` = list(late_decay = list(from = "03-01", factor = 0.7)),
  `decay 0.5` = list(late_decay = list(from = "03-01", factor = 0.5)),
  `off after 1 Mar` = list(off_after = "03-01")
)
grid <- tidyr::expand_grid(rate = names(RATES), decay = names(DECAYS))
# Cached under cache/calibration/experiments/runs; the no-decay runs are E14/E16's.
cals <- purrr::pmap(grid, function(rate, decay) {
  cfg <- c(utils::modifyList(base_cfg, RATES[[rate]]), DECAYS[[decay]])
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
fc <- purrr::map2(cals, seq_len(nrow(grid)), function(cal, i) {
  cal$forecasts %>% mutate(rate = grid$rate[i], decay = grid$decay[i])
}) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), rate = factor(rate, levels = names(RATES)), decay = factor(decay, levels = names(DECAYS))) %>%
  filter(season %in% LIVE, location != "US")
DECAY_COLS <- setNames(c("black", "#56B4E9", "#0072B2", "#D55E00", "#009E73"), names(DECAYS))
```

```{r month-view-cfgs}
mv <- cal_month_view(setNames(cals, paste(grid$rate, grid$decay, sep = ", ")), seasons = LIVE)
# The plots show sqrt 0.018 with each decay setting; the tables cover all three trackers.
mv_plot_cfgs <- paste(names(RATES)[3], names(DECAYS), sep = ", ")
mv_cols <- setNames(DECAY_COLS, mv_plot_cfgs)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Mean offset at the median over time

Calibrated minus base median in admissions per 100k, averaged over states at
each round, h2.

```{r offset-time, fig.height=7}
pop <- wi$rate_scales %>% select(location, scale)
fc %>%
  filter(level == 0.5, horizon == 2) %>%
  left_join(pop, by = "location") %>%
  group_by(rate, decay, season, reference_date) %>%
  summarize(off = mean((value_cal - value_base) / scale), .groups = "drop") %>%
  ggplot(aes(reference_date, off, colour = decay)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line() +
  geom_point(size = 0.6) +
  facet_grid(rate ~ season, scales = "free") +
  scale_colour_manual(values = DECAY_COLS, name = NULL) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = "reference date", y = "mean offset at the median, per 100k") +
  theme(legend.position = "bottom")
```
