---
title: "E17: shrinking the offset late in the season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: does shrinking the offset from 1 March, or switching calibration
off from 1 March, change WIS and coverage compared with the same tracker
without it?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (all 53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay. Cold start, no burn-in season. States only.
- Three constant-rate trackers: rate 0.018 and rate 0.032 per 100k (E14),
  and sqrt 0.018 (E16). Each crossed with five settings:
  - **no decay**.
  - **`late_decay`** with factor 0.85, 0.7 or 0.5: from 1 March to 30 June
    the stored offset is multiplied by the factor at each round before it is
    played. Learning continues, and because offsets carry across seasons,
    the shrink carries into the next season.
  - **`off_after = "03-01"`**: from 1 March to 30 June the base is played,
    those rounds produce no updates, and the offset carries unchanged into
    the next season.
- The 1 March date was chosen after E16, on these same seasons.
  (`fast_decay` and `season_policy` are other ways to shrink offsets; they
  are not run here.)

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
base_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
SQRT_BEST <- 0.018
RATES <- list(
  `rate 0.018` = list(lr = 0.018),
  `rate 0.032` = list(lr = 0.032),
  sqrt = list(lr = SQRT_BEST, transform = "sqrt")
)
names(RATES)[3] <- paste("sqrt", SQRT_BEST)
DECAYS <- list(
  `no decay` = list(),
  `decay 0.85` = list(late_decay = list(from = "03-01", factor = 0.85)),
  `decay 0.7` = list(late_decay = list(from = "03-01", factor = 0.7)),
  `decay 0.5` = list(late_decay = list(from = "03-01", factor = 0.5)),
  `off after 1 Mar` = list(off_after = "03-01")
)
grid <- tidyr::expand_grid(rate = names(RATES), decay = names(DECAYS))
# Cached under cache/calibration/experiments/runs; the no-decay runs are E14/E16's.
cals <- purrr::pmap(grid, function(rate, decay) {
  cfg <- c(utils::modifyList(base_cfg, RATES[[rate]]), DECAYS[[decay]])
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
fc <- purrr::map2(cals, seq_len(nrow(grid)), function(cal, i) {
  cal$forecasts %>% mutate(rate = grid$rate[i], decay = grid$decay[i])
}) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), rate = factor(rate, levels = names(RATES)), decay = factor(decay, levels = names(DECAYS))) %>%
  filter(season %in% LIVE, location != "US")
DECAY_COLS <- setNames(c("black", "#56B4E9", "#0072B2", "#D55E00", "#009E73"), names(DECAYS))
```

# 1. Overall, by season

WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage bias (lower is better), states only.

```{r head}
head_m <- fc %>%
  group_by(rate, decay) %>%
  group_modify(\(d, k) ws_scores(d)) %>%
  ungroup() %>%
  select(rate, decay, season, horizon, which, wis, cal_err) %>%
  tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
  mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base, season = factor(season, levels = c(LIVE, "both"), labels = c(LIVE, "both seasons")))
base_row <- head_m %>%
  filter(rate == names(RATES)[1], decay == "no decay") %>%
  transmute(config = "base", season, horizon, cell = sprintf("%+.1f / %.3f", 0, cal_err_base))
bind_rows(base_row, head_m %>% transmute(config = paste(rate, decay, sep = ", "), season, horizon, cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal))) %>%
  mutate(config = factor(config, levels = unique(config))) %>%
  transmute(config, col = paste0(season, " h", horizon), cell) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  knitr::kable()
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r head-month}
mv <- cal_month_view(setNames(cals, paste(grid$rate, grid$decay, sep = ", ")), seasons = LIVE)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

The same WIS reductions as a plot, one panel per season and horizon.

```{r head-plot, fig.height=5.5}
ggplot(head_m, aes(rate, wis_pct, fill = decay)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ paste0("h", horizon)) +
  scale_fill_manual(values = DECAY_COLS, name = NULL) +
  labs(x = NULL, y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 20, hjust = 1))
```

# 2. Month by month

WIS reduction % (positive = calibrated WIS lower than base) by calendar month of the reference date, one figure per
constant rate; rows are seasons, columns horizons. Grey bars are each
month's share of the season's base WIS. Clipped at −40%.

```{r month, results="asis", fig.height=5.5}
month_m <- ws_month_scores(fc %>% mutate(config = interaction(rate, decay, sep = "|"))) %>%
  tidyr::separate(config, c("rate", "decay"), sep = "[|]") %>%
  mutate(decay = factor(decay, levels = names(DECAYS)), h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
for (r in names(RATES)) {
  cat("\n\n## ", r, "\n\n", sep = "")
  d <- month_m %>% filter(rate == r)
  share <- d %>% filter(decay == "no decay")
  print(
    ggplot(d, aes(month, pct, colour = decay, group = decay)) +
      geom_col(data = share, aes(month, share), inherit.aes = FALSE, fill = "grey85", width = 0.7) +
      geom_hline(yintercept = 0, colour = "grey40") +
      geom_line() +
      geom_point(size = 1.1) +
      facet_grid(season ~ h, scales = "free_x", space = "free_x") +
      scale_colour_manual(values = DECAY_COLS, name = NULL) +
      coord_cartesian(ylim = c(-40, 30)) +
      labs(x = "month of reference date", y = "WIS reduction % (positive = calibrated\nWIS lower than base); grey: % of season base WIS", title = r) +
      theme(legend.position = "bottom")
  )
}
```

## Share of outcomes below the median, March–May

0.5 is calibrated; above 0.5 the median is too high. h2, states only.

```{r month-below}
month_m %>%
  filter(h == "h2", month %in% c("Mar", "Apr", "May")) %>%
  transmute(rate, season, month, decay, v = sprintf("%+.1f (%.2f)", pct, below_med_cal)) %>%
  tidyr::pivot_wider(names_from = decay, values_from = v) %>%
  left_join(
    month_m %>% filter(h == "h2", decay == "no decay") %>% transmute(rate, season, month, `base below median` = round(below_med_base, 2)),
    by = c("rate", "season", "month")
  ) %>%
  arrange(rate, season, month) %>%
  knitr::kable(caption = "h2: WIS reduction %, positive = calibrated WIS lower than base (share of outcomes below the calibrated median)")
```

# 3. Mean offset at the median over time

Calibrated minus base median in admissions per 100k, averaged over states at
each round, h2.

```{r offset-time, fig.height=7}
pop <- wi$rate_scales %>% select(location, scale)
fc %>%
  filter(level == 0.5, horizon == 2) %>%
  left_join(pop, by = "location") %>%
  group_by(rate, decay, season, reference_date) %>%
  summarize(off = mean((value_cal - value_base) / scale), .groups = "drop") %>%
  ggplot(aes(reference_date, off, colour = decay)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line() +
  geom_point(size = 0.6) +
  facet_grid(rate ~ season, scales = "free") +
  scale_colour_manual(values = DECAY_COLS, name = NULL) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = "reference date", y = "mean offset at the median, per 100k") +
  theme(legend.position = "bottom")
```

```{r month-view-cfgs}
# sqrt 0.018 with each decay setting; section 2 has the WIS months for the rate-scale trackers.
mv_plot_cfgs <- paste(names(RATES)[3], names(DECAYS), sep = ", ")
mv_cols <- setNames(DECAY_COLS, mv_plot_cfgs)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Findings

There are no uncertainty estimates; a WIS gap under 1 point or a month-avg
coverage gap under 0.01 is read as no clear difference. The cold start leaves
Nov–Dec 2024 equal to the base for every config. States only, h0/h1/h2/h3.

- **Decays vs no decay, pooled WIS:** sqrt 0.018 loses 0.4–0.5 at h0 and is
  within 0.2 at h1–h3; rate 0.018 loses 0.5–0.7 at h0 and 0.2–0.3 at h1–h3;
  rate 0.032 loses 0.2–0.5 at h0 and at most 0.2 at h1–h3. No strong result.
  For sqrt the order flips by season: in 2024-25 the decays are 0.1–0.3
  ahead at h1–h3; in 2025-26 they are 0.9–1.3 behind at h0.
- **`off_after` vs no decay, pooled WIS:** sqrt 0.018 +2.3/+1.2/+1.2/+1.3 vs
  +2.2/+0.8/+0.6/+0.6 (0.1–0.7 apart); rate 0.032 −0.1 to +0.5; rate 0.018
  −0.5 to 0. No pooled gap reaches 1 point. The sqrt gain is mostly 2025-26
  (+3.4/+1.6/+1.9/+2.6 vs +3.0/+1.0/+1.1/+1.5). In 2024-25 the rate-scale
  trackers are lower at h0 with the switch-off. The date was tuned
  in-sample.
- **March and spring at h2:** the decays shrink the March loss (sqrt
  2025-26: −27.9 to −10.7/−6.3/−2.9; `off_after` −0.0), and also remove the
  April–May gains (sqrt May 2025-26: +17.3 to +3.3/−0.0/−0.1; `off_after`
  +0.0).
- **Month-avg coverage (both seasons):** sqrt 0.018 no decay
  0.094/0.116/0.143/0.169, decay 0.85 0.103/0.123/0.149/0.173, `off_after`
  0.104/0.121/0.150/0.170, base 0.112/0.127/0.153/0.177: gaps ≤ 0.010, no
  clear difference. For rate 0.032, decay 0.85 is 0.005–0.007 lower than no
  decay overall and 0.014–0.024 lower in 2025-26. Season-pooled, no decay is
  0.032–0.043 lower for sqrt, but pooling lets ramp and post-peak errors
  cancel.
- **Carry-over:** `off_after` carries the pre-March offset into the next
  ramp. In 2025-26 that ramp under-predicted too (Nov 2025 h0 month bias
  0.065 for `off_after` vs 0.096–0.103 for the decays), which a season that
  ramps differently would not repeat.
