---
title: "E17: shrinking the offset late in the season"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

E16 found that the constant-rate trackers lose WIS in March of both seasons:
the offsets learned on the ramp up still point up after the peak, while
outcomes fall below the base median. This notebook asks whether **shrinking
the offset toward zero from a calendar date onward** removes those losses
without giving up the gains elsewhere.

**Existing options.** None does exactly this:

- `fast_decay` shrinks the offset every round of the year, so it also
  weakens the ramp-up correction.
- `season_policy` (`carry`, `reset`, `shrink`) acts only at the season
  boundary in the autumn, after the spring has been scored.
- `off_after` is the abrupt version: from a calendar date to July it plays
  the base forecast and stops learning. It is run here as a comparison.

So this notebook uses a new option, `late_decay = list(from = "MM-DD",
factor = f)`: from that date until July, the offset is multiplied by `f`
at each round before it is played. Learning continues, so new outcomes can
still push the offset.

**Setup.** As E14/E16: flu, the clean `windowed_seasonal` replay, h0–h3,
learning from data as published each week and re-run from scratch each week,
one week of extra revision delay, no burn-in, states only. Three constant
rates: rate scale 0.018 and 0.032 per 100k (E14) and sqrt scale
0.018 (E16's best). For each, five settings: no decay, `late_decay` from
1 March with factor 0.85, 0.7 or 0.5 per round, and `off_after = "03-01"`.
The tracker carries offsets across seasons, so a decay in spring 2025 also
changes where 2025-26 starts.

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so **positive
= calibrated WIS lower than base**; "L1 coverage bias" (lower is better) is
the mean over the 23 levels of |empirical coverage − level|.

# Findings

Numbers are from the render of 1 October 2026, states only, WIS reduction %
(positive = calibrated WIS lower than base) / L1 coverage bias (lower is better), h0/h1/h2/h3.

- **The decay removes the March loss.** For sqrt 0.018 at h2, March 2025-26
  goes from −27.9% to −10.7 / −6.3 / −2.9% (factor 0.85 / 0.7 / 0.5), and
  the calibrated median comes back toward the base's (69–73% of outcomes
  below it, base 67%). The rate-scale trackers behave the same way.
- **But it costs more than it saves.** It also removes the April–May gains,
  and because the offsets carry across seasons, it hands 2025-26 an offset of
  almost zero: the next season starts cold again. 2025-26 at h0 drops from
  +3.0 to +1.7–2.1 (sqrt 0.018) and its coverage bias triples (0.021–0.024
  to 0.066–0.078). Pooled over both seasons, every decay setting is worse
  than no decay on coverage and no better on WIS for all three rates.
- **`off_after = "03-01"` does better on WIS.** It plays the base from 1
  March and freezes the offsets, which then carry into the next season
  unchanged. Pooled WIS for sqrt 0.018 goes from +2.2/+0.8/+0.6/+0.6 to
  +2.3/+1.2/+1.2/+1.3, and both seasons gain at h1–h3 (2025-26:
  +3.4/+1.6/+1.9/+2.6). Rate 0.032 gains at h2–h3 too (+0.8/+0.9 pooled vs
  +0.5/+0.4). The cost is the October 2025 start (about −10% at h2/h3 on a
  month with under 1% of the WIS).
- **Any spring switch-off gives back most of the coverage gain.** With
  `off_after`, pooled coverage bias for sqrt 0.018 is 0.094–0.116 (no decay
  0.059–0.078, base 0.105–0.127). Spring rounds then play the base, and
  the over-prediction after the peak no longer cancels the under-prediction
  on the ramp (E16).
- **Answer.** The offsets do go stale after the peak, but shrinking them
  toward zero by the calendar is not a free fix. The gradual decay as built
  here loses because it also wipes the offset the next season starts from.
  `off_after` in March is the better WIS choice and the worse coverage choice.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
base_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
SQRT_BEST <- 0.018
RATES <- list(
  `rate 0.018` = list(lr = 0.018),
  `rate 0.032` = list(lr = 0.032),
  sqrt = list(lr = SQRT_BEST, transform = "sqrt")
)
names(RATES)[3] <- paste("sqrt", SQRT_BEST)
DECAYS <- list(
  `no decay` = list(),
  `decay 0.85` = list(late_decay = list(from = "03-01", factor = 0.85)),
  `decay 0.7` = list(late_decay = list(from = "03-01", factor = 0.7)),
  `decay 0.5` = list(late_decay = list(from = "03-01", factor = 0.5)),
  `off after 1 Mar` = list(off_after = "03-01")
)
grid <- tidyr::expand_grid(rate = names(RATES), decay = names(DECAYS))
# Cached under cache/calibration/experiments/runs; the no-decay runs are E14/E16's.
cals <- purrr::pmap(grid, function(rate, decay) {
  cfg <- c(utils::modifyList(base_cfg, RATES[[rate]]), DECAYS[[decay]])
  do.call(cal_run_cached, c(list(wi$fc, wi$truth), cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
})
fc <- purrr::map2(cals, seq_len(nrow(grid)), function(cal, i) {
  cal$forecasts %>% mutate(rate = grid$rate[i], decay = grid$decay[i])
}) %>%
  bind_rows() %>%
  mutate(season = season_of(reference_date), rate = factor(rate, levels = names(RATES)), decay = factor(decay, levels = names(DECAYS))) %>%
  filter(season %in% LIVE, location != "US")
DECAY_COLS <- setNames(c("black", "#56B4E9", "#0072B2", "#D55E00", "#009E73"), names(DECAYS))
```

# 1. Overall, by season

WIS reduction % (positive = calibrated WIS lower than base) / L1 coverage bias (lower is better), states only.

```{r head}
head_m <- fc %>%
  group_by(rate, decay) %>%
  group_modify(\(d, k) ws_scores(d)) %>%
  ungroup() %>%
  select(rate, decay, season, horizon, which, wis, cal_err) %>%
  tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
  mutate(wis_pct = 100 * (wis_base - wis_cal) / wis_base, season = factor(season, levels = c(LIVE, "both"), labels = c(LIVE, "both seasons")))
base_row <- head_m %>%
  filter(rate == names(RATES)[1], decay == "no decay") %>%
  transmute(config = "base", season, horizon, cell = sprintf("%+.1f / %.3f", 0, cal_err_base))
bind_rows(base_row, head_m %>% transmute(config = paste(rate, decay, sep = ", "), season, horizon, cell = sprintf("%+.1f / %.3f", wis_pct, cal_err_cal))) %>%
  mutate(config = factor(config, levels = unique(config))) %>%
  transmute(config, col = paste0(season, " h", horizon), cell) %>%
  tidyr::pivot_wider(names_from = col, values_from = cell) %>%
  knitr::kable()
```

L1 coverage bias, season-pooled against month-average (see "Month view" below):

```{r head-month}
mv <- cal_month_view(setNames(cals, paste(grid$rate, grid$decay, sep = ", ")), seasons = LIVE)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

The same WIS reductions as a plot, one panel per season and horizon.

```{r head-plot, fig.height=5.5}
ggplot(head_m, aes(rate, wis_pct, fill = decay)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_grid(season ~ paste0("h", horizon)) +
  scale_fill_manual(values = DECAY_COLS, name = NULL) +
  labs(x = NULL, y = "WIS reduction % (positive = calibrated WIS lower than base)") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 20, hjust = 1))
```

# 2. Month by month

WIS reduction % (positive = calibrated WIS lower than base) by calendar month of the reference date, one figure per
constant rate; rows are seasons, columns horizons. Grey bars are each
month's share of the season's base WIS. Clipped at −40%.

```{r month, results="asis", fig.height=5.5}
month_m <- ws_month_scores(fc %>% mutate(config = interaction(rate, decay, sep = "|"))) %>%
  tidyr::separate(config, c("rate", "decay"), sep = "[|]") %>%
  mutate(decay = factor(decay, levels = names(DECAYS)), h = factor(paste0("h", horizon), levels = paste0("h", 0:3)))
for (r in names(RATES)) {
  cat("\n\n## ", r, "\n\n", sep = "")
  d <- month_m %>% filter(rate == r)
  share <- d %>% filter(decay == "no decay")
  print(
    ggplot(d, aes(month, pct, colour = decay, group = decay)) +
      geom_col(data = share, aes(month, share), inherit.aes = FALSE, fill = "grey85", width = 0.7) +
      geom_hline(yintercept = 0, colour = "grey40") +
      geom_line() +
      geom_point(size = 1.1) +
      facet_grid(season ~ h, scales = "free_x", space = "free_x") +
      scale_colour_manual(values = DECAY_COLS, name = NULL) +
      coord_cartesian(ylim = c(-40, 30)) +
      labs(x = "month of reference date", y = "WIS reduction % (positive = calibrated\nWIS lower than base); grey: % of season base WIS", title = r) +
      theme(legend.position = "bottom")
  )
}
```

## Share of outcomes below the median, March–May

0.5 is calibrated; above 0.5 the median is too high. h2, states only.

```{r month-below}
month_m %>%
  filter(h == "h2", month %in% c("Mar", "Apr", "May")) %>%
  transmute(rate, season, month, decay, v = sprintf("%+.1f (%.2f)", pct, below_med_cal)) %>%
  tidyr::pivot_wider(names_from = decay, values_from = v) %>%
  left_join(
    month_m %>% filter(h == "h2", decay == "no decay") %>% transmute(rate, season, month, `base below median` = round(below_med_base, 2)),
    by = c("rate", "season", "month")
  ) %>%
  arrange(rate, season, month) %>%
  knitr::kable(caption = "h2: WIS reduction %, positive = calibrated WIS lower than base (share of outcomes below the calibrated median)")
```

# 3. Mean offset at the median over time

Calibrated minus base median in admissions per 100k, averaged over states at
each round, h2.

```{r offset-time, fig.height=7}
pop <- wi$rate_scales %>% select(location, scale)
fc %>%
  filter(level == 0.5, horizon == 2) %>%
  left_join(pop, by = "location") %>%
  group_by(rate, decay, season, reference_date) %>%
  summarize(off = mean((value_cal - value_base) / scale), .groups = "drop") %>%
  ggplot(aes(reference_date, off, colour = decay)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line() +
  geom_point(size = 0.6) +
  facet_grid(rate ~ season, scales = "free") +
  scale_colour_manual(values = DECAY_COLS, name = NULL) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = "reference date", y = "mean offset at the median, per 100k") +
  theme(legend.position = "bottom")
```

```{r month-view-cfgs}
# sqrt 0.018 with each decay setting; section 2 has the WIS months for the rate-scale trackers.
mv_plot_cfgs <- paste(names(RATES)[3], names(DECAYS), sep = ", ")
mv_cols <- setNames(DECAY_COLS, mv_plot_cfgs)
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
