---
title: "E18: a bridge from REF-op to sqrt constant 0.018"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: REF-op (prod's tracker) and sqrt constant 0.018 (E16) differ in
several settings. How do WIS and coverage change as one setting at a time is
changed from one to the other?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Steps 0–3 run with the 2023-24 season present (the HHS-stitched backfill
  from E13); steps 4–6 and the references are cold.
- Scores over the two live seasons, each on its own rounds; states only (US
  excluded).

Each row changes one setting from the row above:

| step | change | config after the change |
|---|---|---|
| 0 REF-op | — | sqrt, adaptive mult 0.03 floor 1e-3, window 20, carry; slow term (mult 0.003) warm-started from the 2023-24 burn-in; leak 0.1 on the fast term |
| 1 no warm start | `slow_init = NULL` | the slow term starts at zero |
| 2 no slow term | `lr_slow = NULL` | single fast term with the leak |
| 3 no leak | `fast_decay = 0` | sqrt, adaptive, mult 0.03, window 20, carry, with the 2023-24 season present |
| 4 no burn-in season | drop the 2023-24 rounds | as 3, cold: the eta pool starts empty |
| 5 per-100k units | `scales` = population / 100k | as 4, on the square root of admissions per 100k |
| 6 sqrt constant 0.018 (E16) | `lr = 0.018` | sqrt per 100k, constant 0.018 |

The 2023-24 season does two things: with `slow_init` it sets the slow term's
starting value (step 1 removes this), and its residuals fill the adaptive
learning rate's window, so eta is sized from the first live round (step 4
removes this). Burn-in outcomes never step the offsets. Step 5 changes only
units, so that step 6 changes only the learning rate.

Outside the chain: a branch, sqrt 0.018 warm, which is step 6 with the 2023-24 season present and `slow_init =
"burn_in_quantile"` (the warm start of step 0, here on the single term).

Steps 0, 3, 4 and 6 are the cached runs of E13 (REF-op warm), E02
(`reference + burn-in`), E01/E02 (`reference`) and E16 (sqrt 0.018), so they
are not independent checks of those notebooks.

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better. "Month-avg" coverage
bias is computed within each calendar month and then averaged.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
fc_burn_in <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Each config is written as the experiment that first ran it writes it, so cached runs are reused.
refop_warm <- ws_configs(wi$rate_scales, burn_in = TRUE)[["REF-op warm"]]
sqrt_adaptive <- list(
  ref = "paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3),
  burn_in_seasons = character(0), slow_init = NULL, lr_window = 20, season_policy = "carry"
)
const_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
STEPS <- list(
  `0 REF-op` = list(cfg = refop_warm, burn_in = TRUE),
  `1 no warm start` = list(cfg = list(ref = "op", burn_in_seasons = "2023-2024", slow_init = NULL), burn_in = TRUE),
  `2 no slow term` = list(cfg = list(ref = "op", burn_in_seasons = "2023-2024", slow_init = NULL, lr_slow = NULL), burn_in = TRUE),
  `3 no leak` = list(cfg = utils::modifyList(sqrt_adaptive, list(burn_in_seasons = "2023-2024")), burn_in = TRUE),
  `4 no burn-in season` = list(cfg = sqrt_adaptive, burn_in = FALSE),
  `5 per-100k units` = list(cfg = c(sqrt_adaptive, list(scales = wi$rate_scales)), burn_in = FALSE),
  `6 sqrt constant 0.018 (E16)` = list(cfg = utils::modifyList(const_cfg, list(lr = 0.018, transform = "sqrt")), burn_in = FALSE)
)
# Outside the chain: step 6 with a warm start.
REFS <- list(
  `branch: sqrt 0.018 warm` = list(cfg = utils::modifyList(STEPS[[7]]$cfg, list(burn_in_seasons = "2023-2024", slow_init = "burn_in_quantile")), burn_in = TRUE)
)
run <- function(spec) {
  fc <- if (spec$burn_in) fc_burn_in else wi$fc
  do.call(cal_run_cached, c(list(fc, wi$truth), spec$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
cals <- purrr::map(c(STEPS, REFS), run)
```

```{r metrics}
SEASONS <- c(LIVE, "both")
STEP_COLS <- setNames(c("#000000", "#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#0072B2", "#D55E00", "#AA4499"), names(c(STEPS, REFS)))
step_m <- purrr::imap(cals, function(cal, nm) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date)) %>% filter(location != "US")
  ws_scores(fc) %>%
    select(season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(step = nm, wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}) %>%
  bind_rows() %>%
  mutate(
    step = factor(step, levels = names(c(STEPS, REFS))),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
mv <- cal_month_view(cals, seasons = LIVE)
```

```{r month-view-cfgs}
# Five configs, so the month plots stay readable; the tables below them cover every config.
mv_plot_cfgs <- c("0 REF-op", "1 no warm start", "4 no burn-in season", "6 sqrt constant 0.018 (E16)", "branch: sqrt 0.018 warm")
mv_cols <- STEP_COLS
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# What each step changes

The change each step makes to the step before it, both seasons pooled,
states only. WIS: change in WIS reduction % (positive = the step lowers
WIS). Coverage: change in L1 coverage bias, season-pooled and month-average
(negative = the step improves coverage).

```{r deltas}
wis_both <- step_m %>%
  filter(season == "both seasons", step %in% names(STEPS)) %>%
  transmute(step, h, v = wis_pct)
cov_both <- mv$cov %>%
  filter(season == "both") %>%
  transmute(step = config, h = factor(paste0("h", horizon), levels = paste0("h", 0:3)), pooled = pooled_cal, mavg = mavg_cal)
delta <- wis_both %>%
  left_join(cov_both, by = c("step", "h")) %>%
  arrange(h, step) %>%
  group_by(h) %>%
  mutate(d_wis = v - lag(v), d_pooled = pooled - lag(pooled), d_mavg = mavg - lag(mavg)) %>%
  ungroup() %>%
  filter(step != names(STEPS)[1])
knitr::kable(
  delta %>%
    transmute(step, h, cell = sprintf("%+.1f / %+.3f / %+.3f", d_wis, d_pooled, d_mavg)) %>%
    tidyr::pivot_wider(names_from = h, values_from = cell) %>%
    arrange(step),
  caption = "Both seasons, states only: change from the previous step in WIS reduction % / season-pooled L1 coverage bias / month-avg L1 coverage bias"
)
```

# State by state: REF-op against sqrt 0.018

Per state, pooled over h0–h3: WIS reduction % of REF-op against sqrt 0.018
(above the line, REF-op is better).

```{r state-scatter, fig.height=5}
PAIR <- c("0 REF-op", "6 sqrt constant 0.018 (E16)")
state_wis <- cal_state_wis(cals[PAIR])
state_pct <- state_wis %>%
  filter(config %in% PAIR) %>%
  group_by(config, season, location) %>%
  summarize(pct = 100 * (sum(wis_base) - sum(wis_cal)) / sum(wis_base), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = config, values_from = pct)
ggplot(state_pct, aes(.data[[PAIR[2]]], .data[[PAIR[1]]])) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(alpha = 0.8) +
  facet_wrap(~season) +
  labs(x = "sqrt 0.018: WIS reduction %, h0–h3", y = "REF-op: WIS reduction %, h0–h3")
```
