---
title: "E18: a bridge from REF-op to sqrt constant 0.018"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: REF-op (prod's tracker) and sqrt constant 0.018 (E16) differ in
several settings. How do WIS and coverage change as one setting at a time is
changed from one to the other?

Setup:

- Disease: flu. Base forecaster: the clean `windowed_seasonal` replay (53
  locations, the 55 hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Steps 0–3 run with the 2023-24 season present (the HHS-stitched backfill
  from E13); steps 4–6 and the references are cold.
- Scores over the two live seasons, each on its own rounds; states only (US
  excluded).

Each row changes one setting from the row above:

| step | change | config after the change |
|---|---|---|
| 0 REF-op | — | sqrt, adaptive mult 0.03 floor 1e-3, window 20, carry; slow term (mult 0.003) warm-started from the 2023-24 burn-in; leak 0.1 on the fast term |
| 1 no warm start | `slow_init = NULL` | the slow term starts at zero |
| 2 no slow term | `lr_slow = NULL` | single fast term with the leak |
| 3 no leak | `fast_decay = 0` | sqrt, adaptive, mult 0.03, window 20, carry, with the 2023-24 season present |
| 4 no burn-in season | drop the 2023-24 rounds | as 3, cold: the eta pool starts empty |
| 5 per-100k units | `scales` = population / 100k | as 4, on the square root of admissions per 100k |
| 6 sqrt constant 0.018 (E16) | `lr = 0.018` | sqrt per 100k, constant 0.018 |

The 2023-24 season does two things: with `slow_init` it sets the slow term's
starting value (step 1 removes this), and its residuals fill the adaptive
learning rate's window, so eta is sized from the first live round (step 4
removes this). Burn-in outcomes never step the offsets. Step 5 changes only
units, so that step 6 changes only the learning rate.

Two rate-scale constant rates are shown as references outside the chain:
rate 0.032 (E14) and rate 0.056 (E09's knee).

Steps 0, 3, 4 and 6 are the cached runs of E13 (REF-op warm), E02
(`reference + burn-in`), E01/E02 (`reference`) and E16 (sqrt 0.018), so they
are not independent checks of those notebooks.

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better. "Month-avg" coverage
bias is computed within each calendar month and then averaged.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
fc_burn_in <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Each config is written as the experiment that first ran it writes it, so cached runs are reused.
refop_warm <- ws_configs(wi$rate_scales, burn_in = TRUE)[["REF-op warm"]]
sqrt_adaptive <- list(
  ref = "paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3),
  burn_in_seasons = character(0), slow_init = NULL, lr_window = 20, season_policy = "carry"
)
const_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
STEPS <- list(
  `0 REF-op` = list(cfg = refop_warm, burn_in = TRUE),
  `1 no warm start` = list(cfg = list(ref = "op", burn_in_seasons = "2023-2024", slow_init = NULL), burn_in = TRUE),
  `2 no slow term` = list(cfg = list(ref = "op", burn_in_seasons = "2023-2024", slow_init = NULL, lr_slow = NULL), burn_in = TRUE),
  `3 no leak` = list(cfg = utils::modifyList(sqrt_adaptive, list(burn_in_seasons = "2023-2024")), burn_in = TRUE),
  `4 no burn-in season` = list(cfg = sqrt_adaptive, burn_in = FALSE),
  `5 per-100k units` = list(cfg = c(sqrt_adaptive, list(scales = wi$rate_scales)), burn_in = FALSE),
  `6 sqrt constant 0.018 (E16)` = list(cfg = utils::modifyList(const_cfg, list(lr = 0.018, transform = "sqrt")), burn_in = FALSE)
)
# Rate-scale constant rates for comparison, outside the chain: E14's best and E09's knee, as E14 runs them.
e14_const <- ws_configs(wi$rate_scales, burn_in = TRUE)[["E11 constant 0.1"]]
REFS <- list(
  `ref: rate 0.032 (E14)` = list(cfg = utils::modifyList(e14_const, list(lr = 0.032)), burn_in = FALSE),
  `ref: rate 0.056 (E09 knee)` = list(cfg = utils::modifyList(e14_const, list(lr = 0.056)), burn_in = FALSE)
)
run <- function(spec) {
  fc <- if (spec$burn_in) fc_burn_in else wi$fc
  do.call(cal_run_cached, c(list(fc, wi$truth), spec$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
cals <- purrr::map(c(STEPS, REFS), run)
```

```{r metrics}
SEASONS <- c(LIVE, "both")
STEP_COLS <- setNames(c("#000000", "#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#0072B2", "#D55E00", "#999999", "#882255"), names(c(STEPS, REFS)))
step_m <- purrr::imap(cals, function(cal, nm) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date)) %>% filter(location != "US")
  ws_scores(fc) %>%
    select(season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(step = nm, wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}) %>%
  bind_rows() %>%
  mutate(
    step = factor(step, levels = names(c(STEPS, REFS))),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
mv <- cal_month_view(cals, seasons = LIVE)
step_table <- function(col, fmt, base_col = NULL) {
  rows <- step_m %>% mutate(step = as.character(step), v = .data[[col]])
  if (!is.null(base_col)) {
    rows <- bind_rows(step_m %>% filter(step == names(STEPS)[1]) %>% mutate(step = "base", v = .data[[base_col]]), rows)
  }
  rows %>%
    mutate(step = factor(step, levels = unique(step))) %>%
    arrange(season, h) %>%
    transmute(step, col = paste(season, h), cell = sprintf(fmt, v)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(step)
}
```

# 1. Season scores per step

States only. Each season is scored on its own rounds; "both seasons" pools
them. The tracker runs over both seasons, so 2025-26 starts from the offsets
2024-25 left behind.

## WIS reduction % (positive = calibrated WIS lower than base)

```{r table-wis}
knitr::kable(step_table("wis_pct", "%+.1f"), caption = "WIS reduction % (positive = calibrated WIS lower than base), states only")
```

## L1 coverage bias (lower is better), season-pooled

```{r table-cov}
knitr::kable(step_table("cal_err_cal", "%.3f", base_col = "cal_err_base"), caption = "L1 coverage bias (lower is better), season-pooled, states only")
```

## L1 coverage bias (lower is better), season-pooled and month-average

```{r table-month-cov}
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

# 2. What each step changes

The change each step makes to the step before it, both seasons pooled,
states only. WIS: change in WIS reduction % (positive = the step lowers
WIS). Coverage: change in L1 coverage bias, season-pooled and month-average
(negative = the step improves coverage).

```{r deltas}
wis_both <- step_m %>%
  filter(season == "both seasons", step %in% names(STEPS)) %>%
  transmute(step, h, v = wis_pct)
cov_both <- mv$cov %>%
  filter(season == "both") %>%
  transmute(step = config, h = factor(paste0("h", horizon), levels = paste0("h", 0:3)), pooled = pooled_cal, mavg = mavg_cal)
delta <- wis_both %>%
  left_join(cov_both, by = c("step", "h")) %>%
  arrange(h, step) %>%
  group_by(h) %>%
  mutate(d_wis = v - lag(v), d_pooled = pooled - lag(pooled), d_mavg = mavg - lag(mavg)) %>%
  ungroup() %>%
  filter(step != names(STEPS)[1])
knitr::kable(
  delta %>%
    transmute(step, h, cell = sprintf("%+.1f / %+.3f / %+.3f", d_wis, d_pooled, d_mavg)) %>%
    tidyr::pivot_wider(names_from = h, values_from = cell) %>%
    arrange(step),
  caption = "Both seasons, states only: change from the previous step in WIS reduction % / season-pooled L1 coverage bias / month-avg L1 coverage bias"
)
```

```{r delta-plot, fig.height=6.5}
delta_long <- delta %>%
  transmute(step, h,
    `WIS reduction %, change\n(positive = step lowers WIS)` = d_wis,
    `month-avg L1 coverage bias,\nchange (negative = better)` = d_mavg,
    `season-pooled L1 coverage bias,\nchange (negative = better)` = d_pooled
  ) %>%
  tidyr::pivot_longer(-c(step, h), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = unique(metric)), step = factor(step, levels = rev(names(STEPS)[-1])))
ggplot(delta_long, aes(value, step, fill = step)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = STEP_COLS) +
  facet_grid(h ~ metric, scales = "free_x") +
  labs(x = "change from the previous step (both seasons, states only)", y = NULL)
```

```{r month-view-cfgs}
mv_cols <- STEP_COLS
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Findings

There are no uncertainty estimates; a WIS gap under 1 point or a month-avg
coverage gap under 0.01 is read as no clear difference. The cold steps equal
the base through Nov–Dec 2024; only the warm steps act there. States only,
h0/h1/h2/h3, both seasons pooled unless stated.

- **Endpoints:** REF-op +2.8/+1.1/+0.9/+1.1, sqrt constant 0.018
  +2.2/+0.8/+0.6/+0.6 (0.3–0.6 apart, no clear difference). The order flips
  by season: REF-op is ahead in 2024-25 (+3.2/+1.9/+1.6/+1.6 vs
  +1.7/+0.6/+0.3/+0.1), sqrt 0.018 in 2025-26 (+3.0/+1.0/+1.1/+1.5 vs
  +2.2/−0.1/−0.0/+0.2).
- **Warm start (step 1):** −0.9/−0.5/−0.6/−1.0 pooled when removed. In
  2024-25 it is worth 0.9–2.0 points; in 2025-26 removing it gains 0.2–0.7
  at h1–h3.
- **Slow term without warm start (step 2):** within ±0.2. No strong result.
- **Leak (step 3):** removing it costs 0.6–1.2 at h1–h3 and gains 0.4 at h0.
- **Burn-in season's eta warm-up (step 4):** −0.1 to −0.7, in 2024-25 only.
  No strong result.
- **Per-100k units (step 5):** identical to step 4 in every reported digit.
  The adaptive rate is scale invariant when the 1e-3 floor doesn't bind; a
  direct check (2026-10-02) confirmed `scales` is applied.
- **Constant 0.018 vs cold adaptive (step 6):** +0.3/+1.0/+1.7/+2.3, ahead at
  h2–h3 in both seasons.
- **Month-avg coverage:** no step changes it by more than 0.008. REF-op
  0.095/0.121/0.150/0.175, sqrt 0.018 0.094/0.116/0.143/0.169, base
  0.112/0.127/0.153/0.177. Season-pooled bias moves by up to 0.06 per step,
  but pooling lets ramp and post-peak errors cancel.
- **References:** rate 0.032 +1.3/+0.7/+0.5/+0.4 and rate 0.056
  +1.3/+0.5/+0.3/+0.0 pooled. Month-avg coverage: rate 0.032 equals the base;
  rate 0.056 is 0.008–0.014 above it.
