---
title: "E18: a bridge from REF-op to sqrt constant 0.018"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

REF-op (prod's tracker) and sqrt constant 0.018 (E16's best) differ in
several settings at once. This notebook walks from one to the other,
changing one setting per step, to see which piece of REF-op does the work.

**Setup.** As E13/E16: the clean `windowed_seasonal` replay for flu (53
locations, the 55 hub rounds, h0–h3), learning from data as published each
week and re-run from scratch each week as prod does, one week of extra
revision delay. Scores are over the two live seasons; results are states
only (US excluded).

**The steps.** Each row changes one setting from the row above.

| step | change | config after the change |
|---|---|---|
| 0 REF-op | — | sqrt, adaptive mult 0.03 floor 1e-3, window 20, carry; slow term (mult 0.003) warm-started from the 2023-24 burn-in; leak 0.1 on the fast term |
| 1 no warm start | `slow_init = NULL` | the slow term starts at zero |
| 2 no slow term | `lr_slow = NULL` | single fast term with the leak |
| 3 no leak | `fast_decay = 0` | sqrt, adaptive, mult 0.03, window 20, carry, with the 2023-24 season present |
| 4 no burn-in season | drop the 2023-24 rounds | as 3, cold: the eta pool starts empty |
| 5 per-100k units | `scales` = population / 100k | as 4, on the square root of admissions per 100k |
| 6 sqrt constant 0.018 (E16) | `lr = 0.018` | E16's best: sqrt per 100k, constant 0.018 |

**Why steps 4 and 5.** The 2023-24 burn-in season does two things. With
`slow_init` it sets the slow term's starting value (the warm start). Without
it, its residuals still fill the adaptive learning rate's window, so eta is
sized from the first live round instead of growing from an empty pool; the
offsets themselves stay at zero, since burn-in outcomes never step. So the
burn-in season stays present through step 3, and dropping it is step 4: step
1 isolates the warm start, step 4 the eta warm-up (E13 follow-up 2). E16's
constant rate runs on the square root of the per-100k rate. The adaptive
rate is almost unchanged by that rescaling (it scales with the residuals;
only the 1e-3 floor does not), but it is still a change, so it is step 5,
and step 6 changes only the learning rate.

Steps 0, 3, 4 and 6 are the cached runs of E13 (REF-op warm), E02
("reference + burn-in"), E01/E02 ("reference") and E16 (sqrt 0.018). Step 3
is written as E02 writes it (REF-paper with sqrt and the 1e-3 floor), which
equals REF-op with `slow_init`, `lr_slow` and the leak removed.

"WIS reduction %" is `100 × (WIS_base − WIS_cal) / WIS_base`, so
**positive = calibrated WIS lower than base**. "L1 coverage bias" (lower is
better) is the mean over the 23 levels of |empirical coverage − level|.
"Month-avg" coverage bias is computed within each calendar month and then
averaged, so errors on the ramp and after the peak cannot cancel.

# Findings

Numbers are from the render of 1 October 2026, states only, both seasons
pooled unless named, h0/h1/h2/h3. WIS is WIS reduction % (positive =
calibrated WIS lower than base); coverage is L1 coverage bias (lower is
better).

- **Endpoints.** REF-op +2.8/+1.1/+0.9/+1.1, sqrt constant 0.018
  +2.2/+0.8/+0.6/+0.6. REF-op is ahead overall because of 2024-25
  (+3.2/+1.9/+1.6/+1.6 vs +1.7/+0.6/+0.3/+0.1); sqrt 0.018 is ahead in
  2025-26 (+3.0/+1.0/+1.1/+1.5 vs +2.2/−0.1/−0.0/+0.2). Month-avg coverage
  bias is about the same: 0.095/0.121/0.150/0.175 vs 0.094/0.116/0.143/0.169.
- **WIS: the learning rate moves it most at h1–h3, the warm start at h0.**
  Switching the adaptive rate to constant 0.018 (step 6) adds
  +0.3/+1.0/+1.7/+2.3. Inside REF-op, three pieces each add about a point:
  the warm start (step 1 costs −0.9/−0.5/−0.6/−1.0, nearly all of it in
  2024-25: h3 goes from +1.6 to −0.4), the leak (step 3 costs
  −0.6/−1.0/−1.2 at h1–h3 and gains +0.4 at h0) and, smaller, the burn-in
  season's eta warm-up (step 4 costs −0.1/−0.2/−0.5/−0.7, again in
  2024-25). Without the warm start, the slow term does nothing (step 2:
  within ±0.2).
- **Warm start vs eta warm-up (E13 follow-up 2).** Both matter only in
  2024-25, the first live season. The warm start is worth about twice as
  much WIS as the eta warm-up and shows from the first rounds (the month
  view gains at h2–h3 in Nov–Feb 2024-25). The eta warm-up is about half a
  point at h2–h3 and nothing in 2025-26.
- **Month-level coverage: no step moves it beyond noise.** Every step
  changes month-avg coverage bias by 0.008 or less, under the ~0.01 noise
  floor. The largest are dropping the warm start at h3 (−0.008) and the
  constant rate at h1–h3 (−0.005/−0.007/−0.007).
- **Season-pooled coverage moves a lot, and it is cancellation.** Dropping
  the warm start worsens pooled bias by +0.024/+0.032/+0.044/+0.052,
  dropping the leak improves it by −0.040/−0.042/−0.051/−0.060, and the
  constant rate worsens it by +0.014/+0.023/+0.034/+0.047. None of these
  shows up month by month, so they come from ramp and post-peak errors
  cancelling differently, not from better-calibrated months.
- **Where the WIS steps act: March and April.** The month view shows the
  WIS steps mostly as spring losses at h2–h3. In 2024-25 at h2, March costs
  about −13% with the leak and no slow term (step 2), −23% without the leak
  (step 3) and −31% without the burn-in season (steps 4–5); constant 0.018
  costs about −7%. April is worse for the adaptive trackers without the leak
  (down to the −40% clip). In 2025-26 every step loses 27–40% in March at
  h1–h2, constant 0.018 slightly less.
- **Per-100k units change nothing (step 5).** The adaptive rate scales with
  the residuals, so on the square-root scale only the 1e-3 floor sees the
  units. Calibrated values differ by at most 1.6 admissions and every score
  matches step 4 to the reported digits.
- **Rate-scale constant rates (references, outside the chain).** Rate
  0.032 (E14's best) and 0.056 (E09's knee) give +1.3/+0.7/+0.5/+0.4 and
  +1.3/+0.5/+0.3/0.0, below sqrt 0.018 (+2.2/+0.8/+0.6/+0.6) and REF-op
  (+2.8/+1.1/+0.9/+1.1). Month-avg coverage bias is about the base's
  (0.108–0.177 and 0.112–0.191 vs base 0.112–0.177; 0.056 is worse at
  h2–h3). Both sit at zero through the ramp. On the ensemble (E09) the knee
  looked good after the 2024-25 peak because the ensemble was still
  under-predicting that spring (E07); `windowed_seasonal` over-predicts
  after the peak, so here the same upward offsets cost WIS and coverage in
  March–April.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
wi <- ws_inputs("flu")
fc_burn_in <- bind_rows(ws_burn_in_forecasts("flu"), wi$fc)
# Each config is written as the experiment that first ran it writes it, so cached runs are reused.
refop_warm <- ws_configs(wi$rate_scales, burn_in = TRUE)[["REF-op warm"]]
sqrt_adaptive <- list(
  ref = "paper", transform = "sqrt", lr_args = list(mult = 0.03, floor = 1e-3),
  burn_in_seasons = character(0), slow_init = NULL, lr_window = 20, season_policy = "carry"
)
const_cfg <- ws_configs(wi$rate_scales, burn_in = FALSE)[["E11 constant 0.1"]]
STEPS <- list(
  `0 REF-op` = list(cfg = refop_warm, burn_in = TRUE),
  `1 no warm start` = list(cfg = list(ref = "op", burn_in_seasons = "2023-2024", slow_init = NULL), burn_in = TRUE),
  `2 no slow term` = list(cfg = list(ref = "op", burn_in_seasons = "2023-2024", slow_init = NULL, lr_slow = NULL), burn_in = TRUE),
  `3 no leak` = list(cfg = utils::modifyList(sqrt_adaptive, list(burn_in_seasons = "2023-2024")), burn_in = TRUE),
  `4 no burn-in season` = list(cfg = sqrt_adaptive, burn_in = FALSE),
  `5 per-100k units` = list(cfg = c(sqrt_adaptive, list(scales = wi$rate_scales)), burn_in = FALSE),
  `6 sqrt constant 0.018 (E16)` = list(cfg = utils::modifyList(const_cfg, list(lr = 0.018, transform = "sqrt")), burn_in = FALSE)
)
# Rate-scale constant rates for comparison, outside the chain: E14's best and E09's knee, as E14 runs them.
e14_const <- ws_configs(wi$rate_scales, burn_in = TRUE)[["E11 constant 0.1"]]
REFS <- list(
  `ref: rate 0.032 (E14)` = list(cfg = utils::modifyList(e14_const, list(lr = 0.032)), burn_in = FALSE),
  `ref: rate 0.056 (E09 knee)` = list(cfg = utils::modifyList(e14_const, list(lr = 0.056)), burn_in = FALSE)
)
run <- function(spec) {
  fc <- if (spec$burn_in) fc_burn_in else wi$fc
  do.call(cal_run_cached, c(list(fc, wi$truth), spec$cfg, list(learn = "exact", vintages = wi$vintages, workers = params$workers)))
}
cals <- purrr::map(c(STEPS, REFS), run)
```

```{r metrics}
SEASONS <- c(LIVE, "both")
STEP_COLS <- setNames(c("#000000", "#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#0072B2", "#D55E00", "#999999", "#882255"), names(c(STEPS, REFS)))
step_m <- purrr::imap(cals, function(cal, nm) {
  fc <- cal$forecasts %>% mutate(season = season_of(reference_date)) %>% filter(location != "US")
  ws_scores(fc) %>%
    select(season, horizon, which, wis, cal_err) %>%
    tidyr::pivot_wider(names_from = which, values_from = c(wis, cal_err)) %>%
    mutate(step = nm, wis_pct = 100 * (wis_base - wis_cal) / wis_base)
}) %>%
  bind_rows() %>%
  mutate(
    step = factor(step, levels = names(c(STEPS, REFS))),
    season = factor(season, levels = SEASONS, labels = c(LIVE, "both seasons")),
    h = factor(paste0("h", horizon), levels = paste0("h", 0:3))
  )
mv <- cal_month_view(cals, seasons = LIVE)
step_table <- function(col, fmt, base_col = NULL) {
  rows <- step_m %>% mutate(step = as.character(step), v = .data[[col]])
  if (!is.null(base_col)) {
    rows <- bind_rows(step_m %>% filter(step == names(STEPS)[1]) %>% mutate(step = "base", v = .data[[base_col]]), rows)
  }
  rows %>%
    mutate(step = factor(step, levels = unique(step))) %>%
    arrange(season, h) %>%
    transmute(step, col = paste(season, h), cell = sprintf(fmt, v)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(step)
}
```

# 1. Season scores per step

States only. Each season is scored on its own rounds; "both seasons" pools
them. The tracker runs over both seasons, so 2025-26 starts from the offsets
2024-25 left behind.

## WIS reduction % (positive = calibrated WIS lower than base)

```{r table-wis}
knitr::kable(step_table("wis_pct", "%+.1f"), caption = "WIS reduction % (positive = calibrated WIS lower than base), states only")
```

## L1 coverage bias (lower is better), season-pooled

```{r table-cov}
knitr::kable(step_table("cal_err_cal", "%.3f", base_col = "cal_err_base"), caption = "L1 coverage bias (lower is better), season-pooled, states only")
```

## L1 coverage bias (lower is better), season-pooled and month-average

```{r table-month-cov}
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv), caption = CAL_MONTH_COV_CAPTION)
```

# 2. What each step changes

The change each step makes to the step before it, both seasons pooled,
states only. WIS: change in WIS reduction % (positive = the step lowers
WIS). Coverage: change in L1 coverage bias, season-pooled and month-average
(negative = the step improves coverage).

```{r deltas}
wis_both <- step_m %>%
  filter(season == "both seasons", step %in% names(STEPS)) %>%
  transmute(step, h, v = wis_pct)
cov_both <- mv$cov %>%
  filter(season == "both") %>%
  transmute(step = config, h = factor(paste0("h", horizon), levels = paste0("h", 0:3)), pooled = pooled_cal, mavg = mavg_cal)
delta <- wis_both %>%
  left_join(cov_both, by = c("step", "h")) %>%
  arrange(h, step) %>%
  group_by(h) %>%
  mutate(d_wis = v - lag(v), d_pooled = pooled - lag(pooled), d_mavg = mavg - lag(mavg)) %>%
  ungroup() %>%
  filter(step != names(STEPS)[1])
knitr::kable(
  delta %>%
    transmute(step, h, cell = sprintf("%+.1f / %+.3f / %+.3f", d_wis, d_pooled, d_mavg)) %>%
    tidyr::pivot_wider(names_from = h, values_from = cell) %>%
    arrange(step),
  caption = "Both seasons, states only: change from the previous step in WIS reduction % / season-pooled L1 coverage bias / month-avg L1 coverage bias"
)
```

```{r delta-plot, fig.height=6.5}
delta_long <- delta %>%
  transmute(step, h,
    `WIS reduction %, change\n(positive = step lowers WIS)` = d_wis,
    `month-avg L1 coverage bias,\nchange (negative = better)` = d_mavg,
    `season-pooled L1 coverage bias,\nchange (negative = better)` = d_pooled
  ) %>%
  tidyr::pivot_longer(-c(step, h), names_to = "metric") %>%
  mutate(metric = factor(metric, levels = unique(metric)), step = factor(step, levels = rev(names(STEPS)[-1])))
ggplot(delta_long, aes(value, step, fill = step)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = STEP_COLS) +
  facet_grid(h ~ metric, scales = "free_x") +
  labs(x = "change from the previous step (both seasons, states only)", y = NULL)
```

```{r month-view-cfgs}
mv_cols <- STEP_COLS
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```
