---
title: "E20: warm start and leak on both learning-rate anchors, flu and covid"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
# The month view is knit once per disease, so its chunk labels repeat.
options(knitr.duplicate.label = "allow")
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
```

Question: do the warm start and the leak change WIS and coverage in the same
way on the adaptive learning rate and on the constant one, and on both
diseases? E04 and E05 asked this on the flu ensemble with the adaptive rate
only; E18 found that the leak's effect depends on how large the adaptive
step gets late in the season, so the two anchors may answer differently.

Setup:

- Diseases: flu and covid. Base forecaster: the clean `windowed_seasonal`
  replay (53 locations, the hub rounds). Horizons h0–h3.
- Learning truth: exact (re-run at every round on the data published by then;
  E00), one week of extra delay.
- Warm configs run with the 2023-24 season present (the HHS-stitched backfill
  from E13) and start the offset from that season's error quantiles; cold
  configs start at zero with no burn-in rounds, so they play the base through
  the first live rounds.
- Scores over the two live seasons, each on its own rounds; states only (US
  excluded).

The grid is 2 × 2 × 2 (E18 step 2 shows the slow term does nothing, so it is
left out), plus prod's two configs:

| factor | levels |
|---|---|
| anchor | `adaptive 0.03`: sqrt scale, eta = 0.03 × the 90th percentile of the last 20 rounds' absolute errors, floor 0.001 (E01's elbow). `constant 0.018`: sqrt of admissions per 100k, eta = 0.018 (E16's elbow, picked on flu) |
| start | `cold`: offsets start at zero. `warm`: `slow_init = "burn_in_quantile"` from the 2023-24 season |
| leak | `no leak`. `leak 0.1`: the offset shrinks by 10% each round |
| references | `REF-op warm` (flu prod: adaptive, warm, leak, plus a slow term); `REF-op cold` (covid prod) |

WIS reduction % = 100 × (base − calibrated) / base, so positive = calibrated
WIS lower than base. L1 coverage bias: lower is better. "Month-avg" coverage
bias is computed within each calendar month and then averaged.

```{r load}
source(here::here("scripts/calibration/calibration_ws_replay.R"))
DISEASES <- c(flu = "flu", covid = "covid")
RUNS <- purrr::map(DISEASES, function(d) {
  wi <- ws_inputs(d)
  fc_burn_in <- bind_rows(ws_burn_in_forecasts(d), wi$fc)
  specs <- c(ws_grid_configs(wi$rate_scales), ws_prod_specs(wi$rate_scales))
  list(wi = wi, specs = specs, cals = purrr::map(specs, ws_run_spec, wi = wi, fc_burn_in = fc_burn_in, workers = params$workers))
})
grid_names <- names(ws_grid_configs(RUNS$flu$wi$rate_scales))
split_name <- function(d) {
  d %>%
    mutate(config = as.character(config)) %>%
    tidyr::separate_wider_delim(config, " | ", names = c("anchor", "start", "leak"), cols_remove = FALSE, too_few = "align_start")
}
ANCHOR_COLS <- c(`adaptive 0.03` = "#E69F00", `constant 0.018` = "#0072B2")
h_levels <- paste0("h", 0:3)
```

```{r metrics}
M <- purrr::map(RUNS, function(r) {
  m <- ws_wis_by_season(r$cals)
  mv <- cal_month_view(r$cals, seasons = LIVE)
  cov <- mv$cov %>% transmute(config, season, h = factor(paste0("h", horizon), levels = h_levels), pooled = pooled_cal, mavg = mavg_cal)
  list(m = m, mv = mv, cov = cov)
})
```

# WIS by season and horizon

```{r wis-table, results="asis"}
for (d in names(M)) {
  cat("\n\n", cal_h(2), " ", d, "\n\n", sep = "")
  M[[d]]$m %>%
    transmute(config, col = paste(season, h), cell = sprintf("%+.1f", wis_pct)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    knitr::kable(caption = paste0(d, ": WIS reduction % (positive = calibrated WIS lower than base), states only; `both` pools the seasons")) %>%
    print()
}
```

# What each factor changes, per anchor

The change from adding the warm start (warm − cold) at each leak level, and
from adding the leak (leak − no leak) at each start, per anchor. WIS: change
in WIS reduction % (positive = the factor lowers WIS). Coverage: change in
month-avg L1 coverage bias (negative = the factor improves coverage).

```{r factor-deltas, results="asis"}
contrast <- function(d, factor, hi, lo, held) {
  d %>%
    select(anchor, start, leak, season, h, wis_pct, mavg) %>%
    tidyr::pivot_wider(names_from = all_of(factor), values_from = c(wis_pct, mavg)) %>%
    transmute(
      anchor, season, h,
      factor = paste0("+ ", hi, " (", .data[[held]], ")"),
      d_wis = .data[[paste0("wis_pct_", hi)]] - .data[[paste0("wis_pct_", lo)]],
      d_mavg = .data[[paste0("mavg_", hi)]] - .data[[paste0("mavg_", lo)]]
    )
}
DELTAS <- purrr::imap(M, function(x, d) {
  g <- x$m %>%
    filter(config %in% grid_names) %>%
    select(config, season, h, wis_pct) %>%
    left_join(x$cov, by = c("config", "season", "h")) %>%
    split_name()
  bind_rows(contrast(g, "start", "warm", "cold", "leak"), contrast(g, "leak", "leak 0.1", "no leak", "start")) %>%
    mutate(disease = d)
})
for (d in names(DELTAS)) {
  cat("\n\n", cal_h(2), " ", d, "\n\n", sep = "")
  DELTAS[[d]] %>%
    transmute(anchor, factor, col = paste(season, h), cell = sprintf("%+.1f / %+.3f", d_wis, d_mavg)) %>%
    tidyr::pivot_wider(names_from = col, values_from = cell) %>%
    arrange(anchor, factor) %>%
    knitr::kable(caption = paste0(d, ": change in WIS reduction % / month-avg L1 coverage bias from adding the factor, per anchor and season")) %>%
    print()
}
```

```{r factor-plot, fig.height=8}
ggplot(bind_rows(DELTAS) %>% filter(season != "both"), aes(h, d_wis, colour = anchor, group = anchor)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line() +
  geom_point() +
  facet_grid(disease + season ~ factor) +
  scale_colour_manual(values = ANCHOR_COLS) +
  labs(x = NULL, y = "change in WIS reduction %", colour = NULL) +
  theme(legend.position = "bottom")
```

# Anchor against anchor, state by state

Per state, pooled over h0–h3: WIS reduction % of the adaptive anchor against
the constant one at the same start and leak (above the line, adaptive is
better).

```{r state-scatter, results="asis", fig.height=5.5}
STATE <- purrr::map(RUNS, function(r) {
  cal_state_wis(r$cals[grid_names]) %>%
    group_by(config, season, location) %>%
    summarize(pct = 100 * (sum(wis_base) - sum(wis_cal)) / sum(wis_base), .groups = "drop") %>%
    split_name() %>%
    select(-config) %>%
    tidyr::pivot_wider(names_from = anchor, values_from = pct) %>%
    mutate(cell = paste(start, leak, sep = ", "))
})
for (d in names(STATE)) {
  cat("\n\n", cal_h(2), " ", d, "\n\n", sep = "")
  print(
    ggplot(STATE[[d]], aes(`constant 0.018`, `adaptive 0.03`)) +
      geom_abline(slope = 1, intercept = 0, colour = "grey50") +
      geom_hline(yintercept = 0, colour = "grey85") +
      geom_vline(xintercept = 0, colour = "grey85") +
      geom_point(alpha = 0.8) +
      facet_grid(season ~ cell) +
      labs(title = d, x = "constant 0.018: WIS reduction %, h0–h3", y = "adaptive 0.03: WIS reduction %, h0–h3")
  )
  STATE[[d]] %>%
    group_by(cell, season) %>%
    summarize(adaptive_better = sum(`adaptive 0.03` > `constant 0.018`), states = n(), .groups = "drop") %>%
    transmute(cell, season, v = sprintf("%d / %d", adaptive_better, states)) %>%
    tidyr::pivot_wider(names_from = season, values_from = v) %>%
    knitr::kable(caption = paste0(d, ": states (of those scored) where the adaptive anchor beats the constant one, h0–h3 pooled")) %>%
    print()
}
```

# Month views

The month plots show the no-leak grid and flu prod; the tables keep every
config.

```{r month-views, results="asis"}
mv_plot_cfgs <- c(grid_names[c(1, 3, 5, 7)], "REF-op warm")
mv_cols <- setNames(c("#E69F00", "#D55E00", "#0072B2", "#56B4E9", "#000000"), mv_plot_cfgs)
for (d in names(M)) {
  mv <- M[[d]]$mv
  cal_month_section(d, level = 2L)
}
```
