---
title: "Calibration experiments"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```

These notebooks test a post-hoc calibration step for our FluSight and covid
hub submissions. The step adjusts each submitted quantile forecast by an
additive offset that an online quantile tracker (MultiQT; Ding, Gibbs &
Tibshirani, 2025) learns from past coverage misses. Each notebook changes one
setting of that tracker and compares the result with the uncalibrated
forecast.

The summaries below give sizes, not verdicts. There are no uncertainty
estimates, so a WIS gap under 1 point or a month-avg coverage gap under 0.01
is read as no clear difference.

```{r rows, include=FALSE}
out_dir <- here::here("reports/calibration_experiments")
link <- function(id, file) {
  html <- paste0(file, ".html")
  if (!file.exists(file.path(out_dir, html))) return(paste(id, "(not rendered)"))
  stamp <- format(file.mtime(file.path(out_dir, html)), "%m-%d", tz = "America/Los_Angeles")
  sprintf("[%s](%s) (%s)", id, html, stamp)
}
```

# Notebooks

One notebook per experiment, grouped by topic; the date after each link is
when it was last rendered. The ids match `notes/calibration-ledger.md`.

## Finalists

| | base | question | what the numbers show |
|---|---|---|---|
| `r link("E18", "e18_ref_op_bridge")` | replay | Stepping from REF-op to sqrt constant 0.018 one setting at a time | REF-op vs sqrt 0.018: no strong result (0.3–0.6 points pooled, order flips by season). The warm start helps in 2024-25 only. Constant 0.018 is 1.7–2.3 points above cold adaptive at h2–h3. No step moves month-avg coverage by more than 0.008. sqrt 0.018 warm (branch) is the best pooled (+3.4/+1.6/+1.6/+2.5) but 0.011–0.018 worse than the base on month-avg coverage at h2–h3. REF-op beats sqrt 0.018 cold in 41–45 of 52 states in 2024-25 and 14–18 in 2025-26. |
| `r link("E07b", "e07b_candidates")` | both | Do the operating-point candidates rank the same way on both bases? | The candidate order flips between seasons on both. |
| `r link("E15", "e15_wis_sources")` | replay, cold | Where does the WIS change come from? | Offsets are 2–5% of the 90% interval width. REF-op's gain (+2.0 at h0, about 0 by h3) comes from raising the lower quantiles; a median shift alone matches it within 0.6 points. |

## Flu and covid

| | base | question | what the numbers show |
|---|---|---|---|
| `r link("E10", "e10_covid")` | both diseases, both bases | How do the finalists do on flu and covid? (Part 2: covid ensemble at h−1–h3 with REF-paper.) | Part 2: h−1 improves about 10–11% in both seasons. REF-paper loses 10–19% at h1–h3 in 2025-26. REF-op at h1–h3 is −4 to +3, sign depending on the season. Part 1: covid, nothing beats REF-op cold at h0–h3 and sqrt 0.018 loses 3–4 points at h1–h3; flu, sqrt 0.018 warm leads on both bases by 1–2 points at h1–h3, all in 2024-25 on the ensemble. sqrt 0.018 has the largest covid h−1 gain (+14.9). |
| `r link("E07", "e07_across_eras")` | both | Does calibration act the same on the ensemble and the clean replay? | Same direction on both; REF-op gains 0.7–1.1 points more on the ensemble. |

## Learning rate

| | base | question | what the numbers show |
|---|---|---|---|
| `r link("E01", "e01_eta_settings")` | replay, cold | How do the adaptive learning-rate settings (multiplier, window, carry or reset) change WIS and coverage? | Multipliers 0.1 and 0.3 lose 7–77 points at h1–h3. 0.01 is 0.5–2.4 points above 0.03 at h1–h3; month-avg coverage shows no clear difference. Count scale is 2–5 points below sqrt. Window and carry vs reset: no strong result. |
| `r link("E09", "e09_lr_delay")` | ensemble | With a constant learning rate on the rate scale, how do WIS and coverage change with the rate and with extra learning delay? | WIS peaks near 0.056–0.1 per 100k and falls fast past 0.3. Month-avg coverage is no lower than the base at h1–h3 for any constant rate. Each week of extra delay costs about 1 point at h−1. Stateful learning with no extra delay makes h−1 worse than the base. |
| `r link("E14", "e14_lr_by_season")` | replay, cold | How do WIS and coverage change with a constant learning rate, season by season? | Every rate from 0.0032 to 0.056 is at or above the base in both seasons, within 1.2 points of each other: no strong result for a best rate. From 0.1 WIS falls. Month-avg coverage for rates ≤ 0.032 stays within about 0.01 of the base. |
| `r link("E16", "e16_sqrt_constant_lr")` | replay, cold | Constant learning rate on the square-root scale vs the rate scale | sqrt 0.018 is about 1 point above rate 0.032 at h0 and within 0.2 at h1–h3 pooled. Against REF-op cold the order flips by season. Every tracker loses in March (up to −28% within the month at h2). |

## Tracker structure

| | base | question | what the numbers show |
|---|---|---|---|
| `r link("E03", "e03_scale")` | ensemble | Which scale should the offsets live on: counts, square root, or rate per 100k? | sqrt is 1.4–3.9 points above count pooled, mostly from 2025-26. Coverage: no strong result. Rate is identical to count in every cell. |
| `r link("E04", "e04_offset_structure")` | ensemble | Leak and slow term | The leak is +3.3 to +5.6 at h1–h3 in 2025-26 and −0.4 to −1.3 in 2024-25. The slow term without a warm start: no strong result. |
| `r link("E05", "e05_warm_start")` | ensemble | Warm start from the burn-in season | +1.7 to +3.2 at h2/h3 in 2024-25 for every tracker, within ±0.4 in 2025-26. |
| `r link("E11", "e11_constant_lr")` | ensemble | Does a constant learning rate do better than the adaptive rate? | Constant 0.1 vs adaptive sqrt: the order flips by horizon (1–2 points either way). REF-op has the highest pooled WIS at every horizon, by up to 2 points. On month-avg coverage the constant rate is 0.003–0.022 worse. |

## Late season

| | base | question | what the numbers show |
|---|---|---|---|
| `r link("E02", "e02_eta_variants")` | replay, cold | Spring switch-off, per-level learning rate, seasonal window | Switching off from 1 March or 15 February is 2–4 points above the reference at h1–h3 (dates in-sample; in 2025-26 about half the gain is December carry-over). Per-level about +1. Month-avg coverage: no clear difference. |
| `r link("E17", "e17_late_decay")` | replay, cold | Does shrinking or switching off the offset from 1 March change WIS and coverage? | No strong result: pooled WIS gaps ≤ 0.7 points and month-avg coverage gaps ≤ 0.010. The switch-off's sqrt gain comes mostly from 2025-26; the date was chosen in-sample. |

## Learning truth

| | base | question | what the numbers show |
|---|---|---|---|
| `r link("E00", "e00_vintage_backtest")` | ensemble | Should the tracker learn from finalized data or from data as published? | Exact and final differ by 0.6 points or less, except REF-op h−1 (2.2, 2024-25). Learning once from the first report loses most of the h−1 gain. All other notebooks use the exact (prod) design. |

E12 and E13 (the clean replay without and with a 2023-24 burn-in) have no
notebook; they are `scripts/calibration/calibration_ws_replay.R`.

# How to read the results

- **Horizons** h−1 to h3 are weeks relative to the forecast's reference date.
  h−1 is the latest week NHSN has reported, still open to revision; h0 and
  later have not been reported yet.
- **WIS reduction %** is `100 × (WIS_base − WIS_cal) / WIS_base`, the
  calibrated forecast's weighted interval score against the uncalibrated
  forecast's, pooled over locations. Positive = calibrated WIS lower than
  base.
- **L1 coverage bias** (lower is better) is the mean over the 23 quantile
  levels of |empirical coverage − level|. **Month-avg** computes it within
  each calendar month and averages; season-pooled lets under-prediction on
  the ramp cancel over-prediction after the peak, and usually looks much
  better. Month-level noise is about 0.01.
- **Two bases.** "Ensemble" notebooks calibrate our submitted forecasts
  (h−1 to h3, headline tables over all hub locations). "Replay" notebooks
  calibrate a clean weekly replay of `windowed_seasonal` with no hand edits
  (h0–h3, states only).
- **Scoring.** Every score is against finalized NHSN data over the live
  seasons (2024-25 and 2025-26). A burn-in season (2023-24 for flu) only warms
  up the learning rate and is not scored.
- **Cold start.** Without a warm start, offsets start at zero, and every
  config equals the base through Nov–Dec 2024, so 2024-25 gains are diluted.
- **In-sample choices.** The constant rates and switch-off dates in later
  notebooks were picked on the same two seasons they are scored on.
- **Excluded submissions.** A few submissions were broken by pipeline bugs
  or bad data and are left out, both from what the tracker learns and from
  the scores. Flu: h−1 of the 2025-11-22 round (every quantile was 0) and
  the US forecasts of the 2024-12-14, 2024-12-21 and 2025-01-04 rounds
  (about half the sum of the states). Covid: the whole 2024-11-23 round
  (made from a bad NHSN release that was corrected the next day) and h−1 of
  the 2026-04-25 round (near zero in many states, most likely from a stale
  data fetch).
- **Extra delay.** The tracker can learn from a week's outcome as soon as
  NHSN first reports it (zero extra delay) or wait for revisions. The
  reference configs wait one extra week.
- **Re-run vs stateful.** Prod re-runs the tracker from the first round each
  week on that week's data, so early reports are replaced by their revisions.
  A stateful tracker instead learns each outcome once, from the first value
  it sees. All notebooks use the re-run design; E00 and E09 also show the
  stateful one.

## Reference configs

- **REF-paper**: the paper's tracker on counts. One offset per quantile level,
  updated with an adaptive learning rate (0.03 × the 90th percentile of the
  last 20 rounds' absolute errors), carried across seasons, with one week of
  extra delay.
- **REF-op**: the configuration running in prod. REF-paper on the
  square-root scale, with the offset shrinking 10% per round, a second
  slow-moving offset, and a starting offset taken from the burn-in season's
  errors.
