---
title: "Calibration experiments"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```

These notebooks test a post-hoc calibration step for our FluSight and covid
hub submissions. The step adjusts each submitted quantile forecast by an
additive offset that an online quantile tracker (MultiQT; Ding, Gibbs &
Tibshirani, 2025) learns from past coverage misses. Each notebook changes one
setting of that tracker and compares the result with the uncalibrated
forecast.

There are no uncertainty estimates, so a WIS gap under 1 point or a
month-avg coverage gap under 0.01 is read as no clear difference.
Conclusions are kept in `notes/calibration-ledger.md`.

```{r rows, include=FALSE}
out_dir <- here::here("reports/calibration_experiments")
link <- function(id, file) {
  html <- paste0(file, ".html")
  if (!file.exists(file.path(out_dir, html))) return(paste(id, "(not rendered)"))
  stamp <- format(file.mtime(file.path(out_dir, html)), "%m-%d", tz = "America/Los_Angeles")
  sprintf("[%s](%s) (%s)", id, html, stamp)
}
```

# Notebooks

One notebook per experiment, grouped by topic; the date after each link is
when it was last rendered. The ids match `notes/calibration-ledger.md`.

## Finalists

| | base | question |
|---|---|---|
| `r link("E10", "e10_covid")` | both diseases, both bases | How do the finalists do on flu and covid, and do they rank the same way on the ensemble and the replay? (Part 2: covid ensemble at h−1–h3.) |
| `r link("E18", "e18_ref_op_bridge")` | replay | Which setting accounts for the gap between REF-op and sqrt constant 0.018, stepping from one to the other a setting at a time? |
| `r link("E15", "e15_wis_sources")` | replay, both diseases | Where does each finalist's WIS change come from? |

## Prod configs

`r link("Prod views", "prod_views")`: per-state forecast fans and a gallery
of the largest gains and losses, for flu REF-op warm and covid REF-op cold
on the submitted ensemble.

## Bases

| | base | question |
|---|---|---|
| `r link("E07", "e07_across_eras")` | both | Does calibration act the same on the ensemble and the clean replay? |

## Learning rate

| | base | question |
|---|---|---|
| `r link("E01", "e01_eta_settings")` | replay, cold | How do the adaptive learning-rate settings (multiplier, window, carry or reset) change WIS and coverage? |
| `r link("E09", "e09_lr_delay")` | ensemble | With a constant learning rate on the rate scale, how do WIS and coverage change with the rate and with extra learning delay? |
| `r link("E16", "e16_sqrt_constant_lr")` | replay, cold | How do WIS and coverage change with a constant learning rate, season by season, on the square-root and rate scales? |
| `r link("E19", "e19_tracker_internals")` | replay | Month by month, how large a step does each tracker take, how large an offset does it play, and does the month's WIS change follow from those and the base's bias? How much of 2025-26 is the offset carried in from 2024-25? |

The two learning-rate anchors used by the structure notebooks are E01's
adaptive multiplier 0.03 and E16's constant 0.018 on the sqrt per-100k
scale.

## Tracker structure

E20 asks the warm-start and leak questions on both anchors and both
diseases, on the replay. E03 is the only scale comparison: every E20 cell
runs on the square-root scale. (E04, E05 and E11 asked the leak, warm-start
and constant-rate questions on the ensemble with the adaptive rate; E20
reproduces their directions on the replay, so they were retired on
2026-10-07. Their numbers stay in the ledger.)

| | base | question |
|---|---|---|
| `r link("E20", "e20_structure_grid")` | replay, both diseases | Do the warm start and the leak change WIS and coverage in the same way on the adaptive and the constant learning rate, for flu and for covid? State by state, which anchor wins at the same start and leak? |
| `r link("E03", "e03_scale")` | ensemble | Which scale should the offsets live on: counts, square root, or rate per 100k? |

## Late season

| | base | question |
|---|---|---|
| `r link("E02", "e02_eta_variants")` | replay, cold | Do a spring switch-off, a per-level learning rate or a seasonal window change WIS and coverage? |
| `r link("E17", "e17_late_decay")` | replay, cold | Does shrinking or switching off the offset from 1 March change WIS and coverage? |

## Learning truth

| | base | question |
|---|---|---|
| `r link("E00", "e00_vintage_backtest")` | ensemble | Should the tracker learn from finalized data or from data as published? |

## Data

| | base | question |
|---|---|---|
| `r link("E21", "e21_data_by_season")` | replay data | How do 2024-25 and 2025-26 differ as data: season shape, revisions, the gap between the value the tracker learns from and the final one, and the HHS-stitched 2023-24 burn-in against NHSN? |

E12 and E13 (the clean replay without and with a 2023-24 burn-in) have no
notebook; they are `scripts/calibration/calibration_ws_replay.R`. E14 is
part of E16, and E07b's ranking is part of E10.

# How to read the results

- **Horizons** h−1 to h3 are weeks relative to the forecast's reference date.
  h−1 is the latest week NHSN has reported, still open to revision; h0 and
  later have not been reported yet.
- **WIS reduction %** is `100 × (WIS_base − WIS_cal) / WIS_base`, the
  calibrated forecast's weighted interval score against the uncalibrated
  forecast's, pooled over locations. Positive = calibrated WIS lower than
  base.
- **L1 coverage bias** (lower is better) is the mean over the 23 quantile
  levels of |empirical coverage − level|. **Month-avg** computes it within
  each calendar month and averages; season-pooled lets under-prediction on
  the ramp cancel over-prediction after the peak, and usually looks much
  better. Month-level noise is about 0.01.
- **Two bases.** "Ensemble" notebooks calibrate our submitted forecasts
  (h−1 to h3). "Replay" notebooks calibrate a clean weekly replay of
  `windowed_seasonal` with no hand edits (h0–h3).
- **Scoring.** Every score is states only (US excluded), against finalized
  NHSN data over the live seasons (2024-25 and 2025-26). A burn-in season (2023-24 for flu) only warms
  up the learning rate and is not scored.
- **Cold start.** Without a warm start, offsets start at zero, and every
  config equals the base through Nov–Dec 2024, so 2024-25 gains are diluted.
- **In-sample choices.** The constant rates and switch-off dates in later
  notebooks were picked on the same two seasons they are scored on.
- **Excluded submissions.** A few submissions were broken by pipeline bugs
  or bad data and are left out, both from what the tracker learns and from
  the scores. Flu: h−1 of the 2025-11-22 round (every quantile was 0) and
  the US forecasts of the 2024-12-14, 2024-12-21 and 2025-01-04 rounds
  (about half the sum of the states). Covid: the whole 2024-11-23 round
  (made from a bad NHSN release that was corrected the next day) and h−1 of
  the 2026-04-25 round (near zero in many states, most likely from a stale
  data fetch).
- **Extra delay.** The tracker can learn from a week's outcome as soon as
  NHSN first reports it (zero extra delay) or wait for revisions. The
  reference configs wait one extra week.
- **Re-run vs stateful.** Prod re-runs the tracker from the first round each
  week on that week's data, so early reports are replaced by their revisions.
  A stateful tracker instead learns each outcome once, from the first value
  it sees. All notebooks use the re-run design; E00 and E09 also show the
  stateful one.

## Reference configs

- **REF-paper**: the paper's tracker on counts. One offset per quantile level,
  updated with an adaptive learning rate (0.03 × the 90th percentile of the
  last 20 rounds' absolute errors), carried across seasons, with one week of
  extra delay.
- **REF-op**: the configuration running in prod. REF-paper on the
  square-root scale, with the offset shrinking 10% per round, a second
  slow-moving offset, and a starting offset taken from the burn-in season's
  errors.
