---
title: "Calibration experiments"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```

These notebooks test a post-hoc calibration step for our FluSight and covid
hub submissions. The step adjusts each submitted quantile forecast by an
additive offset that an online quantile tracker (MultiQT; Ding, Gibbs &
Tibshirani, 2025) learns from past coverage misses. Each notebook changes one
setting of that tracker and compares the result with the uncalibrated
submission.

# Notebooks

```{r notebooks, results="asis", echo=FALSE}
src <- list.files(here::here("reports/writeups/calibration_experiments"), pattern = "^e[0-9]+_.*[.]Rmd$", full.names = TRUE)
out_dir <- here::here("reports/calibration_experiments")
for (f in src) {
  html <- sub("[.]Rmd$", ".html", basename(f))
  title <- rmarkdown::yaml_front_matter(f)$title
  rendered <- file.exists(file.path(out_dir, html))
  stamp <- if (rendered) format(file.mtime(file.path(out_dir, html)), "%Y-%m-%d %H:%M %Z", tz = "America/Los_Angeles") else "not rendered"
  cat(sprintf("- %s (%s)\n", if (rendered) sprintf("[%s](%s)", title, html) else title, stamp))
}
```

| | question | short answer |
|---|---|---|
| E00 | Should the tracker learn from finalized data or from data as published at the time? | As published, re-run each week as prod does. Finalized data overstated the earlier gains only a little (about 2 points at h−1). |
| E03 | Which scale should the offsets live on: counts, square root, or rate per 100k? | Square root, at every horizon, mostly by avoiding spring losses. Rate equals counts with the adaptive learning rate. |
| E04 | Should the offset shrink between updates, or have a second slow component? | Shrinking buys WIS at h1–h3 but nearly triples coverage error. A slow component alone does nothing. |
| E05 | Should the offset start from the burn-in season's errors instead of zero? | Yes; it never hurts. With the plain tracker it gives the best-calibrated config so far. |
| E09 | With a constant learning rate on the rate scale, how do coverage and WIS trade off, and how much revision delay is best? | An L-shaped trade-off whose knee matches or beats the paper's adaptive rate at every horizon. Under prod's weekly re-run, less delay is better. |
| E10 | Do the flu results carry over to covid? | Only at h−1 (+10% WIS). The covid base is already calibrated at h1–h3, where calibrating is neutral or harmful. |
| E11 | How does E09's best constant learning rate compare with the adaptive trackers? | Better coverage than the cold adaptive trackers at less WIS gain near the present, and the most stable across seasons. The warm-started adaptive tracker still wins at h0–h1. |
| E14 | On the clean `windowed_seasonal` replay, is there a constant learning rate that is good in both seasons? | Yes, 0.018–0.032 per 100k: better WIS at every horizon in both seasons and lower coverage bias. 2024-25 needs a larger rate for good coverage, which costs WIS. |
| E15 | Why does WIS improve when the calibrated intervals barely move? | The offsets are small (2–5% of the 90% interval width) and point up. The gain comes from shifting the forecast up, mostly at the central levels below the median; width changes cost WIS. |

# How to read the results

- **Horizons** h−1 to h3 are weeks relative to the forecast's reference date.
  h−1 is the latest week NHSN has reported, still open to revision; h0 and
  later have not been reported yet.
- **WIS change %** compares the calibrated forecast's weighted interval score
  with the uncalibrated submission's, pooled over locations. The sign
  convention is stated in each notebook.
- **L1 coverage bias** is the mean over the 23 quantile levels of
  |empirical coverage − level|. 0 is perfect; the uncalibrated flu
  submission sits near 0.1.
- **Scoring.** Every score is against finalized NHSN data, over the live
  seasons (2024-25 and 2025-26 for flu), all hub locations. A burn-in season
  (2023-24 for flu) is used only to warm up the tracker and is not scored.
- **Excluded submissions.** A few submissions were broken by pipeline bugs
  or bad data and are left out, both from what the tracker learns and from
  the scores. Flu: h−1 of the 2025-11-22 round (every quantile was 0) and
  the US forecasts of the 2024-12-14, 2024-12-21 and 2025-01-04 rounds
  (about half the sum of the states). Covid: the whole 2024-11-23 round
  (made from a bad NHSN release that was corrected the next day) and h−1 of
  the 2026-04-25 round (near zero in many states, most likely from a stale
  data fetch).
- **Extra delay.** The tracker can learn from a week's outcome as soon as
  NHSN first reports it (zero extra delay) or wait for revisions. The
  reference configs wait one extra week.
- **Re-run vs stateful.** Prod re-runs the tracker from the first round each
  week on that week's data, so early reports are replaced by their revisions.
  A stateful tracker instead learns each outcome once, from the first value
  it sees. All notebooks use the re-run design; E09 also shows the stateful
  one.

## Reference configs

Each experiment starts from one of two configurations and changes one
setting.

- **REF-paper**: the paper's tracker on counts. One offset per quantile level,
  updated with an adaptive learning rate (0.03 × the 90th percentile of the
  last 20 rounds' absolute errors), carried across seasons, with one week of
  extra delay.
- **REF-op**: the configuration running in prod. REF-paper on the
  square-root scale, with the offset shrinking 10% per round, a second
  slow-moving offset, and a starting offset taken from the burn-in season's
  errors.
