---
title: "Learning truth (E00): learning from vintages instead of finalized truth"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
params:
  hub_dir: "../FluSight-forecast-hub"
  workers: 12
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
# cache = TRUE chunks are keyed on chunk code, not on functions in R/. After
# changing calibration code, clear cache/calibration/experiments/knitr_e00/.
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 10, fig.height = 5,
  message = FALSE, warning = FALSE,
  cache.path = here::here("cache/calibration/experiments/knitr_e00/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

Question: how do results change when the tracker learns from NHSN data as
published, rather than from finalized values?

Setup:

- Disease: flu. Base forecaster: the submitted hub ensemble (spoiled
  submissions dropped).
- References: REF-paper (count scale, one offset per level) and REF-op (sqrt
  scale, leak 0.1, slow term with a burn-in-quantile warm start); see the index.
- Burn-in: 2023-24, used only to warm up the learning rate. Offsets start at
  zero at the first 2024-25 round (REF-op's slow term starts from its warm
  start).
- Seasons scored: 2024-25 and 2025-26, burn-in rounds excluded. Horizons
  h−1 to h3. Headline tables: all hub locations including US; month view:
  states only.
- Scores are always against finalized truth.

Configs are `ref/settle_days/learn`:

| axis | values |
|---|---|
| `ref` | `paper`, `op` |
| `settle_days` | `14` (one week of extra delay, the reference) or `7` (none) |
| `learn` | `final`: latest vintage. `exact`: re-run from the first round at every round on the data published by then, as prod does. `vintage`: one run where each outcome keeps the value it had at the round that revealed it (a stateful tracker that never revisits a step) |

A round with reference date `d` sees data published by `d - 3`. Vintages
start 2024-11-19; earlier as-of dates use that first vintage. "WIS reduction
%" is 100 × (base WIS − calibrated WIS) / base WIS, so positive = calibrated
WIS lower than base. "L1 coverage bias" is lower-is-better.

```{r libs, cache=FALSE}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))
```

```{r load, cache=TRUE}
archive <- get_nhsn_data_archive("flu")
truth <- nhsn_read_truth("flu", archive)
vintages <- nhsn_read_vintages("flu", archive)
hub_fc <- hub_read_forecasts(hub_dir = here::here(params$hub_dir))
```

```{r runs, cache=TRUE, dependson="load"}
grid <- tidyr::expand_grid(ref = c("paper", "op"), settle_days = c(7L, 14L), learn = c("final", "vintage", "exact"))
cals <- purrr::pmap(grid, function(ref, settle_days, learn) {
  cal_run(hub_fc, truth, ref, settle_days = settle_days, learn = learn, vintages = vintages, workers = params$workers)
})
names(cals) <- with(grid, paste(ref, settle_days, learn, sep = "/"))
```

```{r headline}
LIVE <- c("2024-2025", "2025-2026")
head_tbl <- cal_headline(cals) %>%
  tidyr::separate(variant, c("ref", "settle_days", "learn"), sep = "/", remove = FALSE) %>%
  mutate(learn = factor(learn, levels = c("final", "vintage", "exact")))
first_version <- min(vintages$version)
n_clamped <- sum(unique(hub_fc$reference_date) - 3L < first_version &
  hub_label_seasons(unique(hub_fc$reference_date))$season %in% LIVE)
```

`r n_clamped` live rounds are issued before the first vintage (2024-11-19); only
burn-in rounds use that vintage in place of their own.

# Headline

WIS reduction % by horizon (positive = calibrated WIS lower than base):

```{r wis-table}
knitr::kable(cal_wide(head_tbl, "pct_improvement", id = c("ref", "settle_days", "learn")) %>% arrange(ref, settle_days, learn))
```

L1 coverage bias (lower is better; mean over levels of |coverage − level|); the `base` row is
the uncalibrated forecast:

```{r cal-table}
knitr::kable(
  bind_rows(
    cal_wide(head_tbl, "cal_error_cal", 3, id = c("ref", "settle_days", "learn")),
    cal_wide(head_tbl %>% filter(learn == "final", ref == "paper", settle_days == "14") %>% mutate(ref = "base", settle_days = "", learn = ""), "cal_error_base", 3, id = c("ref", "settle_days", "learn"))
  ) %>% arrange(ref, settle_days, learn)
)
```

The same bias for states only, both seasons, season-pooled against
month-average (see "Month view" below). Configs are `ref/settle_days/learn`:

```{r cal-table-month}
mv <- cal_month_view(cals, seasons = LIVE, horizons = NULL)
mv_show_table <- FALSE
knitr::kable(cal_month_cov_table(mv, seasons = "both"), caption = CAL_MONTH_COV_CAPTION)
```

WIS reduction % by season (positive = calibrated WIS lower than base):

```{r season-table}
knitr::kable(
  cal_wide(
    cal_headline(cals, by = "season") %>% filter(season %in% LIVE) %>%
      tidyr::separate(variant, c("ref", "settle_days", "learn"), sep = "/"),
    "pct_improvement", id = c("ref", "settle_days", "learn", "season")
  ) %>% arrange(ref, settle_days, season, learn)
)
```

# Vintage vs exact: difference in the calibrated median

Median over locations of |vintage − exact| in the calibrated median,
relative to the base median, by month, one panel per `ref/settle_days`:

```{r drift, fig.height=6}
drift <- purrr::map(unique(paste(grid$ref, grid$settle_days, sep = "/")), function(cfg) {
  pick <- function(mode) {
    cals[[paste(cfg, mode, sep = "/")]]$forecasts %>%
      filter(level == 0.5, !is_burn_in, !is.na(value_base)) %>%
      select(location, horizon, reference_date, value_base, value_cal)
  }
  inner_join(pick("vintage"), pick("exact"), by = c("location", "horizon", "reference_date", "value_base"), suffix = c("_cheap", "_exact")) %>%
    mutate(config = cfg)
}) %>%
  bind_rows() %>%
  mutate(month = lubridate::floor_date(reference_date, "month")) %>%
  group_by(config, horizon, month) %>%
  summarize(rel = median(abs(value_cal_cheap - value_cal_exact) / pmax(value_base, 1)), .groups = "drop")
ggplot(drift, aes(month, 100 * rel, colour = factor(horizon), group = horizon)) +
  geom_line() +
  geom_point(size = 0.8) +
  facet_wrap(~config) +
  labs(x = NULL, y = "median |vintage − exact| / base median (%)", colour = "horizon")
```

```{r month-view-cfgs}
# Runs drawn in the month view.
mv_plot_cfgs <- c("op/14/final", "op/14/exact", "op/7/vintage", "op/7/exact", "paper/14/final", "paper/14/exact")
```

```{r child=here::here("reports/writeups/calibration_experiments/_month_view.Rmd")}
```

# Findings

WIS reduction % is positive = calibrated WIS lower than base; coverage bias is
lower-is-better. A WIS gap under 1 point or a month-avg coverage gap under
0.01 is read as no clear difference.

- **final vs exact.** REF-op at settle 14: h−1 +7.6 (final) vs +5.4 (exact),
  a 2.2-point gap that comes from 2024-25 (11.2 vs 7.7; 2025-26 2.4 vs 2.1).
  At h0–h3 the gap is 0.5 points or less. REF-paper: under 1 point at every
  horizon (0.3–0.6). Month-avg coverage bias: no clear difference (REF-op
  0.106 vs 0.107 at h−1; REF-paper 0.122 vs 0.115). Season-pooled bias is
  higher under exact (REF-paper h−1 0.044 vs 0.030), but pooling lets ramp
  and post-peak errors cancel.
- **vintage vs exact.** At settle 7 the vintage mode loses most of the h−1
  gain: REF-op 0.0 vs +6.1, REF-paper −1.1 vs +4.5. Its month-avg coverage
  bias at h−1 is worse than the base (REF-op 0.136, REF-paper 0.149, base
  0.134; exact 0.107 and 0.110). At settle 14 the gaps are smaller (REF-op h−1
  +4.9 vs +5.4). At REF-paper h1–h3 the vintage mode has higher WIS reduction
  than exact (settle 14: −0.9/−1.6/−4.2 vs −2.0/−2.7/−5.3).
- **settle 7 vs 14 under exact.** REF-paper: settle 7 is 1.3–2.8 points
  higher at every horizon, in both seasons. REF-op: no clear difference
  pooled (0.0–0.7); in 2025-26 settle 14 is 0.5–0.6 points higher at h1–h3.
  Month-avg coverage: no clear difference at either reference.
- **settle 7 vs 14 under vintage.** settle 14 is higher at h−1/h0 (REF-op
  +4.9/+4.5 vs 0.0/+3.0). At REF-paper h1–h3 the order is the other way
  (−0.1/−0.4/−2.3 at settle 7 vs −0.9/−1.6/−4.2 at settle 14).
- **Month-avg coverage at h2–h3.** Every REF-paper config is at or above the
  base (h3 0.160–0.170 vs 0.151); every REF-op config is below it (h3
  0.135–0.137).
- Offsets start from zero in November 2024 (the burn-in only warms the
  learning rate), so early 2024-25 rounds of the REF-paper runs equal the base.
