---
title: "Learning rate"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M %Z', tz = 'America/Los_Angeles')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
params:
  workers: 14
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.width = 11, fig.height = 5, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(panel.grid.minor = ggplot2::element_blank()))
suppressPackageStartupMessages(source(here::here("R/load_all.R")))
```

How large should the tracker's steps be? E01 varies the adaptive learning rate's settings on the clean replay; E09 sweeps a constant rate (and the extra revision delay) on the submitted ensemble; E14 and E16 sweep a constant rate on the replay, on the rate and square-root scales. Each section is a self-contained experiment with its own runs;
the ids match `notes/calibration-ledger.md`.

```{r e01, results="asis"}
cal_section("_e01_eta_settings.Rmd", list(workers = params$workers))
```

```{r e09, results="asis"}
cal_section("_e09_lr_delay.Rmd", list(workers = params$workers))
```

```{r e14, results="asis"}
cal_section("_e14_lr_by_season.Rmd", list(workers = params$workers))
```

```{r e16, results="asis"}
cal_section("_e16_sqrt_constant_lr.Rmd", list(workers = params$workers))
```
