---
title: "Yeo-Johnson Transformation Testing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
```

## Setup and Data Loading

First, we'll set up the environment and load the necessary data:

```{r setup-env}
# Simple case with keys = geo_value.
jhu <- cases_deaths_subset %>%
  filter(time_value > "2021-01-01", geo_value %in% c("ca", "ny")) %>%
  select(geo_value, time_value, cases)

# Load and prepare the data
data <- jhu
filtered_data <- data %>% select(geo_value, time_value, cases)
```

## Yeo-Johnson Transformation

Let's apply the Yeo-Johnson transformation to our data:

```{r yeo-johnson-transform}
r <- epi_recipe(filtered_data) %>%
  step_YeoJohnson2(cases) %>%
  prep(filtered_data)

# Display the recipe
r

# Inspect the lambda values for each state
r$steps[[1]]$lambdas
```

## Manual Whitening Comparison

Now, let's compare the Yeo-Johnson transformation with a manual whitening approach using quarter root scaling:

```{r manual-whitening}
# Apply the transformation
out1 <- r %>% bake(filtered_data)
out2 <- filtered_data %>%
  mutate(cases = (cases + 0.01)^(1 / 4))
filtered_data %>%
  mutate(cases = log(cases)) %>%
  ggplot(aes(time_value, cases)) +
  geom_line(color = "blue") +
  geom_line(data = out1 %>% mutate(cases = log(cases)),
            aes(time_value, cases), color = "green") +
  geom_line(data = out2 %>% mutate(cases = log(cases)),
            aes(time_value, cases), color = "red") +
  facet_wrap(~geo_value, scales = "free_y") +
  theme_minimal() +
  labs(title = "Yeo-Johnson transformation", x = "Time", y = "Log Cases")

# Test filtered_data not an epi_df still works.
filtered_data_not_epi_df <- filtered_data %>%
  as_tibble()
r %>% bake(filtered_data_not_epi_df)
```

## Workflow Testing

Finally, let's test the workflow with the Yeo-Johnson transformation:

```{r workflow-test}
debugonce(slather.layer_YeoJohnson2)
data <- filtered_data
# Create and fit the workflow
r <- epi_recipe(data) %>%
  step_YeoJohnson2(cases) %>%
  step_epi_lag(cases, lag = 1) %>%
  step_epi_ahead(cases, ahead = 1, role = "outcome") %>%
  step_epi_naomit()
f <- frosting() %>%
  layer_predict() %>%
  layer_threshold(.pred) %>%
  layer_naomit(.pred) %>%
  layer_YeoJohnson2(.pred)

# TODO: This needs to be fixed.
wf <- epi_workflow(r, linear_reg()) %>%
  fit(data) %>%
  add_frosting(f)

forecast(wf)
```

```{r}
library(epipredict)
debugonce(epipredict:::slather.layer_population_scaling)
jhu <- cases_deaths_subset %>%
  filter(time_value > "2021-11-01", geo_value %in% c("ca", "ny")) %>%
  select(geo_value, time_value, cases)

pop_data <- data.frame(states = c("ca", "ny"), value = c(20000, 30000))

r <- epi_recipe(jhu) %>%
  step_population_scaling(
    df = pop_data,
    df_pop_col = "value",
    by = c("geo_value" = "states"),
    cases, suffix = "_scaled"
  ) %>%
  step_epi_lag(cases_scaled, lag = c(0, 7, 14)) %>%
  step_epi_ahead(cases_scaled, ahead = 7, role = "outcome") %>%
  step_epi_naomit()

f <- frosting() %>%
  layer_predict() %>%
  layer_threshold(.pred) %>%
  layer_naomit(.pred) %>%
  layer_population_scaling(.pred,
    df = pop_data,
    by = c("geo_value" = "states"),
    df_pop_col = "value"
  )

wf <- epi_workflow(r, linear_reg()) %>%
  fit(jhu) %>%
  add_frosting(f)

forecast(wf)
```