# Appendix

## Scoring Metrics

:::: {.columns}
::: {.column width="50%"}

- Weighted interval score (WIS)
  - Discrete approximation to CRPS, which sums the colored area in the figure squared
- Absolute error (AE)
  - $|y_t - m|$ where $m$ is the median forecast
- 80% prediction interval coverage (80%PI)
  - Proportion of times the true value falls within the center 80%
:::
::: {.column width="50%"}

```{r wis_definition_intuition, fig.width = 5, fig.height = 5}
df <- data.frame(x = seq(-1, 2, length.out = 3000))

df$y1 <- pnorm(df$x, mean = 0.5, sd = 0.25)
df$y2 <- ifelse(df$x >= 0.75, 1, 0)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y1, color = "CDF of forecast")) +
  geom_line(aes(y = y2, color = "Eventually observed value")) +
  geom_ribbon(aes(ymin = pmin(y1, y2), ymax = pmax(y1, y2), fill = "WIS ~ ∫(F(x) - I(x ≥ .75))²"),
              alpha = 0.3) +
  scale_color_manual(name = "", values = c("CDF of forecast" = "red",
                                          "Eventually observed value" = "blue")) +
  scale_fill_manual(name = "", values = c("WIS ~ ∫(F(x) - I(x ≥ .75))²" = "red")) +
  theme(legend.position = c(0.2, 0.8),
        legend.background = element_rect(fill = "white"),
        legend.text = element_text(size = 6)) +
  labs(x = "x", y = "CDF", title = "WIS example") +
  theme(plot.title = element_text(hjust = 0.5))
```

:::
::::


## Explicit phase definitions: Flu

```{r flu_season_definitions, fig.width = 25, fig.height = 15, fig.align = "center"}
flu_current %>%
  filter(time_value > "2024-11-01") %>%
  autoplot(value, .facet_by = "geo_value") +
  geom_vline(data = flu_within_max, aes(xintercept = first_above)) +
  geom_vline(data = flu_within_max, aes(xintercept = last_above)) +
  facet_wrap(~geo_value, scale = "free") +
  theme(legend.position = "none")
```


## Explicit phase definitions: Covid
```{r covid_phase_definitions, fig.width = 18, fig.height = 9}
no_phase <- covid_within_max %>% arrange(desc(rel_duration)) %>% filter(rel_duration > covid_flat_threshold)
covid_current %>%
  filter(time_value > "2024-11-01", geo_value %nin% no_phase$geo_value) %>%
  autoplot(value, .facet_by = "geo_value") +
  geom_vline(data = covid_within_max, aes(xintercept = first_above)) +
  geom_vline(data = covid_within_max, aes(xintercept = last_above)) +
  facet_wrap(~geo_value, scale = "free") +
  ylim(0, NA) +
  theme(legend.position = "none")
```

## Explicit phase definitions: Covid `flat`
```{r covid_phase_definitions_flat, fig.width = 15, fig.height = 8}
covid_current %>%
  filter(time_value > "2022-06-01") %>%
  filter(geo_value %in% no_phase$geo_value) %>%
  autoplot(value, .facet_by = "geo_value") +
  geom_vline(aes(xintercept = as.Date("2024-11-01"))) +
  ylim(0, NA) +
  theme(legend.position = "none")
```
