# Covid Scores

## Covid Scores: WIS {.smaller}
::: {.notes}
`CMU-TimeSeries` did significantly worse; a large part of this is poor performance on literally the first week
:::
```{r computing_phase_covid}
covid_current <- nhsn_covid_latest
covid_max <- covid_current %>% group_by(geo_value) %>% summarize(max_value = max(value))
covid_within_max <- compute_peak_season(covid_current)

covid_flat_threshold <- 0.6
sanity_check_classifying <- covid_current %>%
  left_join(covid_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(time_value, first_above, last_above, rel_duration, covid_flat_threshold)) %>%
  group_by(geo_value) %>%
  distinct(phase)

repo_forecasters <- c(setdiff(unique(covid_scores$forecaster), our_forecasters), "CMU-TimeSeries")

se_mean_states <- covid_scores %>%
  left_join(covid_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis_rate = wis * 1e5/pop) %>%
  mutate(pop = factor(pop)) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold)) %>%
  group_by(forecaster, geo_value) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster) %>%
  summarize(
    se_mean_wis = sd(mean_wis) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis) / sqrt(n()),
    se_mean_ae = sd(mean_ae) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))


covid_score_summary <- covid_scores %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  group_by(forecaster) %>%
  mutate(
    min_wis = min(wis[wis > 1e-5]),
    min_ae = min(ae_median[ae_median > 1e-5])
  ) %>%
  summarize(
    mean_wis = round(Mean(wis), 2),
    wis_sd = round(sd(wis), 2),
    pop_norm_wis = round(Mean(wis *1e5/pop), 2),
    pop_norm_wis_sd = round(sd(wis * 1e5/pop), 2),
    geomean_wis = round(GeoMean(wis, min_wis), 2),
    mean_ae = round(Mean(ae_median), 2),
    ae_sd = round(sd(ae_median), 2),
    pop_norm_ae = round(Mean(ae_median*1e5/pop), 2),
    pop_norm_ae_sd = round(sd(ae_median * 1e5/pop), 2),
    geomean_ae = round(GeoMean(ae_median, min_ae), 2),
    mean_cov_50 = round(Mean(interval_coverage_50), 2),
    mean_cov_90 = round(Mean(interval_coverage_90), 2),
    n = n()
  ) %>%
  left_join(se_mean_states, by = "forecaster") %>%
  arrange(mean_wis)
```

```{r covid_datatable, fig.height = 60, fig.width = 12, echo=FALSE}
wis_score_order <- covid_score_summary %>% pull(forecaster)
pop_score_order <- covid_score_summary %>% arrange(pop_norm_wis) %>% pull(forecaster)
covid_wis_summary <-
  covid_score_summary %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, mean_wis, se_mean_wis, pop_norm_wis, geomean_wis, se_geomean_wis)

  datatable(
    covid_wis_summary,
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
      pageLength = 25
    )
  ) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$mean_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "pop_norm_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$pop_norm_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "geomean_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$geomean_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```

## Covid Scores: Absolute Error {.smaller}
```{r covid_scores_ae}
covid_ae_summary <- covid_score_summary %>%
  arrange(mean_ae) %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, mean_ae, se_mean_ae, pop_norm_ae, geomean_ae, se_geomean_ae)
datatable(
  covid_ae_summary,
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
    pageLength = 25
  )
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_ae",
    background = styleColorBar(c(0, max(covid_ae_summary$mean_ae)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "pop_norm_ae",
    background = styleColorBar(c(0, max(covid_ae_summary$pop_norm_ae)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "geomean_ae",
    background = styleColorBar(c(0, max(flu_ae_summary$geomean_ae)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```

## Covid Scores: Coverage {.smaller}
```{r covid_coverage_datatable}
covid_coverage_summary <- covid_score_summary %>%
  arrange(mean_ae) %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, mean_cov_50, mean_cov_90, n)
datatable(
  covid_coverage_summary,
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
    pageLength = 25
  )
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline"))
```

## Covid Scores: Aggregated By Forecast Date

```{r covid_agg_by_date, fig.height = 6, fig.width = 12, echo=FALSE}
agg_covid <- covid_scores %>%
  filter(forecast_date > "2024-10-01") %>%
  filter(forecast_date != as.Date("2025-01-25")) %>%
  group_by(forecaster, forecast_date) %>%
  summarize(
    mean_wis = round(Mean(wis), 2),
    wis_sd = round(sd(wis), 2),
    geomean_wis = round(GeoMean(wis), 2),
    mean_ae = round(Mean(ae_median), 2),
    geomean_ae = round(GeoMean(ae_median), 2),
    mean_interval_coverage_90 = round(Mean(interval_coverage_90), 2),
  )

p <- ggplot(agg_covid, aes(x = forecast_date, y = mean_wis, color = forecaster)) +
  geom_line() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Forecast Date", y = "Mean WIS")

style(ggplotly(p), visible = "legendonly", traces = c(1, 6, 8, 9, 10, 11, 12, 13, 14))
```


## Covid Scores: WIS Rate Histogram

::: {.notes}
These scores are rates rather than counts.

Note that again, the score distributions are very similar.
You can see where CMU-TimeSeries performed worse at the beginning of the season in the top left.
This was due to a naive use of the climatological baseline, which completely overestimated values in states that had no season.
:::

```{r covid_score_histogram, fig.height = 13, fig.width = 25, echo=FALSE}
data <- covid_scores %>%
  filter(forecaster %in% c("PSI-PROF", "CovidHub-ensemble", "climate_base", "CMU-TimeSeries", "UMass-ar6_pooled", "windowed_seasonal_nssp"))  %>%
  left_join(covid_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis = wis * 1e5/pop) %>%
  mutate(pop = factor(pop)) %>%
  mutate(forecaster = factor(forecaster, levels = wis_score_order)) %>%
  group_by(forecaster) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold))
data %>%
  ggplot(aes(x = wis, color = pop,  y = ifelse(after_stat(count) > 0, after_stat(count), NA))) +
  geom_histogram(bins = 70) +
  facet_grid(forecaster~phase) +
  labs(title = "Wis score histogram") +
  ylab("count") +
  xlab("wis, population normalized") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_color_viridis_d()
```

## Covid Scores: WIS Rate Histogram log10

```{r covid_score_histogram_log10, fig.height = 13, fig.width = 25, echo=FALSE}
data <- covid_scores %>%
  filter(forecaster %in% c("PSI-PROF", "CovidHub-ensemble", "climate_base", "CMU-TimeSeries", "UMass-ar6_pooled", "windowed_seasonal_nssp"))  %>%
  left_join(covid_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis = wis * 1e5/pop) %>%
  mutate(pop = factor(pop)) %>%
  mutate(forecaster = factor(forecaster, levels = wis_score_order)) %>%
  group_by(forecaster) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold))
data %>%
  ggplot(aes(x = wis, color = pop,  y = ifelse(after_stat(count) > 0, after_stat(count), NA))) +
  geom_histogram(bins = 70) +
  facet_grid(forecaster~phase) +
  labs(title = "Wis score histogram") +
  ylab("count") +
  xlab("wis, population normalized, log10") +
  scale_x_log10() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_color_viridis_d()
```


## Covid Scores: Phase

```{r covid_phase}
covid_include_phase <- covid_scores %>%
  left_join(covid_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold))

covid_phase_se <- covid_include_phase %>%
  group_by(forecaster, geo_value, phase) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster, phase) %>%
  summarize(
    se_mean_wis = sd(mean_wis, na.rm = TRUE) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis, na.rm = TRUE) / sqrt(n()),
    se_mean_ae = sd(mean_ae, na.rm = TRUE) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae, na.rm = TRUE) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))

phase_scores <-
  covid_scores %>%
  left_join(covid_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold)) %>%
  group_by(forecaster, phase) %>%
  summarize(
    wis_sd = round(sd(wis, na.rm = TRUE), 2),
    ae_sd = round(sd(ae_median, na.rm = TRUE), 2),
    across(c(wis, ae_median, interval_coverage_90), \(x) round(Mean(x), 2)),
    n = n(),
    our_forecaster = first(our_forecaster),
    .groups = "drop"
  ) %>%
  left_join(covid_phase_se, by = c("forecaster", "phase"))
sketch <- htmltools::withTags(table(
  class = 'display',
  style = "font-size: 12px;",
  thead(
    tr(
      th(colspan = 2, "forecaster"),
      th(colspan = 2, 'increasing'),
      th(colspan = 2, 'peak'),
      th(colspan = 2, 'decreasing'),
      th(colspan = 2, 'flat')
    ),
    tr(
      list(th("peak_wis_rank"), th("name"),
           lapply(rep(c('wis', 'se_mean_wis'), 4), th)
           )
    )
  )
))
phase_scores_wider <- phase_scores %>%
  select(forecaster, phase, wis, se_mean_wis, our_forecaster) %>%
  pivot_wider(names_from = phase, values_from = c(wis, se_mean_wis)) %>%
  select(forecaster, wis_increasing, se_mean_wis_increasing, wis_peak, se_mean_wis_peak, wis_decreasing, se_mean_wis_decreasing, wis_flat, se_mean_wis_flat) %>%
  arrange(wis_peak)
wis_max <- phase_scores_wider %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(wis_increasing, wis_peak, wis_decreasing) %>%
  max()
datatable(
  phase_scores_wider %>% filter(forecaster %nin% c("seasonal_nssp_latest")),
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "12pt", "'});"),
          "}"),
    pageLength = 25
  ),
  container = sketch
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "wis_increasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_peak",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_decreasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_flat",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```


## Covid: Sample Forecasts
::: {.notes}
80% CI
We're plotting the 80% CI along with the median.
The locations were chosen based on the scores to have a sample of large and small states with large and small (population normalized) WIS.
We've scaled so everything is in rates per 100k so that it's easier to actually compare; even so the peak value varies drastically.
Forecasters we've produced are blue, while forecasters from other teams are red.
They are ordered by `mean_wis` score, best to worst.
:::

```{r covid_plot_sample_forecast, fig.height = 7, fig.width = 12, echo=FALSE}
plot_geos <- c("ca", "dc", "pa", "hi", "tx")
filtered_covid_forecasts <- covid_forecasts %>%
  filter(quantile %in% c(0.1, 0.5, 0.9), geo_value %in% plot_geos) %>%
  filter(forecaster %in% c("CovidHub-ensemble", "CMU-TimeSeries", "UMass-ar6_pooled", "windowed_seasonal_nssp"))

covid_forecast_plt <-
  filtered_covid_forecasts %>%
  filter(forecast_date %in% forecast_weeks_to_plot) %>%
  mutate(forecaster = factor(forecaster, levels = wis_score_order)) %>%
  mutate(our_forecaster = factor(forecaster %in% our_forecasters, levels = c(TRUE, FALSE))) %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(value = value * 1e5/ pop) %>%
  pivot_wider(names_from = quantile, values_from = value) %>%
  ggplot(aes(x = target_end_date)) +
  geom_ribbon(aes(ymin = `0.1`, ymax = `0.9`, color = our_forecaster, fill = our_forecaster, group = forecast_date), alpha = 0.5) +
  geom_line(aes(y = `0.5`, color = our_forecaster, group = forecast_date)) +
  geom_line(
    data = covid_current %>%
      filter(time_value > "2024-11-01", geo_value %in% plot_geos) %>%
      left_join(state_census, by = join_by(geo_value == abbr)) %>%
      mutate(value = value * 1e5/ pop),
    aes(x = time_value, y = value)) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  facet_grid(forecaster ~ geo_value, scale = "free") +
  theme(legend.position = "none") +
  scale_x_date(breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 16)) +
  ylab("rates")

covid_forecast_plt
```

## Covid: Forecasting on latest
::: {.notes}
The point is to demonstrate the upper bound of what accurate nowcasting would get us.
named `seasonal_nssp_latest` compare with `windowed_seasonal_nssp`. Generally outperforms all other models by a good margin.
:::

```{r covid_phase_latest}
datatable(
  phase_scores_wider,
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "12pt", "'});"),
          "}"),
    pageLength = 25
  ),
  container = sketch
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "wis_increasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_peak",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_decreasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_flat",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```
