# The Data

## Recent Seasons Data: Flu

```{r flu_archive, fig.width = 20, fig.height = 10, fig.align = "center"}
nhsn_latest <- nhsn_archive_flu %>%
  epix_as_of_current() %>%
  filter(time_value >= "2020-10-01") %>%
  add_season_info() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi"), season <= "2024/25")
ggplot(nhsn_latest, aes(x = season_week, y = value, color = season, size = season)) +
  geom_line() +
  facet_wrap(~geo_value, scale = "free") +
  theme(legend.text = element_text(size = 12)) +
  scale_size_manual(values = c(rep(0.5, n_distinct(nhsn_latest$season) - 1), 1))
```

::: {.notes}
Here we see the 2024/25 season in pink.
It's the worst season in 15 years and very different from the previous 2 seasons.
Before that, the 2020/21 and the 2021/22 seasons are essentially negligible due to the pandemic (so we've dropped them).

[SciAm article with some speculation on why](https://www.scientificamerican.com/article/why-this-years-flu-season-is-the-worst-in-more-than-a-decade/).
- Late start -> less protection + 2nd wave
- different strain (H3N2 H1N1)
- decreased vaccination (not that low)
- lower levels of immunity from decreased exposure

This bears out somewhat in NSSP, but it is not as extreme an increase, so it may be a change in reporting.
:::

## Recent Seasons Data: Covid

```{r covid_archive, fig.width = 20, fig.height = 10, fig.align = "center"}
nhsn_latest_covid <- nhsn_archive_covid %>%
  epix_as_of_current() %>%
  add_season_info() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi"), season %nin% c("2019/20", "2020/21", "2021/22"), season <= "2024/25")
ggplot(nhsn_latest_covid, aes(x = season_week, y = value, color = season, size = season)) +
  geom_line() +
  facet_wrap(~geo_value, scale = "free") +
  theme(legend.text = element_text(size = 12)) +
  scale_size_manual(values = c(rep(0.5, n_distinct(nhsn_latest_covid$season) - 1), 1))
```

::: {.notes}
Here we've dropped the 2020 through 2022 seasons because they absolutely swamp the past 3 years.
Even so, this year is frequently without a season at all depending on the geo (for instance, see CA and FL and TX); part of the reason for this is the late summer wave, which is not visible in this data.
:::

## Data Revisions: Flu

```{r flu_revision}
nhsn_archive_flu$time_type <- "day"
revision_sum <- nhsn_archive_flu %>% epiprocess::revision_analysis(value, min_waiting_period = NULL)
av_re_spread <- revision_sum$revision_behavior %>%
  group_by(geo_value) %>%
  summarize(rel_spread = mean(rel_spread, na.rm = TRUE)) %>%
  arrange(desc(rel_spread)) %>%
  filter(geo_value %nin% c("vi", "as", "gu"))
nhsn_filtered <- nhsn_archive_flu %>%
  filter(geo_value %in% av_re_spread$geo_value[1:18]) %>%
  filter(time_value >= "2024-11-19")
nhsn_filtered$DT %<>%
  mutate(geo_value = factor(geo_value, levels = av_re_spread$geo_value[1:18]))
```

```{r flu_revision_plot, fig.width = 20, fig.height = 10, fig.align = "center"}
autoplot(nhsn_filtered, "value") + facet_wrap(~geo_value, ncol = 3, scales = "free") + theme(strip.text.x = element_text(size = 8)) +
  labs(title = "States with the largest mean revision")
```

::: {.notes}
Here we see the revision behavior in the states with the largest mean revision.
Note that some revisions are massive: NH, AZ, MN, MO.
:::

## Data Revisions: Covid

```{r covid_revision}
nhsn_archive_covid$time_type <- "day"
revision_sum <- nhsn_archive_covid %>% epiprocess::revision_analysis(value, min_waiting_period = NULL)
av_re_spread <- revision_sum$revision_behavior %>%
  group_by(geo_value) %>%
  summarize(rel_spread = mean(rel_spread, na.rm = TRUE)) %>%
  arrange(desc(rel_spread)) %>%
  filter(geo_value %nin% c("vi", "as", "gu"))
nhsn_filtered <- nhsn_archive_covid %>%
  filter(geo_value %in% av_re_spread$geo_value[1:18]) %>%
  filter(time_value >= "2024-11-19")
nhsn_filtered$DT %<>%
  mutate(geo_value = factor(geo_value, levels = av_re_spread$geo_value[1:18]))
```

```{r covid_revision_plot, fig.width = 20, fig.height = 10, fig.align = "center"}
autoplot(nhsn_filtered, "value") + facet_wrap(~geo_value, ncol = 3, scales = "free") + theme(strip.text.x = element_text(size = 8)) +
  labs(title = "States with the largest mean revision")
```

::: {.notes}
Here we see the revision behavior in the states with the largest mean revision.
Note that some revisions are massive: DC, AZ, MN, MO, HI.
:::
