# Flu Scores

Throughout, `se_mean` is the standard error of the sample mean with each state a separate estimate.

## Flu Scores: WIS {.smaller}

```{r computing_phase}
flu_current <- nhsn_flu_latest
flu_max <- flu_current %>% group_by(geo_value) %>% summarize(max_value = max(value))
flu_flat_threshold <- 0.9
flu_within_max <- compute_peak_season(flu_current)

sanity_check_classifying <- flu_current %>%
  left_join(flu_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(time_value, first_above, last_above, rel_duration, flu_flat_threshold)) %>%
  group_by(geo_value) %>%
  distinct(phase)
repo_forecasters <- c(setdiff(unique(flu_scores$forecaster), our_forecasters), "CMU-TimeSeries")

se_mean_states <- flu_scores %>%
  left_join(flu_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis_rate = wis * 1e5/pop) %>%
  mutate(pop = factor(pop)) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, flu_flat_threshold)) %>%
  group_by(forecaster, geo_value) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster) %>%
  summarize(
    se_mean_wis = sd(mean_wis) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis) / sqrt(n()),
    se_mean_ae = sd(mean_ae) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))

flu_score_summary <- flu_scores %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  group_by(forecaster) %>%
  mutate(
    min_wis = min(wis[wis > 1e-5]),
    min_ae = min(ae_median[ae_median > 1e-5])
  ) %>%
  summarize(
    mean_wis = round(Mean(wis), 2),
    wis_sd = round(sd(wis), 2),
    pop_norm_wis = round(Mean(wis *1e5/pop), 2),
    pop_norm_wis_sd = round(sd(wis * 1e5/pop), 2),
    geomean_wis = round(GeoMean(wis, min_wis), 2),
    mean_ae = round(Mean(ae_median), 2),
    ae_sd = round(sd(ae_median), 2),
    pop_norm_ae = round(Mean(ae_median*1e5/pop), 2),
    pop_norm_ae_sd = round(sd(ae_median * 1e5/pop), 2),
    geomean_ae = round(GeoMean(ae_median, min_ae), 2),
    mean_cov_50 = round(Mean(interval_coverage_50), 2),
    mean_cov_90 = round(Mean(interval_coverage_90), 2),
    n = n()
  ) %>%
  left_join(se_mean_states, by = "forecaster") %>%
  arrange(mean_wis)
```

```{r flu_datatable, fig.height = 60, fig.width = 12, echo=FALSE}
wis_score_order <- flu_score_summary %>% pull(forecaster)
pop_score_order <- flu_score_summary %>% arrange(pop_norm_wis) %>% pull(forecaster)
flu_wis_summary <- flu_score_summary %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, mean_wis, se_mean_wis, pop_norm_wis, geomean_wis, se_geomean_wis)
datatable(
  flu_wis_summary,
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
    pageLength = 25
  )
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$mean_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "pop_norm_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$pop_norm_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "geomean_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$geomean_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```

::: {.notes}
The external forecasters are in bold.
Our submitted forecasts are "CMU-TimeSeries".
The rest are our component models (as they exist now).

Takeaways:
- `windowed_seasonal_nssp` attains very low WIS
- `climate_linear` performs surprisingly well given its simplicity
:::

## Flu Scores: Absolute Error {.smaller}
```{r flu_scores_ae}
flu_ae_summary <- flu_score_summary %>%
  arrange(mean_ae) %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, mean_ae, se_mean_ae, pop_norm_ae, geomean_ae, se_geomean_ae) %>%
  arrange(mean_ae)
datatable(
    flu_ae_summary,
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
      pageLength = 25
    )
  ) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_ae",
    background = styleColorBar(c(0, max(flu_ae_summary$mean_ae)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "pop_norm_ae",
    background = styleColorBar(c(0, max(flu_ae_summary$pop_norm_ae)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "geomean_ae",
    background = styleColorBar(c(0, max(flu_ae_summary$geomean_ae)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```


## Flu Scores: Coverage {.smaller}
```{r flu_coverage_datatable}
flu_coverage_summary <- flu_score_summary %>%
  arrange(mean_ae) %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, mean_cov_50, mean_cov_90, n)
datatable(
  flu_coverage_summary,
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
    pageLength = 25
  )
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline"))
```

## Flu Scores: WIS Rate Histogram

::: {.notes}
These scores are rates rather than counts, and color gives population.
Note there's an implied fat tail in the distribution of scores.
Overall, the distributions of scores are very similar.

In the increasing phase, CMU-TimeSeries has a smaller left tail, but a more concentrated middle, and smaller right tail, so this doesn't show up in the aggregate.
In the peak phase, CMU-TimeSeries has a thinner left tail (meaning we have fewer cases where we're dead on), which is enough to show up in the aggregate.
The decreasing phase is about the same for the compared models.
:::

```{r flu_score_histogram, fig.height = 13, fig.width = 25, echo=FALSE}
flu_scores %>%
  filter(forecaster %in% c("PSI-PROF", "FluSight-ensemble", "climate_base", "CMU-TimeSeries", "UMass-flusion"))  %>%
  left_join(flu_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis = wis * 1e5/pop) %>%
  mutate(pop = factor(pop)) %>%
  mutate(forecaster = factor(forecaster, levels = wis_score_order)) %>%
  group_by(forecaster) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, flu_flat_threshold)) %>%
  ggplot(aes(x = wis, color = pop,  y = ifelse(after_stat(count) > 0, after_stat(count), NA))) +
  geom_histogram(bins = 70) +
  facet_grid(forecaster~phase) +
  labs(title = "Wis score histogram") +
  ylab("count") +
  xlab("wis, population normalized") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_color_viridis_d()
```

## Flu Scores: WIS Rate Histogram log10

```{r flu_score_histogram_log10, fig.height = 13, fig.width = 25, echo=FALSE}
flu_scores %>%
  filter(forecaster %in% c("PSI-PROF", "FluSight-ensemble", "climate_base", "CMU-TimeSeries", "UMass-flusion"))  %>%
  left_join(flu_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis = wis * 1e5/pop) %>%
  mutate(pop = factor(pop)) %>%
  mutate(forecaster = factor(forecaster, levels = wis_score_order)) %>%
  group_by(forecaster) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, flu_flat_threshold)) %>%
  ggplot(aes(x = wis, color = pop,  y = ifelse(after_stat(count) > 0, after_stat(count), NA))) +
  geom_histogram(bins = 70) +
  facet_grid(forecaster~phase) +
  scale_x_log10() +
  labs(title = "Wis score histogram") +
  ylab("count") +
  xlab("wis, population normalized, log10") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_color_viridis_d()
```

## Flu Scores: Phase {.smallest}
::: {.notes}
We classify the season into three phases: increasing, peak, and decreasing.
Increasing is before the first time the value exceeds a threshold.
Decreasing is the last time the value dips below the threshold.
Peak is between these two.
The threshold is chosen at 50% of the max value.

Note that our WIS/AE is in the top 5, but everyone's score is quite similar and washed out.
We performed well in the increasing/decreasing phases, but most of our error came in the peak phase.
:::

```{r flu_phase}
flu_include_phase <- flu_scores %>%
  left_join(flu_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, flu_flat_threshold))
flu_phase_se <- flu_include_phase %>%
  group_by(forecaster, geo_value, phase) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster, phase) %>%
  summarize(
    se_mean_wis = sd(mean_wis) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis) / sqrt(n()),
    se_mean_ae = sd(mean_ae) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))
phase_scores <- flu_include_phase %>%
  group_by(forecaster, phase) %>%
  summarize(
    wis_sd = round(sd(wis, na.rm = TRUE), 2),
    ae_sd = round(sd(ae_median, na.rm = TRUE), 2),
    across(c(wis, ae_median, interval_coverage_90), \(x) round(Mean(x), 2)),
    n = n(),
    our_forecaster = first(our_forecaster),
    .groups = "drop"
  ) %>%
  left_join(flu_phase_se, by = c("forecaster", "phase"))
sketch <- htmltools::withTags(table(
  class = 'display',
  style = "font-size: 11px;",
  thead(
    tr(
      th(colspan = 2, "forecaster"),
      th(colspan = 2, 'increasing'),
      th(colspan = 2, 'peak'),
      th(colspan = 2, 'decreasing')
    ),
    tr(
      list(th("peak_wis_rank"), th("name"),
           lapply(rep(c('wis', "se_mean_wis"), 3), th)
           )
    )
  )
))
phase_scores_wider <- phase_scores %>%
  select(forecaster, phase, wis, se_mean_wis, our_forecaster) %>%
  pivot_wider(names_from = phase, values_from = c(wis, se_mean_wis), names_expand = TRUE)
wis_max <- phase_scores_wider %>% select(wis_increasing, wis_peak, wis_decreasing) %>% max()
phase_scores_wider %>%
  filter(forecaster %nin% c("seasonal_nssp_latest")) %>%
  select(forecaster, wis_increasing, se_mean_wis_increasing, wis_peak, se_mean_wis_peak, wis_decreasing, se_mean_wis_decreasing) %>%
  arrange(wis_peak) %>%
  datatable(
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "11pt", "'});"),
          "}"),
      pageLength = 25
    ),
    container = sketch
  ) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "wis_increasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_peak",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_decreasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```


## Flu: Sample Forecasts
::: {.notes}
80% CI
We're plotting the 80% CI along with the median.
The locations were chosen based on the scores to have a sample of large and small states with large and small (population normalized) WIS.
We've scaled so everything is in rates per 100k so that it's easier to actually compare; even so the peak value varies drastically.
Forecasters we've produced are blue, while forecasters from other teams are red.
They are ordered by `mean_wis` score, best to worst.
:::

```{r flu_plot_sample_forecast, fig.height = 7, fig.width = 16, echo=FALSE}
plot_geos <- c("ca", "dc", "pa", "hi", "tx")
filtered_flu_forecasts <- flu_forecasts %>%
  filter(quantile %in% c(0.1, 0.5, 0.9), geo_value %in% plot_geos) %>%
  filter(forecaster %in% c("UMass-flusion", "CMU-TimeSeries", "PSI-PROF", "windowed_seasonal"))

flu_forecast_plt <-
  filtered_flu_forecasts %>%
  filter(forecast_date %in% forecast_weeks_to_plot) %>%
  mutate(forecaster = factor(forecaster, levels = wis_score_order)) %>%
  mutate(our_forecaster = factor(forecaster %in% our_forecasters, levels = c(TRUE, FALSE))) %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(value = value * 1e5/ pop) %>%
  pivot_wider(names_from = quantile, values_from = value) %>%
  ggplot(aes(x = target_end_date)) +
  geom_ribbon(aes(ymin = `0.1`, ymax = `0.9`, color = our_forecaster, fill = our_forecaster, group = forecast_date), alpha = 0.5) +
  geom_line(aes(y = `0.5`, color = our_forecaster, group = forecast_date)) +
  geom_line(
    data = flu_current %>%
      filter(time_value > "2024-11-01", geo_value %in% plot_geos) %>%
      left_join(state_census, by = join_by(geo_value == abbr)) %>%
      mutate(value = value * 1e5/ pop),
    aes(x = time_value, y = value)) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  facet_grid(forecaster ~ geo_value, scale = "free") +
  theme(legend.position = "none") +
  scale_x_date(breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 16)) +
  ylab("rates")

flu_forecast_plt
```


## Flu: Forecasting on latest
::: {.notes}
The point is to demonstrate the upper bound of what accurate nowcasting would get us.
named `seasonal_nssp_latest` compare with `windowed_seasonal_nssp`. Generally outperforms all other models by a good margin.
:::

```{r flu_phase_on_latest}
phase_scores_wider %>%
  select(forecaster, wis_increasing, se_mean_wis_increasing, wis_peak, se_mean_wis_peak, wis_decreasing, se_mean_wis_decreasing) %>%
  arrange(wis_peak) %>%
  datatable(
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "11pt", "'});"),
          "}"),
      pageLength = 25
    ),
    container = sketch
  ) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(repo_forecasters, rep("900", length(repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "wis_increasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_peak",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "wis_decreasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```
