---
title: "`r params$disease` health check, `r params$forecast_date`"
date: "Rendered: `r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
params:
  disease: ""
  forecast_date: ""
  health: NULL
  coverage: NULL
  messages: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
health <- params$health
coverage <- params$coverage
messages <- params$messages
components <- if (!is.null(coverage)) bind_rows(purrr::map(coverage, "components")) else tibble()
submission <- if (!is.null(coverage)) bind_rows(purrr::map(coverage, "submission")) else tibble()
status_color <- c(ok = "#2e7d32", attention = "#b26a00", fail = "#c62828", error = "#c62828")[[health$status]]
```

Check this before the fan plots. See `notes/prod-health-check.md` for what each
section means.

<div style="padding: 0.8em 1em; border-left: 6px solid `r status_color`; background: #f6f6f6; margin-bottom: 1em;">
**Status: `r toupper(health$status)`**
`r if (length(health$reasons) > 0) paste0("<br>", paste("&bull;", health$reasons, collapse = "<br>")) else "<br>Everything configured was forecast and submitted."`
</div>

## Submission coverage

Locations submitted per signal and horizon, out of those expected. A horizon
missing more than a quarter of its locations fails the run.

```{r}
if (nrow(submission) > 0) {
  submission %>%
    group_by(signal, horizon = ahead) %>%
    summarize(
      expected = n(),
      submitted = sum(submitted),
      missing = paste(sort(geo_value[!submitted]), collapse = ", "),
      .groups = "drop"
    ) %>%
    datatable(rownames = FALSE, options = list(pageLength = 20, dom = "t"))
} else {
  cat("No coverage table: the health_coverage target was not built.")
}
```

## Substituted components

Configured ensemble components that produced no forecast for a location and
horizon. Their weight went to the components that did (listed under
`used_instead`); for example, a missing `revision_aware` at h−1 means
`climate_linear` alone carried h−1.

```{r}
missing <- if (nrow(components) > 0) filter(components, !present) else tibble()
if (nrow(missing) > 0) {
  used <- components %>%
    filter(present) %>%
    group_by(signal, geo_value, ahead) %>%
    summarize(used_instead = paste(sort(forecaster), collapse = " + "), .groups = "drop")
  missing %>%
    left_join(used, by = c("signal", "geo_value", "ahead")) %>%
    group_by(signal, horizon = ahead, component = forecaster, used_instead) %>%
    summarize(
      n_locations = n(),
      mean_share_lost = round(mean(share), 2),
      locations = paste(sort(geo_value), collapse = ", "),
      .groups = "drop"
    ) %>%
    arrange(signal, horizon, component) %>%
    datatable(rownames = FALSE, options = list(pageLength = 20))
} else {
  cat("None: every configured component forecast every location and horizon.")
}
```

## Pipeline errors and warnings

Errors and warnings that `targets` recorded for this round's targets and for
the shared data targets. A shared target's message can come from an earlier
build; `time` says when.

```{r}
if (!is.null(messages) && nrow(messages) > 0) {
  messages %>%
    mutate(message = substr(message, 1, 400), time = format(time, "%Y-%m-%d %H:%M")) %>%
    datatable(rownames = FALSE, options = list(pageLength = 25))
} else {
  cat("None.")
}
```
