---
title: "`r params$disease` h−1 nowcasts"
date: "Rendered: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    code_folding: hide
params:
  forecaster_parameters: ""
  baseline_id: ""
  forecasts: ""
  scores: ""
  truth_data: ""
  disease: ""
editor_options:
  chunk_output_type: console
---

Forecasts of the already-reported week (h−1), from the families that opt into
`g_nowcast_aheads`. Relative WIS is each forecaster's summed WIS over the
(geo, forecast date) pairs it shares with the baseline, divided by the
baseline's; below 1 beats the baseline.

```{r echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE, cache = FALSE)
ggplot2::theme_set(ggplot2::theme_bw())
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
```

```{r}
scores <- params$scores
baseline_id <- params$baseline_id
base_scores <- scores %>%
  filter(forecaster == baseline_id) %>%
  select(geo_value, forecast_date, target_end_date, wis_base = wis)
paired <- scores %>% inner_join(base_scores, by = c("geo_value", "forecast_date", "target_end_date"))
```

**Baseline:** `r baseline_id` (`revision_ratio_nowcast`).
**Forecast dates:** `r paste(range(scores$forecast_date), collapse = " to ")`.

### Overall

```{r}
summary_table <- paired %>%
  group_by(forecaster) %>%
  summarize(
    n = n(),
    rel_wis = round(sum(wis) / sum(wis_base), 3),
    mean_wis = round(mean(wis), 2),
    mean_ae = round(mean(ae), 2),
    coverage_50 = round(mean(coverage_50), 2),
    coverage_90 = round(mean(coverage_90), 2),
    .groups = "drop"
  ) %>%
  left_join(
    params$forecaster_parameters %>% select(id, family) %>% distinct(),
    by = c("forecaster" = "id")
  ) %>%
  relocate(forecaster, family) %>%
  arrange(rel_wis)
datatable(summary_table)
```

### Over time {.tabset}

#### Relative WIS by forecast date

```{r}
paired %>%
  group_by(forecaster, forecast_date) %>%
  summarize(rel_wis = sum(wis) / sum(wis_base), .groups = "drop") %>%
  ggplot(aes(forecast_date, rel_wis, color = forecaster)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line() +
  scale_y_log10() +
  labs(x = "forecast date", y = "relative WIS (log scale)")
```

#### Coverage by forecast date

```{r}
scores %>%
  group_by(forecaster, forecast_date) %>%
  summarize(`50%` = mean(coverage_50), `90%` = mean(coverage_90), .groups = "drop") %>%
  pivot_longer(c(`50%`, `90%`), names_to = "interval", values_to = "coverage") %>%
  ggplot(aes(forecast_date, coverage, color = forecaster)) +
  geom_hline(aes(yintercept = as.numeric(sub("%", "", interval)) / 100), linetype = "dashed") +
  geom_line() +
  facet_wrap(~interval, ncol = 1) +
  labs(x = "forecast date", y = "empirical coverage")
```

### Relative WIS by location

```{r fig.height = 10}
paired %>%
  filter(forecaster != baseline_id) %>%
  group_by(forecaster, geo_value) %>%
  summarize(rel_wis = sum(wis) / sum(wis_base), .groups = "drop") %>%
  ggplot(aes(rel_wis, reorder(geo_value, rel_wis), color = forecaster)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  labs(x = "relative WIS (log scale)", y = NULL)
```

### Nowcasts against finalized values

Median with 50% and 90% intervals, by target week. The line is the finalized value.

```{r fig.height = 12}
plot_geos <- params$truth_data %>%
  filter(geo_value %in% unique(params$forecasts$geo_value)) %>%
  group_by(geo_value) %>%
  summarize(total = sum(true_value, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total, n = 9) %>%
  pull(geo_value)
bands <- params$forecasts %>%
  filter(geo_value %in% plot_geos) %>%
  mutate(q = case_when(
    abs(quantile - 0.05) < 1e-6 ~ "q05", abs(quantile - 0.25) < 1e-6 ~ "q25",
    abs(quantile - 0.5) < 1e-6 ~ "q50", abs(quantile - 0.75) < 1e-6 ~ "q75",
    abs(quantile - 0.95) < 1e-6 ~ "q95"
  )) %>%
  filter(!is.na(q)) %>%
  pivot_wider(id_cols = c(forecaster, geo_value, target_end_date), names_from = q, values_from = prediction)
truth <- params$truth_data %>%
  filter(geo_value %in% plot_geos, between(target_end_date, min(bands$target_end_date), max(bands$target_end_date)))
ggplot(bands, aes(target_end_date)) +
  geom_ribbon(aes(ymin = q05, ymax = q95, fill = forecaster), alpha = 0.15) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = forecaster), alpha = 0.3) +
  geom_point(aes(y = q50, color = forecaster), size = 0.8) +
  geom_line(data = truth, aes(y = true_value), color = "black") +
  facet_wrap(~geo_value, scales = "free_y", ncol = 3) +
  labs(x = "target week", y = NULL)
```
