---
title: Online calibration of flu forecasts — what we are seeing
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
params:
  hub_dir: "../FluSight-forecast-hub"
  rep_min_pop: 5000000
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
# cache = TRUE chunks are keyed on chunk code, NOT on functions in R/ -- after
# changing calibration code, clear cache/calibration/knitr_cache_findings_flu/.
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 10, fig.height = 6,
  message = FALSE, warning = FALSE, cache = FALSE,
  cache.path = here::here("cache/calibration/knitr_cache_findings_flu/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

This notebook collects the main findings of the MultiQT calibration work
(`notes/CALIBRATION.md` has the full record and the method). In short: per
(location, horizon) series, an additive offset over the 23 hub quantile levels
is learned online from past coverage misses and added to the base forecast.
The questions:

1. **Which way is the base biased?** On NHSN, low, in every season, for both
   `windowed_seasonal` and the submitted ensemble. The ILI+ comparison is on
   hold: the ILI+ replay of `windowed_seasonal` turns out to be broken
   (section 1.2).
2. **Why do short and long horizons behave so differently?** An outcome is
   only learned from once it has settled, `h + 2` rounds after the forecast,
   so the h3 offset is always at least five rounds stale.
3. **What does a leak on the fast term buy?** It stops a stale offset from
   persisting into the part of the season where it has the wrong sign.
4. **What does the sqrt transform buy?** Offsets that stay closer to
   proportionate between trough and peak.

All tracker runs use `lr_mult = 0.03`, `lr_window = 20`, `season_policy =
"carry"`, and 2023-24 as the burn-in season. Scores are on counts against the
latest NHSN vintage, over all hub locations (US included) unless stated. "WIS
change %" is positive when calibration improves on the base forecast. Burn-in
rounds are excluded from every score.

```{r libs}
# Not cached: attaches packages and defines functions every render.
suppressPackageStartupMessages(source(here::here("R/load_all.R")))
source(here::here("scripts/calibration/calibration_ili_backfill.R"))
source(here::here("scripts/calibration/calibration_ws_backfill.R"))
```

```{r load, cache=TRUE}
nhsn_arch <- get_nhsn_data_archive("flu")
nhsn_truth <- nhsn_read_truth("flu", nhsn_arch)
hub_fc <- hub_read_forecasts(hub_dir = here::here(params$hub_dir))
ili <- ili_read_pseudo_hub()
ws <- ws_read_pseudo_hub()

geo_map <- get_population_data() %>%
  distinct(state_id, .keep_all = TRUE) %>%
  transmute(geo_value = state_id, abbrev = toupper(state_id), location = state_code, population)
```

```{r prep, cache=TRUE, dependson="load"}
LIVE_SEASONS <- c("2024-2025", "2025-2026")

# windowed_seasonal replay on the hub's submission rounds only, so seasons,
# gaps and round counts match the ensemble's.
ws_fc <- ws %>%
  filter(reference_date %in% unique(hub_fc$reference_date)) %>%
  select(-as_of_policy)
# Locations both bases cover, for the like-for-like comparison in section 5.
common_loc <- intersect(unique(ws_fc$location), unique(hub_fc$location))

ili_rounds <- hub_label_seasons(unique(ili$forecasts$reference_date))
ili_fc <- ili$forecasts %>%
  filter(reference_date %in% ili_rounds$round_date[!ili_rounds$season %in% c("2010-2011", "2020-2021", "2021-2022", "2023-2024")])

in_season <- function(d) as.integer(format(d, "%m")) %in% c(10:12, 1:5)
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))
```

```{r runs, cache=TRUE, dependson="prep"}
run <- function(fc, truth, ...) {
  args <- modifyList(
    list(
      burn_in_seasons = "2023-2024", transform = "sqrt", lr_window = 20, season_policy = "carry",
      lr_args = list(mult = 0.03, floor = 1e-3), progress = FALSE
    ),
    list(...)
  )
  suppressMessages(do.call(calibrate_hub_forecasts, c(list(fc, truth), args)))
}
operating <- list(slow_init = "burn_in_quantile", lr_slow = list(mult = 0.003), fast_decay = 0.1)

# The submitted ensemble, all hub locations: the method variants.
ens <- list(
  count = run(hub_fc, nhsn_truth, transform = "identity"),
  count_leaky = run(hub_fc, nhsn_truth, transform = "identity", fast_decay = 0.1),
  sqrt = run(hub_fc, nhsn_truth),
  sqrt_leaky = run(hub_fc, nhsn_truth, fast_decay = 0.1),
  warm_static = run(hub_fc, nhsn_truth, slow_init = "burn_in_quantile", lr_args = list(mult = 1e-6, floor = 0)),
  warm_leaky = run(hub_fc, nhsn_truth, slow_init = "burn_in_quantile", fast_decay = 0.1),
  operating = do.call(run, c(list(hub_fc, nhsn_truth), operating))
)
# Ensemble (A) and windowed_seasonal (B) on their common locations.
ens_common <- hub_fc %>% filter(location %in% common_loc)
ws_common <- ws_fc %>% filter(location %in% common_loc)
ab <- list(
  A_single = run(ens_common, nhsn_truth),
  A_leaky = run(ens_common, nhsn_truth, fast_decay = 0.1),
  A_operating = do.call(run, c(list(ens_common, nhsn_truth), operating)),
  B_single = run(ws_common, nhsn_truth),
  B_leaky = run(ws_common, nhsn_truth, fast_decay = 0.1),
  B_operating = do.call(run, c(list(ws_common, nhsn_truth), operating))
)

variant_labels <- c(
  count = "count space, single term (the paper)",
  count_leaky = "count space, leaky fast (decay 0.1)",
  sqrt = "sqrt, single term",
  sqrt_leaky = "sqrt, leaky fast (decay 0.1)",
  warm_static = "sqrt, warm start only (no tracking)",
  warm_leaky = "sqrt, warm start + leaky fast",
  operating = "sqrt, warm start + slow 0.003 + leaky fast (operating point)"
)
```

```{r score-helpers}
wis_table <- function(cals, labels = names(cals), by = character(0), seasons = NULL) {
  purrr::imap(cals, function(cal, nm) {
    left_join(
      hub_quantile_loss(cal, by = by),
      hub_coverage_summary(cal, by = by) %>% select(all_of(c(by, "horizon")), "cal_error_base", "cal_error_cal"),
      by = c(by, "horizon")
    ) %>% mutate(variant = nm)
  }) %>%
    bind_rows() %>%
    { if (!is.null(seasons)) filter(., season %in% seasons) else . } %>%
    mutate(variant = factor(variant, levels = names(cals), labels = labels))
}
wide <- function(tbl, col, digits = 1, id = "variant") {
  tbl %>%
    select(all_of(id), horizon, value = all_of(col)) %>%
    mutate(value = round(value, digits)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = value, names_prefix = "h")
}
median_rows <- function(fc, truth, base) {
  rounds <- hub_label_seasons(unique(fc$reference_date)) %>% select(reference_date = round_date, season)
  fc %>%
    filter(level == 0.5) %>%
    inner_join(truth, by = c("location", "target_end_date")) %>%
    inner_join(rounds, by = "reference_date") %>%
    mutate(base = base)
}
season_of <- function(d) {
  y <- as.integer(format(d, "%Y")) - (as.integer(format(d, "%m")) < 8)
  paste0(y, "-", y + 1)
}
fan_rows <- function(fc, loc, seasons, every = 3L) {
  rounds <- hub_label_seasons(unique(fc$reference_date)) %>%
    filter(season %in% seasons) %>%
    group_by(season) %>%
    filter((season_round - 1L) %% every == 0L) %>%
    ungroup() %>%
    select(reference_date = round_date, season)
  fc %>%
    filter(location == loc, level %in% c(0.1, 0.5, 0.9)) %>%
    inner_join(rounds, by = "reference_date") %>%
    mutate(level = paste0("q", level * 100)) %>%
    select(season, reference_date, target_end_date, level, value) %>%
    tidyr::pivot_wider(names_from = level, values_from = value)
}
fan_plot <- function(fans, truth_rows, vintages = NULL, ylab) {
  p <- ggplot(fans, aes(target_end_date, group = reference_date))
  if (!is.null(vintages)) {
    p <- p + geom_line(aes(time_value, value, group = reference_date), data = vintages, color = "orange3", linewidth = 0.4, alpha = 0.8)
  }
  p +
    geom_line(aes(target_end_date, truth), data = truth_rows, inherit.aes = FALSE, color = "black") +
    geom_ribbon(aes(ymin = q10, ymax = q90), fill = "steelblue", alpha = 0.25) +
    geom_line(aes(y = q50), color = "steelblue4") +
    facet_wrap(~season, scales = "free", ncol = 3) +
    labs(x = NULL, y = ylab)
}
```

# 1. The base bias

## 1.1 On NHSN, both bases forecast too low

The tracker's largest gains come from correcting a persistent bias, so the
first question is whether the base has one. Two bases on NHSN:

- **the submitted ensemble** (`CMU-TimeSeries`), read from the hub checkout;
- **`windowed_seasonal`**, flu prod's seasonal component, replayed by
  `scripts/calibration/calibration_ws_backfill.R` with prod's archive, seed
  and substitutions. 2024-25 and 2025-26 are honest as-of replays; 2023-24
  uses finalized data, since no NHSN vintages exist before 2024-11-19.

```{r bias-data}
med <- bind_rows(
  median_rows(ws_fc, nhsn_truth, "windowed_seasonal"),
  median_rows(hub_fc, nhsn_truth, "submitted ensemble")
)
bias_by_season <- med %>%
  group_by(base, season, horizon) %>%
  summarize(above = 100 * mean(truth > value), .groups = "drop")
```

```{r bias-plot, fig.height=4}
ggplot(bias_by_season, aes(season, above, color = factor(horizon), group = horizon)) +
  geom_hline(yintercept = 50, linetype = "dashed") +
  geom_line() +
  geom_point() +
  facet_wrap(~base) +
  scale_color_viridis_d(name = "horizon", end = 0.9) +
  labs(x = NULL, y = "% of forecasts with truth above the median", title = "Above the dashed line: forecast too low")
```

Pooled over all three seasons: the share of forecasts with truth above the
median, and the summed median as a percentage error on summed truth (negative
means too low; this weights the peak weeks).

```{r bias-table}
bias_pooled <- med %>%
  group_by(base, horizon) %>%
  summarize(above = 100 * mean(truth > value), level_err = 100 * (sum(value) / sum(truth) - 1), .groups = "drop")
knitr::kable(
  left_join(
    wide(bias_pooled, "above", 0, id = "base") %>% rename_with(~ paste(.x, "above %"), starts_with("h")),
    wide(bias_pooled, "level_err", 0, id = "base") %>% rename_with(~ paste(.x, "level %"), starts_with("h")),
    by = "base"
  )
)
```

Where in the season the bias sits: the same two numbers at h0 and h3, split
by how high truth is relative to the location's 90th-percentile in-season
level (trough on the left, peak on the right).

```{r bias-bins}
loc_scale <- nhsn_truth %>%
  filter(target_end_date >= as.Date("2023-10-01"), in_season(target_end_date)) %>%
  group_by(location) %>%
  summarize(scale = stats::quantile(truth, 0.9, names = FALSE), .groups = "drop")
bins <- med %>%
  filter(horizon %in% c(0, 3)) %>%
  inner_join(loc_scale, by = "location") %>%
  mutate(level_bin = cut(truth / scale, c(0, 0.1, 0.25, 0.5, 0.75, 1, Inf), right = FALSE)) %>%
  filter(!is.na(level_bin)) %>%
  group_by(base, horizon, level_bin) %>%
  summarize(above = round(100 * mean(truth > value)), level = round(100 * (sum(value) / sum(truth) - 1)), .groups = "drop") %>%
  mutate(cell = sprintf("%d%% / %+d%%", above, level)) %>%
  select(base, horizon, level_bin, cell) %>%
  tidyr::pivot_wider(names_from = level_bin, values_from = cell)
knitr::kable(bins, caption = "truth above median % / level error %, by truth ÷ location's in-season 90th percentile")
```

- Both bases are low in all three NHSN seasons, and the bias grows with
  horizon. At h0 it is a fairly even level shift (roughly 5–20% low across the
  range). At h3 it concentrates at the peak: near the top of the season the
  median is 60–70% low and truth is above it almost every time. That is the
  base failing to anticipate how high the peak goes, not a constant offset.
- `windowed_seasonal` is the less biased of the two at h−1 and the more biased
  of the two by h3. The ensemble's h−1 is a different model
  (`climate_linear`) and is further off there.

### A representative state

The state whose `windowed_seasonal` above-median share (pooled over h0–h3
and both live seasons) is closest to the pooled value, among states with
population over `r format(params$rep_min_pop, big.mark = ",", scientific = FALSE)`.
Every third round is drawn: the median path from h−1 to h3 with an 80% band,
over finalized truth (black). The orange lines are the NHSN vintage each drawn
round saw (its last eight weeks); where the end of an orange line sits below
the black line, the forecaster was working from data that was later revised
up.

```{r rep-state}
pooled <- med %>%
  filter(base == "windowed_seasonal", horizon >= 0, season %in% LIVE_SEASONS) %>%
  summarize(p = mean(truth > value)) %>%
  pull(p)
rep_scores <- med %>%
  filter(base == "windowed_seasonal", horizon >= 0, season %in% LIVE_SEASONS) %>%
  group_by(location) %>%
  summarize(score = abs(mean(truth > value) - pooled), .groups = "drop") %>%
  inner_join(geo_map, by = "location") %>%
  filter(population > params$rep_min_pop) %>%
  arrange(score)
rep_loc <- rep_scores$location[1]
rep_abbrev <- rep_scores$abbrev[1]
rep_geo <- rep_scores$geo_value[1]
```

```{r nhsn-fans, fig.height=4.5}
fans <- fan_rows(ws_fc, rep_loc, LIVE_SEASONS)
truth_rows <- nhsn_truth %>%
  filter(location == rep_loc) %>%
  mutate(season = season_of(target_end_date)) %>%
  filter(season %in% LIVE_SEASONS, in_season(target_end_date))
vintages <- purrr::map(unique(fans$reference_date), function(rd) {
  as_of <- rd - 3L
  if (as_of < min(nhsn_arch$DT$version)) return(NULL)
  epix_as_of(nhsn_arch, as_of) %>%
    as_tibble() %>%
    filter(geo_value == rep_geo, time_value > rd - 7L * 9L) %>%
    transmute(reference_date = rd, season = season_of(rd), time_value, value)
}) %>% bind_rows()
fan_plot(fans, truth_rows, vintages, ylab = "weekly flu admissions") +
  ggtitle(sprintf("windowed_seasonal on NHSN, %s", rep_abbrev))
```

## 1.2 The ILI+ replay is broken, so the ILI+ comparison is on hold

The plan was to show the same forecaster on the ILI+ decade
(`scripts/calibration/calibration_ili_backfill.R`: finalized state ILI+,
snapshot cut one week before the forecast date), where earlier analysis read
it as biased high. That replay does not do what it is meant to. Its h−1
forecast is a forecast of the last week the model observed, so its median
should track truth almost one-for-one. Regressing log median on log truth:

```{r ili-slope}
slopes <- bind_rows(
  median_rows(ili_fc, ili$truth, "windowed_seasonal, ILI+ replay"),
  med %>% filter(base == "windowed_seasonal") %>% mutate(base = "windowed_seasonal, NHSN replay")
) %>%
  filter(truth > 0, value > 0) %>%
  group_by(base, horizon) %>%
  summarize(slope = coef(stats::lm(log(value) ~ log(truth)))[[2]], .groups = "drop")
knitr::kable(wide(slopes, "slope", 2, id = "base"), caption = "slope of log(median) on log(truth); 1 means proportional")
```

A slope of about 0.5 at h−1 means the median is far too high at the trough
and far too low at the peak, even for a week the model has already seen. The plot below
shows the same thing per round: the first point of each fan (h−1) is already
well off the black line. The snapshots themselves are correct (checked
for Indiana on 2018-01-03: the last observed value is in the snapshot and
matches truth), and the run emits epipredict's *"less latency at bake time
than there was at prep time … will discard the most recent data"* warning, so
the suspect is the latency adjustment (`adjust_latency = "extend_lags"` by
default) interacting with the one-week snapshot truncation. The forecasts for
the representative state:

```{r ili-fans, fig.height=4.5}
ili_seasons <- c("2016-2017", "2017-2018", "2018-2019")
fans <- fan_rows(ili_fc, rep_loc, ili_seasons)
truth_rows <- ili$truth %>%
  filter(location == rep_loc) %>%
  mutate(season = season_of(target_end_date)) %>%
  filter(season %in% ili_seasons, in_season(target_end_date))
fan_plot(fans, truth_rows, ylab = "ILI+ (%)") +
  ggtitle(sprintf("windowed_seasonal ILI+ replay, %s: flat medians, huge bands, and h−1 misses an observed week", rep_abbrev))
```

Everything built on this replay is suspect until it is rebuilt: the "bias has
the opposite sign on ILI+" reading, the ILI+ burn-in negative result, and the
ILI-decade-then-NHSN experiment. Those results are not repeated here.

# 2. Staleness: why h−1/h0 and h2/h3 respond differently

The tracker learns from an outcome only once it has settled (14 days after
the target week), so the outcome of an h-step forecast is revealed `h + 2`
rounds after it was issued, and the step at round `t` is driven by forecasts
issued at `t − h − 2` or earlier. The long horizons have the largest reveal
lag *and* are the hardest to forecast, so they also carry most of the base
error:

```{r lag-table}
base_wis <- hub_quantile_loss(ens$sqrt) %>%
  mutate(
    `reveal lag (rounds)` = horizon + 2L,
    `base WIS (ensemble)` = round(2 * loss_base, 1),
    `relative to h0` = round(loss_base / loss_base[horizon == 0], 2)
  ) %>%
  select(horizon, `reveal lag (rounds)`, `base WIS (ensemble)`, `relative to h0`)
knitr::kable(base_wis)
```

## The offset echoes the past

For the median of each series, define the *needed* shift as
`sqrt(truth) − sqrt(base median)` (what an oracle would have added, in the
tracker's working units) and the *played* shift as
`sqrt(calibrated median) − sqrt(base median)`. Below: the correlation between
the played shift at round `t` and the needed shift at round `t − k`, averaged
over locations, for the single-term sqrt tracker.

```{r lag-corr, fig.height=4}
shift_rows <- function(cal) {
  cal$forecasts %>%
    filter(!is_burn_in, level == 0.5, !is.na(truth), !is.na(value_base)) %>%
    mutate(needed = sqrt(truth) - sqrt(value_base), played = sqrt(value_cal) - sqrt(value_base)) %>%
    select(location, horizon, round_index, reference_date, season, value_base, value_cal, needed, played)
}
ks <- 0:12
lag_corr <- shift_rows(ens$sqrt) %>%
  group_by(location, horizon) %>%
  arrange(round_index, .by_group = TRUE) %>%
  reframe(k = ks, r = purrr::map_dbl(ks, function(k) {
    n <- length(played)
    if (n - k < 10) return(NA_real_)
    x <- played[(k + 1):n]
    y <- needed[1:(n - k)]
    if (sd(x) == 0 || sd(y) == 0) return(NA_real_)
    stats::cor(x, y)
  })) %>%
  group_by(horizon, k) %>%
  summarize(r = mean(r, na.rm = TRUE), .groups = "drop")
peaks <- lag_corr %>% group_by(horizon) %>% slice_max(r, n = 1) %>% ungroup()
ggplot(lag_corr, aes(k, r, color = factor(horizon))) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_line() +
  geom_point(data = peaks, size = 3) +
  geom_vline(aes(xintercept = horizon + 2, color = factor(horizon)), data = distinct(lag_corr, horizon), linetype = "dotted") +
  scale_color_viridis_d(name = "horizon", end = 0.9) +
  scale_x_continuous(breaks = ks) +
  labs(
    x = "k (rounds): played shift at t vs needed shift at t − k",
    y = "mean correlation over locations",
    title = "Dotted: reveal lag h + 2. Large dots: best-matching lag."
  )
knitr::kable(
  lag_corr %>%
    filter(k == 0) %>%
    select(horizon, `r at k = 0` = r) %>%
    left_join(peaks %>% select(horizon, `best k` = k, `r at best k` = r), by = "horizon") %>%
    mutate(`reveal lag` = horizon + 2L, across(starts_with("r "), ~ round(.x, 2)))
)
```

- **The current needed shift and the played shift are negatively
  correlated** at every horizon from h0 up, and more so at longer horizons:
  on average the tracker is pushing the wrong way *right now*.
- **The played shift best matches the needed shift of `h + 4` to `h + 6`
  rounds earlier** (h0–h3): the reveal lag plus two to four rounds for the
  tracker to build up an offset. At h3 that is nine rounds, about two months.
  h−1 has almost no structure: its lag is one round.

Share of rounds where the calibrated median moved in the right direction
(same sign as the needed shift), among rounds where it moved the median by
more than 1%:

```{r sign-table}
sign_right <- purrr::imap(ens[c("count", "sqrt", "sqrt_leaky", "warm_static", "operating")], function(cal, nm) {
  shift_rows(cal) %>%
    filter(abs(value_cal - value_base) > 0.01 * value_base) %>%
    group_by(horizon) %>%
    summarize(right = 100 * mean(sign(played) == sign(needed)), .groups = "drop") %>%
    mutate(variant = variant_labels[[nm]])
}) %>% bind_rows()
knitr::kable(wide(sign_right, "right", 0), caption = "% of moves in the right direction")
```

Because the base is low most of the time, "shift up" is right more often than
not, so every variant is above 50%. The static warm start (a constant
upward shift, no tracking) is right as often as any tracker at h0–h3: what
the trackers add is size and coverage, not timing.

## One series through a season

`r rep_abbrev` at h0 and h3 over both live seasons: the needed shift (black)
and the played shift for three variants, in sqrt units.

```{r series-plot, fig.height=6}
series <- purrr::imap(ens[c("sqrt", "sqrt_leaky", "operating")], function(cal, nm) {
  shift_rows(cal) %>% filter(location == rep_loc, horizon %in% c(0, 3)) %>% mutate(variant = variant_labels[[nm]])
}) %>% bind_rows()
needed <- series %>% distinct(horizon, season, reference_date, needed)
ggplot(series, aes(reference_date)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_line(aes(y = needed), data = needed, color = "black", linewidth = 0.8) +
  geom_line(aes(y = played, color = variant)) +
  facet_grid(paste0("h", horizon) ~ season, scales = "free") +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = NULL, y = "shift of the median (sqrt units)") +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

The needed shift is dominated by the base missing the peak: large, fast and
changing sign within a few weeks, more so at h3. None of the trackers
follows it; at this learning rate they only build a small, smooth level. That
is by design (a faster tracker overshoots; see the sweep in
`calibration_qt_flu.Rmd`), and it is why the gain is mostly a level
correction plus coverage rather than a fix for the peak miss.

## Score by horizon

WIS change % over both live seasons, and calibration error (mean absolute
coverage miss over the 23 levels; base in the first row).

```{r horizon-table}
tbl <- wis_table(ens, variant_labels[names(ens)])
knitr::kable(wide(tbl, "pct_improvement"), caption = "WIS change % (positive is better)")
cal_err <- bind_rows(
  tbl %>% filter(variant == variant_labels[["sqrt"]]) %>% mutate(variant = "base", cal_error_cal = cal_error_base),
  tbl %>% mutate(variant = as.character(variant))
)
knitr::kable(wide(cal_err, "cal_error_cal", 3), caption = "Calibration error")
```

- **h−1 and h0 gain under every tracker.** The lag is one to two rounds,
  short enough that recent misses are still informative. At h−1 the
  ensemble's base is `climate_linear`, a weaker model, so there is more to
  correct.
- **h2 and h3 are where the method choices matter.** A tracker alone is
  negative (single term) or roughly neutral (leaky) there; the warm start (a constant correction
  learned from the burn-in season, which needs no timely feedback) is what
  makes them positive.

# 3. The leak: letting stale offsets decay

With `fast_decay = 0.1`, the fast term is multiplied by 0.9 before each
step, so a correction that new outcomes do not refresh fades (half-life about
6.6 rounds). The effect by calendar month of the reference date:

```{r by-month}
by_month <- function(cal) {
  cal$forecasts %>%
    filter(!is_burn_in, !is.na(truth), !is.na(value_base)) %>%
    mutate(month = format(reference_date, "%b")) %>%
    group_by(horizon, month) %>%
    summarize(
      base = sum(pinball(truth, value_base, level)),
      cal = sum(pinball(truth, value_cal, level)),
      .groups = "drop"
    ) %>%
    group_by(horizon) %>%
    mutate(pct = 100 * (base - cal) / base, share = 100 * base / sum(base)) %>%
    ungroup()
}
month_levels <- month.abb[c(10:12, 1:5)]
month_tbl <- purrr::imap(ens[c("sqrt", "sqrt_leaky", "warm_leaky", "operating")], function(cal, nm) {
  by_month(cal) %>% mutate(variant = variant_labels[[nm]])
}) %>%
  bind_rows() %>%
  filter(month %in% month_levels) %>%
  mutate(month = factor(month, levels = month_levels))

ggplot(month_tbl %>% filter(horizon %in% c(0, 3)), aes(month, pct, color = variant, group = variant)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_line() +
  geom_point() +
  facet_wrap(~ paste0("h", horizon)) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  coord_cartesian(ylim = c(-60, 30)) +
  labs(x = "month of reference date", y = "WIS change % (clipped at −60)") +
  theme(legend.position = "bottom", legend.direction = "vertical")
knitr::kable(
  month_tbl %>%
    filter(horizon == 3) %>%
    mutate(pct = round(pct, 1)) %>%
    select(variant, month, pct) %>%
    arrange(month) %>%
    tidyr::pivot_wider(names_from = month, values_from = pct) %>%
    bind_rows(
      month_tbl %>%
        filter(horizon == 3, variant == variant_labels[["sqrt"]]) %>%
        transmute(variant = "share of base WIS %", month, pct = round(share, 1)) %>%
        arrange(month) %>%
        tidyr::pivot_wider(names_from = month, values_from = pct)
    ),
  caption = "h3: WIS change % by month, and each month's share of the base WIS"
)
```

- **The losses are in March–April and October, which carry little WIS.** In
  March–April the offset built while chasing the winter peak is still in
  place when the base has already come down; in October the offset carried
  over from the previous spring has not been unwound yet (the reveal lag
  means the first live outcomes arrive weeks into the season).
- **The leak alone cuts those losses by two thirds or more** (h3: March −17%
  against −47% for the single-term tracker, October −14% against −48%, April
  +9% against −36%), at the cost of coverage:
  a correction that must be refreshed every round cannot hold a level through
  a five-round reveal lag (calibration error in the table above: the leaky
  variant sits between single-term and uncalibrated).
- **The warm start wins in December–January**, the months that carry most of
  the base WIS (+4% at h3 against +1–2% without it), and so wins overall at
  h1–h3 even though part of the spring loss comes back (March −28 to −33%). The warm start is a constant,
  so it is too large once the season comes down; the leak cannot take it
  away because it only decays the fast term.

# 4. The sqrt transform

In sqrt space the tracker sees `sqrt(truth)` and `sqrt(base quantiles)`,
learns and projects offsets there, and the played quantiles are squared back
(clamped at zero) and scored on counts. The coverage gradient is unchanged,
since a monotone map preserves which side of a quantile the truth falls on;
what changes is the size of a step. A fixed sqrt-space offset `d` on a base
median `b` moves it by `2 d sqrt(b) + d²` counts, so it grows with the square
root of the level instead of staying a fixed count.

The plot shows the played median shift relative to the base median, split by
where the base median sits within the location's season (base median divided
by that location's 90th-percentile in-season truth).

```{r relative-shift, fig.height=4}
rel <- purrr::imap(ens[c("count", "sqrt", "count_leaky", "sqrt_leaky")], function(cal, nm) {
  cal$forecasts %>%
    filter(!is_burn_in, level == 0.5, !is.na(value_base), value_base > 0) %>%
    inner_join(loc_scale, by = "location") %>%
    mutate(
      level_bin = cut(value_base / scale, c(0, 0.05, 0.1, 0.2, 0.4, 0.7, Inf), right = FALSE),
      rel_shift = abs(value_cal - value_base) / value_base,
      variant = variant_labels[[nm]]
    )
}) %>% bind_rows()
rel_summary <- rel %>%
  filter(horizon %in% c(0, 3)) %>%
  group_by(variant, horizon, level_bin) %>%
  summarize(med = 100 * median(rel_shift), .groups = "drop")
ggplot(rel_summary, aes(level_bin, med, color = variant, group = variant)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ paste0("h", horizon)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Paired", name = NULL, direction = -1) +
  labs(
    x = "base median / location's in-season 90th percentile (trough → peak)",
    y = "median |shift| / base median (%, log)"
  ) +
  theme(legend.position = "bottom", legend.direction = "vertical")
knitr::kable(
  rel_summary %>%
    group_by(variant, horizon) %>%
    summarize(`trough ÷ peak` = round(first(med) / last(med), 1), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = horizon, values_from = `trough ÷ peak`, names_prefix = "h"),
  caption = "Relative shift at the trough bin divided by the relative shift at the peak bin"
)
```

```{r transform-table}
tt <- wis_table(ens[c("count", "sqrt", "count_leaky", "sqrt_leaky")], variant_labels[c("count", "sqrt", "count_leaky", "sqrt_leaky")])
knitr::kable(wide(tt, "pct_improvement"), caption = "WIS change %, both live seasons")
knitr::kable(wide(tt, "cal_error_cal", 3), caption = "Calibration error")
```

- **In count space the relative shift spans a much wider range** between
  trough and peak than in sqrt space: an offset sized on peak residuals is
  far too large for the trough. Sqrt narrows that range but does not remove
  it, and it makes the relative shift at the peak somewhat larger.
- **Sqrt is better on WIS at every horizon**, with or without the leak. Its
  calibration error is similar (slightly better at h1–h3 for the single-term
  tracker, slightly worse with the leak).
- **The transform does not address staleness.** A stale sqrt-space offset is
  still wrong-signed, just more modestly sized. The leak and the transform
  are complementary: count space with a leak is comparable to sqrt without
  one at h1–h3.

# 5. The same tracker on two bases

The ensemble (A) and `windowed_seasonal` (B) calibrated the same way, on the
locations both cover. `windowed_seasonal` is less miscalibrated to start with,
so there is less to gain.

```{r ab-table}
ab_tbl <- wis_table(ab, names(ab), by = "season", seasons = LIVE_SEASONS) %>%
  mutate(variant = as.character(variant))
knitr::kable(wide(ab_tbl, "pct_improvement", id = c("variant", "season")) %>% arrange(season, variant), caption = "WIS change %")
ab_err <- bind_rows(
  ab_tbl %>% filter(variant %in% c("A_single", "B_single")) %>% mutate(variant = paste0(substr(variant, 1, 1), "_base"), cal_error_cal = cal_error_base),
  ab_tbl
)
knitr::kable(wide(ab_err, "cal_error_cal", 3, id = c("variant", "season")) %>% arrange(season, variant), caption = "Calibration error")
```

- The operating point helps both bases in both seasons at h0–h3, with small
  gains on `windowed_seasonal` (under 4%).
- On `windowed_seasonal` in 2025-26 the single-term tracker has far better
  calibration error than the operating point (0.009–0.014 against
  0.019–0.045) for slightly less WIS gain: the leak is the price of the
  spring fix, paid in coverage.
- Whether the warm start transfers to a *different data source* was the
  question the ILI+ replay was meant to answer; that has to wait for the
  rebuild (section 1.2). Within NHSN, the warm start comes from the same
  base's burn-in season, and it helps both.

# Summary

| question | answer |
|---|---|
| Is there a stable bias to correct? | On NHSN yes: both bases forecast too low in every season, worse at longer horizons and, at h3, concentrated at the peak. The ILI+ comparison waits on a rebuilt replay. |
| Why do h2/h3 behave differently from h−1/h0? | The offset is at least `h + 2` rounds stale. Right now it is negatively correlated with the needed shift, and it best matches the needed shift of `h + 4` to `h + 6` rounds earlier. The long horizons are also the hardest and carry most of the base WIS. |
| What does the leak do? | Cuts the spring and October losses from stale offsets by two thirds or more, at the cost of coverage. |
| What does the warm start do? | Holds the persistent low bias as a constant: gains where the WIS mass is (December–January), which the trackers alone cannot at h2/h3. Its spring losses come back, since the leak does not touch it. |
| What does sqrt do? | Narrows the range of relative offset size between trough and peak, and is better on WIS than count space at every horizon. It does not fix staleness. |
