---
talk-title: "Delphi Forecasting Season 2024-25"
talk-short-title: "Forecasting 2025"
talk-subtitle: "State level quantile forecasting of flu and covid hospitalizations (NHSN dataset)"
author: "Dmitry Shemetov & David Weber"
other-authors: "help from Logan Brooks, Nat DeFries, Daniel McDonald, and Ryan Tibshirani."
repo-address: ""
date: "2026-05-01"
format:
  revealjs:
    theme: [default, themer.scss]
    css: tachyons-minimal.css
    logo: "gfx/delphi.jpg"
    chalkboard: false
    width: 1280
    height: 720
    html-math-method: mathjax
    slide-number: true
    fig-format: svg
    fig-align: center
    fig-width: 5.33
    fig-height: 3
    scrollable: false
    history: false
    link-external-icon: true
    link-external-newwindow: true
    code-overflow: wrap
    lang: en-CA
    footer: "{{< meta talk-short-title >}} --- {{< meta repo-address >}}"
    output-file: "delphi_forecasting_2025.html"
    self-contained: true
---

<!-- Set any of the above to "" to omit them -->

<!-- The below is _titleslide.qmd from before -->

---
---

\DeclareMathOperator*{\minimize}{minimize}

```{r setup}
#| include: false
primary <- "#a8201a"
secondary <- "#f9c80e"
tertiary <- "#2a76dd"
fourth_colour <- "#311847"
library(epiprocess)
suppressMessages(library(tidyverse))
x <- archive_cases_dv_subset
x_latest <- epix_as_of(x, max_version = max(x$DT$version))
self_max = max(x$DT$version)
versions = seq(as.Date("2020-06-01"), self_max - 1, by = "1 month")
snapshots_all <- map_dfr(versions, function(v) {
  epix_as_of(x, max_version = v) %>% mutate(version = v)}) %>%
  bind_rows(x_latest %>% mutate(version = self_max)) %>%
  mutate(latest = version == self_max)
snapshots <- snapshots_all %>%
  filter(geo_value %in% c("ca", "fl"))
```

```{r cover-art}
#| include: false
snapshots_all |>
  arrange(geo_value, version, time_value) |>
  # filter(!latest) |>
  ggplot(aes(x = time_value, y = percent_cli)) +
  geom_line(
    aes(color = factor(version), group = interaction(geo_value, version))
  ) +
  # geom_vline(aes(color = factor(version), xintercept = version), lty = 3,
  #           size = 0.5) +
  scale_x_date(minor_breaks = "month", labels = NULL) +
  labs(x = "", y = "") +
  theme_void() +
  coord_cartesian(xlim = as.Date(c("2020-10-01", NA)), ylim = c(-5, NA)) +
  scale_color_viridis_d(option = "B", end = .8) +
  theme(legend.position = "none", panel.background = element_blank()) +
  geom_line(
    data = snapshots %>% filter(latest),
    aes(x = time_value, y = percent_cli, group = geo_value),
    inherit.aes = FALSE, color = primary)
```

::: flex
::: w-30

:::
::: w-70

## {{< meta talk-title >}} {background-image="gfx/cover-art-1.svg" background-position="bottom"}

::: {.notes}
Just one project, the grad students are up to other things, such as nowcasting tools, real case count estimates, etc.
:::
### {{< meta talk-subtitle >}}

<br>

#### {{< meta author >}}
{{< meta other-authors >}}

{{< meta date >}}
:::
:::

<!-- The above is _titleslide.qmd from before -->

```{r useful_functions}
compute_peak_season <- function(data_current, threshold = 0.5, start_of_year = as.Date("2024-11-01")) {
  season_length <- data_current %>% pull(time_value) %>% max() - start_of_year
  data_current %>%
    filter(time_value > start_of_year) %>%
    group_by(geo_value) %>%
    mutate(max_val = max(value)) %>%
    filter(value >= threshold * max_val) %>%
    summarize(first_above = min(time_value), last_above = max(time_value)) %>%
    mutate(
      duration = last_above - first_above,
      rel_duration = as.integer(duration) / as.integer(season_length))
}
classify_phase <- function(time_value, first_above, last_above, rel_duration, threshold) {
  case_when(
    rel_duration > threshold ~ "flat",
    time_value < first_above ~ "increasing",
    time_value > last_above ~ "decreasing",
    .default = "peak"
  ) %>% factor(levels = c("increasing", "peak", "decreasing", "flat"))
}
```

```{r data_setup, include=FALSE}
library(aws.s3)
library(qs2)
library(tidyverse)
library(epidatr)
library(epipredict)
library(epidatasets)
theme_set(theme_bw())
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev = "png"
)
ggplot2::theme_set(ggplot2::theme_bw())
here::i_am("scripts/reports/delphi_forecasting_2025/delphi_forecasting_2025.qmd")
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
Mean <- function(x) mean(x, na.rm = TRUE)
GeoMean <- function(x, offset = 0) exp(Mean(log(x + offset)))
our_forecasters <- c("linear", "windowed_seasonal", "windowed_seasonal_nssp", "climate_base", "climate_geo_agged", "climate_linear", "ensemble_windowed", "retro_submission", "CMU-TimeSeries", "seasonal_nssp_latest", "ensemble_mix")
rename_forecasters <- function(df) {
  dplyr::mutate(df, forecaster = dplyr::case_match(
    forecaster,
    "windowed_seasonal_extra_sources" ~ "windowed_seasonal_nssp",
    "ensemble_linclim_windowed_seasonal" ~ "retro_submission",
    "ens_ar_only" ~ "ensemble_windowed",
    .default = forecaster
  ))
}

cache_dir <- file.path(getwd(), "_cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
cache_key <- format(as.Date(cut(Sys.Date(), "week")), "%Y-%m-%d")
cached_load <- function(name, fn) {
  path <- file.path(cache_dir, paste0(name, "_", cache_key, ".rds"))
  if (file.exists(path)) readRDS(path) else { result <- fn(); saveRDS(result, path); result }
}
s3_read <- function(object, bucket = "forecasting-team-data") {
  aws.s3::s3read_using(qs2::qs_read, object = object, bucket = bucket)
}

nhsn_archive_flu   <- cached_load("nhsn_archive_flu",   \() s3_read("2024/flu_hosp_prod/objects/nhsn_archive_data"))
nhsn_archive_covid <- cached_load("nhsn_archive_covid", \() s3_read("2024/covid_hosp_prod/objects/nhsn_archive_data"))
flu_forecasts      <- cached_load("flu_forecasts",      \() bind_rows(
  s3_read("2024/flu_hosp_prod/objects/local_forecasts_and_ensembles_nhsn"),
  s3_read("2024/flu_hosp_prod/objects/external_forecasts_full") %>%
    filter(target == "wk inc flu hosp" | is.na(target))
) %>% rename_forecasters())
flu_scores         <- cached_load("flu_scores",         \() s3_read("2024/flu_hosp_prod/objects/scores_nhsn") %>% rename_forecasters() %>% mutate(our_forecaster = forecaster %in% our_forecasters, target_end_date = forecast_date + ahead * 7L))
forecast_dates     <- cached_load("forecast_dates",     \() s3_read("2024/flu_hosp_prod/objects/forecast_dates"))
covid_scores       <- cached_load("covid_scores",       \() s3_read("2024/covid_hosp_prod/objects/scores_nhsn") %>% rename_forecasters() %>% mutate(our_forecaster = forecaster %in% our_forecasters, target_end_date = forecast_date + ahead * 7L))
covid_forecasts    <- cached_load("covid_forecasts",    \() bind_rows(
  s3_read("2024/covid_hosp_prod/objects/local_forecasts_and_ensembles_nhsn"),
  s3_read("2024/covid_hosp_prod/objects/external_forecasts_full")
) %>%
ungroup() %>%
filter(target == "wk inc covid hosp" | is.na(target)) %>%
rename_forecasters())

covid_current <- nhsn_archive_covid %>%
  epix_as_of_current() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi"))
covid_max <- covid_current %>% group_by(geo_value) %>% summarize(max_value = max(value))
covid_within_max <- compute_peak_season(covid_current)

forecast_week <- flu_scores$forecast_date %>% unique()
forecast_weeks_to_plot <- c(seq.Date(min(forecast_week), max(forecast_week), by = 3*7), as.Date("2025-01-18"), as.Date("2025-02-01"))
forecast_weeks_to_plot %in% (flu_scores$forecast_date %>% unique())
forecast_weeks_to_plot %in% (covid_scores$forecast_date %>% unique())

flu_current <- nhsn_archive_flu %>%
  epix_as_of_current() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi"))
```


# Major Forecasting Challenges

- **non-stationarity** - varying vaccination rates, new strains, changing mobility patterns, superspreader events, etc.
- **large seasonal variation (flu)** - season start, duration, and peak size vary
- **lack of seasonal component (covid)** - seasonal pattern in the data is weak or inconsistent (post-pandemic)
- **data quality issues** - initial observations are incomplete, regular reporting delays, reporting errors, etc.
- **few seasons to train on** - NHSN data starts in 2020, the first two seasons are excluded because they were anomalous due to the pandemic, so effectively only 3 seasons of data are available for training

::: {.notes}
- non stationarity
- seasonal variation
- non-seasonal
- data quality
- few seasons: for Flu, you can augment this using older data sources, which we do. The data isn't on the same scale however, so you have to be careful.
:::

# The Data
Target: NHSN State level hospitalizations

## Recent Seasons Data: Flu

```{r flu_archive, fig.width = 20, fig.height = 10, fig.align = "center"}
nhsn_latest <- nhsn_archive_flu %>%
  epix_as_of_current() %>%
  # Cut out the tiny fraction of 2019/20 data (there's like 58 data points)
  filter(time_value >= "2020-10-01") %>%
  add_season_info() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi")) %>%
  dplyr::rename(counts = value)
n_flu_seasons <- n_distinct(nhsn_latest$season)
ggplot(nhsn_latest, aes(x = season_week, y = counts, color = season, size = season)) +
  geom_line() +
  facet_wrap(~geo_value, scale = "free") +
  theme(legend.text = element_text(size = 12)) +
  scale_size_manual(values = c(rep(0.5, n_flu_seasons - 1), 1))
```

::: {.notes}
Here we see the 2024/25 season in pink.
It's the worst season in 15 years and very different from the previous 2 seasons.
Before that, the 2020/21 and the 2021/22 seasons are essentially negligible due to the pandemic (so we've dropped them).

[SciAm article with some speculation on why](https://www.scientificamerican.com/article/why-this-years-flu-season-is-the-worst-in-more-than-a-decade/).
- Late start -> less protection + 2nd wave
- different strain (H3N2 H1N1)
- decreased vaccination (not that low)
- lower levels of immunity from decreased exposure

This bears out somewhat in NSSP, but it is not as extreme an increase, so it may be a change in reporting.
:::

## Recent Seasons Data: Covid

```{r covid_archive, fig.width = 20, fig.height = 10, fig.align = "center"}
nhsn_latest_covid <- nhsn_archive_covid %>%
  epix_as_of_current() %>%
  add_season_info() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi"), season %nin% c("2019/20", "2020/21", "2021/22")) %>%
  rename(counts = value)
n_covid_seasons <- n_distinct(nhsn_latest_covid$season)
ggplot(nhsn_latest_covid, aes(x = season_week, y = counts, color = season, size = season)) +
  geom_line() +
  facet_wrap(~geo_value, scale = "free") +
  theme(legend.text = element_text(size = 12)) +
  scale_size_manual(values = c(rep(0.5, n_covid_seasons - 1), 1))
```

::: {.notes}
Here we've dropped the 2020 through 2022 seasons because they absolutely swamp the past 3 years.
Even so, this year is frequently without a season at all depending on the geo (for instance, see CA and FL and TX); part of the reason for this is the late summer wave, which is not visible in this data.
:::

## Data Revisions: Flu

```{r flu_revision}
nhsn_archive_flu$time_type <- "day"
revision_sum <- nhsn_archive_flu %>% epiprocess::revision_analysis(value, min_waiting_period = NULL)
av_re_spread <- revision_sum$revision_behavior %>%
  group_by(geo_value) %>%
  summarize(rel_spread = mean(rel_spread, na.rm = TRUE)) %>%
  arrange(desc(rel_spread)) %>%
  filter(geo_value %nin% c("vi", "as", "gu"))
nhsn_filtered <- nhsn_archive_flu %>%
  filter(geo_value %in% av_re_spread$geo_value[1:18]) %>%
  filter(time_value >= "2024-11-19")
nhsn_filtered$DT %<>%
  mutate(geo_value = factor(geo_value, levels = av_re_spread$geo_value[1:18]))
```

```{r flu_revision_plot, fig.width = 20, fig.height = 10, fig.align = "center"}
autoplot(nhsn_filtered, "value") + facet_wrap(~geo_value, ncol = 3, scales = "free") + theme(strip.text.x = element_text(size = 10)) +
  labs(title = "States with the largest mean revision")
```

::: {.notes}
Revision behavior in the states with the largest mean revision, ordered by mean revision.
Note that some revisions are massive: NH, AZ, MN, MO.

There's ~2 kinds of revision; the predictable upwards that takes a couple of weeks to filter in (nm), and massive on-time revisions (az).
:::

## Data Revisions: Covid

```{r covid_revision}
nhsn_archive_covid$time_type <- "day"
revision_sum <- nhsn_archive_covid %>% epiprocess::revision_analysis(value, min_waiting_period = NULL)
av_re_spread <- revision_sum$revision_behavior %>%
  group_by(geo_value) %>%
  summarize(rel_spread = mean(rel_spread, na.rm = TRUE)) %>%
  arrange(desc(rel_spread)) %>%
  filter(geo_value %nin% c("vi", "as", "gu"))
nhsn_filtered <- nhsn_archive_covid %>%
  filter(geo_value %in% av_re_spread$geo_value[1:18]) %>%
  filter(time_value >= "2024-11-19")
nhsn_filtered$DT %<>%
  mutate(geo_value = factor(geo_value, levels = av_re_spread$geo_value[1:18]))
```

```{r covid_revision_plot, fig.width = 20, fig.height = 10, fig.align = "center"}
autoplot(nhsn_filtered, "value") + facet_wrap(~geo_value, ncol = 3, scales = "free") + theme(strip.text.x = element_text(size = 8)) +
  labs(title = "States with the largest mean revision")
```

::: {.notes}
Here we see the revision behavior in the states with the largest mean revision.
Note that some revisions are massive: DC, AZ, MN, MO, HI.
:::

# Our Models

Ensembles of simpler forecasters

## Our Models: Baselines

::: {.notes}
Our typical forecast submission is an ensemble of several models.
These assume the data is untrustworthy in some way. (linear throws out old data, climate_base throws out historical data).

Contrast with the current baseline which just forecasts today's value forwards.
:::

- `linear`: linearly extrapolate the last 4 weeks, quantiles from the fit error
- `climate_base`: a climatological model, quantiles estimated from a 7 week period centered on the `target_date`'s season week from historical data for that geo

- `climate_linear`: a weighted ensemble of the climatological and linear, depending on the ahead and quantile

## Our Models: Baselines
```{r computing_phase}
flu_current <- nhsn_archive_flu %>%
  epix_as_of_current() %>%
  filter(geo_value %nin% c("as", "gu", "mp", "vi"))
flu_max <- flu_current %>% group_by(geo_value) %>% summarize(max_value = max(value))
flu_flat_threshold <- 0.9
flu_within_max <- compute_peak_season(flu_current)

sanity_check_classifying <- flu_current %>%
  left_join(flu_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(time_value, first_above, last_above, rel_duration, flu_flat_threshold)) %>%
  group_by(geo_value) %>%
  distinct(phase)
flu_repo_forecasters <- c(setdiff(unique(flu_scores$forecaster), our_forecasters), "CMU-TimeSeries")

se_mean_states <- flu_scores %>%
  left_join(flu_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis_rate = wis * 1e5/pop) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, flu_flat_threshold)) %>%
  group_by(forecaster, geo_value) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median),
    mean_rate_wis = Mean(wis * 1e5/pop)
  ) %>%
  group_by(forecaster) %>%
  summarize(
    se_mean_wis = sd(mean_wis) / sqrt(n()),
    se_rate_wis = sd(mean_rate_wis) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis) / sqrt(n()),
    se_mean_ae = sd(mean_ae) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))
flu_score_summary <- flu_scores %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  group_by(forecaster) %>%
  mutate(
    min_wis = min(wis[wis > 1e-5]),
    min_ae = min(ae_median[ae_median > 1e-5])
  ) %>%
  summarize(
    mean_wis = round(Mean(wis), 2),
    wis_sd = round(sd(wis), 2),
    pop_norm_wis = round(Mean(wis *1e5/pop), 2),
    pop_norm_wis_sd = round(sd(wis * 1e5/pop), 2),
    geomean_wis = round(GeoMean(wis, min_wis), 2),
    #nWISzero = sum(wis < 1e-5),
    mean_ae = round(Mean(ae_median), 2),
    ae_sd = round(sd(ae_median), 2),
    pop_norm_ae = round(Mean(ae_median*1e5/pop), 2),
    pop_norm_ae_sd = round(sd(ae_median * 1e5/pop), 2),
    geomean_ae = round(GeoMean(ae_median, min_ae), 2),
    #nAEzero = sum(ae_median < 1e-5),
    mean_cov_50 = round(Mean(interval_coverage_50), 2),
    mean_cov_90 = round(Mean(interval_coverage_90), 2),
    n = n()
  ) %>%
  left_join(se_mean_states, by = "forecaster") %>%
  arrange(mean_wis)
wis_score_order <- flu_score_summary %>% pull(forecaster)


#flu_scores %>% group_by(geo_value,)
```

```{r demo_climate_linear, fig.height = 7, fig.width = 16, echo=FALSE}
plot_geos <- c("pa", "mi")
demo_start <- as.Date("2024-11-15")
demo_date <- as.Date("2025-02-19")
demo_end <- as.Date("2025-04-15")
filtered_flu_forecasts <-
  flu_forecasts %>%
  filter(quantile %in% c(0.05, 0.1, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9, 0.95), geo_value %in% plot_geos) %>%
  filter(forecaster %in% c("climate_base", "linear", "climate_linear"))

flu_forecast_plt <-
  filtered_flu_forecasts %>%
  filter(forecast_date == demo_date) %>%
  mutate(forecaster = factor(forecaster, levels = c("linear", "climate_base", "climate_linear"))) %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(value = value * 1e5 / pop) %>%
  pivot_wider(names_from = quantile, values_from = value) %>%
  ggplot(aes(x = target_end_date)) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, color = NA, fill = forecaster, group = forecast_date), alpha = 0.2) +
  geom_ribbon(aes(ymin = `0.1`, ymax = `0.9`, color = NA, fill = forecaster, group = forecast_date), alpha = 0.2) +
  geom_ribbon(aes(ymin = `0.25`, ymax = `0.75`, color = NA, fill = forecaster, group = forecast_date), alpha = 0.2) +
  geom_ribbon(aes(ymin = `0.4`, ymax = `0.6`, color = NA, fill = forecaster, group = forecast_date), alpha = 0.2) +
  geom_line(aes(y = `0.5`, color = forecaster, group = forecast_date), size = 1.2) +
  geom_line(
    data = flu_current %>%
      filter(time_value > demo_start, time_value < demo_end, geo_value %in% plot_geos) %>%
      left_join(state_census, by = join_by(geo_value == abbr)) %>%
      mutate(value = value * 1e5 / pop),
    aes(x = time_value, y = value)) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  facet_grid(geo_value ~ forecaster, scale = "free") +
  theme(legend.position = "none") +
  scale_x_date(breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 16)) +
  ylab("rates")

flu_forecast_plt
```

:::{.notes}
`climate_linear` is ~linear at the start, and ~ climate at the end.
Chosen to clearly represent behavior, atypically good.
:::

## Our Models: Autoregressive

`windowed_seasonal`: an AR model trained on data from the same seasonal window in previous years

$$x_{t+h} = a_0 + a_1 \cdot x_{t-7} + a_2 \cdot x_{t-14}$$

- $x$ is hospitalization rates, $t$ is the forecast date, $h$ is the horizon in days
- built using [epipredict](https://cmu-delphi.github.io/epipredict/)
- training data: 3 weeks before now through 5 weeks after the target date (-3/+8 from now)

```{r windowed_seasonal_illustration, fig.width=25, fig.height=3.2, echo=FALSE, out.extra='class="r-stretch"'}
demo_week <- 10
backward_window <- 3
forward_window <- 5
max_ahead <- 3
anomalous <- c("2020/21", "2021/22")
seasons <- c("2020/21", "2021/22", "2022/23", "2023/24", "2024/25", "2025/26")
week_min <- -5
week_max <- 26

pal <- RColorBrewer::brewer.pal(8, "Set2")
target_week <- demo_week + max_ahead
window_left  <- demo_week - backward_window
window_right <- target_week + forward_window

x_breaks <- c(seq(week_min + 1, 0, by = 2), seq(1, week_max, by = 5))
x_labels <- ifelse(x_breaks <= 0, as.character(52 + x_breaks), as.character(x_breaks))

tidyr::expand_grid(
  season = factor(seasons, levels = rev(seasons)),
  season_week = week_min:week_max
) %>%
  dplyr::mutate(category = dplyr::case_when(
    season %in% anomalous                                          ~ "excluded (anomalous)",
    season == "2025/26" & season_week == demo_week                                    ~ "forecast date",
    season == "2025/26" & season_week == target_week                                  ~ "target date",
    season == "2025/26" & season_week > demo_week                                     ~ "unobserved",
    season == "2025/26" & ((season_week == demo_week - 1) | (season_week == demo_week - 2))                                ~ "predictor",
    season == "2025/26" & season_week >= window_left & season_week <= demo_week - 2   ~ "training window",
    season == "2025/26" & ((season_week == window_left - 1) | (season_week == window_left - 2))                              ~ "predictor",
    season == "2025/26"                                                               ~ "observed and discarded",
    season_week >= window_left & season_week <= window_right       ~ "training window",
    !(season %in% anomalous) & ((season_week == window_left - 1) | (season_week == window_left - 2)) ~ "predictor",
    TRUE                                                       ~ "observed and discarded"
  )) %>%
  ggplot(aes(x = season_week, y = season, fill = category)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_vline(xintercept = window_left - 0.5,
             linetype = "dashed", color = "#2ca25f", linewidth = 1.2) +
  geom_vline(xintercept = window_right + 0.5,
             linetype = "dashed", color = "#2ca25f", linewidth = 1.2) +
  geom_vline(xintercept = demo_week,
             linetype = "solid", color = pal[2], linewidth = 1.2) +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "black", linewidth = 0.8) +
  annotate("text", x = -2.5,      y = 6.7,  label = "← prior yr", size = 7,   color = "grey40") +
  annotate("text", x = demo_week,   y = -0.15, label = "now",   size = 6.5, color = pal[2], vjust = 1) +
  annotate("text", x = target_week, y = -0.15, label = "+3 wk", size = 6.5, color = pal[4], vjust = 1) +
  scale_fill_manual(
    name = "",
    values = c(
      "training window"       = pal[1],
      "forecast date"         = pal[2],
      "target date"           = pal[4],
      "predictor"             = pal[3],
      "observed and discarded" = "#d9d9d9",
      "unobserved"            = "#f0f0f0",
      "excluded (anomalous)"  = "#bdbdbd"
    )
  ) +
  scale_x_continuous(breaks = x_breaks, labels = x_labels, name = "Week") +
  coord_cartesian(clip = "off") +
  labs(y = "") +
  theme_bw() +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "grey50"),
    legend.text = element_text(size = 16),
    legend.key.size = unit(0.7, "cm"),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 17),
    plot.margin = margin(t = 5, r = 5, b = 30, l = 5)
  )
```

::: {.notes}
The "training window" approach to handling seasonality gave a large improvement over a plain AR model.
:::

## Our Models: Autoregressive
For forecasting NSSP for either flu or covid, since it is a rate to begin with, we just use that 2 term autoregressive model, using all states as training data.

## Our Models: Autoregressive — Data Preprocessing

NHSN reports **counts**; we convert to **rates per 100k** before forecasting

For Flu, we also augment with ILI+ and FluSURV data (inspired by Flusion), then whiten on a per-geo+source basis:

- NHSN usable data: **2 seasons** (2022/23–2023/24, excluding pandemic years)
- ILI+ extends this to **~25 seasons** (back to 1997)
- FluSURV extends this another **~11 seasons** (back to 2009/10)

To make these comparable, we normalize in several ways:


1. **Variance stabilization** — 4th root transform
2. **Scaling** — divide by the inter-quantile range (5th–95th percentile)
3. **Centering** — subtract so that the median is zero

::: {.notes}
The augmented data sources improved the model score.
We did not see much improvement from adding more lags or tweaking the whitening settings.
Covid uses rates directly across all states; no whitening.
:::

## Our Models: Autoregressive with Exogenous Predictors

`windowed_seasonal_nssp`: adds NSSP emergency department visits as an exogenous predictor

$$x_{t+h} = a_0 + a_1 x_{t-7} + a_2 x_{t-14} + a_3 y_{t-7} + a_4 y_{t-14}$$

- $y$ is NSSP (ED visit rates) — a leading indicator for hospitalizations
- no whitening, no augmented sources (NSSP doesn't go back far enough)
- same formulation for both Flu and Covid

::: {.notes}
Also tried Google Symptoms and NWSS state-level aggregations — no meaningful improvement.
ED visits lead hospitalizations by ~1 week, so the signal is real but modest.
:::

## Our Models: Autoregressive — Fitting

We use **quantile regression** via [epipredict](https://cmu-delphi.github.io/epipredict/reference/quantile_reg.html), which wraps the standard [quantreg](https://rdrr.io/cran/quantreg/) package:

- Each quantile level $\tau$ is fit as a **separate weighted linear program**:

$$\minimize_{\beta} \sum_i w_i \cdot L(y_i,\, x_i^\top \beta)$$

- The **pinball loss** penalises under- and over-shoots asymmetrically:

$$L(y, t) := \begin{cases} (1-\tau)(t - y), & y < t \\ \tau(y - t), & y \geq t \end{cases}$$

::: {.notes}
This is the standard quantile regression / pinball loss / check function setup.
The weighted LP formulation means no closed-form solution, but it's fast enough in practice.
Directly optimises the WIS scoring metric used for evaluation.
:::

## Our Models: Ensemble

Our submission (`CMU-TimeSeries`) is a **weighted ensemble** of the component models, with geo-level weights

- Weights are (nominally) set manually each week based on performance and data quality
- **2024/25**: initially relied heavily on `climate_linear` — the season started late and we were unsure of data quality. Eventually transitioned back to AR forecasters
- **2025/26**: almost exclusively `windowed_seasonal_nssp` (NSSP exogenous), with a small contribution from `windowed_seasonal` (the whitening AR) for Flu

::: {.notes}
The geo-level weights let us down-weight or exclude forecasters for specific states where data quality is poor or a model is known to misbehave.
The shift from climate_linear to nssp reflects the NSSP data becoming more reliable and the season being more "normal".
:::

## Infrastructure

**Pipeline**: weekly automated runs via [`targets`](https://docs.ropensci.org/targets/) + [`epipredict`](https://cmu-delphi.github.io/epipredict/)

- Each model variant is a target; `targets` skips recomputation when inputs haven't changed
- Forecasts and scores uploaded to S3; rendered notebooks for review each week
- Use of make to handle shared environmental variables, submitting after manual checks and any adjustments


::: {.notes}
The targets pipeline means the whole run is reproducible and incremental — adding a new forecast date only reruns the affected targets.
The exploration grid was run as a backtest over 2023/24 and 2024/25; the best-performing configs (by WIS) informed the production ensemble choices.
We ended up with windowed_seasonal_nssp as the clear winner once NSSP data quality stabilised.
:::

## Model Selection
**Exploration**: backtest over the 2023/24 season of versioned data with several model variants

- lags, up to 5 weeks back
- Exogenous sources: NSSP, NWSS, NWSS region, Veteran's affairs, google symptoms
- Seasonal augmentation methods: 
  + windowed (various sizes, {3, 5, 7} weeks × forward ∈ {1, 3, 5} weeks)
  + Aux signals:
    - seasonal sinusoid
    - PCA
    - climatological
- Residual whitening: on/off

## Flu: Sample Forecasts Rate Scale
::: {.notes}
80% CI
Locations large/small and large/small WIS.
We've scaled so everything is in rates per 100k; even after that the peak value varies drastically.
Forecasters we've produced are blue, while forecasters from other teams are red; PSI-PROF is a SIR model, UMass-flusion is an AR using pre-2020 data.
Something about the Flu data in particular leads to constantly decreasing forecasts.
:::

```{r flu_plot_sample_forecast, fig.height = 7, fig.width = 16, echo=FALSE}
plot_geos <- c("ca", "dc", "pa", "hi", "tx")
flu_season_start <- if (any(flu_forecasts$forecast_date >= "2025-11-01")) "2025-11-01" else "2024-11-01"
truth_season_start <- if (flu_season_start == "2025-11-01") "2025-11-01" else "2024-11-01"

# external forecasters present in 2025/26 with good geo coverage,
# ranked by 2024/25 WIS where available
flu_external_to_plot <- c("UMass-flusion", "FluSight-ensemble")
flu_plot_forecasters <- c("CMU-TimeSeries", "windowed_seasonal", "windowed_seasonal_nssp", flu_external_to_plot)

to_wednesday <- function(d) floor_date(d, "weeks", week_start = 7) + 3
season_max_fd <- max(flu_forecasts$forecast_date[flu_forecasts$forecast_date >= flu_season_start])
target_wednesdays <- seq.Date(
  to_wednesday(as.Date(flu_season_start)),
  to_wednesday(season_max_fd),
  by = 3 * 7
)
flu_forecast_plt <-
  flu_forecasts %>%
  filter(quantile %in% c(0.1, 0.5, 0.9), geo_value %in% plot_geos,
         forecaster %in% flu_plot_forecasters,
         forecast_date >= flu_season_start,
         forecast_date <= target_end_date,
         to_wednesday(forecast_date) %in% target_wednesdays) %>%
  mutate(our_forecaster = factor(forecaster %in% our_forecasters, levels = c(TRUE, FALSE))) %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(value = value * 1e5 / pop) %>%
  pivot_wider(names_from = quantile, values_from = value) %>%
  ggplot(aes(x = target_end_date)) +
  geom_ribbon(aes(ymin = `0.1`, ymax = `0.9`, color = our_forecaster, fill = our_forecaster, group = forecast_date), alpha = 0.5) +
  geom_line(aes(y = `0.5`, color = our_forecaster, group = forecast_date)) +
  geom_line(
    data = flu_current %>%
      filter(time_value > truth_season_start, geo_value %in% plot_geos) %>%
      left_join(state_census, by = join_by(geo_value == abbr)) %>%
      mutate(value = value * 1e5 / pop),
    aes(x = time_value, y = value)) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  ggh4x::facet_grid2(forecaster ~ geo_value, scales = "free", independent = "y") +
  theme(legend.position = "none") +
  scale_x_date(
    breaks = "1 month",
    labels = \(x) ifelse(format(x, "%m") == "01", format(x, "%b\n%Y"), format(x, "%b"))
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 16)) +
  ylab("rates per 100k")

flu_forecast_plt
```

## Covid: Sample Forecasts Rate Scale
::: {.notes}
80% CI
Covid doesn't have the decreasing forecasters problem. (Though PA has some revision troubles like many locations).
:::

```{r computing_phase_covid}
covid_flat_threshold <- 0.6
sanity_check_classifying <- covid_current %>%
  left_join(covid_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(time_value, first_above, last_above, rel_duration, covid_flat_threshold)) %>%
  group_by(geo_value) %>%
  distinct(phase)

covid_repo_forecasters <- c(setdiff(unique(covid_scores$forecaster), our_forecasters), "CMU-TimeSeries")

covid_se_mean_states <- covid_scores %>%
  left_join(covid_within_max, by = "geo_value") %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(wis_rate = wis * 1e5/pop) %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold)) %>%
  group_by(forecaster, geo_value) %>%
  summarize(
    mean_wis = Mean(wis),
    mean_rate_wis = Mean(wis_rate),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster) %>%
  summarize(
    se_mean_wis = sd(mean_wis) / sqrt(n()),
    se_mean_rate_wis = sd(mean_rate_wis) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis) / sqrt(n()),
    se_mean_ae = sd(mean_ae) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))


covid_score_summary <- covid_scores %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  group_by(forecaster) %>%
  mutate(
    min_wis = min(wis[wis > 1e-5]),
    min_ae = min(ae_median[ae_median > 1e-5])
  ) %>%
  summarize(
    mean_wis = round(Mean(wis), 2),
    wis_sd = round(sd(wis), 2),
    pop_norm_wis = round(Mean(wis *1e5/pop), 2),
    pop_norm_wis_sd = round(sd(wis * 1e5/pop), 2),
    geomean_wis = round(GeoMean(wis, min_wis), 2),
    mean_ae = round(Mean(ae_median), 2),
    ae_sd = round(sd(ae_median), 2),
    pop_norm_ae = round(Mean(ae_median*1e5/pop), 2),
    pop_norm_ae_sd = round(sd(ae_median * 1e5/pop), 2),
    geomean_ae = round(GeoMean(ae_median, min_ae), 2),
    mean_cov_50 = round(Mean(interval_coverage_50), 2),
    mean_cov_90 = round(Mean(interval_coverage_90), 2),
    n = n()
  ) %>%
  left_join(covid_se_mean_states, by = "forecaster") %>%
  arrange(mean_wis)
covid_wis_score_order <- covid_score_summary %>% pull(forecaster)
```

```{r covid_plot_sample_forecast, fig.height = 7, fig.width = 16, echo=FALSE}
plot_geos <- c("ca", "dc", "pa", "hi", "tx")
covid_season_start <- if (any(covid_forecasts$forecast_date >= "2025-11-01")) "2025-11-01" else "2024-11-01"
covid_truth_season_start <- if (max(covid_current$time_value) >= as.Date("2025-11-01")) "2025-11-01" else "2024-11-01"

covid_2526_coverage <- covid_forecasts %>%
  filter(forecast_date >= covid_season_start, !forecaster %in% our_forecasters) %>%
  group_by(forecaster) %>%
  summarize(n_geos = n_distinct(geo_value)) %>%
  filter(n_geos >= 40)
covid_2425_wis <- covid_scores %>%
  filter(!forecaster %in% our_forecasters) %>%
  group_by(forecaster) %>%
  summarize(mean_wis = mean(wis, na.rm = TRUE))
covid_external_to_plot <- covid_2526_coverage %>%
  left_join(covid_2425_wis, by = "forecaster") %>%
  arrange(coalesce(mean_wis, Inf)) %>%
  pull(forecaster) %>% head(2)
covid_plot_forecasters <- c("CMU-TimeSeries", "windowed_seasonal_nssp", "windowed_seasonal", "CovidHub-ensemble", "OHT_JHU-nbxd")

covid_season_max_fd <- max(covid_forecasts$forecast_date[covid_forecasts$forecast_date >= covid_season_start])
covid_target_wednesdays <- seq.Date(
  to_wednesday(as.Date(covid_season_start)),
  to_wednesday(covid_season_max_fd),
  by = 3 * 7
)

covid_forecast_plt <-
  covid_forecasts %>%
  filter(quantile %in% c(0.1, 0.5, 0.9), geo_value %in% plot_geos,
         forecaster %in% covid_plot_forecasters,
         forecast_date >= covid_season_start,
         forecast_date <= target_end_date,
         to_wednesday(forecast_date) %in% covid_target_wednesdays,
         (target == "wk inc covid hosp") | is.na(target)) %>%
  mutate(our_forecaster = factor(forecaster %in% our_forecasters, levels = c(TRUE, FALSE))) %>%
  left_join(state_census, by = join_by(geo_value == abbr)) %>%
  mutate(value = value * 1e5 / pop) %>%
  pivot_wider(names_from = quantile, values_from = value) %>%
  ggplot(aes(x = target_end_date)) +
  geom_ribbon(aes(ymin = `0.1`, ymax = `0.9`, color = our_forecaster, fill = our_forecaster, group = forecast_date), alpha = 0.5) +
  geom_line(aes(y = `0.5`, color = our_forecaster, group = forecast_date)) +
  geom_line(
    data = covid_current %>%
      filter(time_value > covid_truth_season_start, geo_value %in% plot_geos) %>%
      left_join(state_census, by = join_by(geo_value == abbr)) %>%
      mutate(value = value * 1e5 / pop),
    aes(x = time_value, y = value)) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  ggh4x::facet_grid2(forecaster ~ geo_value, scales = "free", independent = "y") +
  theme(legend.position = "none") +
  scale_x_date(
    breaks = "1 month",
    labels = \(x) ifelse(format(x, "%m") == "01", format(x, "%b\n%Y"), format(x, "%b"))
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 16)) +
  ylab("rates per 100k")

covid_forecast_plt
```

## Scoring Metrics {visibility="hidden"}

:::: {.columns}
::: {.column width="50%"}

- **Weighted interval score (WIS)**
  - Discrete approximation to CRPS, which sums the colored area in the figure squared
  - Roughly, the number of mis-forecasted hospitalizations, weighted by probability squared.
- 80% prediction interval coverage (80%PI)
  - Proportion of times the true value falls within the center 80% probability interval
:::
::: {.column width="50%"}

```{r wis_definition_intuition, fig.width = 5, fig.height = 5}
df <- data.frame(x = seq(-1, 2, length.out = 3000))

# Calculate y values for both functions
df$y1 <- pnorm(df$x, mean = 0.5, sd = 0.25)  # Normal CDF
df$y2 <- ifelse(df$x >= 0.75, 1, 0)        # Step function

# Plot
ggplot(df, aes(x = x)) +
  # Plot the functions with color aesthetic
  geom_line(aes(y = y1, color = "CDF of forecast")) +
  geom_line(aes(y = y2, color = "Eventually observed value")) +
  # Add the ribbon between the functions
  geom_ribbon(aes(ymin = pmin(y1, y2), ymax = pmax(y1, y2), fill = "WIS ~ ∫(F(x) - I(x ≥ .75))²"),
              alpha = 0.3) +
  # Customize colors and fills
  scale_color_manual(name = "", values = c("CDF of forecast" = "red",
                                          "Eventually observed value" = "blue")) +
  scale_fill_manual(name = "", values = c("WIS ~ ∫(F(x) - I(x ≥ .75))²" = "red")) +
  # Add legend inside plot
  theme(legend.position = c(0.2, 0.8),  # Adjust these values to position the legend
        legend.background = element_rect(fill = "white"),
        legend.text = element_text(size = 6)) +
  labs(x = "x", y = "CDF", title = "WIS example") +
  theme(plot.title = element_text(hjust = 0.5))
```

:::
::::


## Flu Scores: WIS {visibility="hidden"}

:::{.notes}
mean\_wis gives the average wis across all weeks, aheads, and geos.
mean\_rate\_wis converts the errors to a rate scale before averaging, which increases the importance of getting Wyoming correct relative to California.

Throughout, `se_mean` is the standard error of the sample mean with each state a separate estimate.

Our submitted forecasts are "CMU-TimeSeries".
The rest are our component models (as they exist now).

Takeaways:
- most forecasters are statistically indistinguishable
- `windowed_seasonal_nssp` attains very low WIS
- `climate_linear` performs surprisingly well given its simplicity
:::

```{r flu_datatable, fig.height = 60, fig.width = 12, echo=FALSE}
flu_score_non_plot_forecasters <- c("retro_submission", "ensemble_windowed", "climate_geo_agged")
pop_score_order <- flu_score_summary %>% arrange(pop_norm_wis) %>% pull(forecaster)
flu_wis_summary <- flu_score_summary %>%
  select(forecaster, mean_wis, se_mean_wis, mean_rate_wis = pop_norm_wis, se_rate_wis) %>%
  filter(forecaster %nin% flu_score_non_plot_forecasters)
datatable(
  flu_wis_summary,
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
    pageLength = 25
  )
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(flu_repo_forecasters, rep("900", length(flu_repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$mean_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_rate_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$mean_rate_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```

Forecasters in **bold** are actually submitted to the hub. Others are our experimental models.

Related Flusight-hub dashboard [here](https://reichlab.io/flusight-dashboard/eval.html): https://reichlab.io/flusight-dashboard/eval.html

## Flu Scores: Phase {visibility="hidden"}
::: {.notes}
We classify the season into three phases: increasing, peak, and decreasing.
Increasing is before the first time the value exceeds a threshold.
Decreasing is the last time the value dips below the threshold.
Peak is between these two.
The threshold is chosen at 50% of the max value.

Note that our WIS/AE is in the top 5, but everyone's score is quite similar and washed out.
We performed well in the increasing/decreasing phases, but most of our error came in the peak phase.
:::

```{r flu_phase}
flu_include_phase <- flu_scores %>%
  left_join(flu_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, flu_flat_threshold))
flu_phase_se <- flu_include_phase %>%
  group_by(forecaster, geo_value, phase) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster, phase) %>%
  summarize(
    se_mean_wis = sd(mean_wis) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis) / sqrt(n()),
    se_mean_ae = sd(mean_ae) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))
phase_scores <- flu_include_phase %>%
  group_by(forecaster, phase) %>%
  summarize(
    wis_sd = round(sd(wis, na.rm = TRUE), 2),
    ae_sd = round(sd(ae_median, na.rm = TRUE), 2),
    across(c(wis, ae_median, interval_coverage_90), \(x) round(Mean(x), 2)),
    n = n(),
    our_forecaster = first(our_forecaster),
    .groups = "drop"
  ) %>%
  left_join(flu_phase_se, by = c("forecaster", "phase")) %>%
  rename(mean_wis = wis)
sketch <- htmltools::withTags(table(
  class = 'display',
  style = "font-size: 11px;",
  thead(
    tr(
      th(colspan = 2, "forecaster"),
      th(colspan = 2, 'increasing'),
      th(colspan = 2, 'peak'),
      th(colspan = 2, 'decreasing')
    ),
    tr(
      list(th("peak_wis_rank"), th("name"),
           lapply(rep(c('mean_wis', "se_mean_wis"), 3), th)
           )
    )
  )
))
phase_scores_wider <- phase_scores %>%
  select(forecaster, phase, mean_wis, se_mean_wis, our_forecaster) %>%
  pivot_wider(names_from = phase, values_from = c(mean_wis, se_mean_wis))
wis_max <- phase_scores_wider %>% select(mean_wis_increasing, mean_wis_peak, mean_wis_decreasing) %>% max()
phase_scores_wider %>%
  select(forecaster, mean_wis_increasing, se_mean_wis_increasing, mean_wis_peak, se_mean_wis_peak, mean_wis_decreasing, se_mean_wis_decreasing) %>%
  arrange(mean_wis_peak) %>%
  filter(forecaster %nin% flu_score_non_plot_forecasters) %>%
  datatable(
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "11pt", "'});"),
          "}"),
      pageLength = 25
    ),
    container = sketch
  ) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(flu_repo_forecasters, rep("900", length(flu_repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_wis_increasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_wis_peak",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_wis_decreasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )

```
Forecasters in **bold** are actually submitted to the hub. Others are our experimental models.

Related Flusight-hub dashboard [here](https://reichlab.io/flusight-dashboard/eval.html): https://reichlab.io/flusight-dashboard/eval.html

## Covid Scores: WIS {visibility="hidden"}
::: {.notes}
`CMU-TimeSeries` did significantly worse; a large part of this is poor performance on first few weeks when we were using the baseline models.
:::

```{r covid_datatable, fig.height = 60, fig.width = 12, echo=FALSE}
wis_score_order <- covid_score_summary %>% pull(forecaster)
pop_score_order <- covid_score_summary %>% arrange(pop_norm_wis) %>% pull(forecaster)
covid_score_non_plot_forecasters <- c("ensemble_windowed", "ensemble_mix", "climate_geo_agged")
covid_wis_summary <-
  covid_score_summary %>%
  filter(forecaster %nin% covid_score_non_plot_forecasters) %>%
  select(forecaster, mean_wis, se_mean_wis, mean_rate_wis =pop_norm_wis, se_mean_rate_wis)

  datatable(
    covid_wis_summary,
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "10pt", "'});"),
          "}"),
      pageLength = 25
    )
  ) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(covid_repo_forecasters, rep("900", length(covid_repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$mean_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_rate_wis",
    background = styleColorBar(c(0, max(flu_wis_summary$mean_rate_wis)), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```

Forecasters in **bold** are actually submitted to the hub. Others are our experimental models.

Related covidhub-dashboard [here](https://reichlab.io/covidhub-dashboard/eval.html): https://reichlab.io/covidhub-dashboard/eval.html

## Covid Scores: Aggregated By Forecast Date {visibility="hidden"}
:::{.notes}
Too much inside baseball to be worth including.
:::
```{r, fig.height = 6, fig.width = 12, echo=FALSE}
agg_covid <- covid_scores %>%
  filter(forecaster %nin% c("climate_geo_agged", "linear")) %>%
  filter(forecast_date > "2024-10-01") %>%
  filter(forecast_date != as.Date("2025-01-25")) %>%
  group_by(forecaster, forecast_date) %>%
  summarize(
    mean_wis = round(Mean(wis), 2),
    wis_sd = round(sd(wis), 2),
    geomean_wis = round(GeoMean(wis), 2),
    mean_ae = round(Mean(ae_median), 2),
    geomean_ae = round(GeoMean(ae_median), 2),
    mean_interval_coverage_90 = round(Mean(interval_coverage_90), 2),
  )

p <- ggplot(agg_covid, aes(x = forecast_date, y = mean_wis, color = forecaster)) +
  geom_line() +
  #geom_ribbon(aes(ymin = mean_wis - wis_sd, ymax = mean_wis + wis_sd, fill = forecaster), alpha = 0.1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Forecast Date", y = "Mean WIS")

style(ggplotly(p), visible = "legendonly", traces = c(1, 6, 8, 9, 10, 11, 12, 13, 14))
```


## Covid Scores: Phase {visibility="hidden"}

```{r covid_phase}
covid_include_phase <- covid_scores %>%
  left_join(covid_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold))

covid_phase_se <- covid_include_phase %>%
  group_by(forecaster, geo_value, phase) %>%
  summarize(
    mean_wis = Mean(wis),
    geo_mean_wis = GeoMean(wis),
    mean_ae = Mean(ae_median),
    geomean_ae = GeoMean(ae_median)
  ) %>%
  group_by(forecaster, phase) %>%
  summarize(
    se_mean_wis = sd(mean_wis, na.rm = TRUE) / sqrt(n()),
    se_geomean_wis = sd(geo_mean_wis, na.rm = TRUE) / sqrt(n()),
    se_mean_ae = sd(mean_ae, na.rm = TRUE) / sqrt(n()),
    se_geomean_ae = sd(geomean_ae, na.rm = TRUE) / sqrt(n())
  ) %>%
  mutate(across(starts_with("se"), \(x) round(x, 2)))

phase_scores <-
  covid_scores %>%
  left_join(covid_within_max, by = "geo_value") %>%
  mutate(phase = classify_phase(target_end_date, first_above, last_above, rel_duration, covid_flat_threshold)) %>%
  group_by(forecaster, phase) %>%
  summarize(
    wis_sd = round(sd(wis, na.rm = TRUE), 2),
    ae_sd = round(sd(ae_median, na.rm = TRUE), 2),
    across(c(wis, ae_median, interval_coverage_90), \(x) round(Mean(x), 2)),
    n = n(),
    our_forecaster = first(our_forecaster),
    .groups = "drop"
  ) %>%
  left_join(covid_phase_se, by = c("forecaster", "phase")) %>%
  rename(mean_wis = wis)
sketch <- htmltools::withTags(table(
  class = 'display',
  style = "font-size: 12px;",
  thead(
    tr(
      th(colspan = 2, "forecaster"),
      th(colspan = 2, 'increasing'),
      th(colspan = 2, 'peak'),
      th(colspan = 2, 'decreasing'),
      th(colspan = 2, 'flat')
    ),
    tr(
      list(th("peak_wis_rank"), th("name"),
           lapply(rep(c('mean_wis', 'se_mean_wis'), 4), th)
           )
    )
  )
))
phase_scores_wider <- phase_scores %>%
  select(forecaster, phase, mean_wis, se_mean_wis, our_forecaster) %>%
  pivot_wider(names_from = phase, values_from = c(mean_wis, se_mean_wis)) %>%
  select(forecaster, mean_wis_increasing, se_mean_wis_increasing, mean_wis_peak, se_mean_wis_peak, mean_wis_decreasing, se_mean_wis_decreasing, mean_wis_flat, se_mean_wis_flat) %>%
  arrange(mean_wis_peak)
wis_max <- phase_scores_wider %>% 
  filter(forecaster %nin% covid_score_non_plot_forecasters) %>%
  select(mean_wis_increasing, mean_wis_peak, mean_wis_decreasing) %>%
  max()
datatable(
  phase_scores_wider %>% filter(forecaster %nin% c("seasonal_nssp_latest")),
  fillContainer = FALSE,
  options = list(
    initComplete = htmlwidgets::JS(
          "function(settings, json) {",
          paste0("$(this.api().table().container()).css({'font-size': '", "12pt", "'});"),
          "}"),
    pageLength = 25
  ),
  container = sketch
) %>%
  formatStyle("forecaster", target = c("cell"),
              fontWeight = styleEqual(covid_repo_forecasters, rep("900", length(covid_repo_forecasters))),
              textDecoration = styleEqual("CMU-TimeSeries", "underline")) %>%
  formatStyle(
    "mean_wis_increasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_wis_peak",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_wis_decreasing",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  formatStyle(
    "mean_wis_flat",
    background = styleColorBar(c(0, wis_max), 'lightblue'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  )
```

Forecasters in **bold** are actually submitted to the hub. Others are our experimental models.

Related covidhub-dashboard [here](https://reichlab.io/covidhub-dashboard/eval.html): https://reichlab.io/covidhub-dashboard/eval.html

# Future Work

- **"Always decreasing" bias** — many forecasters underpredict during rising phases, especially for flu; root cause is data bias, possible fix is phase-specific models
- **Metrocasting** — city/county level forecasting for NSSP

# Future Work

- **Nowcasting / revision anticipation** — NHSN data is substantially revised in the weeks after initial release; a nowcasting model that anticipates these revisions could recover ~10% of forecast error

```{r future_work_revision, fig.width = 20, fig.height = 7, fig.align = "center", echo = FALSE}
nhsn_archive_flu$time_type <- "day"
revision_sum <- nhsn_archive_flu %>% epiprocess::revision_analysis(value, min_waiting_period = NULL)
av_re_spread <- revision_sum$revision_behavior %>%
  group_by(geo_value) %>%
  summarize(rel_spread = mean(rel_spread, na.rm = TRUE)) %>%
  arrange(desc(rel_spread)) %>%
  filter(geo_value %nin% c("vi", "as", "gu"))
nhsn_filtered <- nhsn_archive_flu %>%
  filter(geo_value %in% av_re_spread$geo_value[1:12]) %>%
  filter(time_value >= "2024-11-19")
nhsn_filtered$DT %<>%
  mutate(geo_value = factor(geo_value, levels = av_re_spread$geo_value[1:12]))
autoplot(nhsn_filtered, "value") +
  facet_wrap(~geo_value, ncol = 4, scales = "free") +
  theme(strip.text.x = element_text(size = 10)) +
  labs(title = "Flu: states with the largest mean revision (2024/25 season)")
```


# Appendix {visibility="hidden"}

## Explicit phase definitions: Flu {visibility="hidden"}

```{r flu_season_definitions, fig.width = 25, fig.height = 15, fig.align = "center"}
flu_current %>%
  filter(time_value > "2024-11-01") %>%
  autoplot(value, .facet_by = "geo_value") +
  geom_vline(data = flu_within_max, aes(xintercept = first_above)) +
  geom_vline(data = flu_within_max, aes(xintercept = last_above)) +
  facet_wrap(~geo_value, scale = "free") +
  theme(legend.position = "none")
```


## Explicit phase definitions: Covid {visibility="hidden"}
```{r, fig.width = 18, fig.height = 9}
no_phase <- covid_within_max %>% arrange(desc(rel_duration)) %>% filter(rel_duration > covid_flat_threshold)
covid_current %>%
  filter(time_value > "2024-11-01", geo_value %nin% no_phase$geo_value) %>%
  autoplot(value, .facet_by = "geo_value") +
  geom_vline(data = covid_within_max, aes(xintercept = first_above)) +
  geom_vline(data = covid_within_max, aes(xintercept = last_above)) +
  facet_wrap(~geo_value, scale = "free") +
  ylim(0, NA) +
  theme(legend.position = "none")
```

## Explicit phase definitions: Covid `flat` {visibility="hidden"}
```{r, fig.width = 15, fig.height = 8}
covid_current %>%
  filter(time_value > "2022-06-01") %>%
  filter(geo_value %in% no_phase$geo_value) %>%
  autoplot(value, .facet_by = "geo_value") +
  geom_vline(aes(xintercept = as.Date("2024-11-01"))) +
  ylim(0, NA) +
  theme(legend.position = "none")
```
