---
title: Evaluation of Covid climatological model 11-20-2024
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
params:
  forecaster_set: 1
editor_options:
  chunk_output_type: console
---

$$\\[.4in]$$

```{r echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(MMWRweek)
library(readr)
library(rlang)
library(magrittr)
library(aws.s3)
library(data.table)
library(dplyr)
library(DT)
library(ggplot2)
library(plotly)
library(readr)
library(stringr)
library(tidyr)
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
forecast_date <- as.Date("2024-11-20")
epi_data <- readr::read_csv("https://data.cdc.gov/resource/ua7e-t2fy.csv?$limit=20000&$select=weekendingdate,jurisdiction,totalconfc19newadm")
epi_data <- epi_data %>%
  mutate(
    epiweek = epiweek(weekendingdate),
    epiyear = epiyear(weekendingdate)
  ) %>%
  left_join(
    (.) %>%
      distinct(epiweek, epiyear) %>%
      mutate(
        season = convert_epiweek_to_season(epiyear, epiweek),
        season_week = convert_epiweek_to_season_week(epiyear, epiweek)
      ),
    by = c("epiweek", "epiyear")
  )
epi_data <- epi_data %>%
  mutate(time_value = as.Date(weekendingdate), geo_value = tolower(jurisdiction), nhsn = totalconfc19newadm) %>%
  select(-weekendingdate, -jurisdiction, -totalconfc19newadm)
ahead <- 0
aheads <- -1:3
window_size <- 3
recent_window <- 3
geos_to_plot <- c("mn", "ca", "pa", "usa", "ri") #epi_data %>% pull(geo_value) %>% unique() %>% setdiff(c("as", "mp", "vi")) # 
all_geos_to_plot <- epi_data %>% pull(geo_value) %>% unique() %>% setdiff(c("as", "mp", "vi"))
plot_quantiles <- c(0.6, 0.75, 0.95, 0.99)
quantile_alphas <- c(0.9, 0.6, 0.4, 0.3)
min_plot_quantiles <- c(0.75, 0.90)
min_quantile_alphas <- c(0.9, 0.3)
forecast_week <- epiweek(forecast_date)
last_date_data <- epi_data %>%
  pull(time_value) %>%
  max()
relevant_period <- epi_data %>% select(epiweek, time_value, season) %>%distinct %>% filter(
      (abs(forecast_week + min(aheads) - epiweek) <= window_size) |
      (abs(forecast_week + min(aheads) - epiweek) <= window_size) |
      (last_date_data - time_value <= recent_window * 7),
      season != "2024/25"
    )
relevant_period %<>% group_by(season) %>% summarise(start = min(time_value), stop = max(time_value))
```

The model used is specified [here](https://github.com/cmu-delphi/exploration-tooling/tree/c36c7ddf9a0568426fef7ce5ef0ef6712e6faedf/R/forecasters/climatological_model.R).
For flu, the data for the 2020/21 and 2021/22 seasons are dropped, while for covid we've done versions both including and excluding.

# Comparing a couple of methods {.tabset}
* `quantile_basic` uses base R's quantile's, each geo separately, using a window of 7 weeks (3 before and after the target date), and the 3 weeks before the last week of data
* `epipred quantile` is more or less the same, but using the quantile extrapolation in epipredict after fitting the quantiles .25, .5, and .75. The main difference is that the tails are more extreme and can be outside the range of previously seen values for that location.
* `geo aggregated` is a global fit, that first puts everything on a rate scale, and then does a single global forecast, and then returns to the original scales. It is just using base R quantiles.

The intervals here are the 75th, 95th, and 99th quantiles. Some notes:

* Compared with Flu, there are far more locations that are much higher than the original value


```{r, results = "hide"}
quantile_basic <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, disease = "covid")) %>%
    bind_rows() %>%
    mutate(geo_type = "state", forecaster = "quantile basic")

quantile_7 <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, quant_type = 7, disease = "covid")) %>%
    bind_rows() %>%
    mutate(geo_type = "state", forecaster = "quantile_7") 

epipred_extrapolation <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, quantile_method = "epipredict", disease = "covid")) %>% bind_rows() %>%
    mutate(geo_type = "state", forecaster = "epipred quantile") 

geo_aggregate <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, geo_agg = TRUE, disease = "covid")) %>% bind_rows() %>%
    mutate(geo_type = "state", forecaster = "geo aggregated")

all_results <-  bind_rows(
  geo_aggregate,
  quantile_basic,
  epipred_extrapolation
)
```
## Small set of states
```{r}
small_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% geos_to_plot)

```{r}
the_plot <-
  all_results %>%
  filter(geo_value %in% geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = small_truth_data, quantiles = min_plot_quantiles, alphas = min_quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 1500, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```
## Small set of states, more quantiles
```{r}
small_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% geos_to_plot)

```{r}
the_plot <-
  all_results %>%
  filter(geo_value %in% geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = small_truth_data, quantiles = plot_quantiles, alphas = quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 1500, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

## Full set of states
```{r}

full_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% all_geos_to_plot)
the_plot <-
  all_results %>%
  filter(geo_value %in% all_geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = full_truth_data, quantiles = min_plot_quantiles, alphas = min_quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 5000, width = 1700) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

## Full set of states, more quantiles
```{r}

full_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% all_geos_to_plot)
the_plot <-
  all_results %>%
  filter(geo_value %in% all_geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = full_truth_data, quantiles = plot_quantiles, alphas = quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 5000, width = 1700) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

# Dropping 2020/21 and 2021/22 {.tabset}
Given that those 2 years are clearly larger than whatever we're settling into in 2022 and beyond, we should try a model that doesn't include those years, which may be closer to something used this year
```{r, results = "hide"}
quantile_basic <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead)) %>%
    bind_rows() %>%
    mutate(geo_type = "state", forecaster = "quantile basic")

quantile_7 <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, quant_type = 7)) %>%
    bind_rows() %>%
    mutate(geo_type = "state", forecaster = "quantile_7") 

epipred_extrapolation <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, quantile_method = "epipredict")) %>% bind_rows() %>%
    mutate(geo_type = "state", forecaster = "epipred quantile") 

geo_aggregate <- lapply(-1:3, \(ahead) climatological_model(epi_data, forecast_date, ahead, geo_agg = TRUE)) %>% bind_rows() %>%
    mutate(geo_type = "state", forecaster = "geo aggregated")

all_results <-  bind_rows(
  geo_aggregate,
  quantile_basic,
  epipred_extrapolation
)
```
## Small set of states
```{r}
small_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% geos_to_plot)

```{r}
the_plot <-
  all_results %>%
  filter(geo_value %in% geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = small_truth_data, quantiles = min_plot_quantiles, alphas = min_quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 1500, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```
## Small set of states, more quantiles
```{r}
small_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% geos_to_plot)

```{r}
the_plot <-
  all_results %>%
  filter(geo_value %in% geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = small_truth_data, quantiles = plot_quantiles, alphas = quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 1500, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

## Full set of states
```{r}

full_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% all_geos_to_plot)
the_plot <-
  all_results %>%
  filter(geo_value %in% all_geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = full_truth_data, quantiles = min_plot_quantiles, alphas = min_quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 5000, width = 1700) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

## Full set of states, more quantiles
```{r}

full_truth_data <- epi_data %>%
  mutate(value = nhsn, target_end_date = time_value, forecaster = "nhsn") %>%
  filter(geo_value %in% all_geos_to_plot)
the_plot <-
  all_results %>%
  filter(geo_value %in% all_geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = full_truth_data, quantiles = plot_quantiles, alphas = quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 5000, width = 1700) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

# Geo Mean Averaging {.tabset}
Since the dynamics are exponential, averaging on the log scale is probably a better method. That way, if the national is predicting something an order of magnitude higher that won't swamp the local information. This is the same set using the geomean instead of the arithmetic mean.

* `"all three"`the mean of all 3
* `"epipred quantile"`the mean of the geo aggregated and the `quantile basic` above
* `"quantile_basic"`the mean of the geo aggregated and the `epipredict quantile` above

```{r, echo=FALSE, result = 'hide'}
geomean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
mean_climatological <-
  all_results %>%
  filter((forecaster == "geo aggregated") | (forecaster == "epipred quantile")) %>%
  # for some reason, the quantiles have different float values
  mutate(quantile = round(quantile, digits = 2)) %>%
  group_by(geo_value, forecast_date, target_end_date, geo_type, quantile) %>%
  summarise(value = geomean(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(forecaster = "epipred quantile") %>%
  bind_rows(
    all_results %>%
    # for some reason, the quantiles have different float values
    mutate(quantile = round(quantile, digits = 2)) %>%
    group_by(geo_value, forecast_date, target_end_date, geo_type, quantile) %>%
    summarise(value = geomean(value, na.rm = TRUE), .groups = "drop") %>%
    mutate(forecaster = "quantile basic"),
    all_results %>%
    filter((forecaster == "geo aggregated") | (forecaster == "quantile basic")) %>%
    # for some reason, the quantiles have different float values
    mutate(quantile = round(quantile, digits = 2)) %>%
    group_by(geo_value, forecast_date, target_end_date, geo_type, quantile) %>%
    summarise(value = geomean(value, na.rm = TRUE), .groups = "drop") %>%
    mutate(forecaster = "all three")
  )
```

## Small set of states
```{r}
the_plot <- mean_climatological %>%
  filter(geo_value %in% geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = small_truth_data, quantiles = min_plot_quantiles, alphas = min_quantile_alphas, relevant_period = relevant_period)
ggplotly(the_plot, tooltip = "text", height = 1500, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```
## Small set of states, more quantiles
```{r}
the_plot <- mean_climatological %>%
  filter(geo_value %in% geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = small_truth_data, quantiles = plot_quantiles, alphas = quantile_alphas, relevant_period = relevant_period)
ggplotly(the_plot, tooltip = "text", height = 1500, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

## Full set of states
```{r}
the_plot <- mean_climatological %>%
  filter(geo_value %in% all_geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = full_truth_data, quantiles = min_plot_quantiles, alphas = min_quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 5000, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```


## Full set of states, more quantiles
```{r}
the_plot <- mean_climatological %>%
  filter(geo_value %in% all_geos_to_plot) %>%
  plot_forecasts(forecast_date, geo_type = "state", truth_data = full_truth_data, quantiles = plot_quantiles, alphas = quantile_alphas, relevant_period = relevant_period)

ggplotly(the_plot, tooltip = "text", height = 5000, width = 1600) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```
