---
title: Evaluation of Hospitalization Forecasters 2024-2025
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
params:
  forecaster_set: 1
editor_options:
  chunk_output_type: console
---

$$\\[.4in]$$

```{r echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
library(aws.s3)
library(data.table)
library(dplyr)
library(DT)
library(ggplot2)
library(plotly)
library(readr)
library(stringr)
library(tidyr)
```

```{r}
# params <- list(
#     forecaster_set = 5
# )
# Load the table of parameter combinations
s3load("flu_2023_forecaster_parameter_combinations.rds", bucket = "forecasting-team-data")

# Select forecasters for this notebook
cmu_forecasters <- forecaster_parameter_combinations_[[params$forecaster_set]]$id
outside_forecasters <- c("FluSight-baseline", "FluSight-ensemble")

# Load scores and filter them, get global variables
s3load(object = "flu_2023_joined_scores.rds", bucket = "forecasting-team-data")
scores <- flu_2023_joined_scores %>%
  filter(forecaster %in% c(cmu_forecasters, outside_forecasters))
forecast_dates <- scores %>%
  pull(forecast_date) %>%
  unique()
forecasters <- c(cmu_forecasters, outside_forecasters)
aheads <- scores %>%
  pull(ahead) %>%
  unique()
forecaster_family <- forecaster_parameter_combinations_[[params$forecaster_set]]$forecaster %>% unique()
base_forecaster_name <- "FluSight-baseline"

# Define aggregation functions
Mean <- function(x) mean(x, na.rm = TRUE)
GeoMean <- function(x, offset = 0) exp(Mean(log(x + offset)))
```

### Notebook Information

#### Forecaster Family

`r forecaster_family`

#### Target Dates

`r forecast_dates`

#### Forecaster Parameter Mapping and Overall Scores

The table is sorted by descending WIS and contains all the forecasters in this notebook.

```{r}
# Display the table
param_table <- forecaster_parameter_combinations_[[params$forecaster_set]] %>%
  select(-any_of(c("forecaster", "keys_to_ignore", "pop_scaling"))) %>%
  {
    if ("n_training" %in% colnames(.)) {
      (.) %>% mutate(n_training = as.character(n_training))
    } else {
      .
    }
  } %>%
  full_join(
    scores %>%
      group_by(forecaster) %>%
      summarize(
        mean_wis = round(mean(wis, na.rm = TRUE), 2),
        mean_ae = round(mean(ae, na.rm = TRUE), 2),
        mean_coverage_80 = round(mean(coverage_80, na.rm = TRUE), 2)
      ) %>%
      rename(
        id = forecaster
      )
  ) %>%
  arrange(mean_wis)
datatable(param_table)
```

$$\\[.07in]$$

### Score Plots {.tabset}

- The WIS plots are relative to the FluSight-baseline (which isn't plotted, but is implicitly the 1 line). These plots are aggregated with the geometric mean.
- The absolute error plots are also relative to the FluSight-baseline, but are aggregated with the arithmetic mean.
- The dashed black line in all plots is the FluSight-ensemble.

#### WIS by Forecast Date

```{r}
var <- "wis"
group_cols <- c("forecaster", "forecast_date", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := GeoMean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (
  df %>%
    filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>%
    nrow() > 0
) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

facets.label <- str_glue("{aheads} days ahead")
names(facets.label) <- aheads
subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df %>% filter(forecaster != "FluSight-ensemble"),
  aes(x = forecast_date, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_line(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = forecast_date, y = !!sym(var)),
    color = "black", linetype = 2
  ) +
  geom_point(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = forecast_date, y = !!sym(var)),
    color = "black", shape = 21, fill = "white"
  ) +
  facet_grid(rows = vars(ahead)) +
  facet_wrap(~ahead, nrow = 4, labeller = labeller(ahead = facets.label)) +
  scale_y_log10() +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Forecast Dates", y = "Geometric Mean WIS")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### WIS by Ahead

```{r}
var <- "wis"
group_cols <- c("forecaster", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := GeoMean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (df %>% filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>% nrow() > 0) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df %>% filter(forecaster != "FluSight-ensemble"),
  aes(x = ahead, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_line(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = ahead, y = !!sym(var)),
    color = "black", linetype = 2
  ) +
  geom_point(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = ahead, y = !!sym(var)),
    color = "black", shape = 21, fill = "white"
  ) +
  scale_y_log10() +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Days ahead", y = "Geometric Mean WIS")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### Absolute Error by Forecast Date

```{r}
var <- "ae"
group_cols <- c("forecaster", "forecast_date", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (
  df %>%
    filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>%
    nrow() > 0
) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

facets.label <- str_glue("{aheads} days ahead")
names(facets.label) <- aheads
subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df %>% filter(forecaster != "FluSight-ensemble"),
  aes(x = forecast_date, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_line(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = forecast_date, y = !!sym(var)),
    color = "black", linetype = 2
  ) +
  geom_point(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = forecast_date, y = !!sym(var)),
    color = "black", shape = 21, fill = "white"
  ) +
  facet_grid(rows = vars(ahead)) +
  facet_wrap(~ahead, nrow = 4, labeller = labeller(ahead = facets.label)) +
  scale_y_log10() +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Forecast Dates", y = "Geometric Mean WIS")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### Absolute Error by Ahead

```{r}
var <- "ae"
group_cols <- c("forecaster", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (df %>% filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>% nrow() > 0) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df %>% filter(forecaster != "FluSight-ensemble"),
  aes(x = ahead, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_line(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = ahead, y = !!sym(var)),
    color = "black", linetype = 2
  ) +
  geom_point(
    data = normalized_df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = ahead, y = !!sym(var)),
    color = "black", shape = 21, fill = "white"
  ) +
  scale_y_log10() +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Days ahead", y = "Geometric Mean WIS")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### % Coverage by Forecast Date

```{r}
var <- "coverage_80"
group_cols <- c("forecaster", "forecast_date", "ahead")

# Aggregate metric across groups
df <- scores %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

facets.label <- str_glue("{aheads} days ahead")
names(facets.label) <- aheads
subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  df %>% filter(forecaster != "FluSight-ensemble"),
  aes(x = forecast_date, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_line(
    data = df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = forecast_date, y = !!sym(var)),
    color = "black", linetype = 2
  ) +
  geom_point(
    data = df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = forecast_date, y = !!sym(var)),
    color = "black", shape = 21, fill = "white"
  ) +
  geom_hline(yintercept = .8, linetype = 1, color = "black") +
  facet_grid(rows = vars(ahead)) +
  facet_wrap(~ahead, nrow = 4, labeller = labeller(ahead = facets.label)) +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Forecast Dates", y = "Mean 80% Coverage")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### % Coverage by Ahead

```{r}
var <- "coverage_80"
id_cols <- c("forecaster", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(id_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(id_cols)) %>%
  filter(ahead >= 0)

subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  df %>% filter(forecaster != "FluSight-ensemble"),
  aes(x = ahead, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_line(
    data = df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = ahead, y = !!sym(var)),
    color = "black", linetype = 2
  ) +
  geom_point(
    data = df %>% filter(forecaster == "FluSight-ensemble"),
    aes(x = ahead, y = !!sym(var)),
    color = "black", shape = 21, fill = "white"
  ) +
  geom_hline(yintercept = .8, linetype = 1, color = "black") +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Days ahead", y = "Mean 80% Coverage")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

### Fan plots

Fan plots showing the 80% prediction intervals for the forecasts made by the CMU forecasters and the outside forecasters. The black line is the truth data.

```{r}
# Load the forecasts and the truth data
s3load(object = "flu_2023_joined_forecasts.rds", bucket = "forecasting-team-data")
s3load(object = "flu_2023_truth_data.rds", bucket = "forecasting-team-data")

# We plot a subset of the dates and geos for the fan plot
plot_dates <- seq.Date(as.Date("2023-10-07"), by = "4 weeks", length.out = 8)
geo_vals <- c("ca", "fl", "pa", "tx")
forecast_subset <- flu_2023_joined_forecasts %>%
  filter(
    forecaster %in% c(cmu_forecasters, outside_forecasters),
    geo_value %in% geo_vals,
    forecast_date %in% plot_dates
  ) %>%
  mutate(quantile = as.character(quantile)) %>%
  pivot_wider(names_from = "quantile", values_from = "prediction") %>%
  mutate(ahead = as.numeric(target_end_date - forecast_date)) %>%
  inner_join(
    param_table %>% rename(forecaster = id) %>% select(forecaster, mean_wis)
  ) %>%
  arrange(mean_wis)

p <- ggplot(
  data = forecast_subset,
  aes(x = target_end_date, group = forecast_date)
) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = forecast_date), alpha = 0.3) +
  geom_line(aes(y = `0.5`, color = forecast_date), linetype = 2L) +
  geom_point(aes(y = `0.5`, color = forecast_date), size = 0.75) +
  geom_line(
    data = flu_2023_truth_data %>% filter(geo_value %in% geo_vals),
    aes(x = target_end_date, y = true_value),
    inherit.aes = FALSE, na.rm = TRUE,
    color = "black", linetype = 1
  ) +
  facet_grid(forecaster ~ geo_value) +
  labs(x = "Reference Date", y = "Forecasts")

ggplotly(p, tooltip = "text", height = 3000, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```
