---
title: "Prod/Explore Comparison: `r params$disease`"
date: "Rendered: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    code_folding: hide
    toc: true
params:
  scores: ""
  forecasts: ""
  truth_data: ""
  asof_data: ""
  disease: ""
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
body { display: block; max-width: 1280px !important; margin-left: auto; margin-right: auto; }
body .main-container { max-width: 1280px !important; width: 1280px !important; }
```

```{r echo=FALSE}
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE, cache = FALSE)
ggplot2::theme_set(ggplot2::theme_bw())
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
```

```{r setup}
if (params$disease == "flu") {
  base_forecaster_name <- "FluSight-baseline"
} else {
  base_forecaster_name <- "CovidHub-baseline"
}

Mean <- function(x) mean(x, na.rm = TRUE)
GeoMean <- function(x, offset = 0) exp(Mean(log(x + offset)))

scores <- params$scores %>%
  filter(ahead >= 0) %>%
  left_join(select(epidatasets::state_census, abbr, pop), by = join_by(geo_value == abbr)) %>%
  mutate(wis_rate = wis * 1e5 / pop)

forecast_dates <- sort(unique(scores$forecast_date))
```

Forecasts made `r format(min(forecast_dates), "%Y-%m-%d")` to `r format(max(forecast_dates), "%Y-%m-%d")`.

Scores relative to `r base_forecaster_name` (rel < 1 = better than baseline).
Prod models shown with their actual IDs; explore models show best-in-class per family.

```{r score_functions}
scores_by_state <- function(sc) {
  sc %>%
    group_by(forecaster, geo_value) %>%
    summarize(mean_wis = Mean(wis), mean_wis_rate = Mean(wis_rate), .groups = "drop")
}

se_mean_states <- function(sc) {
  scores_by_state(sc) %>%
    group_by(forecaster) %>%
    summarize(
      se_mean_wis = round(sd(mean_wis, na.rm = TRUE) / sqrt(n()), 2),
      se_mean_wis_rate = round(sd(mean_wis_rate, na.rm = TRUE) / sqrt(n()), 2),
      .groups = "drop"
    )
}

score_summary <- function(sc) {
  baseline_wis <- sc %>% filter(forecaster == base_forecaster_name) %>% summarize(b = Mean(wis)) %>% pull(b)
  baseline_wis_rate <- sc %>% filter(forecaster == base_forecaster_name) %>% summarize(b = Mean(wis_rate)) %>% pull(b)
  baseline_geomean <- sc %>%
    filter(forecaster == base_forecaster_name) %>%
    summarize(b = GeoMean(wis, min(wis[wis > 1e-5]))) %>%
    pull(b)
  sc %>%
    group_by(forecaster) %>%
    summarize(
      mean_wis = round(Mean(wis), 2),
      mean_wis_rate = round(Mean(wis_rate), 2),
      geomean_wis = round(GeoMean(wis, min(wis[wis > 1e-5])), 2),
      mean_cov_50 = round(Mean(coverage_50), 2),
      mean_cov_90 = round(Mean(coverage_90), 2),
      n = n(),
      .groups = "drop"
    ) %>%
    left_join(se_mean_states(sc), by = "forecaster") %>%
    mutate(
      rel_wis = round(mean_wis / baseline_wis, 3),
      rel_wis_rate = round(mean_wis_rate / baseline_wis_rate, 3),
      rel_geomean_wis = round(geomean_wis / baseline_geomean, 3)
    ) %>%
    arrange(rel_geomean_wis) %>%
    select(forecaster, geomean_wis, rel_geomean_wis, mean_wis, se_mean_wis, rel_wis,
           mean_wis_rate, se_mean_wis_rate, rel_wis_rate, mean_cov_50, mean_cov_90, n)
}

datatable_function <- function(summary_df) {
  DT::datatable(
    summary_df,
    fillContainer = FALSE,
    options = list(
      initComplete = htmlwidgets::JS(
        "function(settings, json) {",
        "$(this.api().table().container()).css({'font-size': '10pt'});",
        "}"
      ),
      pageLength = 30
    )
  ) %>%
    DT::formatStyle("forecaster", target = "cell",
      textDecoration = DT::styleEqual("CMU-TimeSeries", "underline")) %>%
    DT::formatStyle("geomean_wis",
      background = DT::styleColorBar(c(0, max(summary_df$geomean_wis, na.rm = TRUE)), "lightblue"),
      backgroundSize = "98% 88%", backgroundRepeat = "no-repeat", backgroundPosition = "center") %>%
    DT::formatStyle("mean_wis",
      background = DT::styleColorBar(c(0, max(summary_df$mean_wis, na.rm = TRUE)), "lightblue"),
      backgroundSize = "98% 88%", backgroundRepeat = "no-repeat", backgroundPosition = "center") %>%
    DT::formatStyle("mean_wis_rate",
      background = DT::styleColorBar(c(0, max(summary_df$mean_wis_rate, na.rm = TRUE)), "lightblue"),
      backgroundSize = "98% 88%", backgroundRepeat = "no-repeat", backgroundPosition = "center")
}
```

# Overall summary

```{r overall_table}
summary_df <- score_summary(scores)
forecasters_ordered <- summary_df$forecaster
datatable_function(summary_df)
```

# Scores by ahead {.tabset}

```{r ahead_prep}
baseline_by_ahead <- scores %>%
  filter(forecaster == base_forecaster_name) %>%
  summarize(baseline_wis = Mean(wis), baseline_ae = Mean(ae), .by = ahead)
```

## WIS relative to baseline

```{r wis_by_ahead, fig.height=5, fig.width=8}
plot_df <- scores %>%
  summarize(mean_wis = Mean(wis), .by = c(forecaster, ahead)) %>%
  left_join(baseline_by_ahead %>% select(ahead, baseline_wis), by = "ahead") %>%
  mutate(rel_wis = mean_wis / baseline_wis) %>%
  filter(forecaster != base_forecaster_name, !is.na(rel_wis))

ggplot(plot_df, aes(x = ahead, y = rel_wis, color = forecaster)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(n.breaks = 10) +
  labs(x = "Days ahead", y = "Mean WIS / Baseline WIS") +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, title = NULL))
```

## AE relative to baseline

```{r ae_by_ahead, fig.height=5, fig.width=8}
plot_df <- scores %>%
  summarize(mean_ae = Mean(ae), .by = c(forecaster, ahead)) %>%
  left_join(baseline_by_ahead %>% select(ahead, baseline_ae), by = "ahead") %>%
  mutate(rel_ae = mean_ae / baseline_ae) %>%
  filter(forecaster != base_forecaster_name, !is.na(rel_ae))

ggplot(plot_df, aes(x = ahead, y = rel_ae, color = forecaster)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(n.breaks = 10) +
  labs(x = "Days ahead", y = "Mean AE / Baseline AE") +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, title = NULL))
```

# Calibration curves

For each nominal quantile level q, the empirical fraction of observations where the true value fell below the predicted q-th quantile.
A perfectly calibrated forecaster lies on the diagonal.

```{r calibration, fig.height=5, fig.width=8}
calibration <- params$forecasts %>%
  filter(forecaster %in% forecasters_ordered) %>%
  left_join(
    params$truth_data %>% select(geo_value, target_end_date, true_value),
    by = c("geo_value", "target_end_date")
  ) %>%
  drop_na(true_value) %>%
  group_by(forecaster, quantile) %>%
  summarize(empirical = mean(true_value <= value, na.rm = TRUE), .groups = "drop")

ggplot(calibration, aes(x = quantile, y = empirical, color = forecaster)) +
  geom_line() + geom_point(size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Nominal quantile", y = "Empirical coverage") +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, title = NULL))
```

# Forecast coverage

Number of distinct aheads forecasted per model per week.
Expected maximum is 5 (aheads 0–4 weeks).

```{r coverage_heatmap, fig.height=max(2, length(forecasters_ordered)*0.4 + 1), fig.width=12}
coverage_df <- params$forecasts %>%
  filter(forecaster %in% forecasters_ordered) %>%
  mutate(ahead = as.integer(target_end_date - forecast_date)) %>%
  summarize(n_aheads = n_distinct(ahead), .by = c(forecaster, forecast_date))

ggplot(coverage_df, aes(x = forecast_date, y = factor(forecaster, levels = rev(forecasters_ordered)), fill = n_aheads)) +
  geom_tile() +
  scale_fill_viridis_c(name = "# aheads", limits = c(0, 5), breaks = 0:5) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Fan plots {.tabset}

80% prediction intervals (0.1–0.9 quantiles) for a subset of forecast dates. Black line is finalized data; coloured lines show data available at each forecast date.

```{r fan_setup}
available_fc_dates <- sort(unique(params$forecasts$forecast_date))
plot_dates <- if (length(forecast_dates) > 0) {
  intersect(
    forecast_dates[seq(1, length(forecast_dates), by = 4)],
    available_fc_dates
  ) %>% as.Date(origin = "1970-01-01")
} else {
  as.Date(character(0))
}

fan_plot <- function(geo_vals) {
  if (length(plot_dates) == 0 || length(forecasters_ordered) == 0) return(invisible(NULL))
  fc_sub <- params$forecasts %>%
    filter(forecaster %in% forecasters_ordered, geo_value %in% geo_vals, forecast_date %in% plot_dates) %>%
    select(forecaster, geo_value, forecast_date, target_end_date, quantile, value) %>%
    mutate(
      quantile = as.character(quantile),
      date_label = factor(format(forecast_date, "%b %d"), levels = format(sort(unique(forecast_date)), "%b %d"))
    ) %>%
    pivot_wider(names_from = quantile, values_from = value)
  if (nrow(fc_sub) == 0) {
    message("No forecast data for these states in this season.")
    return(invisible(NULL))
  }
  truth_sub <- params$truth_data %>%
    filter(
      geo_value %in% geo_vals,
      target_end_date >= min(plot_dates) - 12 * 7,
      target_end_date <= max(plot_dates) + 5 * 7
    )
  asof_sub <- if (is.data.frame(params$asof_data) && nrow(params$asof_data) > 0) {
    params$asof_data %>%
      filter(
        geo_value %in% geo_vals,
        forecast_date %in% plot_dates,
        time_value >= min(plot_dates) - 12 * 7,
        time_value <= max(plot_dates)
      ) %>%
      mutate(date_label = factor(
        format(forecast_date, "%b %d"),
        levels = levels(fc_sub$date_label)
      ))
  } else {
    tibble::tibble()
  }
  n_dates <- length(unique(fc_sub$date_label))
  ggplot(fc_sub, aes(x = target_end_date, group = date_label)) +
    geom_ribbon(aes(ymin = `0.1`, ymax = `0.9`, fill = date_label), alpha = 0.3) +
    geom_line(aes(y = `0.5`, color = date_label), linetype = 2L) +
    geom_point(aes(y = `0.5`, color = date_label), size = 0.75) +
    { if (nrow(asof_sub) > 0)
        geom_line(data = asof_sub, aes(x = time_value, y = value, color = date_label, group = date_label),
                  alpha = 0.6, linewidth = 0.5)
    } +
    geom_line(data = truth_sub, aes(x = target_end_date, y = true_value), inherit.aes = FALSE, color = "black", linewidth = 0.4) +
    scale_fill_viridis_d(name = "forecast date", direction = -1) +
    scale_color_viridis_d(name = "forecast date", direction = -1) +
    scale_y_continuous(n.breaks = 6) +
    facet_grid(factor(forecaster, levels = forecasters_ordered) ~ geo_value, scales = "free_y") +
    labs(x = NULL, y = "Weekly hospitalizations") +
    theme(
      legend.position = "bottom",
      strip.text.y = element_text(size = 7)
    ) +
    guides(fill = guide_legend(nrow = ceiling(n_dates / 5), title.position = "top"),
           color = guide_legend(nrow = ceiling(n_dates / 5), title.position = "top"))
}
```

## Large states

```{r fan_large, fig.height=max(2, length(forecasters_ordered)*2), fig.width=10}
fan_plot(c("ca", "fl", "pa", "tx", "ny"))
```

## Medium states

```{r fan_medium, fig.height=max(2, length(forecasters_ordered)*2), fig.width=10}
fan_plot(c("mn", "co", "ky", "md", "or"))
```

## Small states

```{r fan_small, fig.height=max(2, length(forecasters_ordered)*2), fig.width=10}
fan_plot(c("vt", "de", "ri", "nm", "mt"))
```
