---
title: Baseline Forecaster Fan Plots 2024-2025
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
params:
  disease: "flu"
editor_options:
  chunk_output_type: console
---

$$\\[.4in]$$

```{r echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

Simple model:

- For each geo, replace zero weeks with NA.
- For each geo, regress the last few weeks (population scaled) on the week number (possibly log transformed).
- Any location that had fewer than 2 recent non-zero weeks gets the pop-scaled median (or mean) as their point forecast.
- Use pooled residuals to get PIs for everything and then resample for future aheads.
- Get back to the original scale.

# Linear Extrapolation Model for `r params$disease` {.tabset}

```{r}
params <- list(disease = "flu")
source("R/load_all.R")
library(ggplot2)
library(plotly)

if (params$disease == "flu") {
  epi_data <- readr::read_csv("https://data.cdc.gov/resource/ua7e-t2fy.csv?$limit=20000&$select=weekendingdate,jurisdiction,totalconfflunewadm", show_col_types = FALSE) %>%
    rename(time_value = weekendingdate, value = totalconfflunewadm, geo_value = jurisdiction)
} else if (params$disease == "covid") {
  epi_data <- readr::read_csv("https://data.cdc.gov/resource/ua7e-t2fy.csv?$limit=20000&$select=weekendingdate,jurisdiction,totalconfc19newadm", show_col_types = FALSE) %>%
    rename(time_value = weekendingdate, value = totalconfc19newadm, geo_value = jurisdiction)
} else {
  stop("Invalid disease")
}
today <- Sys.Date()
epi_data <- epi_data %>%
  filter(geo_value %nin% c("AS", "USA", "VI", "PR", "MP", "GU")) %>%
  mutate(time_value = as.Date(time_value), geo_value = tolower(geo_value)) %>%
  mutate(
    target_end_date = time_value,
    value = ifelse(value == 0, NA, value)
  ) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df(as_of = today)

aheads <- -1:3
forecast_date <- today
subset_geos <- epi_data$geo_value %>% unique()
subset_geos <- c("ca", "ny", "fl", "tx", "pa")
```

## Fit On Raw Data

```{r}
quantile_forecast <- map(aheads, ~ forecaster_baseline_linear(epi_data, .x, log = FALSE)) %>%
  bind_rows() %>%
  filter(geo_value %in% subset_geos) %>%
  mutate(forecaster = "baseline_linear")
truth_data <- epi_data %>%
  filter(geo_value %in% subset_geos, time_value > forecast_date - 365 * 2)
train_fit <- quantile_forecast %>%
  filter(is.na(quantile)) %>%
  mutate(time_value = target_end_date)
quantile_forecast <- quantile_forecast %>% filter(!is.na(quantile))
train_data <- epi_data %>%
  filter(geo_value %in% subset_geos, time_value > forecast_date - 30) %>%
  summarize(start = min(time_value), stop = max(time_value))

# Bad forecast filters
filter_geos <- filter_forecast_geos(quantile_forecast, truth_data)
```

Filter geo recommendations: `r paste(filter_geos, collapse = ", ")`.

(Based on throwing out forecasts that trend down, forecasts where the median exceeds all prior peaks, and forecasts where the 75th percentile exceeds all prior peaks at 2 ahead.)

```{r}
p <- plot_forecasts(quantile_forecast, forecast_date, truth_data = truth_data, relevant_period = train_data) +
  geom_point(data = train_fit, mapping = aes(x = time_value, y = value), size = 0.25)

ggplotly(p, tooltip = "text", height = 5000, width = 1700)
```

## Fit On Log Data

```{r}
quantile_forecast <- map(aheads, ~ forecaster_baseline_linear(epi_data, .x, log = TRUE)) %>%
  bind_rows() %>%
  filter(geo_value %in% subset_geos) %>%
  mutate(forecaster = "baseline_linear")
truth_data <- epi_data %>%
  filter(geo_value %in% subset_geos, time_value > forecast_date - 365 * 2)
train_fit <- quantile_forecast %>%
  filter(is.na(quantile)) %>%
  mutate(time_value = target_end_date)
quantile_forecast <- quantile_forecast %>% filter(!is.na(quantile))
train_data <- epi_data %>%
  filter(geo_value %in% subset_geos, time_value > forecast_date - 30) %>%
  summarize(start = min(time_value), stop = max(time_value))

filter_geos <- filter_forecast_geos(quantile_forecast, truth_data)
```

Filter geo recommendations: `r paste(filter_geos, collapse = ", ")`.

(Based on throwing out forecasts that trend down, forecasts where the median exceeds all prior peaks, and forecasts where the 75th percentile exceeds all prior peaks at 2 ahead.)

```{r}
p <- plot_forecasts(quantile_forecast, forecast_date, truth_data = truth_data, relevant_period = train_data) +
  geom_point(data = train_fit, mapping = aes(x = time_value, y = value), size = 0.25)

ggplotly(p, tooltip = "text", height = 5000, width = 1700)
```
