---
title: "Baseline Nowcast Evaluation: COVID NHSN"
date: "Rendered: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
format:
  html:
    code-fold: true
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup}
#| message: false
#| warning: false
knitr::opts_chunk$set(fig.align = "center", warning = FALSE, message = FALSE, cache = TRUE)
ggplot2::theme_set(ggplot2::theme_bw())

library(baselinenowcast)
library(scoringutils)
library(furrr)
source(here::here("R/load_all.R"))
```

It may be worth noting that they have literally already [done some nowcasting for nssp](https://baselinenowcast.epinowcast.org/articles/nssp_nowcast.html)! They do seem to be using line-list data, so it's a different nssp source than the one we typically forecast.

TODO: may also be worth checking out [NoBS](https://cran.r-project.org/web//packages/NobBS/NobBS.pdf) and/or [epi-nowcast](https://www.epinowcast.org/)

## Data

Load the NHSN COVID archive (cached to avoid repeat API calls).

```{r load-data}
nhsn_archive <- get_nhsn_data_archive("covid")

align_version_to_wednesday <- function(dt) {
  dt |>
    as_tibble() |>
    mutate(
      version_wed = lubridate::floor_date(version, "week", week_start = 3L),
      is_wednesday = lubridate::wday(version, week_start = 1L) == 3L
    ) |>
    group_by(geo_value, time_value, version_wed) |>
    arrange(desc(is_wednesday)) |>
    slice(1) |>
    ungroup() |>
    mutate(version = version_wed) |>
    select(-version_wed, -is_wednesday)
}
nhsn_archive <- nhsn_archive$DT %>%
  align_version_to_wednesday() %>%
  mutate(time_value = lubridate::floor_date(time_value, "week", week_start = 3L)) %>%
  distinct(geo_value, time_value, version, .keep_all = TRUE) %>%
  as_epi_archive(compactify = TRUE)

cat("Archive spans", format(min(nhsn_archive$DT$time_value)), "to",
    format(max(nhsn_archive$DT$time_value)), "\n")
cat("Version range:", format(min(nhsn_archive$DT$version)), "to",
    format(max(nhsn_archive$DT$version)), "\n")
cat("Geos:", length(unique(nhsn_archive$DT$geo_value)), "\n")
```

## Data Transformation

The archive stores `value` as the revised count for a reference week at each `version` date. `baselinenowcast` expects a reporting triangle where each cell is the **incremental** count added at each delay step. We compute that as the first-difference within each (geo, time_value) group.

```{r transform-fn}
#' Align archive versions onto Wednesdays using ceiling_date.
#'
#' Within each (geo_value, time_value, ceiled version) group, prefers the row
#' whose version was already a Wednesday; otherwise keeps the first.

#' Convert a subset of the NHSN epi_archive to reporting triangle long format.
#'
#' Aligns versions to Wednesdays, filters to a single geo and all versions up
#' to nowcast_date, then computes incremental counts per reference week.
incremental_data <- nhsn_archive$DT |>
  filter(time_value >= min(version)) |>
  group_by(geo_value, time_value) |>
  arrange(geo_value, time_value, version) |>
  mutate(count = as.integer(value - lag(value, default = 0L))) |>
  ungroup() |>
  rename(reference_date = time_value, report_date = version) |>
  select(geo_value, reference_date, report_date, count)
make_reporting_long <- function(archive_dt, geo, nowcast_date) {
  archive_dt |>
    filter(geo_value == geo, version <= nowcast_date) |>
    filter(time_value >= min(version)) |>
    arrange(time_value, version) |>
    group_by(time_value) |>
    mutate(count = as.integer(value - lag(value, default = 0L))) |>
    ungroup() |>
    rename(reference_date = time_value, report_date = version) |>
    select(reference_date, report_date, count)
}
```

### Explore the reporting structure

Before nowcasting, inspect the revision behaviour for national data using `epiprocess::revision_analysis`.

```{r explore-revisions}
national_archive <- nhsn_archive$DT |>
  filter(geo_value == "us") |>
  as_epi_archive(compactify = TRUE)
national_archive$time_type <- "day"

revision_sum <- national_archive |>
  revision_analysis(value, min_waiting_period = NULL)

print(revision_sum, quick_revision = as.difftime(7, units = "days"))
```

```{r explore-revision-plots}
#| fig-width: 11
#| fig-height: 8
rev_tib <- revision_sum$revision_behavior |>
  mutate(lag_near_latest_days = as.numeric(lag_near_latest, units = "days")) |> filter(time_value >= min(nhsn_archive$DT$version))

p1 <- ggplot(rev_tib, aes(time_value, rel_spread)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(se = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Relative spread over time",
       subtitle = "spread / max_value per reference week",
       x = NULL, y = "Relative spread")

p2 <- ggplot(rev_tib, aes(n_revisions)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of revision counts",
       x = "Number of revisions", y = "Count")

p3 <- ggplot(rev_tib, aes(lag_near_latest_days)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Days until value settles within 20% of final",
       x = "lag_near_latest (days)", y = "Count")

cowplot::plot_grid(p1, p2, p3, ncol = 1)
```


```{r geos_by_latency}
#| fig-width: 11
#| fig-height: 8
recent_archive <- nhsn_archive
recent_archive$time_type <- "day"
revision_sum <- recent_archive |>
  epiprocess::revision_analysis(value, min_waiting_period = NULL)

revision_sum$revision_behavior |>
  arrange(desc(rel_spread)) |>
  filter(
    geo_value %nin% c("vi", "as", "gu"),
    n_revisions > 1,
    time_value > "2025-05-01") |>
  print(n=30)
```

## Single Nowcast Example

Run a single nowcast for the national level at a specific date to verify the workflow end-to-end.

```{r single-nowcast}
example_nowcast_date <- as.Date("2026-02-11")
geo <- "us"
max_delay_weeks <- 4L # 4 weeks

example_long <- incremental_data |>
  filter(report_date <= example_nowcast_date, geo_value == geo) |>
  select(-geo_value)

rep_tri <- example_long |>
  as_reporting_triangle(delays_unit="weeks") |>
  truncate_to_quantile(0.5)

summary(rep_tri)
```

```{r single-nowcast-run}
set.seed(42)
nowcast_draws <- baselinenowcast(
  rep_tri,
  scale_factor = 3,
  prop_delay = 0.5,
  draws = 500
)

nowcast_draws
```

```{r single-nowcast-plot}
#| fig-width: 10
#| fig-height: 5
# Final values (most recent version per reference week)
final_values <- nhsn_archive |>
  epix_as_of_current() |>
  filter(geo_value == geo) |>
  select(reference_date = time_value, final_value = value)

nowcast_summary <- nowcast_draws |>
  group_by(reference_date) |>
  summarise(
    median = median(pred_count),
    q05 = quantile(pred_count, 0.05),
    q25 = quantile(pred_count, 0.25),
    q75 = quantile(pred_count, 0.75),
    q95 = quantile(pred_count, 0.95)
  ) |>
  left_join(final_values, by = "reference_date")

# Latest reported (as of nowcast_date) for comparison
as_of_values <- nhsn_archive |>
  epix_as_of(example_nowcast_date) |>
  filter(geo_value == geo) |>
  select(reference_date = time_value, as_of_value = value)

plot_data <- nowcast_summary |>
  left_join(as_of_values, by = "reference_date")
## first_plot_date <- plot_data |> filter((final_value - as_of_value) / pmax(final_value, as_of_value) >0.01) |> pull(reference_date) |> min()
## plot_data <- plot_data |> filter(reference_date > first_plot_date)

ggplot(plot_data, aes(reference_date)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median, colour = "Nowcast median")) +
  geom_point(aes(y = final_value, colour = "Final value")) +
  geom_point(aes(y = as_of_value, colour = "As-of value"), shape = 1) +
  geom_vline(xintercept = example_nowcast_date, linetype = "dashed", alpha = 0.5) +
  scale_colour_manual(values = c(
    "Nowcast median" = "steelblue",
    "Final value" = "black",
    "As-of value" = "tomato"
  )) +
  labs(
    title = glue::glue("Baseline nowcast as of {example_nowcast_date} (national)"),
    subtitle = "90% and 50% CI shaded; dashed line = nowcast date",
    x = "Reference week", y = "COVID hospitalisations", colour = NULL
  )
```

So this has the unfortunate potential of royally messing up your data by predicting medians that are Just Wrong. Focusing just on the tail where ther's at least a 1% difference between the 5th and 95th quantiles:


```{r plotting-latest-only}
first_plot_date <- plot_data |> filter(q95 > q05) |> pull(reference_date) |> min()
plot_data <- plot_data |> filter(reference_date > first_plot_date - 7)

ggplot(plot_data, aes(reference_date)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median, colour = "Nowcast median")) +
  geom_point(aes(y = final_value, colour = "Final value")) +
  geom_point(aes(y = as_of_value, colour = "As-of value"), shape = 1) +
  geom_vline(xintercept = example_nowcast_date, linetype = "dashed", alpha = 0.5) +
  scale_colour_manual(values = c(
    "Nowcast median" = "steelblue",
    "Final value" = "black",
    "As-of value" = "tomato"
  )) +
  labs(
    title = glue::glue("Baseline nowcast as of {example_nowcast_date} (national)"),
    subtitle = "90% and 50% CI shaded; dashed line = nowcast date",
    x = "Reference week", y = "COVID hospitalisations", colour = NULL
  )
```
### Triangle Plot

```{r triangle-plot}
#| fig-width: 10
#| fig-height: 5
# Final values (most recent version per reference week)
triangle_df <- as.data.frame(rep_tri) |>
  mutate(time = as.numeric(factor(reference_date)))

plot_triangle <- ggplot(
  triangle_df,
  aes(x = delay, y = time, fill = count)
) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Reporting triangle", x = "Delay", y = "Time") +
  theme_bw() +
  scale_y_reverse()
```

```{r show-triangle-plot}
plot_triangle
```

## Retrospective Evaluation

Evaluate nowcast accuracy across multiple past dates by iterating over a sequence of nowcast dates.

```{r retro-setup}
retro_nowcast_dates <- seq(
  from = min(nhsn_archive$DT$version),
  to = max(nhsn_archive$DT$version),
  by = "1 week"
)

# Only keep dates where we have enough archive history
min_version <- min(nhsn_archive$DT$version)
retro_nowcast_dates <- retro_nowcast_dates[retro_nowcast_dates >= min_version + 35 * 7]

cat("Running retrospective nowcasts for", length(retro_nowcast_dates), "dates\n")
```

```{r retro-run}
#| cache: true
#| dependson: ["retro-setup", "transform-fn", "load-data"]
nowcast_date <- as.Date("2026-02-04")
geo <- "ak"
run_retro_nowcast <- function(nowcast_date, incremental_data, geo,
                              truncate_quantile = 0.5, draws = 500L, eval_window_weeks = 7L) {
  long_data <- incremental_data |>
    filter(geo_value == geo, report_date <= nowcast_date) |>
    select(-geo_value)

  has_negatives <- any(long_data$count < 0)

  rep_tri_trunc <- tryCatch(
    suppressMessages(
      as_reporting_triangle(long_data, delays_unit = "weeks") |>
        truncate_to_quantile(truncate_quantile)
    ),
    error = function(e) {NULL}
  )
  if (is.null(rep_tri_trunc) || nrow(as.data.frame(rep_tri_trunc)) == 0) return(NULL)

  set.seed(as.integer(nowcast_date))
  draws_df <- tryCatch(
    baselinenowcast(rep_tri_trunc, scale_factor = 3, prop_delay = 0.5, draws = draws),
    error = function(e) {NULL})
  if (is.null(draws_df)) return(NULL)

  draws_df |>
    filter(nowcast_date - reference_date < eval_window_weeks * 7) |>
    group_by(reference_date) |>
    summarise(
      median = median(pred_count),
      q05 = quantile(pred_count, 0.05),
      q25 = quantile(pred_count, 0.25),
      q75 = quantile(pred_count, 0.75),
      q95 = quantile(pred_count, 0.95)
    ) |>
    mutate(nowcast_date = nowcast_date, geo_value = geo, has_negatives = has_negatives)
}


geo_values <- unique(nhsn_archive$DT$geo_value)

retro_grid <- tidyr::expand_grid(
  nowcast_date = retro_nowcast_dates,
  geo = geo_values
)

plan(multisession, workers = parallel::detectCores() - 1)
retro_nowcasts <- furrr::future_pmap(
  retro_grid,
  run_retro_nowcast,
  incremental_data = incremental_data,
  .options = furrr_options(seed = TRUE)
) |>
  purrr::compact() |>
  bind_rows()
plan(sequential)

cat("Total nowcast dates to evaluate:", nrow(retro_nowcasts), "\n")
cat("Reference dates covered:", n_distinct(retro_nowcasts$reference_date), "\n")
```

```{r retro-score}
#| cache: true
#| dependson: ["retro-run", "load-data"]
final_values_retro <- nhsn_archive |>
  epix_as_of_current() |>
  select(geo_value, reference_date = time_value, final_value = value)

calc_naive_wis <- function(as_of_date) {
  nhsn_archive |>
    epix_as_of(as_of_date) |>
    rename(reference_date = time_value, as_of_value = value) |>
    left_join(final_values_retro, by = join_by(geo_value, reference_date)) |>
    mutate(nowcast_date = as_of_date, naive_wis = abs(as_of_value - final_value)) |>
    select(geo_value, nowcast_date, reference_date, naive_wis) |>
    filter(naive_wis > 0)
}
# Naive WIS = absolute error of the as-of value (WIS of a point forecast)
naive_wis <- map(retro_nowcasts$nowcast_date |> unique(), calc_naive_wis) |> list_rbind()

scores <- retro_nowcasts |>
  left_join(final_values_retro, by = join_by(geo_value, reference_date)) |>
  filter(!is.na(final_value)) |>
  pivot_longer(
    cols = c(q05, q25, median, q75, q95),
    names_to = "quantile_level",
    values_to = "predicted"
  ) |>
  mutate(
    quantile_level = case_match(
      quantile_level,
      "q05" ~ 0.05, "q25" ~ 0.25, "median" ~ 0.5, "q75" ~ 0.75, "q95" ~ 0.95
    ),
    model = "baselinenowcast"
  ) |>
  rename(observed = final_value) |>
  as_forecast_quantile(
    forecast_unit = c("model", "nowcast_date", "reference_date", "geo_value"),
    observed = "observed",
    predicted = "predicted",
    quantile_level = "quantile_level"
  ) |>
  score() |>
  left_join(naive_wis, by = join_by(geo_value, nowcast_date, reference_date)) |>
  mutate(
    relative_wis = wis / naive_wis,
    horizon_weeks = difftime(nowcast_date, reference_date, units = "weeks")
  )

scores |>
  summarise(
    across(c(wis, naive_wis, relative_wis, overprediction, underprediction, dispersion),
           ~ round(mean(.x, na.rm = TRUE), 2))
  )
```

### Score by horizon

Horizon = number of weeks between the reference date and the nowcast date (smaller = harder to nowcast).

```{r score-by-horizon}
#| fig-width: 9
#| fig-height: 4
scores |>
  group_by(horizon_weeks) |>
  summarise(mean_relative_wis = mean(relative_wis, na.rm = TRUE), n = n()) |>
  ggplot(aes(horizon_weeks, mean_relative_wis)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "tomato") +
  scale_x_continuous(breaks = \(x) seq(floor(min(x)), ceiling(max(x)))) +
  labs(title = "Mean relative WIS by nowcast horizon",
       subtitle = "WIS / naive WIS (as-of value); dashed = no improvement over nominal",
       x = "Horizon (weeks)", y = "Mean relative WIS")
```

### Coverage

Empirical coverage of 50% and 90% credible intervals.

```{r coverage}
retro_nowcasts |>
  left_join(final_values_retro, by = join_by(geo_value, reference_date)) |>
  filter(!is.na(final_value)) |>
  summarise(
    `50% coverage` = mean(final_value >= q25 & final_value <= q75),
    `90% coverage` = mean(final_value >= q05 & final_value <= q95),
    n = n()
  )
```

### Nowcasts for most-revised weeks

For the reference weeks with the largest relative revision, show how the nowcast evolved across successive nowcast dates compared to the final value.

```{r high-revision-nowcasts}
#| cache: true
#| dependson: ["geos_by_latency", "retro-run", "retro-score", "load-data"]
#| fig-width: 12
#| fig-height: 10
revision_sum$revision_behavior |> print(width=10000)
top_revised <- revision_sum$revision_behavior |>
  filter(
    geo_value %nin% c("vi", "as", "gu"),
    min_value > 10
  ) |>
  semi_join(retro_nowcasts, by = join_by(geo_value, time_value == reference_date)) |>
  slice_max(rel_spread, n = 6) |>
  select(geo_value, reference_date = time_value)

as_of_retro <- retro_nowcasts$nowcast_date |>
  unique() |>
  purrr::map(\(d) nhsn_archive |> epix_as_of(d) |> mutate(nowcast_date = d)) |>
  list_rbind() |>
  rename(reference_date = time_value, as_of_value = value) |>
  select(geo_value, nowcast_date, reference_date, as_of_value)

high_rev_summary <- retro_nowcasts |>
  inner_join(top_revised, by = join_by(geo_value, reference_date)) |>
  left_join(final_values_retro, by = join_by(geo_value, reference_date)) |>
  left_join(as_of_retro, by = join_by(geo_value, nowcast_date, reference_date))

ggplot(high_rev_summary, aes(nowcast_date)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median, colour = "Nowcast median")) +
  geom_point(aes(y = as_of_value, colour = "As-of value"), shape = 1, size = 1.5) +
  geom_hline(aes(yintercept = final_value, colour = "Final value"), linetype = "dashed") +
  scale_colour_manual(values = c(
    "Nowcast median" = "steelblue",
    "As-of value" = "tomato",
    "Final value" = "black"
  )) +
  facet_wrap(~ geo_value + reference_date, scales = "free", ncol = 2) +
  labs(
    title = "Nowcast evolution for 6 most-revised (geo, reference week) pairs",
    subtitle = "x-axis = nowcast date; dashed = final value; points = as-of value at that date",
    x = "Nowcast date", y = "COVID hospitalisations", colour = NULL
  )
```

### Nowcast vs final over time


Plotting the latest nowcast for each nowcast_date; this will be the day which has the worst quality data, and the one we're most interested in getting right.

The other option would be to choose the last nowcast that is ever made for any given day, which may actually be asking what we nowcast months later.

```{r retro-plot}
#| cache: true
#| dependson: ["high-revision-nowcasts", "retro-run", "retro-score"]
#| fig-width: 24
#| fig-height: 20
#| column: screen
retro_summary <- retro_nowcasts |>
  group_by(geo_value, nowcast_date) |>
  slice_max(reference_date, n = 1) |>
  ungroup() |>
  left_join(final_values_retro, by = join_by(geo_value, reference_date)) |>
  left_join(as_of_retro, by = join_by(geo_value, nowcast_date, reference_date))
plot_ex <- retro_summary |> filter(geo_value %nin% c("vi", "as", "gu")) |> pull(geo_value) |> unique()
ggplot(retro_summary |> filter(geo_value %in% plot_ex), aes(reference_date)) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              alpha = 0.5, fill = "steelblue") +
  geom_ribbon(aes(ymin = q05, ymax = q95),
              alpha = 0.2, fill = "steelblue") +
  geom_point(aes(y = median, colour = "Nowcast median"), size = 1) +
  geom_point(aes(y = as_of_value, colour = "As-of value"), shape = 1, size = 1) +
  geom_line(
    data = semi_join(final_values_retro, retro_summary, by = join_by(geo_value, reference_date)) |> filter(geo_value %in% plot_ex),
    aes(y = final_value, colour = "Final value")
  ) +
  facet_wrap(~ geo_value, scales = "free_y", ncol=4) +
  scale_colour_manual(values = c(
    "Nowcast median" = "steelblue",
    "As-of value" = "tomato",
    "Final value" = "black"
  )) +
  theme(text = element_text(size = 9), plot.title = element_text(size = 11)) +
  labs(
    title = "Retrospective nowcast medians vs final values",
    subtitle = "90% CI shown; latest nowcast per nowcast date",
    x = "Reference week", y = "COVID hospitalisations", colour = NULL
  )
```
since the 95% being as large as it is obscures the values, here's a capped version to get a clearer plot.
```{r retro-plot-clipped}
#| cache: true
#| dependson: ["retro-plot", "retro-score"]
#| fig-width: 24
#| fig-height: 20
#| column: screen
retro_clipped <- retro_summary |>
  filter(geo_value %in% plot_ex) |>
  group_by(geo_value) |>
  mutate(y_max = max(q75, final_value, na.rm = TRUE)) |>
  ungroup() |>
  mutate(across(c(q05, q25, median, q75, q95, as_of_value), \(x) pmin(x, y_max)))

ggplot(retro_clipped, aes(reference_date)) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              alpha = 0.5, fill = "steelblue") +
  geom_ribbon(aes(ymin = q05, ymax = q95),
              alpha = 0.2, fill = "steelblue") +
  geom_point(aes(y = median, colour = "Nowcast median"), size = 1) +
  geom_point(aes(y = as_of_value, colour = "As-of value"), shape = 1, size = 1) +
  geom_line(
    data = semi_join(final_values_retro, retro_clipped, by = join_by(geo_value, reference_date)),
    aes(y = final_value, colour = "Final value")
  ) +
  facet_wrap(~ geo_value, scales = "free_y", ncol=4) +
  scale_colour_manual(values = c(
    "Nowcast median" = "steelblue",
    "As-of value" = "tomato",
    "Final value" = "black"
  )) +
  theme(text = element_text(size = 9), plot.title = element_text(size = 11)) +
  labs(
    title = "Retrospective nowcast medians vs final values (clipped to q75)",
    subtitle = "y-axis capped at max(q75, final_value) per geo; outlier nowcasts extend beyond",
    x = "Reference week", y = "COVID hospitalisations", colour = NULL
  )
```
