---
title: MultiQT calibration gallery — per-forecast panels at the chosen operating point
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 2
params:
  burn_in_seasons: "2023-2024"
  season_policy: "carry"
  lr_window: 20
  lr_mult: 0.03
  n_panels: 150
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
# cache = TRUE chunks are keyed on chunk code + cache.extra, NOT on functions
# in R/ -- after changing calibration code, clear cache/calibration/knitr_cache_gallery/.
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 10, fig.height = 6,
  message = FALSE, warning = FALSE, cache = FALSE,
  cache.path = here::here("cache/calibration/knitr_cache_gallery/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r load, cache=TRUE, cache.extra=list(params)}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))

forecasts <- hub_read_forecasts()
truth <- hub_read_truth()
rounds <- hub_label_seasons(unique(forecasts$reference_date))

burn_in <- if (nzchar(params$burn_in_seasons)) {
  strsplit(params$burn_in_seasons, ",")[[1]]
} else {
  character(0)
}

cal <- calibrate_hub_forecasts(
  forecasts, truth,
  burn_in_seasons = burn_in,
  season_policy = params$season_policy,
  lr_window = params$lr_window,
  lr_args = list(mult = params$lr_mult),
  progress = FALSE
)
```

Companion to `calibration_qt.Rmd` (the parameter-sweep record; see
`notes/CALIBRATION.md` for the project state). Parameters are fixed at the
operating point the sweep selected — `lr_mult = `r params$lr_mult``,
`lr_window = `r params$lr_window``, `season_policy = "`r params$season_policy`"`
— and the rest of the document is a ranked gallery of individual forecasts,
one panel per (location, round), for the two live seasons.

# Headline at the fixed operating point

Calibration error is the mean over the 23 quantile levels of
`|geo-pooled empirical coverage at a - a|`; WIS uses the standard
factor-of-two convention. The burn-in season is excluded.

```{r headline}
bind_rows(
  left_join(
    hub_coverage_summary(cal),
    select(hub_quantile_loss(cal), horizon, loss_base, loss_cal, loss_pct = pct_improvement),
    by = "horizon"
  ) %>% mutate(season = "all"),
  left_join(
    hub_coverage_summary(cal, by = "season"),
    select(hub_quantile_loss(cal, by = "season"),
      season, horizon, loss_base, loss_cal, loss_pct = pct_improvement
    ),
    by = c("season", "horizon")
  )
) %>%
  transmute(
    season, horizon,
    `cal err (base)` = round(cal_error_base, 4),
    `cal err (calibrated)` = round(cal_error_cal, 4),
    `cal err improvement %` = round(pct_improvement, 1),
    `WIS (base)` = round(2 * loss_base, 1),
    `WIS (calibrated)` = round(2 * loss_cal, 1),
    `WIS improvement %` = round(loss_pct, 1)
  ) %>%
  arrange(season, horizon)
```

# Data vintages for the panels

Each panel shows the NHSN data **as of the Wednesday before the reference
date** (`reference_date - 3`, roughly what the forecaster saw at submission
time) alongside the finalized truth. Vintages come from a frozen copy of the
NHSN archive (`cache/calibration/nhsn_archive_flu.parquet`; provenance in
`notes/CALIBRATION.md`). No vintages exist before 2024-11-19, which still
covers every live-season round.

```{r snapshots, cache=TRUE, dependson="load"}
archive <- nanoparquet::read_parquet(
  here::here("cache/calibration/nhsn_archive_flu.parquet")
) %>%
  as_epi_archive(compactify = TRUE)

geo_map <- get_population_data() %>%
  distinct(state_id, .keep_all = TRUE) %>%
  transmute(state_id, abbrev = toupper(state_id), location = state_code)

live_rounds <- rounds %>% filter(!season %in% burn_in)
min_version <- min(archive$DT$version)

snapshots <- purrr::map(live_rounds$round_date, function(rd) {
  as_of <- rd - 3L
  if (as_of < min_version) {
    return(NULL)
  }
  epix_as_of(archive, as_of) %>%
    as_tibble() %>%
    inner_join(geo_map, by = c("geo_value" = "state_id")) %>%
    transmute(reference_date = rd, location, time_value, value)
}) %>% bind_rows()
```

# Ranking

Panels are ordered by the size of the correction: the mean absolute
displacement between calibrated and base quantiles (over the 5 horizons x 23
levels of a round), divided by the round's mean base median so small states
rank on relative shift rather than raw admissions. The denominator is floored
at 20 admissions — without the floor the entire top of the list is
end-of-season rounds where the base median is ~1 and any leftover offset is a
huge ratio. Even with the floor, season edges (May, October) dominate the top:
the tracker carries offsets sized for mid-season signal into weeks with almost
none. That is the first gallery finding, not an artifact of the metric.
Captions carry the raw numbers: MAD, the WIS change of that single forecast
(positive = calibration improved it), and whether the finalized truth escaped
the base 90% interval at any horizon.

```{r ranking}
pinball <- function(truth, value, level) {
  ifelse(truth >= value, level * (truth - value), (1 - level) * (value - truth))
}

live <- cal$forecasts %>%
  filter(!is_burn_in, !is.na(value_base), !is.na(value_cal))

escapes <- live %>%
  filter(level %in% c(0.05, 0.95), !is.na(truth)) %>%
  select(location, reference_date, horizon, level, value_base, truth) %>%
  tidyr::pivot_wider(names_from = level, values_from = value_base, names_prefix = "q") %>%
  group_by(location, reference_date) %>%
  summarize(escape90 = any(truth < q0.05 | truth > q0.95), .groups = "drop")

panel_stats <- live %>%
  group_by(season, location, reference_date) %>%
  summarize(
    mad = mean(abs(value_cal - value_base)),
    base_med = mean(value_base[level == 0.5]),
    wis_base = 2 * mean(pinball(truth, value_base, level)[!is.na(truth)]),
    wis_cal = 2 * mean(pinball(truth, value_cal, level)[!is.na(truth)]),
    .groups = "drop"
  ) %>%
  mutate(
    mad_rel = mad / pmax(base_med, 20),
    wis_pct = if_else(wis_base > 0, 100 * (1 - wis_cal / wis_base), NA_real_)
  ) %>%
  left_join(escapes, by = c("location", "reference_date")) %>%
  left_join(select(geo_map, location, abbrev), by = "location") %>%
  arrange(desc(mad_rel))

panel_stats %>%
  transmute(
    season, abbrev, reference_date,
    mad = round(mad, 1), mad_rel = round(mad_rel, 3),
    wis_pct = round(wis_pct, 1), escape90
  )
```

# Gallery: top `r params$n_panels` panels

How to read a panel: black is the NHSN data as of the forecast Wednesday
(dashed vertical line); gray is the finalized series. Blue is the submitted
(base) forecast, orange the calibrated one; ribbons are the 50% and 90%
central intervals, the line is the median, across horizons -1 to 3.

```{r panel-fn}
fan_cols <- c(base = "#0072B2", calibrated = "#D55E00")
ribbon_levels <- c(
  "0.05" = "lo90", "0.25" = "lo50", "0.5" = "med",
  "0.75" = "hi50", "0.95" = "hi90"
)

panel_plot <- function(loc, ref_date, caption) {
  win_start <- ref_date - 84L
  win_end <- ref_date + 28L
  fin <- truth %>%
    filter(location == loc, target_end_date >= win_start, target_end_date <= win_end)
  snap <- snapshots %>%
    filter(reference_date == ref_date, location == loc, time_value >= win_start)
  fan <- cal$forecasts %>%
    filter(
      location == loc, reference_date == ref_date,
      level %in% c(0.05, 0.25, 0.5, 0.75, 0.95)
    ) %>%
    mutate(
      target_end_date = reference_date + 7L * horizon,
      level = ribbon_levels[as.character(level)]
    ) %>%
    select(target_end_date, level, base = value_base, calibrated = value_cal) %>%
    tidyr::pivot_longer(c(base, calibrated), names_to = "which") %>%
    tidyr::pivot_wider(names_from = level, values_from = value)

  ggplot() +
    geom_line(data = fin, aes(target_end_date, truth), colour = "grey65") +
    geom_point(data = fin, aes(target_end_date, truth), colour = "grey65", size = 0.7) +
    geom_ribbon(
      data = fan, aes(target_end_date, ymin = lo90, ymax = hi90, fill = which),
      alpha = 0.16
    ) +
    geom_ribbon(
      data = fan, aes(target_end_date, ymin = lo50, ymax = hi50, fill = which),
      alpha = 0.3
    ) +
    geom_line(data = fan, aes(target_end_date, med, colour = which)) +
    geom_line(data = snap, aes(time_value, value), colour = "black") +
    geom_point(data = snap, aes(time_value, value), colour = "black", size = 0.7) +
    geom_vline(xintercept = ref_date - 3L, linetype = "dashed", colour = "grey40") +
    scale_fill_manual(values = fan_cols, aesthetics = c("fill", "colour")) +
    labs(x = NULL, y = "admissions", colour = NULL, fill = NULL, subtitle = caption)
}
```

```{r gallery, results='asis', fig.width=9, fig.height=4}
top <- head(panel_stats, params$n_panels)
for (i in seq_len(nrow(top))) {
  p <- top[i, ]
  cat(sprintf(
    "\n\n### %d. %s — %s (%s)\n\n",
    i, p$abbrev, format(p$reference_date), p$season
  ))
  caption <- sprintf(
    "MAD %.1f admissions (base median %.1f) | WIS %s | truth escaped base 90%%: %s",
    p$mad, p$base_med,
    if (is.na(p$wis_pct)) "n/a" else sprintf("%+.1f%%", p$wis_pct),
    if (isTRUE(p$escape90)) "yes" else "no"
  )
  print(panel_plot(p$location, p$reference_date, caption))
}
```
