---
title: MultiQT calibration — whole-season views of CA, across method variants
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 2
params:
  burn_in_seasons: "2023-2024"
  season_policy: "carry"
  lr_window: 20
  lr_mult: 0.03
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
# cache = TRUE chunks are keyed on chunk code + cache.extra, NOT on functions
# in R/ -- after changing calibration code, clear cache/calibration/knitr_cache_seasons/.
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 10, fig.height = 6,
  message = FALSE, warning = FALSE, cache = FALSE,
  cache.path = here::here("cache/calibration/knitr_cache_seasons/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

This notebook takes the operating point chosen in `notes/CALIBRATION.md`
(`lr_mult = 0.03`, `lr_window = 20`, carry, burn-in 2023-24) and looks at one
state (CA) across whole seasons rather than one panel at a time
(`calibration_qt_gallery.Rmd` has the per-forecast gallery view), to understand
the method's behavior in depth before widening again. Method variants compared:
the baseline, turning calibration off after March 1 or February 15, per-level
eta, geo-pooled eta, and a seasonal eta window.

```{r load, cache=TRUE}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))

forecasts <- hub_read_forecasts()
truth <- hub_read_truth()
rounds <- hub_label_seasons(unique(forecasts$reference_date))
burn_in <- strsplit(params$burn_in_seasons, ",")[[1]]

geo_map <- get_population_data() %>%
  distinct(state_id, .keep_all = TRUE) %>%
  transmute(state_id, abbrev = toupper(state_id), location = state_code)
pop <- get_population_data() %>%
  distinct(state_code, .keep_all = TRUE) %>%
  transmute(location = state_code, population)

archive <- nanoparquet::read_parquet(
  here::here("cache/calibration/nhsn_archive_flu.parquet")
) %>%
  as_epi_archive(compactify = TRUE)

live_rounds <- rounds %>% filter(!season %in% burn_in)
min_version <- min(archive$DT$version)
snapshots <- purrr::map(live_rounds$round_date, function(rd) {
  as_of <- rd - 3L
  if (as_of < min_version) {
    return(NULL)
  }
  epix_as_of(archive, as_of) %>%
    as_tibble() %>%
    inner_join(geo_map, by = c("geo_value" = "state_id")) %>%
    transmute(reference_date = rd, location, time_value, value)
}) %>% bind_rows()

states <- tibble(location = "06", abbrev = "CA")
seasons <- c("2024-2025", "2025-2026")
```

```{r variants, cache=TRUE, dependson="load"}
base_args <- list(
  burn_in_seasons = burn_in, season_policy = params$season_policy,
  lr_window = params$lr_window, lr_args = list(mult = params$lr_mult),
  progress = FALSE
)
variant_args <- list(
  baseline = list(),
  off_march = list(off_after = "03-01"),
  off_feb = list(off_after = "02-15"),
  per_level = list(lr_args = list(mult = params$lr_mult, per_level = TRUE)),
  geo_pooled = list(lr_geo_pool = pop),
  seasonal = list(lr_window = 10, lr_seasonal = list(half_width_weeks = 5), season_policy = "reset")
)
variant_titles <- c(
  baseline = "Baseline",
  off_march = "Calibration off after March 1",
  off_feb = "Calibration off after February 15",
  per_level = "Per-level eta",
  geo_pooled = "Geo-pooled eta",
  seasonal = "Seasonal eta window + reset"
)

cals <- purrr::map(variant_args, function(v) {
  do.call(calibrate_hub_forecasts, c(list(forecasts, truth), modifyList(base_args, v)))
})
```

# Summary across variants

WIS % improvement over the base forecast (all seasons, burn-in excluded), per
horizon.

```{r variant-summary}
purrr::imap(cals, function(cal, nm) {
  hub_quantile_loss(cal) %>% transmute(variant = variant_titles[[nm]], horizon, pct = round(pct_improvement, 1))
}) %>%
  bind_rows() %>%
  tidyr::pivot_wider(names_from = horizon, values_from = pct, names_prefix = "h")
```

```{r helpers}
headline_table <- function(cal) {
  bind_rows(
    left_join(
      hub_coverage_summary(cal),
      select(hub_quantile_loss(cal), horizon, loss_base, loss_cal, loss_pct = pct_improvement),
      by = "horizon"
    ) %>% mutate(season = "all"),
    left_join(
      hub_coverage_summary(cal, by = "season"),
      select(hub_quantile_loss(cal, by = "season"),
        season, horizon, loss_base, loss_cal, loss_pct = pct_improvement
      ),
      by = c("season", "horizon")
    )
  ) %>%
    transmute(
      season, horizon,
      `cal err (base)` = round(cal_error_base, 4),
      `cal err (calibrated)` = round(cal_error_cal, 4),
      `cal err improvement %` = round(pct_improvement, 1),
      `WIS (base)` = round(2 * loss_base, 1),
      `WIS (calibrated)` = round(2 * loss_cal, 1),
      `WIS improvement %` = round(loss_pct, 1)
    ) %>%
    arrange(season, horizon)
}

fan_cols <- c(base = "#0072B2", calibrated = "#D55E00")

# One plot per (state, season): whole-season fan (odd season_rounds only, h 0..3;
# h -1 dropped -- with h -1 in, 5 horizons x ~13 rounds x 2 fans was unreadable)
# plus vintage patchwork strips underneath, plus an eta strip.
season_plot <- function(cal, loc, season) {
  rr <- cal$rounds %>% filter(season == !!season, season_round %% 2 == 1)
  ref_dates <- sort(rr$round_date)

  fin <- truth %>%
    filter(location == loc, target_end_date >= min(ref_dates) - 21L, target_end_date <= max(ref_dates) + 28L)

  fan <- cal$forecasts %>%
    filter(location == loc, reference_date %in% ref_dates, horizon %in% 0:3, level %in% c(0.1, 0.5, 0.9)) %>%
    mutate(
      target_end_date = reference_date + 7L * horizon,
      lvl = c(`0.1` = "lo", `0.5` = "med", `0.9` = "hi")[as.character(level)]
    ) %>%
    select(reference_date, target_end_date, lvl, base = value_base, calibrated = value_cal) %>%
    tidyr::pivot_longer(c(base, calibrated), names_to = "which") %>%
    tidyr::pivot_wider(names_from = lvl, values_from = value)

  # Vintage patchwork: each plotted round's as-of snapshot paints only the
  # fortnight since the previous plotted round (like backtesting.Rmd's vignette).
  vint_windows <- tibble(ref = ref_dates) %>%
    mutate(
      win_start = dplyr::lag(ref, default = ref[1] - 14L) - 3L,
      win_end = ref - 3L
    )
  vint <- purrr::pmap(vint_windows, function(ref, win_start, win_end) {
    snapshots %>%
      filter(reference_date == ref, location == loc, time_value >= win_start, time_value <= win_end)
  }) %>% bind_rows()

  main <- ggplot() +
    geom_line(data = fin, aes(target_end_date, truth), colour = "grey65") +
    geom_point(data = fin, aes(target_end_date, truth), colour = "grey65", size = 0.7) +
    geom_ribbon(
      data = filter(fan, which == "base"),
      aes(target_end_date, ymin = lo, ymax = hi, fill = which, group = reference_date),
      alpha = 0.2
    ) +
    geom_line(
      data = filter(fan, which == "calibrated"),
      aes(target_end_date, lo, colour = which, group = reference_date), linewidth = 0.3
    ) +
    geom_line(
      data = filter(fan, which == "calibrated"),
      aes(target_end_date, hi, colour = which, group = reference_date), linewidth = 0.3
    ) +
    geom_line(
      data = fan,
      aes(target_end_date, med, colour = which, group = interaction(reference_date, which))
    ) +
    geom_line(data = vint, aes(time_value, value), colour = "black") +
    geom_point(data = vint, aes(time_value, value), colour = "black", size = 0.6) +
    geom_vline(xintercept = ref_dates - 3L, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
    scale_fill_manual(values = fan_cols, aesthetics = c("fill", "colour")) +
    labs(x = NULL, y = "admissions", colour = NULL, fill = NULL) +
    theme(legend.position = "top")

  eta <- cal$series %>%
    filter(location == loc, season == !!season, horizon %in% 0:3) %>%
    # lr is 0 at rounds with no reveal (gaps); blank those rather than plot dips.
    mutate(lr = if_else(lr > 0, lr, NA_real_)) %>%
    ggplot(aes(round_date, lr, colour = factor(horizon))) +
    geom_line(na.rm = TRUE) +
    geom_point(size = 0.6, na.rm = TRUE) +
    labs(x = NULL, y = "eta", colour = "horizon") +
    theme(legend.position = "top")

  title <- paste0(geo_map$abbrev[match(loc, geo_map$location)], " — ", season)
  cowplot::plot_grid(
    cowplot::ggdraw() + cowplot::draw_label(title, fontface = "bold"),
    cowplot::plot_grid(main, eta, ncol = 1, rel_heights = c(3, 1), align = "v", axis = "lr"),
    ncol = 1, rel_heights = c(0.08, 1)
  )
}

# oh_table.R's internals table, generalized to any location/horizon/season.
oh_table <- function(cal, loc, h, season) {
  cal$forecasts %>%
    filter(location == loc, horizon == h, season == !!season, level %in% c(0.05, 0.25, 0.5, 0.75, 0.95)) %>%
    mutate(level = paste0("q", level)) %>%
    select(reference_date, level, lr_level, truth, hidden, offset, value_base, value_cal) %>%
    tidyr::pivot_wider(
      names_from = level,
      values_from = c(hidden, offset, value_base, value_cal, lr_level),
      names_glue = "{level}_{.value}"
    ) %>%
    mutate(
      eta = round(q0.5_lr_level, 3),
      across(c(truth, matches("hidden|offset|value_base|value_cal")), ~ round(.x))
    ) %>%
    select(
      reference_date, eta, truth,
      matches("^q.*_hidden$"), matches("^q.*_offset$"),
      matches("^q.*_value_base$"), matches("^q.*_value_cal$")
    ) %>%
    arrange(reference_date)
}

internals_block <- function(cal, season) {
  for (i in seq_len(nrow(states))) {
    loc <- states$location[i]
    abbrev <- states$abbrev[i]
    cat(sprintf("\n\n<details><summary>%s internals (h=2)</summary>\n\n", abbrev))
    print(knitr::kable(oh_table(cal, loc, 2L, season)))
    cat("\n\n</details>\n\n")
  }
}
```

The whole-season fans plot horizons 0–3 only (dropping −1) and only every
other round. Blue ribbon and line: base 80% band and median; orange lines:
calibrated 10%, 50%, 90% quantiles. Black: the NHSN vintage each forecast saw,
painted over the fortnight before its reference date (dashed); grey: finalized
truth. The strip below is eta per horizon (blank where a round revealed no
outcome). Read it as "where do the medians diverge, and does the orange band
contain the grey truth better than the blue", not round-by-round.

```{r sections, results='asis'}
for (nm in names(cals)) {
  cal <- cals[[nm]]
  cat(sprintf("\n\n# %s\n\n", variant_titles[[nm]]))

  desc <- switch(nm,
    baseline = "The chosen operating point: lr_mult = 0.03, lr_window = 20, carry, isotonic
      projection. Offsets update every round including the spring tail and carry
      across the off-season gap.",
    off_march = "Rounds from March 1 through June play the base forecast unchanged and take
      no gradient step; the offset learned through the peak is frozen and carried into the
      next season rather than decaying toward zero on low-signal spring data. First variant
      that was WIS-neutral-or-better at every horizon (April 1 still paid at h1-h3).",
    off_feb = "Same switch-off from February 15: does cutting even earlier start costing
      calibration error at the peak, or is the post-peak descent already out of regime?",
    per_level = "Each of the 23 quantile levels gets its own eta from its own residual
      history, instead of one eta shared (via the level mean) across all levels.
      Expect the tails to adapt at a different rate than the median.",
    geo_pooled = "Eta is computed once per (horizon, round) from per-capita residuals
      pooled across all 53 locations, then rescaled by each location's population and
      floored at 0.1 in count space. Eta is not pooled across horizons in any variant --
      each (location, horizon) series keeps its own. Expect low-count states (VT, WY) to
      benefit most, since their own residual history is too sparse to size eta well.",
    seasonal = "lr_window drops to 10 rounds but is padded with a +/-5 week window around
      the same season-week in earlier seasons, so eta at the start of a ramp is sized by
      last year's ramp rather than last year's trough. Offsets are reset to zero at the
      season boundary (with carry, the small early-season eta left last spring's offsets
      in place for longer)."
  )
  cat(desc, "\n\n")

  cat("## Headline\n\n")
  print(knitr::kable(headline_table(cal)))

  for (season in seasons) {
    cat(sprintf("\n\n## %s\n\n", season))
    for (i in seq_len(nrow(states))) {
      loc <- states$location[i]
      cat(sprintf("\n\n### %s — %s\n\n", states$abbrev[i], season))
      print(season_plot(cal, loc, season))
    }
  }

  cat("\n\n## Internals\n\n")
  for (season in seasons) {
    cat(sprintf("\n\n### %s\n\n", season))
    internals_block(cal, season)
  }
}
```
