---
title: Online quantile calibration (MultiQT) on submitted FluSight forecasts
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
params:
  burn_in_seasons: "2023-2024"
  season_policy: "carry"
  lr_window: 50
  lr_mult: 0.1
  roll_window: 20
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
# cache = TRUE chunks are keyed on chunk code + cache.extra, NOT on functions
# in R/ -- after changing calibration code, clear cache/calibration/knitr_cache/.
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 10, fig.height = 6,
  message = FALSE, warning = FALSE, cache = FALSE,
  cache.path = here::here("cache/calibration/knitr_cache/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r load, cache=TRUE, cache.extra=list(params)}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))

forecasts <- hub_read_forecasts()
truth <- hub_read_truth()
rounds <- hub_label_seasons(unique(forecasts$reference_date))

burn_in <- if (nzchar(params$burn_in_seasons)) {
  strsplit(params$burn_in_seasons, ",")[[1]]
} else {
  character(0)
}

cal <- calibrate_hub_forecasts(
  forecasts, truth,
  burn_in_seasons = burn_in,
  season_policy = params$season_policy,
  lr_window = params$lr_window,
  lr_args = list(mult = params$lr_mult),
  progress = FALSE
)
```

We calibrate the forecasts CMU-TimeSeries submitted to FluSight. https://github.com/cdcepi/FluSight-forecast-hub The hub
ships one consistent truth series (`target-data/target-hospital-admissions.csv`)
covering every season. The method is MultiQT (Ding, Gibbs & Tibshirani 2025); our
port is `R/calibration/qt.R`, verified against the authors' Python in
`tests/testthat/test-qt.R` (same input, same output).

Data facts:

```{r data-shape}
forecasts %>%
  left_join(select(rounds, reference_date = round_date, season), by = "reference_date") %>%
  group_by(season) %>%
  summarize(
    rounds = n_distinct(reference_date),
    locations = n_distinct(location),
    horizons = paste(sort(unique(horizon)), collapse = ","),
    cell_completeness = sprintf(
      "%.0f%%", 100 * (n() / 23) / (n_distinct(reference_date) * 265)
    ),
    .groups = "drop"
  )
```

Notes:

1. Waiting for revisions costs us `horizon + 2` rounds. `reference_date` and
   `target_end_date` are both Saturdays with
   `target_end_date = reference_date + 7 * horizon`, so waiting 14 days for NHSN
   to settle puts the delay at exactly `h + 2` rounds. The offset applied at
   horizon 3 is therefore always 5 rounds stale.
2. 2023-24 is incomplete: horizon -1 exists in only 5 of 28 rounds and
   locations drop in and out, so it is used as a burn-in season. Its residuals
   warm up the learning rate, but no gradient step is taken and the offsets
   stay at zero.
3. Per-series coverage cannot be measured at the extreme levels: ~27 rounds
   per season minus delay means the 1% and 99% levels see ~0.3 expected
   exceedances per series. Every coverage number below is pooled over the 53
   locations.

# Score results

Coverage improves at some cost in sharpness. Below we show that the sharpness
cost can be reduced by changing the learning rate.

```{r headline}
headline <- left_join(
  hub_coverage_summary(cal),
  select(hub_quantile_loss(cal), horizon, loss_base, loss_cal, loss_pct = pct_improvement),
  by = "horizon"
)
headline %>%
  transmute(
    horizon,
    n_obs,
    `cal err (base)` = round(cal_error_base, 4),
    `cal err (calibrated)` = round(cal_error_cal, 4),
    `cal err improvement %` = round(pct_improvement, 1),
    `WIS (base)` = round(2 * loss_base, 1),
    `WIS (calibrated)` = round(2 * loss_cal, 1),
    `WIS improvement %` = round(loss_pct, 1)
  )
```

Calibration error here is the mean over the 23 quantile levels of
`|geo-pooled empirical coverage at a - a|`. WIS is reported with the standard
factor-of-two convention (2 x the level-averaged pinball loss on this level
set), so it is directly comparable to hub scores; improvement percentages are
identical either way. The two metrics move in opposite directions, and the WIS
cost grows with the horizon. This is consistent with the `h + 2` staleness: a
stale additive offset in count space does the most damage where the seasonal
ramp is steep relative to the offset.

## By season

```{r headline-by-season}
left_join(
  hub_coverage_summary(cal, by = "season"),
  select(hub_quantile_loss(cal, by = "season"),
    season, horizon, loss_base, loss_cal, loss_pct = pct_improvement
  ),
  by = c("season", "horizon")
) %>%
  transmute(
    season, horizon,
    `cal err (base)` = round(cal_error_base, 4),
    `cal err (calibrated)` = round(cal_error_cal, 4),
    `cal err improvement %` = round(pct_improvement, 1),
    `WIS (base)` = round(2 * loss_base, 1),
    `WIS (calibrated)` = round(2 * loss_cal, 1),
    `WIS improvement %` = round(loss_pct, 1)
  ) %>%
  arrange(season, horizon)
```

The burn-in season (2023-24) is excluded throughout. In the per-season view,
season 2 is the tracker's first live season; season 3 shows the effect of a
second season of accumulated state (under `carry` across the off-season gap).

# 1. What feeds the learning rate

`eta_t = max(floor, mult * quantile_0.9(|Y - Yhat|))` over a window of recent
observed rounds, pooled across all 23 quantile levels -- one step size per
(series, round). The residual distribution below therefore determines the
learning rate, up to the multiplier.

```{r resid-series, fig.height=7}
resid <- cal$forecasts %>%
  filter(!is.na(truth), !is.na(value_base)) %>%
  mutate(abs_resid = abs(truth - value_base)) %>%
  group_by(location, horizon, reference_date, season, is_burn_in) %>%
  summarize(q90 = quantile(abs_resid, 0.9), .groups = "drop")

show_geos <- c("06", "36", "48", "50", "US")
resid %>%
  filter(location %in% show_geos) %>%
  mutate(location = recode(location,
    "06" = "CA", "36" = "NY", "48" = "TX", "50" = "VT", "US" = "US"
  )) %>%
  ggplot(aes(reference_date, q90, colour = factor(horizon))) +
  geom_line() +
  facet_wrap(~location, scales = "free_y", ncol = 1) +
  scale_y_log10() +
  labs(
    title = "90th percentile of |truth - base forecast|, per round",
    subtitle = "The quantity the adaptive learning rate is a fraction of. Log scale.",
    x = NULL, y = "q90 |residual| (log)", colour = "horizon"
  )
```

The residual scale swings by 1-2 orders of magnitude within a season, so an
additive offset learned near the peak is badly sized for the trough, and vice
versa.

## Does the window length matter?

`lr_window` sets how many of the most recent observed rounds feed the adaptive
learning rate; `Inf` is an expanding window over the full history.

```{r lr-window, cache=TRUE, dependson="load"}
lr_by_window <- purrr::map(c(8, 20, 50, Inf), function(w) {
  purrr::map(show_geos, function(loc) {
    series <- forecasts %>% filter(location == loc, horizon == 2)
    if (nrow(series) == 0) {
      return(NULL)
    }
    Yhat <- hub_series_matrix(series, rounds$round_date, 23)
    ted <- rounds$round_date + 14L
    Y <- truth$truth[match(paste(loc, ted), paste(truth$location, truth$target_end_date))]
    learnable <- !is.na(Y) & !apply(is.na(Yhat), 2, any)
    delay <- qt_delay_from_dates(rounds$round_date, ted, 14L)
    delay <- lapply(delay, function(i) i[learnable[i]])
    res <- qt_track(Y, Yhat, HUB_QUANTILE_LEVELS,
      delay = delay, lr_window = w, nonneg = TRUE
    )
    tibble(
      location = loc, round_date = rounds$round_date,
      lr = res$lr, window = as.character(w)
    )
  }) %>% bind_rows()
}) %>% bind_rows()

lr_by_window %>%
  filter(lr > 0) %>%
  mutate(
    location = recode(location,
      "06" = "CA", "36" = "NY", "48" = "TX", "50" = "VT", "US" = "US"
    ),
    window = factor(window, levels = c("8", "20", "50", "Inf"))
  ) %>%
  ggplot(aes(round_date, lr, colour = window)) +
  geom_line() +
  facet_wrap(~location, scales = "free_y", ncol = 1) +
  scale_y_log10() +
  labs(
    title = "Learning-rate trajectory by window length (horizon 2)",
    subtitle = "A short window tracks the season; an expanding one (Inf) ratchets up and stays.",
    x = NULL, y = "eta (log)", colour = "lr_window"
  )
```

The expanding window is monotone by construction: once a peak-season residual
enters the pool it never leaves, so `eta` rises to peak scale and stays there
through the following low season. A short window decays back. On 83 rounds the
difference in outcome is small (below), but it would likely grow over more
seasons.

# 2. Actual vs nominal coverage, per location

```{r coverage-plot, fig.height=6}
cov_by_geo <- hub_coverage(cal, by = "location")

cov_long <- cov_by_geo %>%
  select(location, horizon, level, coverage_base, coverage_cal) %>%
  tidyr::pivot_longer(
    c(coverage_base, coverage_cal),
    names_to = "which", values_to = "coverage"
  ) %>%
  mutate(which = factor(
    if_else(which == "coverage_base", "pre-calibration", "post-calibration"),
    levels = c("pre-calibration", "post-calibration")
  ))

cov_long %>%
  filter(horizon %in% c(0, 2)) %>%
  ggplot(aes(level, coverage, group = location)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  geom_line(alpha = 0.28, colour = "steelblue4") +
  facet_grid(horizon ~ which, labeller = label_both) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "Empirical vs nominal coverage, one line per location",
    subtitle = "Dashed line is perfect calibration. Each line is one of 53 locations, over all non-burn-in rounds.",
    x = "nominal level", y = "empirical coverage"
  )
```

```{r coverage-pooled}
hub_coverage(cal) %>%
  select(horizon, level, coverage_base, coverage_cal) %>%
  tidyr::pivot_longer(c(coverage_base, coverage_cal),
    names_to = "which", values_to = "coverage"
  ) %>%
  mutate(which = if_else(which == "coverage_base", "pre", "post")) %>%
  ggplot(aes(level, coverage, colour = which)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~horizon, labeller = label_both, nrow = 1) +
  coord_equal() +
  labs(
    title = "Geo-pooled coverage, pre and post calibration",
    x = "nominal level", y = "empirical coverage", colour = NULL
  )
```

The per-location spread is not visible in the pooled number: it shows whether
the pooled improvement is a per-series effect or an averaging artefact, and it
bears on whether the offsets themselves should be pooled across geos.

# 3. Calibration error against quantile loss

Both axes are rolling averages over the last `r params$roll_window` rounds, pooled
over locations. Moving down is better calibrated; moving left is sharper. A
trajectory that goes down and to the right means calibration is being bought
with sharpness.

```{r tradeoff, fig.height=6}
tradeoff <- hub_rolling_tradeoff(cal, window = params$roll_window)

tradeoff %>%
  select(horizon, reference_date, cal_error_base, cal_error_cal, qloss_base, qloss_cal) %>%
  tidyr::pivot_longer(
    -c(horizon, reference_date),
    names_to = c(".value", "which"), names_pattern = "(.*)_(base|cal)$"
  ) %>%
  mutate(which = if_else(which == "base", "pre-calibration", "post-calibration")) %>%
  ggplot(aes(qloss, cal_error, colour = which)) +
  geom_path(alpha = 0.7) +
  geom_point(aes(size = as.numeric(reference_date)), alpha = 0.5, show.legend = FALSE) +
  scale_size_continuous(range = c(0.4, 2.2)) +
  facet_wrap(~horizon, scales = "free_x", labeller = label_both) +
  labs(
    title = "Rolling calibration error vs rolling quantile loss",
    subtitle = "Point size increases with time. Down = better calibrated, left = sharper.",
    x = "rolling mean pinball loss (proportional to WIS)", y = "rolling calibration error", colour = NULL
  )
```

```{r tradeoff-time, fig.height=5}
tradeoff %>%
  select(horizon, reference_date, cal_error_base, cal_error_cal) %>%
  tidyr::pivot_longer(-c(horizon, reference_date),
    names_to = "which", values_to = "cal_error"
  ) %>%
  mutate(which = if_else(which == "cal_error_base", "pre", "post")) %>%
  ggplot(aes(reference_date, cal_error, colour = which)) +
  geom_line() +
  facet_wrap(~horizon, labeller = label_both, nrow = 1) +
  labs(
    title = "Rolling calibration error over time",
    subtitle = "How fast the tracker converges after the burn-in season, and what the off-season gap does to it.",
    x = NULL, y = "rolling calibration error", colour = NULL
  )
```

# 4. Interval widths

The WIS cost has to show up somewhere in the quantiles; central interval widths
are the direct measure. For three central intervals -- 20% (0.4 to 0.6), 60%
(0.2 to 0.8), 80% (0.1 to 0.9) -- we compare the calibrated width to the base
width, per (round, location, horizon). Widths are geo-pooled by mean, matching
the implicit weighting of the WIS aggregate (large-count series dominate both).

```{r interval-widths}
interval_defs <- tibble::tribble(
  ~interval, ~lo, ~hi,
  "20% (0.4-0.6)", 0.4, 0.6,
  "60% (0.2-0.8)", 0.2, 0.8,
  "80% (0.1-0.9)", 0.1, 0.9
)

width_rows <- cal$forecasts %>%
  filter(!is_burn_in, !is.na(value_base), !is.na(value_cal)) %>%
  select(reference_date, season, location, horizon, level, value_base, value_cal)

widths <- purrr::pmap(interval_defs, function(interval, lo, hi) {
  inner_join(
    width_rows %>% filter(level == lo) %>%
      select(-level) %>% rename(base_lo = value_base, cal_lo = value_cal),
    width_rows %>% filter(level == hi) %>%
      select(-level) %>% rename(base_hi = value_base, cal_hi = value_cal),
    by = c("reference_date", "season", "location", "horizon")
  ) %>%
    transmute(
      reference_date, season, location, horizon,
      interval,
      width_base = base_hi - base_lo,
      width_cal = cal_hi - cal_lo
    )
}) %>%
  bind_rows() %>%
  mutate(interval = factor(interval, levels = interval_defs$interval))
```

```{r width-time, fig.height=6}
widths %>%
  group_by(interval, horizon, reference_date) %>%
  summarize(
    base = mean(width_base), calibrated = mean(width_cal),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(c(base, calibrated), names_to = "which", values_to = "width") %>%
  ggplot(aes(reference_date, width, colour = which)) +
  geom_line() +
  facet_grid(interval ~ horizon, scales = "free_y", labeller = label_both) +
  labs(
    title = "Geo-pooled mean central interval width, base vs calibrated",
    subtitle = "Mean over the 53 locations, per round. The gap between the lines is the sharpness cost.",
    x = NULL, y = "mean interval width (admissions)", colour = NULL
  )
```

```{r width-change, fig.height=4}
widths %>%
  group_by(interval, horizon) %>%
  summarize(
    pct_change = 100 * (mean(width_cal) / mean(width_base) - 1),
    .groups = "drop"
  ) %>%
  ggplot(aes(horizon, pct_change, colour = interval)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Change in mean interval width from calibration",
    subtitle = "Positive means the calibrated intervals are wider than the base intervals.",
    x = "horizon", y = "width change (%)", colour = "central interval"
  )
```

Wider intervals where the base under-covers are the intended correction; the
question is whether the widening is concentrated in the intervals and horizons
where the coverage plots (section 2) show a deficit, or spread indiscriminately.

# 5. Sensitivity: window or multiplier?

The learning rate has two knobs: how far back the residual pool reaches
(`lr_window`) and what fraction of its 90th percentile becomes the step size
(`lr_mult`). The factorial sweep below shows which one matters: if the lines
for different windows sit on top of each other while the multiplier moves
everything, the multiplier is the tuning parameter and the window is a detail
(and vice versa).

```{r sweep, cache=TRUE, dependson="load"}
sweep_grid <- tidyr::expand_grid(
  season_policy = c("carry", "reset"),
  lr_window = c(8, 20, 50, Inf),
  lr_mult = c(0.3, 0.1, 0.03, 0.01)
)
sweep <- purrr::pmap(sweep_grid, function(season_policy, lr_window, lr_mult) {
  s <- calibrate_hub_forecasts(forecasts, truth,
    burn_in_seasons = burn_in, season_policy = season_policy,
    lr_window = lr_window, lr_args = list(mult = lr_mult), progress = FALSE
  )
  overall <- left_join(
    hub_coverage_summary(s),
    select(hub_quantile_loss(s), horizon, loss_pct = pct_improvement),
    by = "horizon"
  ) %>%
    mutate(season = "all")
  by_season <- left_join(
    hub_coverage_summary(s, by = "season"),
    select(hub_quantile_loss(s, by = "season"),
      season, horizon,
      loss_pct = pct_improvement
    ),
    by = c("season", "horizon")
  )
  bind_rows(overall, by_season) %>%
    mutate(season_policy = season_policy, lr_window = lr_window, lr_mult = lr_mult)
}) %>% bind_rows()

# Keep the sweep around for analysis outside the notebook.
dir.create(here::here("cache/calibration"), showWarnings = FALSE, recursive = TRUE)
saveRDS(sweep, here::here("cache/calibration/sweep.rds"))
```

```{r sweep-cal-error, fig.height=5}
sweep %>%
  filter(season == "all") %>%
  ggplot(aes(lr_mult, cal_error_cal, colour = factor(lr_window))) +
  geom_line() +
  geom_point(size = 1.6) +
  scale_x_log10() +
  facet_grid(season_policy ~ horizon, labeller = label_both) +
  labs(
    title = "Calibration error by multiplier, one line per window",
    subtitle = "Separated lines would indicate a window effect; sloped lines indicate a multiplier effect.",
    x = "lr mult (log)", y = "calibration error (calibrated)",
    colour = "lr_window"
  )
```

```{r sweep-wis, fig.height=5}
sweep %>%
  filter(season == "all") %>%
  ggplot(aes(lr_mult, loss_pct, colour = factor(lr_window))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_line() +
  geom_point(size = 1.6) +
  scale_x_log10() +
  facet_grid(season_policy ~ horizon, labeller = label_both) +
  labs(
    title = "WIS improvement by multiplier, one line per window",
    subtitle = "Dashed line is WIS parity with the raw submission; above it calibration has no WIS cost.",
    x = "lr mult (log)", y = "WIS improvement (%)",
    colour = "lr_window"
  )
```

The target region is the combination of low calibration error and WIS at or
above parity.

```{r sweep-corner, fig.height=5}
sweep %>%
  filter(season == "all", season_policy == "carry") %>%
  ggplot(aes(cal_error_cal, loss_pct,
    colour = factor(lr_mult), shape = factor(lr_window)
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 3, alpha = 0.85) +
  facet_wrap(~horizon, scales = "free_x", labeller = label_both, nrow = 1) +
  labs(
    title = "Calibration error vs WIS change (season_policy = carry)",
    subtitle = "Upper-left is the target region: well calibrated and no worse on WIS.",
    x = "calibration error (calibrated)", y = "WIS change (%)",
    colour = "lr mult", shape = "lr_window"
  )
```

If the expanding window's monotone behavior matters, its effect should grow
with history: season 2 starts with at most one season in the pool, season 3
with two. The per-season view checks that.

```{r sweep-by-season, fig.height=5}
sweep %>%
  filter(season != "all", season_policy == "carry", horizon == 2) %>%
  ggplot(aes(lr_mult, cal_error_cal, colour = factor(lr_window))) +
  geom_line() +
  geom_point(size = 1.6) +
  scale_x_log10() +
  facet_wrap(~season, labeller = label_both) +
  labs(
    title = "Calibration error by season (horizon 2, season_policy = carry)",
    subtitle = "Does the window matter more once multiple seasons of residuals are available?",
    x = "lr mult (log)", y = "calibration error (calibrated)",
    colour = "lr_window"
  )
```

```{r sweep-table}
sweep %>%
  filter(season == "all") %>%
  transmute(
    season_policy, lr_window, lr_mult, horizon,
    cal_error_cal = round(cal_error_cal, 4),
    wis_pct = round(loss_pct, 1)
  ) %>%
  arrange(horizon, season_policy, lr_window, lr_mult)
```

# 6. Offset trajectories

```{r offsets, fig.height=6}
cal$forecasts %>%
  filter(location %in% show_geos, horizon == 2, level %in% c(0.025, 0.5, 0.975)) %>%
  mutate(location = recode(location,
    "06" = "CA", "36" = "NY", "48" = "TX", "50" = "VT", "US" = "US"
  )) %>%
  ggplot(aes(reference_date, offset, colour = factor(level))) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_line() +
  facet_wrap(~location, scales = "free_y", ncol = 1) +
  labs(
    title = "Realised offset (played - base), horizon 2",
    subtitle = "Zero through the burn-in season by construction. With season_policy='carry' the offset is held across the off-season gap.",
    x = NULL, y = "offset (admissions)", colour = "level"
  )
```

```{r offset-scale}
cal$forecasts %>%
  filter(!is_burn_in, !is.na(value_base), value_base > 0) %>%
  group_by(horizon, level) %>%
  summarize(
    median_abs_offset = median(abs(offset)),
    median_rel = median(abs(offset) / pmax(value_base, 1)),
    .groups = "drop"
  ) %>%
  ggplot(aes(level, median_rel, colour = factor(horizon))) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Median |offset| as a fraction of the base forecast",
    subtitle = "How large a correction the tracker is applying, and how that scales with horizon.",
    x = "quantile level", y = "median |offset| / base", colour = "horizon"
  )
```
