---
title: MultiQT calibration — whole-season views of four states, across method variants
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 2
params:
  burn_in_seasons: "2023-2024"
  season_policy: "carry"
  lr_window: 20
  lr_mult: 0.03
  lr_floor: 0.001
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
# cache = TRUE chunks are keyed on chunk code + cache.extra, NOT on functions
# in R/ -- after changing calibration code, clear cache/calibration/knitr_cache_seasons/.
knitr::opts_chunk$set(
  fig.align = "center", fig.width = 10, fig.height = 6,
  message = FALSE, warning = FALSE, cache = FALSE,
  cache.path = here::here("cache/calibration/knitr_cache_seasons/")
)
knitr::opts_knit$set(root.dir = here::here())
ggplot2::theme_set(ggplot2::theme_bw())
```

This notebook takes the sqrt-space operating point from `notes/CALIBRATION.md`
(`transform = "sqrt"`, `lr_mult = 0.03`, `lr_window = 20`, carry, burn-in
2023-24) and looks at four
states across whole seasons rather than one panel at a time
(`calibration_qt_gallery.Rmd` has the per-forecast gallery view). The states
are chosen by how much the baseline calibration changed them over the two live
seasons, pooled over horizons −1…3: the best and worst by WIS change, and the
best and worst by absolute error of the median. If one state is best (or
worst) on both, the next-ranked distinct state is taken so four panels are
shown. Method variants compared: the sqrt single-term tracker (baseline), the
paper's count-space tracker, a static conformal warm start from the burn-in
season with no tracking, the warm start plus a leaky fast tracker, the warm
start plus slow and leaky fast terms (the current operating point), and
switching off after February 15 as the earlier reference.

```{r load, cache=TRUE}
suppressPackageStartupMessages(source(here::here("R/load_all.R")))

forecasts <- hub_read_forecasts()
truth <- hub_read_truth()
rounds <- hub_label_seasons(unique(forecasts$reference_date))
burn_in <- strsplit(params$burn_in_seasons, ",")[[1]]

geo_map <- get_population_data() %>%
  distinct(state_id, .keep_all = TRUE) %>%
  transmute(state_id, abbrev = toupper(state_id), location = state_code)
pop <- get_population_data() %>%
  distinct(state_code, .keep_all = TRUE) %>%
  transmute(location = state_code, population)

archive <- nanoparquet::read_parquet(
  here::here("cache/calibration/nhsn_archive_flu.parquet")
) %>%
  as_epi_archive(compactify = TRUE)

live_rounds <- rounds %>% filter(!season %in% burn_in)
min_version <- min(archive$DT$version)
snapshots <- purrr::map(live_rounds$round_date, function(rd) {
  as_of <- rd - 3L
  if (as_of < min_version) {
    return(NULL)
  }
  epix_as_of(archive, as_of) %>%
    as_tibble() %>%
    inner_join(geo_map, by = c("geo_value" = "state_id")) %>%
    transmute(reference_date = rd, location, time_value, value)
}) %>% bind_rows()

seasons <- c("2024-2025", "2025-2026")
```

```{r variants, cache=TRUE, dependson="load"}
base_args <- list(
  burn_in_seasons = burn_in, season_policy = params$season_policy,
  lr_window = params$lr_window, transform = "sqrt",
  lr_args = list(mult = params$lr_mult, floor = params$lr_floor),
  progress = FALSE
)
variant_args <- list(
  baseline = list(),
  count_space = list(transform = "identity", lr_args = list(mult = params$lr_mult)),
  warm_static = list(slow_init = "burn_in_quantile", lr_args = list(mult = 1e-6, floor = 0)),
  warm_fast = list(slow_init = "burn_in_quantile", fast_decay = 0.1),
  warm_slow_fast = list(slow_init = "burn_in_quantile", lr_slow = list(mult = 0.003), fast_decay = 0.1),
  off_feb = list(off_after = "02-15")
)
variant_titles <- c(
  baseline = "Sqrt-space tracker (baseline)",
  count_space = "Count-space tracker (the paper's method)",
  warm_static = "Conformal warm start only, no tracking",
  warm_fast = "Warm start + leaky fast tracker",
  warm_slow_fast = "Warm start + slow term + leaky fast tracker (operating point)",
  off_feb = "Calibration off after February 15"
)

cals <- purrr::map(variant_args, function(v) {
  do.call(calibrate_hub_forecasts, c(list(forecasts, truth), modifyList(base_args, v)))
})
```

```{r select-states, dependson="variants"}
pinball <- function(y, q, tau) ifelse(y >= q, tau * (y - q), (1 - tau) * (q - y))

by_state <- cals$baseline$forecasts %>%
  filter(!is_burn_in, !is.na(truth)) %>%
  group_by(location) %>%
  summarize(
    wis_pct = 100 * (1 - sum(pinball(truth, value_cal, level)) / sum(pinball(truth, value_base, level))),
    ae_pct = 100 * (1 - sum(abs(truth - value_cal)[level == 0.5]) / sum(abs(truth - value_base)[level == 0.5])),
    .groups = "drop"
  ) %>%
  inner_join(select(geo_map, location, abbrev), by = "location")

# Best/worst by WIS then AE, skipping a state already taken so four are shown.
pick <- function(ranked, taken) ranked$location[!ranked$location %in% taken][1]
roles <- list(
  `best WIS` = arrange(by_state, desc(wis_pct)),
  `worst WIS` = arrange(by_state, wis_pct),
  `best AE` = arrange(by_state, desc(ae_pct)),
  `worst AE` = arrange(by_state, ae_pct)
)
taken <- character(0)
for (r in names(roles)) taken <- c(taken, pick(roles[[r]], taken))
states <- tibble(location = taken, role = names(roles)) %>%
  left_join(by_state, by = "location") %>%
  mutate(label = sprintf("%s (%s: WIS %+.1f%%, AE %+.1f%%)", abbrev, role, wis_pct, ae_pct))
```

# Which states, and why

Per-state change from baseline calibration over both live seasons, all
horizons pooled (positive is an improvement). The four panels below are the
extremes of these two columns.

```{r state-table}
knitr::kable(states %>% transmute(state = abbrev, role, `WIS change %` = round(wis_pct, 1), `AE change %` = round(ae_pct, 1)))
```

# Summary across variants

WIS % improvement over the base forecast (all seasons, burn-in excluded), per
horizon.

```{r variant-summary}
purrr::imap(cals, function(cal, nm) {
  hub_quantile_loss(cal) %>% transmute(variant = variant_titles[[nm]], horizon, pct = round(pct_improvement, 1))
}) %>%
  bind_rows() %>%
  tidyr::pivot_wider(names_from = horizon, values_from = pct, names_prefix = "h")
```

## Where in the season the baseline gains and loses

Baseline variant, pooled over all locations and both live seasons, by calendar
month of the reference date. `WIS change %` is per month and horizon; `share of
base WIS %` is how much of the season's total base WIS that month carries, so a
large percentage loss in April costs little in absolute terms while a small
gain in December or January is worth a lot. `median shift %` is the calibrated
median minus the base median, summed over the month, as a percentage of summed
truth: the direction and size of the correction the tracker was applying.

```{r by-month}
by_month <- cals$baseline$forecasts %>%
  filter(!is_burn_in, !is.na(truth)) %>%
  mutate(month = format(reference_date, "%b"), month_n = as.integer(format(reference_date, "%m"))) %>%
  group_by(horizon, month, month_n) %>%
  summarize(
    base = sum(pinball(truth, value_base, level)),
    cal = sum(pinball(truth, value_cal, level)),
    shift = 100 * sum((value_cal - value_base)[level == 0.5]) / sum(truth[level == 0.5]),
    .groups = "drop"
  ) %>%
  group_by(horizon) %>%
  mutate(pct = 100 * (base - cal) / base, share = 100 * base / sum(base)) %>%
  ungroup() %>%
  mutate(month = factor(month, levels = unique(month[order((month_n - 7) %% 12)])))

show_wide <- function(col, digits = 1) {
  by_month %>%
    select(horizon, month, value = all_of(col)) %>%
    mutate(value = round(value, digits)) %>%
    tidyr::pivot_wider(names_from = horizon, values_from = value, names_prefix = "h") %>%
    arrange(month)
}
knitr::kable(show_wide("pct"), caption = "WIS change % (calibrated vs base)")
knitr::kable(show_wide("share"), caption = "Share of base WIS %")
knitr::kable(show_wide("shift"), caption = "Median shift % of truth")
```

```{r helpers}
headline_table <- function(cal) {
  bind_rows(
    left_join(
      hub_coverage_summary(cal),
      select(hub_quantile_loss(cal), horizon, loss_base, loss_cal, loss_pct = pct_improvement),
      by = "horizon"
    ) %>% mutate(season = "all"),
    left_join(
      hub_coverage_summary(cal, by = "season"),
      select(hub_quantile_loss(cal, by = "season"),
        season, horizon, loss_base, loss_cal, loss_pct = pct_improvement
      ),
      by = c("season", "horizon")
    )
  ) %>%
    transmute(
      season, horizon,
      `cal err (base)` = round(cal_error_base, 4),
      `cal err (calibrated)` = round(cal_error_cal, 4),
      `cal err improvement %` = round(pct_improvement, 1),
      `WIS (base)` = round(2 * loss_base, 1),
      `WIS (calibrated)` = round(2 * loss_cal, 1),
      `WIS improvement %` = round(loss_pct, 1)
    ) %>%
    arrange(season, horizon)
}

fan_cols <- c(base = "#0072B2", calibrated = "#D55E00")

# One plot per (state, season): whole-season fan (odd season_rounds only, h 0..3;
# h -1 dropped -- with h -1 in, 5 horizons x ~13 rounds x 2 fans was unreadable)
# plus vintage patchwork strips underneath, plus an eta strip.
season_plot <- function(cal, loc, season) {
  rr <- cal$rounds %>% filter(season == !!season, season_round %% 2 == 1)
  ref_dates <- sort(rr$round_date)

  fin <- truth %>%
    filter(location == loc, target_end_date >= min(ref_dates) - 21L, target_end_date <= max(ref_dates) + 28L)

  fan <- cal$forecasts %>%
    filter(location == loc, reference_date %in% ref_dates, horizon %in% 0:3, level %in% c(0.1, 0.5, 0.9)) %>%
    mutate(
      target_end_date = reference_date + 7L * horizon,
      lvl = c(`0.1` = "lo", `0.5` = "med", `0.9` = "hi")[as.character(level)]
    ) %>%
    select(reference_date, target_end_date, lvl, base = value_base, calibrated = value_cal) %>%
    tidyr::pivot_longer(c(base, calibrated), names_to = "which") %>%
    tidyr::pivot_wider(names_from = lvl, values_from = value)

  # Vintage patchwork: each plotted round's as-of snapshot paints only the
  # fortnight since the previous plotted round (like backtesting.Rmd's vignette).
  vint_windows <- tibble(ref = ref_dates) %>%
    mutate(
      win_start = dplyr::lag(ref, default = ref[1] - 14L) - 3L,
      win_end = ref - 3L
    )
  vint <- purrr::pmap(vint_windows, function(ref, win_start, win_end) {
    snapshots %>%
      filter(reference_date == ref, location == loc, time_value >= win_start, time_value <= win_end)
  }) %>% bind_rows()

  main <- ggplot() +
    geom_line(data = fin, aes(target_end_date, truth), colour = "grey65") +
    geom_point(data = fin, aes(target_end_date, truth), colour = "grey65", size = 0.7) +
    geom_ribbon(
      data = filter(fan, which == "base"),
      aes(target_end_date, ymin = lo, ymax = hi, fill = which, group = reference_date),
      alpha = 0.2
    ) +
    geom_line(
      data = filter(fan, which == "calibrated"),
      aes(target_end_date, lo, colour = which, group = reference_date), linewidth = 0.3
    ) +
    geom_line(
      data = filter(fan, which == "calibrated"),
      aes(target_end_date, hi, colour = which, group = reference_date), linewidth = 0.3
    ) +
    geom_line(
      data = fan,
      aes(target_end_date, med, colour = which, group = interaction(reference_date, which))
    ) +
    geom_line(data = vint, aes(time_value, value), colour = "black") +
    geom_point(data = vint, aes(time_value, value), colour = "black", size = 0.6) +
    geom_vline(xintercept = ref_dates - 3L, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
    scale_fill_manual(values = fan_cols, aesthetics = c("fill", "colour")) +
    labs(x = NULL, y = "admissions", colour = NULL, fill = NULL) +
    theme(legend.position = "top")

  eta <- cal$series %>%
    filter(location == loc, season == !!season, horizon %in% 0:3) %>%
    # lr is 0 at rounds with no reveal (gaps); blank those rather than plot dips.
    mutate(lr = if_else(lr > 0, lr, NA_real_)) %>%
    ggplot(aes(round_date, lr, colour = factor(horizon))) +
    geom_line(na.rm = TRUE) +
    geom_point(size = 0.6, na.rm = TRUE) +
    labs(x = NULL, y = "eta", colour = "horizon") +
    theme(legend.position = "top")

  title <- paste0(geo_map$abbrev[match(loc, geo_map$location)], " — ", season)
  cowplot::plot_grid(
    cowplot::ggdraw() + cowplot::draw_label(title, fontface = "bold"),
    cowplot::plot_grid(main, eta, ncol = 1, rel_heights = c(3, 1), align = "v", axis = "lr"),
    ncol = 1, rel_heights = c(0.08, 1)
  )
}

# oh_table.R's internals table, generalized to any location/horizon/season.
oh_table <- function(cal, loc, h, season) {
  cal$forecasts %>%
    filter(location == loc, horizon == h, season == !!season, level %in% c(0.05, 0.25, 0.5, 0.75, 0.95)) %>%
    mutate(level = paste0("q", level)) %>%
    select(reference_date, level, lr_level, truth, hidden, offset, value_base, value_cal) %>%
    tidyr::pivot_wider(
      names_from = level,
      values_from = c(hidden, offset, value_base, value_cal, lr_level),
      names_glue = "{level}_{.value}"
    ) %>%
    mutate(
      eta = round(q0.5_lr_level, 3),
      across(c(truth, matches("hidden|offset|value_base|value_cal")), ~ round(.x))
    ) %>%
    select(
      reference_date, eta, truth,
      matches("^q.*_hidden$"), matches("^q.*_offset$"),
      matches("^q.*_value_base$"), matches("^q.*_value_cal$")
    ) %>%
    arrange(reference_date)
}

internals_block <- function(cal, season) {
  for (i in seq_len(nrow(states))) {
    loc <- states$location[i]
    cat(sprintf("\n\n<details><summary>%s internals (h=2)</summary>\n\n", states$label[i]))
    print(knitr::kable(oh_table(cal, loc, 2L, season)))
    cat("\n\n</details>\n\n")
  }
}
```

The whole-season fans plot horizons 0–3 only (dropping −1) and only every
other round. Blue ribbon and line: base 80% band and median; orange lines:
calibrated 10%, 50%, 90% quantiles. Black: the NHSN vintage each forecast saw,
painted over the fortnight before its reference date (dashed); grey: finalized
truth. The strip below is eta per horizon (blank where a round revealed no
outcome). Read it as "where do the medians diverge, and does the orange band
contain the grey truth better than the blue", not round-by-round.

```{r sections, results='asis'}
for (nm in names(cals)) {
  cal <- cals[[nm]]
  cat(sprintf("\n\n# %s\n\n", variant_titles[[nm]]))

  desc <- switch(nm,
    baseline = "The single-term tracker run on sqrt(counts): offsets and eta live in sqrt
      units, so a correction learned at the peak is smaller in counts when applied in the
      trough. lr_mult = 0.03, lr_window = 20, carry, isotonic projection; offsets update
      every round including the spring tail. The eta strip is in sqrt units.",
    count_space = "The paper's method as originally run: additive offsets in count space, same
      multiplier and window. Kept for reference; the eta strip is in counts.",
    warm_static = "No online tracking at all (eta is effectively zero). The offset per (horizon,
      level) is the conformal offset implied by the 2023-24 burn-in season pooled over
      locations -- the level-quantile of truth minus the base quantile, in sqrt units -- and
      it is applied unchanged to both live seasons. Shows how much of the gain is the
      persistent low bias alone.",
    warm_fast = "The conformal warm start plus the sqrt-space tracker with a 10% per-round leak
      on its offset: a correction that is not refreshed by new outcomes decays back toward
      the warm-start level instead of persisting through the reveal lag.",
    warm_slow_fast = "The current operating point: warm start, a slow term that keeps updating
      the persistent level with a tiny step (mult 0.003, expanding window, never decays,
      carries across seasons), and the leaky fast tracker on top. The internals table's
      hidden column is slow + fast.",
    off_feb = "The earlier reference: rounds from February 15 through June play the base
      forecast unchanged and take no gradient step; the offset learned through the peak is
      frozen and carried into the next season."
  )
  cat(desc, "\n\n")

  cat("## Headline\n\n")
  print(knitr::kable(headline_table(cal)))

  for (season in seasons) {
    cat(sprintf("\n\n## %s\n\n", season))
    for (i in seq_len(nrow(states))) {
      loc <- states$location[i]
      cat(sprintf("\n\n### %s — %s\n\n", states$label[i], season))
      print(season_plot(cal, loc, season))
    }
  }

  cat("\n\n## Internals\n\n")
  for (season in seasons) {
    cat(sprintf("\n\n### %s\n\n", season))
    internals_block(cal, season)
  }
}
```
