---
title: "Why are the COVID 2024-2025 season scores so bad on the first day?"
date: "compiled on `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
body {
  display: block;
  max-width: 1280px !important;
  margin-left: auto;
  margin-right: auto;
}

body .main-container {
  max-width: 1280px !important;
  width: 1280px !important;
}
```

$$\\[.4in]$$

```{r echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_template$set(figure1 = list(fig.height = 4, fig.width=4))
source(here::here("R/load_all.R"))
```

The scores on the first forecast day for COVID are all unusually bad, bad enough we initially thought it was a bug.
This notebook is a demonstration that these forecasts are somewhat reasonable given the context, and separated out because it would otherwise be distracting.
The primary reason is that it has an unusual amount of revision.
In comparison, the flu season delayed their initial forecast by a day, which allowed a new data revision to be used, which explains why flu doesn't have this problem.

# Revision Behavior

First, getting the necessary archive, forecasts, and scores, and plotting the versions around the first forecast day `2024-11-20` (the week of `2024-11-23`):

```{r, fig.height = 15, fig.width = 15}
covid_scores <- qs2::qs_read(here::here("covid_hosp_prod", "objects", "scores"))
covid_forecasts <- qs2::qs_read(here::here("covid_hosp_prod", "objects", "joined_forecasts_and_ensembles"))
covid_archive <- qs2::qs_read(here::here("covid_hosp_prod", "objects", "nhsn_archive_data"))
text_size <- 6
covid_archive$DT %>%
  filter(time_value < as.Date("2024-11-23") + 4*7, time_value > "2024-09-01") %>%
  as_epi_archive() %>%
  autoplot(.versions = c(as.Date("2024-11-20"), covid_archive$versions_end)) +
  geom_vline(aes(xintercept = as.Date("2024-11-23")))
```

Most locations have a significantly different version on `2024-11-20`, some by as much as 4 times the final version.

```{r}
# Building a function to plot the different forecasters
plot_problem_day <- function(forecaster, text_size = 6) {
  covid_scores %>% filter(forecast_date == "2024-11-23") %>% arrange(wis)
covid_forecasts %>% filter(forecaster =="climate_base") %>% filter(forecast_date == "2024-11-23")
  cmu_timeseries_fc <- covid_forecasts %>% filter(forecaster ==.env$forecaster) %>% filter(forecast_date == "2024-11-23")
  cmu_timeseries_wide <- cmu_timeseries_fc %>%
    pivot_wider(names_from = "quantile", values_from = "value")
  covid_archive$DT %>% filter(time_value < as.Date("2024-11-23") + 4*7, time_value > "2024-09-01") %>% as_epi_archive() %>% autoplot() +
    geom_vline(aes(xintercept = as.Date("2024-11-23"))) +
    geom_ribbon(data = cmu_timeseries_wide, aes(x = target_end_date, ymin = `0.1`, ymax = `0.9`), alpha = 0.3) +
    geom_ribbon(data = cmu_timeseries_wide, aes(x = target_end_date, ymin = `0.25`, ymax = `0.75`), alpha = 0.3) +
    geom_line(data = cmu_timeseries_wide, aes(x = target_end_date, y = `0.5`)) +
    facet_wrap(~geo_value, scale = "free")
}
```

### Windowed seasonal
```{r, fig.height = 15, fig.width = 15}
plot_problem_day("windowed_seasonal")
```

which is extrapolating out in a straight line from trends that are reporting artifacts.

### Covidhub-baseline
Covidhub baseline forms a good sanity check, since it is forecasting out from the versioned data (which explains why it is ~as  bad as `windowed_seasonal_extra_sources`)

```{r, fig.height = 15, fig.width = 15}
plot_problem_day("CovidHub-baseline")
```

### Linear
```{r, fig.height = 15, fig.width = 15}
plot_problem_day("linear")
```

which is extrapolating out in a straight line from trends that are reporting artifacts.

### Climate
```{r, fig.height = 15, fig.width = 15}
plot_problem_day("climate_base")
```

This one is wrong simply because this season was unusually low at this point.

### CMU-TimeSeries
This is a bit odd, since the forecaster we were using at the time was a simple average of the linear and climate forecasters, and so is off because of a combination of the reasons the linear and climate forecasts are off.

```{r, fig.height = 15, fig.width = 15}
plot_problem_day("CMU-TimeSeries")
```
