---
title: Investigating NHSN Data for Season 2024-2025
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

$$\\[.4in]$$

```{r echo=FALSE}
knitr::opts_chunk$set(
    fig.align = "center",
    message = FALSE,
    warning = FALSE,
    cache = TRUE
)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(MMWRweek)
library(readr)
source(here::here("R", "aux_data_utils.R"))
```

The data is coming from this repository: https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/ua7e-t2fy.

We are focused solely on the column `totalconfflunewadm` as the number of new confirmed flu admissions.

# Data Overview for Large States

```{r}
convert_epiweek_to_season <- function(epiyear, epiweek) {
    # Convert epiweek to season
    update_inds <- epiweek <= 39
    epiyear <- ifelse(update_inds, epiyear - 1, epiyear)

    season <- paste0(epiyear, "/", substr((epiyear + 1), 3, 4))
    return(season)
}

epiweeks_in_year <- function(year) {
    last_week_of_year <- seq.Date(as.Date(paste0(year, "-12-24")),
        as.Date(paste0(year, "-12-31")),
        by = 1
    )
    return(max(as.numeric(MMWRweek::MMWRweek(last_week_of_year)$MMWRweek)))
}

convert_epiweek_to_season_week <- function(epiyear, epiweek, season_start = 39) {
    season_week <- epiweek - 39
    update_inds <- season_week <= 0
    if (!any(update_inds)) {
        # none need to be updated
        return(season_week)
    }
    # last year's # of epiweeks determines which week in the season we're at at
    # the beginning of the year
    season_week[update_inds] <- season_week[update_inds] +
        sapply(epiyear[update_inds] - 1, epiweeks_in_year)

    return(season_week)
}

df <- readr::read_csv("https://data.cdc.gov/resource/ua7e-t2fy.csv?$limit=20000&$select=weekendingdate,jurisdiction,totalconfflunewadm")
df <- df %>% add_season_info()
df %>%
    filter(jurisdiction %in% c("CA", "FL", "TX", "NY", "PA", "MN", "NM")) %>%
    ggplot(aes(x = weekendingdate, y = totalconfflunewadm, color = jurisdiction)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions by Jurisdiction",
        x = "Week Ending Date",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# State Season Comparison {.tabset}

## California

```{r}
df %>%
    filter(jurisdiction == "CA") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in CA by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Florida

```{r}
df %>%
    filter(jurisdiction == "FL") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in FL by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Texas

```{r}
df %>%
    filter(jurisdiction == "TX") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in TX by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## New York

```{r}
df %>%
    filter(jurisdiction == "NY") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in NY by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Pennsylvania

```{r}
df %>%
    filter(jurisdiction == "PA") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in PA by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Minnesota

```{r}
df %>%
    filter(jurisdiction == "MN") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in MN by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## New Mexico

```{r}
df %>%
    filter(jurisdiction == "NM") %>%
    ggplot(aes(x = season_week, y = totalconfflunewadm, color = season)) +
    geom_line() +
    theme_minimal() +
    labs(
        title = "New Confirmed Flu Admissions in NM by Season",
        x = "Season Week",
        y = "Total Confirmed Flu Admissions"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Comparing with Old NHSN Data

The old NHSN data can be found here: https://healthdata.gov/Hospital/COVID-19-Reported-Patient-Impact-and-Hospital-Capa/g62h-syeh.

The relevant column is `previous_day_admission_influenza_confirmed`.

```{r}
# You can install this with `pak::pkg_install("cmu-delphi/epiprocess")
library(epiprocess)
library(plotly)
library(DT)
library(tidyr)

hhs_arch <- read_csv("https://healthdata.gov/resource/g62h-syeh.csv?$limit=90000&$select=date,state,previous_day_admission_influenza_confirmed") %>%
    mutate(
        geo_value = tolower(state),
        time_value = as.Date(date) - 1L,
        hhs = previous_day_admission_influenza_confirmed
    ) %>%
    select(geo_value, time_value, hhs) %>%
    as_epi_df() %>%
    group_by(geo_value) %>%
    epi_slide_sum(
        hhs,
        na.rm = TRUE,
        .window_size = 7L,
        .ref_time_values = seq.Date(as.Date("2020-01-04"), as.Date("2024-04-24"), by = 7),
    ) %>%
    mutate(hhs = slide_value_hhs) %>%
    select(-slide_value_hhs)

to_compare <- df %>%
    mutate(time_value = as.Date(weekendingdate), geo_value = tolower(jurisdiction), nhsn = totalconfflunewadm) %>%
    select(-weekendingdate, -jurisdiction, -totalconfflunewadm) %>%
    full_join(hhs_arch, by = join_by(geo_value, time_value))
p <- to_compare %>%
    mutate(diffed = nhsn - hhs, rel_diff = diffed / hhs) %>%
    filter(abs(diffed) > 10) %>%
    ggplot(aes(x = time_value, y = diffed, color = geo_value)) +
    geom_point() +
    labs(
        title = "new value minus original value, filtered to the difference in counts > 10"
    )

ggplotly(p, tooltip = "text", height = 800, width = 1000)
```

```{r}
p <- to_compare %>%
    mutate(diffed = nhsn - hhs, rel_diff = diffed / hhs) %>%
    filter(abs(diffed) > 10) %>%
    ggplot(aes(x = time_value, y = rel_diff, color = geo_value)) +
    geom_point() +
    labs(
        title = "relative version of the previous, dividing by the original value"
    )

ggplotly(p, tooltip = "text", height = 800, width = 1000)
```

```{r}
p <- to_compare %>%
    mutate(diffed = nhsn - hhs, rel_diff = diffed / hhs) %>%
    filter(abs(diffed) > 10) %>%
    group_by(geo_value) %>%
    summarise(median_diff = median(abs(diffed)), median_rel_diff = median(abs(rel_diff))) %>%
    ggplot(aes(x = geo_value, weight = median_diff)) +
    geom_bar() +
    labs(
        title = "Median absolute difference across states"
    )

ggplotly(p, tooltip = "text", height = 800, width = 1000)
```

```{r}
to_compare %>%
    mutate(diffed = nhsn - hhs, rel_diff = diffed / hhs) %>%
    filter(abs(diffed) > 10) %>%
    datatable()
```

## Directly plotting the worst contenders

```{r}
p <- to_compare %>%
    rename(old_source = hhs, new_source = nhsn) %>%
    pivot_longer(names_to = "data_source", cols = c("old_source", "new_source")) %>%
    filter(geo_value %in% c("ca", "tx", "la", "mi", "in")) %>%
    ggplot(aes(x = time_value, y = value, color = data_source)) +
    geom_line() +
    facet_wrap(~geo_value, scales = "free")

ggplotly(p, tooltip = "text", height = 800, width = 1000)
```
