---
title: Overall Comparison of Hospitalization Forecasters
date: "Rendered: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    code_folding: hide
params:
  forecaster_parameters: ""
  forecasts: ""
  scores: ""
  truth_data: ""
  disease: ""
  dataset: ""
  outside_forecaster_subset: ""
  baseline_forecaster: ""
editor_options:
  chunk_output_type: console
---

$$\\[.4in]$$

```{r echo=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
ggplot2::theme_set(ggplot2::theme_bw())
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
```

```{r}
# params <- list(
#   forecaster_parameters = tar_read(forecaster_parameter_combinations),
#   forecasts = tar_read(joined_forecasts),
#   scores = tar_read(joined_scores),
#   truth_data = tar_read(hhs_evaluation_data),
#   disease = "flu"
# )

# set the baseline and ensemble names
if (all(params$outside_forecaster_subset != "")) {
  base_forecaster_name <- params$outside_forecaster_subset[[1]]
  if (params$baseline_forecaster != "") {
    base_forecaster_name <- params$baseline_forecaster
  }
  ensemble_forecaster_name <- params$outside_forecaster_subset[[2]]
} else if (params$disease == "flu") {
  base_forecaster_name <- "FluSight-baseline"
  ensemble_forecaster_name <- "FluSight-ensemble"
} else {
  base_forecaster_name <- "COVIDhub-baseline"
  ensemble_forecaster_name <- "COVIDhub_CDC-ensemble"
}

# Select forecasters for this notebook
outside_forecasters <- c(base_forecaster_name, ensemble_forecaster_name)
id_to_forecaster <- params$forecaster_parameters %>%
  purrr::imap(\(x, idx) {
    x %>%
      mutate(family_name = idx) %>%
      select(id, forecaster, family_name)
  }) %>%
  bind_rows() %>%
  rename(forecaster_function = forecaster)

if (params$dataset == "nhsn") {
  min_score <- 20
} else {
  min_score <- 0
}

overall_rating <- params$scores %>%
  summarize(mean_score = mean(wis), .by = forecaster) %>%
  arrange(mean_score) %>%
  filter(mean_score > min_score) %>%
  left_join(id_to_forecaster, by = join_by(forecaster == id)) %>%
  mutate(
    family_name = if_else(is.na(family_name), "external", family_name),
    forecaster_function = if_else(is.na(forecaster_function), "external", forecaster_function)
  )
best_in_class <- overall_rating %>%
  slice_min(mean_score, by = family_name) %>%
  pull(forecaster)
forecasters_to_keep <- c(best_in_class, outside_forecasters) %>% unique()
# Load forecasts and scores
scores <- params$scores %>%
  filter(forecaster %in% forecasters_to_keep)
forecast_dates <- scores %>%
  pull(forecast_date) %>%
  unique()
forecasters <- best_in_class
aheads <- scores %>%
  pull(ahead) %>%
  unique()

# Define aggregation functions
Mean <- function(x) mean(x, na.rm = TRUE)
GeoMean <- function(x, offset = 0) exp(Mean(log(x + offset)))
```

### Notebook Information

Best examples from each family of forecasters.

#### Disease

`r params$disease`

#### Forecast Dates

`r forecast_dates`

#### Forecaster Parameter Mapping and Overall Scores

The table is sorted by ascending WIS and contains all the forecasters in this notebook.

```{r}
# Display the table
if (params$disease == "flu") {
  ignore_keys <- c("forecaster", "keys_to_ignore", "pop_scaling")
} else {
  ignore_keys <- c("forecaster", "keys_to_ignore")
}

param_table <- params$forecaster_parameters %>%
  map(\(x) x %>% filter(id %in% forecasters_to_keep)) %>%
  bind_rows() %>%
  left_join(id_to_forecaster, by = join_by(id)) %>%
  select(-any_of(ignore_keys)) %>%
  full_join(
    scores %>%
      group_by(forecaster) %>%
      summarize(
        mean_ae = round(Mean(ae), 3),
        geomean_ae = round(GeoMean(ae), 3),
        mean_wis = round(Mean(wis), 3),
        geomean_wis = round(GeoMean(wis), 3),
        mean_coverage_90 = round(Mean(coverage_90), 3),
      ) %>%
      rename(id = forecaster)
  ) %>%
  arrange(mean_wis) %>%
  relocate(family = family_name, id, mean_ae, geomean_ae, mean_wis, geomean_wis, mean_coverage_90)
datatable(param_table)
```

$$\\[.07in]$$

### Score Plots {.tabset}

- The WIS plots are relative to the baseline. These plots are aggregated with the geometric mean.
- The absolute error plots are also relative to the baseline and are aggregated with the arithmetic mean.
- The dashed black line in all plots is the ensemble.

#### WIS by Forecast Date (GeoMean)

```{r}
var <- "wis"
group_cols <- c("forecaster", "forecast_date", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := GeoMean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (
  df %>%
    filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>%
    nrow() > 0
) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

facets.label <- str_glue("{aheads} days ahead")
names(facets.label) <- aheads
subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df,
  aes(x = forecast_date, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_hline(yintercept = 1, linetype = 1, color = "black") +
  facet_grid(rows = vars(ahead)) +
  facet_wrap(~ahead, nrow = 4, labeller = labeller(ahead = facets.label)) +
  scale_x_date(breaks = "1 month", date_labels = "%b\n%y") +
  scale_y_log10(n.breaks = 10) +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Forecast Dates", y = "Geometric Mean WIS")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### WIS by Ahead

```{r}
var <- "wis"
group_cols <- c("forecaster", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := GeoMean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (df %>% filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>% nrow() > 0) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df,
  aes(x = ahead, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_hline(yintercept = 1, linetype = 1, color = "black") +
  scale_x_date(breaks = "1 month", date_labels = "%b\n%y") +
  scale_y_log10(n.breaks = 20) +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Days ahead", y = "Geometric Mean WIS")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### Absolute Error by Forecast Date

```{r}
var <- "ae"
group_cols <- c("forecaster", "forecast_date", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (
  df %>%
    filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>%
    nrow() > 0
) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

facets.label <- str_glue("{aheads} days ahead")
names(facets.label) <- aheads
subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df,
  aes(x = forecast_date, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_hline(yintercept = 1, linetype = 1, color = "black") +
  facet_grid(rows = vars(ahead)) +
  scale_x_date(breaks = "1 month", date_labels = "%b\n%y") +
  facet_wrap(~ahead, nrow = 4, labeller = labeller(ahead = facets.label)) +
  scale_y_log10(n.breaks = 10) +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Forecast Dates", y = "Arithmetic Mean Absolute Error")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### Absolute Error by Ahead

```{r}
var <- "ae"
group_cols <- c("forecaster", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(group_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

# Make sure we don't divide by zero
if (df %>% filter(forecaster == base_forecaster_name & near(!!sym(var), 0)) %>% nrow() > 0) {
  warning("scale_by_forecaster will divide by zero in column ", var)
}

# Normalize the metric by the baseline
normalized_df <- df %>%
  pivot_wider(names_from = forecaster, names_prefix = var, values_from = !!sym(var)) %>%
  mutate(across(starts_with(var), ~ .x / !!sym(paste0(var, base_forecaster_name)))) %>%
  pivot_longer(cols = starts_with(var), names_to = "forecaster", values_to = var) %>%
  mutate(forecaster = stringr::str_remove(forecaster, var)) %>%
  filter(forecaster != base_forecaster_name)

subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  normalized_df,
  aes(x = ahead, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_hline(yintercept = 1, linetype = 1, color = "black") +
  scale_y_log10(n.breaks = 25) +
  scale_x_date(breaks = "1 month", date_labels = "%b\n%y") +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Days ahead", y = "Arithmetic Mean Absolute Error")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### % Coverage by Forecast Date

```{r}
var <- "coverage_90"
group_cols <- c("forecaster", "forecast_date", "ahead")

# Aggregate metric across groups
df <- scores %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(group_cols)) %>%
  filter(ahead >= 0)

facets.label <- str_glue("{aheads} days ahead")
names(facets.label) <- aheads
subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  df,
  aes(x = forecast_date, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_hline(yintercept = .8, linetype = 1, color = "black") +
  facet_grid(rows = vars(ahead)) +
  facet_wrap(~ahead, nrow = 4, labeller = labeller(ahead = facets.label)) +
  scale_color_discrete() +
  scale_x_date(breaks = "1 month", date_labels = "%b\n%y") +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Forecast Dates", y = "Arithmetic Mean 90% Coverage")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

#### % Coverage by Ahead

```{r}
var <- "coverage_90"
id_cols <- c("forecaster", "ahead")

# Aggregate metric across groups
df <- scores %>%
  select(all_of(c(id_cols, var))) %>%
  drop_na(!!sym(var)) %>%
  summarize(!!var := Mean(!!sym(var)), .by = all_of(id_cols)) %>%
  filter(ahead >= 0)

subtitle <- sprintf(
  "Forecasts made over %s to %s",
  format(min(forecast_dates), "%B %d, %Y"),
  format(max(forecast_dates), "%B %d, %Y")
)
p <- ggplot(
  df,
  aes(x = ahead, y = !!sym(var))
) +
  geom_line(aes(color = forecaster, group = forecaster)) +
  geom_point(aes(color = forecaster, group = forecaster)) +
  geom_hline(yintercept = .8, linetype = 1, color = "black") +
  scale_color_discrete() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = subtitle, x = "Days ahead", y = "Arithmetic Mean 90% Coverage")

ggplotly(p, tooltip = "text", height = 800, width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```


### Fan plots {.tabset}

Fan plots showing the 80% prediction intervals for the forecasts made by the CMU forecasters and the outside forecasters. The black line is the truth data.

```{r}
plot_dates <- forecast_dates[seq(1, length(forecast_dates), by = 4)]
plot_fan_plot <- function(geo_vals) {
    forecast_subset <- params$forecasts %>%
    filter(
        geo_value %in% geo_vals,
        forecast_date %in% plot_dates
    ) %>%
    mutate(quantile = as.character(quantile)) %>%
    pivot_wider(names_from = "quantile", values_from = "prediction") %>%
    mutate(ahead = as.numeric(target_end_date - forecast_date))

    p <- ggplot(
    data = forecast_subset,
    aes(x = target_end_date, group = forecast_date)
    ) +
      geom_ribbon(aes(ymin = `0.1`, ymax = `0.8`, fill = forecast_date), alpha = 0.3) +
      geom_line(aes(y = `0.5`, color = forecast_date), linetype = 2L) +
      geom_point(aes(y = `0.5`, color = forecast_date), size = 0.75) +
      geom_line(
      data = params$truth_data %>%
        filter(
          geo_value %in% geo_vals,
          target_end_date >= min(plot_dates),
          target_end_date <= max(plot_dates) + 5 * 7
        ),
      aes(x = target_end_date, y = true_value),
      inherit.aes = FALSE, na.rm = TRUE,
      color = "black", linetype = 1
    ) +
      scale_y_continuous(n.breaks = 15) +
      scale_x_date(breaks = "1 month", date_labels = "%b\n%y") +
      facet_grid(factor(forecaster, levels = param_table$id) ~ geo_value, scales = "free") +
      labs(x = "Reference Date", y = "Weekly Sums of Hospitalizations", title = "Monthly Forecasts and Truth Data") +
      theme(legend.position = "none")
    ggplotly(p, tooltip = "text", height = 300 * length(param_table$id), width = 1000) %>%
  layout(hoverlabel = list(bgcolor = "white"))
}
```

#### Large States

```{r, fig.height = 40}
# We plot a subset of the dates and geos for the fan plot
plot_fan_plot(c("ca", "fl", "pa", "tx", "ny"))
```

#### Medium States

```{r}
# We plot a subset of the dates and geos for the fan plot
plot_fan_plot(c("mn", "co", "ky", "md"))
```

#### Small States

```{r}
# We plot a subset of the dates and geos for the fan plot
plot_fan_plot(c("nv", "vt", "de", "ri", "nm"))
```
