---
title: "TimesFM vs. R forecasters (`r params$target`)"
author: Delphi Forecast Team
date: "Rendered: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    code_folding: hide
    toc: True
params:
  target: "nhsn"
  start_date: "2025-10-01"
  end_date: "2026-07-08"
  forecasters: !r c("windowed_seasonal", "windowed_seasonal_extra_sources", "ensemble_mix", "climate_linear", "linear")
  exclude_geos: !r c("us", "as", "gu", "mp", "vi")
---

```{r setup, echo=TRUE}
knitr::opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
suppressPackageStartupMessages(source(here::here("R", "load_all.R")))
ggplot2::theme_set(ggplot2::theme_bw())
Mean <- function(x) mean(x, na.rm = TRUE)
q9_levels <- seq(0.1, 0.9, by = 0.1)
```

# TimesFM vs. R forecasters: `r params$target` {.tabset}

This report backtests Google's TimesFM 3.0 foundation model against the repo's R
forecasters on the `r params$target` target, over forecast dates
`r params$start_date` to `r params$end_date`. The main comparison point is
`windowed_seasonal`, the repo's workhorse flu forecaster.

## Setup & load

```{r load, echo=TRUE}
r_forecasts_path <- here::here("cache", "tsfm", "r_forecasts.parquet")
timesfm_forecasts_path <- here::here("cache", "tsfm", "timesfm_forecasts.parquet")
truth_path <- here::here("cache", "tsfm", "truth.parquet")
population_path <- here::here("cache", "tsfm", "population.parquet")
schedule_path <- here::here("cache", "tsfm", "forecast_schedule.parquet")

r_forecasts_raw <- arrow::read_parquet(r_forecasts_path)
truth_raw <- arrow::read_parquet(truth_path)
population <- arrow::read_parquet(population_path)
forecast_schedule <- arrow::read_parquet(schedule_path)

start_date <- as.Date(params$start_date)
end_date <- as.Date(params$end_date)

# TimesFM forecasts may not exist yet (backtest still running); degrade gracefully.
have_timesfm <- file.exists(timesfm_forecasts_path)
if (have_timesfm) {
  timesfm_forecasts_raw <- arrow::read_parquet(timesfm_forecasts_path)
  timesfm_forecaster_ids <- timesfm_forecasts_raw %>%
    filter(target == params$target) %>%
    distinct(forecaster) %>%
    pull(forecaster) %>%
    stringr::str_subset("^timesfm3_")
} else {
  cli::cli_alert_warning("{timesfm_forecasts_path} not found; comparing R forecasters only.")
  timesfm_forecasts_raw <- tibble::tibble()
  timesfm_forecaster_ids <- character(0)
}

wanted_forecasters <- c(params$forecasters, timesfm_forecaster_ids)

filter_common <- function(df) {
  df %>%
    filter(target == params$target) %>%
    filter(forecast_date >= start_date, forecast_date <= end_date) %>%
    filter(geo_value %nin% params$exclude_geos) %>%
    filter(forecaster %in% wanted_forecasters)
}

r_forecasts <- filter_common(r_forecasts_raw)
timesfm_forecasts <- if (have_timesfm) filter_common(timesfm_forecasts_raw) else timesfm_forecasts_raw

all_forecasts <- bind_rows(r_forecasts, timesfm_forecasts)

# Warn (don't error) about any requested forecaster missing entirely from the data.
present_forecasters <- unique(all_forecasts$forecaster)
missing_requested <- setdiff(params$forecasters, present_forecasters)
if (length(missing_requested) > 0) {
  cli::cli_alert_warning("Requested forecaster(s) not found in data, dropping: {missing_requested}")
}
compare_forecasters <- intersect(wanted_forecasters, present_forecasters)
all_forecasts <- all_forecasts %>% filter(forecaster %in% compare_forecasters)

truth <- truth_raw %>%
  filter(source == params$target) %>%
  filter(geo_value %nin% params$exclude_geos) %>%
  select(geo_value, target_end_date, true_value = value)

cli::cli_alert_info("Comparing forecasters: {compare_forecasters}")
```

```{r common-tasks, echo=TRUE}
# Restrict to the common set of (geo_value, forecast_date, target_end_date) tasks
# present for every forecaster being compared, so scores are apples-to-apples.
task_cols <- c("geo_value", "forecast_date", "target_end_date")
tasks_by_forecaster <- all_forecasts %>%
  distinct(forecaster, geo_value, forecast_date, target_end_date)

n_forecasters <- length(compare_forecasters)
common_tasks <- tasks_by_forecaster %>%
  count(geo_value, forecast_date, target_end_date, name = "n_forecasters_present") %>%
  filter(n_forecasters_present == n_forecasters) %>%
  select(all_of(task_cols))

n_all_tasks <- tasks_by_forecaster %>% distinct(across(all_of(task_cols))) %>% nrow()
n_common_tasks <- nrow(common_tasks)
cli::cli_alert_info(
  "Common tasks: {n_common_tasks} of {n_all_tasks} distinct (geo, forecast_date, target_end_date) triples ({n_all_tasks - n_common_tasks} dropped)."
)

all_forecasts_common <- all_forecasts %>%
  semi_join(common_tasks, by = task_cols)
```

```{r scoring, echo=TRUE}
# Two scoring variants: all 23 quantiles ("full") and the 9 quantiles TimesFM
# natively produces ("q9"). The remaining 14 hub quantiles for TimesFM are
# tail-extrapolated, so scores_q9 is the fairer head-to-head.
to_scoring_input <- function(df) {
  df %>%
    rename(model = forecaster, prediction = value)
}

# Like evaluate_predictions() in R/scoring.R, but tolerates quantile sets that
# lack the 0.25/0.75 pair (the 9-quantile variant), reporting NA coverage there.
score_quantiles <- function(forecasts, truth_data) {
  forecast_obj <- forecasts %>%
    arrange(model, geo_value, target_end_date, forecast_date, quantile) %>%
    group_by(model, geo_value, target_end_date, forecast_date) %>%
    mutate(prediction = sort(prediction)) %>%
    ungroup() %>%
    left_join(truth_data, by = c("geo_value", "target_end_date")) %>%
    filter(!is.na(true_value)) %>%
    scoringutils::as_forecast_quantile(
      quantile_level = "quantile",
      observed = "true_value",
      predicted = "prediction",
      forecast_unit = c("model", "geo_value", "forecast_date", "target_end_date")
    )
  scores <- forecast_obj %>%
    scoringutils::score(metrics = scoringutils::get_metrics(.)) %>%
    as_tibble()
  for (col in c("interval_coverage_50", "interval_coverage_90")) {
    if (!col %in% names(scores)) scores[[col]] <- NA
  }
  scores %>%
    select(model, geo_value, forecast_date, target_end_date, wis, ae = ae_median,
           coverage_50 = interval_coverage_50, coverage_90 = interval_coverage_90)
}

score_variant <- function(df) {
  score_quantiles(to_scoring_input(df), truth) %>%
    rename(forecaster = model) %>%
    mutate(ahead_wk = as.numeric(target_end_date - forecast_date - 3) / 7)
}

scores_full <- score_variant(all_forecasts_common)
scores_q9 <- score_variant(all_forecasts_common %>% filter(quantile %in% q9_levels))

# Population-normalized WIS (nhsn only; nssp is already a rate).
if (params$target == "nhsn") {
  scores_full <- scores_full %>%
    left_join(population, by = "geo_value") %>%
    mutate(wis_rate = wis * 1e5 / population)
  scores_q9 <- scores_q9 %>%
    left_join(population, by = "geo_value") %>%
    mutate(wis_rate = wis * 1e5 / population)
}
```

Ahead -1 is a nowcast of the already-reported week (`target_end_date` before
`forecast_date`); TimesFM emits the (already-known) observed value there, which
makes it look artificially strong, so most of the summaries below are also
shown restricted to `ahead_wk >= 0`.

## Headline table

```{r headline, echo=TRUE}
headline_table <- function(scores, label) {
  scores %>%
    filter(ahead_wk >= 0) %>%
    group_by(forecaster) %>%
    summarize(
      mean_wis = Mean(wis),
      mean_ae = Mean(ae),
      coverage_50 = Mean(coverage_50),
      coverage_90 = Mean(coverage_90),
      n = n(),
      .groups = "drop"
    ) %>%
    mutate(
      rel_wis_vs_windowed_seasonal = mean_wis / mean_wis[forecaster == "windowed_seasonal"][1],
      scoring = label
    ) %>%
    arrange(mean_wis)
}

headline <- bind_rows(
  headline_table(scores_full, "full (23 quantiles)"),
  headline_table(scores_q9, "q9 (9 quantiles)")
)

headline %>%
  mutate(across(c(mean_wis, mean_ae, coverage_50, coverage_90, rel_wis_vs_windowed_seasonal), ~ round(.x, 3))) %>%
  select(scoring, forecaster, mean_wis, rel_wis_vs_windowed_seasonal, mean_ae, coverage_50, coverage_90, n) %>%
  knitr::kable()
```

## WIS by ahead

```{r wis-by-ahead-table, echo=TRUE}
by_ahead <- function(scores, label) {
  scores %>%
    group_by(forecaster, ahead_wk) %>%
    summarize(mean_wis = Mean(wis), n = n(), .groups = "drop") %>%
    mutate(scoring = label)
}

wis_by_ahead <- bind_rows(
  by_ahead(scores_full, "full"),
  by_ahead(scores_q9, "q9")
)

wis_by_ahead %>%
  mutate(mean_wis = round(mean_wis, 3)) %>%
  arrange(scoring, ahead_wk, mean_wis) %>%
  knitr::kable()
```

```{r wis-by-ahead-plot, echo=TRUE, fig.width=12, fig.height=5}
wis_by_ahead %>%
  ggplot(aes(x = ahead_wk, y = mean_wis, color = forecaster)) +
  geom_line() +
  geom_point() +
  facet_wrap(~scoring, scales = "free_y") +
  labs(x = "ahead (weeks)", y = "mean WIS", title = "Mean WIS by ahead")
```

## WIS over time

```{r wis-over-time, echo=TRUE, fig.width=12, fig.height=5}
scores_q9 %>%
  filter(ahead_wk >= 0) %>%
  group_by(forecaster, forecast_date) %>%
  summarize(mean_wis = Mean(wis), .groups = "drop") %>%
  ggplot(aes(x = forecast_date, y = mean_wis, color = forecaster)) +
  geom_line() +
  geom_point() +
  scale_y_log10() +
  labs(x = "forecast date", y = "mean WIS (log scale)", title = "Mean WIS over time (q9 scores, ahead >= 0)")
```

## By geo

```{r by-geo, echo=TRUE, fig.width=10, fig.height=8}
if (length(timesfm_forecaster_ids) > 0) {
  focus_timesfm <- timesfm_forecaster_ids[1]

  by_geo <- scores_q9 %>%
    filter(ahead_wk >= 0, forecaster %in% c("windowed_seasonal", focus_timesfm)) %>%
    group_by(forecaster, geo_value) %>%
    summarize(mean_wis = Mean(wis), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = forecaster, values_from = mean_wis) %>%
    filter(!is.na(.data[["windowed_seasonal"]]), !is.na(.data[[focus_timesfm]])) %>%
    mutate(rel_wis = .data[[focus_timesfm]] / .data[["windowed_seasonal"]]) %>%
    arrange(rel_wis)

  by_geo %>%
    mutate(geo_value = forcats::fct_reorder(geo_value, rel_wis)) %>%
    ggplot(aes(x = geo_value, y = rel_wis)) +
    geom_col() +
    geom_hline(yintercept = 1, linetype = "dashed") +
    coord_flip() +
    labs(
      x = "geo", y = glue::glue("relative WIS ({focus_timesfm} / windowed_seasonal)"),
      title = "Relative WIS by geo (ahead >= 0, q9 scores)"
    )

  cli::cli_h3("10 best geos (lowest relative WIS)")
  by_geo %>% head(10) %>% knitr::kable()

  cli::cli_h3("10 worst geos (highest relative WIS)")
  by_geo %>% tail(10) %>% arrange(desc(rel_wis)) %>% knitr::kable()
} else {
  cli::cli_alert_info("No timesfm forecaster available; skipping by-geo comparison.")
}
```

## Coverage

Nominal coverage targets are 0.5 and 0.9.

```{r coverage, echo=TRUE}
scores_full %>%
  filter(ahead_wk >= 0) %>%
  group_by(forecaster, ahead_wk) %>%
  summarize(
    coverage_50 = round(Mean(coverage_50), 3),
    coverage_90 = round(Mean(coverage_90), 3),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(ahead_wk, forecaster) %>%
  knitr::kable()
```

## Fan charts

```{r fan-charts, echo=TRUE, fig.width=14, fig.height=8}
fan_geos <- c("us", "ca", "tx", "ny", "pa", "fl")
fan_geos <- intersect(fan_geos, unique(all_forecasts$geo_value))

if (length(timesfm_forecaster_ids) > 0 && length(fan_geos) > 0) {
  focus_timesfm <- timesfm_forecaster_ids[1]
  fan_forecasters <- c("windowed_seasonal", focus_timesfm)

  fan_wide <- all_forecasts %>%
    filter(
      geo_value %in% fan_geos,
      forecaster %in% fan_forecasters,
      (as.numeric(target_end_date - forecast_date - 3) / 7) %in% 0:3,
      quantile %in% c(0.05, 0.25, 0.5, 0.75, 0.95)
    ) %>%
    mutate(quantile_label = paste0("q", quantile)) %>%
    select(forecaster, geo_value, forecast_date, target_end_date, quantile_label, value) %>%
    tidyr::pivot_wider(names_from = quantile_label, values_from = value)

  fan_truth <- truth %>%
    filter(geo_value %in% fan_geos, target_end_date >= start_date - 28, target_end_date <= end_date + 28)

  fan_wide %>%
    ggplot(aes(x = target_end_date, group = forecast_date)) +
    geom_ribbon(aes(ymin = q0.05, ymax = q0.95), fill = "steelblue", alpha = 0.15) +
    geom_ribbon(aes(ymin = q0.25, ymax = q0.75), fill = "steelblue", alpha = 0.25) +
    geom_line(aes(y = q0.5), color = "steelblue", linewidth = 0.3, alpha = 0.8) +
    geom_line(
      data = fan_truth, aes(x = target_end_date, y = true_value, group = NULL),
      color = "black", linewidth = 0.5, inherit.aes = FALSE
    ) +
    facet_grid(forecaster ~ geo_value, scales = "free_y") +
    labs(x = "target end date", y = "value", title = "Fan charts: windowed_seasonal vs. TimesFM (aheads 0-3)")
} else {
  cli::cli_alert_info("No timesfm forecaster or no matching geos available; skipping fan charts.")
}
```

## Notes

- TimesFM natively produces only the 9 quantile levels 0.1-0.9; the remaining
  14 hub quantiles (0.01, 0.025, 0.05, 0.975, 0.99, and the extreme tails) are
  tail-extrapolated rather than model-native. `scores_full` includes this
  extrapolation; `scores_q9` restricts to the 9 native levels for a fairer
  comparison.
- Ahead -1 (`target_end_date` before `forecast_date`) is a nowcast of the
  already-reported week; TimesFM tends to emit close to the observed value
  there, which is not a meaningful test of forecasting skill. Headline and
  summary tables restrict to `ahead_wk >= 0` for this reason.
- Both sides are scored against as-of (versioned) data available at forecast
  time, not finalized truth revised after the fact -- except the truth table
  itself, which is finalized.
- The R forecasts are the actual real-time production runs; the TimesFM
  forecasts are a replay against the archive, run after the fact. This is not
  a fully apples-to-apples operational comparison (e.g. data availability
  timing may differ slightly).
- TimesFM 3.0's released weights are licensed for non-commercial use only,
  which is a consideration for any operational (as opposed to exploratory) use.
